### 2022-05-02 Manuscript.Rmd
Add simulation functions

---
title: "Non-convergence in iterative imputation"
author: "H.I. Oberman"
output: pdf_document
bibliography: references.bib  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(purrr)
library(mice)
miceadds::source.all("R")
```

# Abstract

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

# Introduction

- Theoretical background

- Research question

# Methods

- Simulation design (ADEMP)

- Simulation checklist (Table \ref{table:check})

Pseudo-code:

```
for (data geneneration conditions) {
  generate complete data
  for (amputation conditions) {
    ampute complete data
    for (imputation conditions) {
      impute incomplete data
      analyze imputed data
      }}}
```

\begin{table}[ht]
\caption{Checklist for reporting on imputation methodology evaluations.}
\label{table:check}
\begin{tabular}{ll}
% \hline
%   & Suggested aspects to report                                                       \\
\hline
1 & Simulation scope                                                                   \\ \cline{2-2}
  & Aim                                                                                \\
  & Design (incl. pseudo-code or flow diagram)                                         \\
  & Number of simulation repetitions                                                   \\ \hline
2 & Comparative truth                                                                  \\ \cline{2-2}
  & Data-generating mechanism (model-based or design-based)                            \\
  & Sampling variance                                                                  \\
%  & Data characteristics (number of observations, number of variables, variable types, \\
%  & coherence between variables incl. data structures e.g. clustering)                 \\
  & Data characteristics (incl. multivariate relations and structures e.g. clustering) \\
  & Estimand                                                                           \\ \hline 
3 & Induced missingness                                                                \\ \cline{2-2}
  & Missingness mechanism (incl. type or functional form of the missing data   model)  \\
  & Missingness pattern (incl. missingness proportion)                                 \\ \hline 
4 & Applied methods                                                                    \\ \cline{2-2}
  & Imputation methods (incl. parameters e.g. the number of imputations)               \\
  & Analytic methods (incl. calculation of standard errors e.g. pooling   rules)       \\
  & Reference method (e.g. complete case analysis)                                    \\ \hline 
5 & Imputation evaluation                                                              \\ \cline{2-2}
  & Imputation-generating process (e.g. algorithmic non-convergence)                   \\
  & Imputation model fit (e.g. posterior predictive checks)                            \\
  & Distributional characteristics (e.g. plausibility of imputed values)               \\ \hline 
6 & Performance evaluation                                                             \\ \cline{2-2}
  & Statistical properties (e.g. confidence validity)                                  \\
  & Comparative performance (e.g. predictive accuracy)                                 \\ \hline
\end{tabular}
\end{table}

```{r eval=FALSE, include=FALSE}
dat <- generation()
amp <- amputation(dat)
imp <- imputation(amp)
est <- evaluation(imp)
```

# Results

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

# Discussion

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

# References

### 2022-05-02 R/DGM.R

@@ -1,0 +1,36 @@
+generation <- function(r = 0.4, n = 500, p = 2) {
+  # generate data based on variance-covariance matrix
+  vcov <- matrix(r, p, p)
+  diag(vcov) <- 1
+  dat <- data.frame(
+    mvtnorm::rmvnorm(
+      n,
+      mean = rep(0, p),
+      sigma = vcov
+    )) %>% setNames(paste0("X", 1:p))
+  # output
+  out <- list(
+    n_obs = n,
+    n_var = p,
+    corr = r,
+    dat = dat)
+  return(out)
+}
+
+amputation <- function(datlist, mech = "MAR", type = "MID", prop = 0.25) {
+  # induce missingness
+  amp <- mice::ampute(datlist$dat, 
+                      prop = prop,
+                      mech = mech,
+                      type = type)
+  # output
+  out <- list(
+    n_obs = datlist$n_obs,
+    n_var = datlist$n_var,
+    corr = datlist$corr,
+    mis_mech = mech,
+    mis_type = type,
+    mis_prop = prop,
+    amp = amp)
+  return(out)
+}

### 2022-05-02 R/methods.R

@@ -1,0 +1,46 @@
+imputation <- function(amplist, meth = "norm", m = 5, it = 5) {
+  # for CCA
+  if(meth == "none") {
+    imp <- na.omit(amplist$amp$amp)
+    average <- mean(imp$X1)
+    rmse_cell <- NA
+  } 
+  # for actual imputation methods
+  if (meth != "none") {
+    if (meth == "mean" | meth == "norm.predict") {
+      imp <- mice::mice(
+        amplist$amp$amp,
+        m = 1,
+        method = meth,
+        maxit = 1,
+        print = FALSE
+      )
+    } else {
+      imp <- mice::mice(
+        amplist$amp$amp,
+        m = m,
+        method = meth,
+        maxit = it,
+        print = FALSE
+      )
+    }
+    average <- mice::complete(imp, "long") %>% .$X1 %>% mean()
+    rmse_cell <- apply(imp$imp$X1 - amplist$amp$data$X1[imp$where[, "X1"]], 2, rmse) %>% mean()
+  }
+  # output
+  out <- list(
+    n_obs = amplist$n_obs,
+    n_var = amplist$n_var,
+    corr = amplist$corr,
+    mis_mech = amplist$mis_mech,
+    mis_type = amplist$mis_type,
+    mis_prop = amplist$mis_prop,    
+    imp_meth = meth,
+    n_imp = m,
+    n_it = it,
+    average = average,
+    rmse_cell = rmse_cell,
+    imp = imp
+  )
+  return(out)
+}

### 2022-05-02 R/performance.R

@@ -1,0 +1,36 @@
+evaluation <- function(implist){
+  # for CCA
+  if (is.data.frame(implist$imp)) {
+    fit <- lm(X1 ~ ., implist$imp)
+    rmse_pred <- rmse(fit$residuals)
+  }
+  # for actual imputation methods
+  if (mice::is.mids(implist$imp)) {
+    mira <- mice::complete(implist$imp, "all") %>% 
+      purrr::map(~lm(X1 ~ ., .))
+    fit <- suppressWarnings(mice::pool(mira))
+    rmse_pred <- purrr::map_dbl(mira, ~rmse(.$residuals)) %>% mean()
+  }
+  # get estimated regression coefficients
+  est <- fit %>% broom::tidy(conf.int = TRUE) %>%
+    .[2, c("estimate", "conf.low", "conf.high")] 
+  # output
+  out <- data.frame(
+    n_obs = implist$n_obs,
+    n_var = implist$n_var,
+    corr = implist$corr,
+    mis_mech = implist$mis_mech,
+    mis_type = implist$mis_type,
+    mis_prop = implist$mis_prop,    
+    imp_meth = implist$imp_meth,
+    n_imp = implist$n_imp,
+    n_it = implist$n_it,
+    average = implist$average, 
+    coeff = est$estimate, 
+    ci_ll = est$conf.low,
+    ci_ul = est$conf.high,
+    rmse_pred = rmse_pred, 
+    rmse_cell = implist$rmse_cell
+  )
+  return(out)
+}

### 2022-05-02 R/simulation.R

@@ -1,0 +1,126 @@
+## simulation parameters ##
+
+# conditions
+n_obs = c(50, 500, 5000) 
+corr = c(0, .4, .8)
+mis_mech = c("MCAR", "MAR", "MNAR")
+mis_type = c("LEFT", "RIGHT", "MID", "TAIL")
+mis_prop = c(.1, .25, .5)
+n_imp = c(1, 5)
+imp_meth = c("mean", "norm.predict", "norm") 
+n_it = c(1, 5, 10)
+
+# combine into lists for mapping
+dat_par <- expand.grid(n_obs = n_obs, corr = corr)
+amp_par <- expand.grid(dat_nr = 1:nrow(dat_par), mis_mech = mis_mech, mis_type = mis_type, mis_prop = mis_prop, stringsAsFactors = FALSE)
+imp_par <- expand.grid(amp_nr = 1:nrow(amp_par), imp_meth = imp_meth, n_imp = n_imp, n_it = n_it, stringsAsFactors = FALSE)
+
+
+## simulation functions ##
+
+# generate data
+generation <- function(n, r){
+  dat <- data.frame(mvtnorm::rmvnorm(
+    n = n,
+    mean = c(10, 10),
+    sigma = matrix(c(1, r, r, 1), 2, 2)
+  )) %>% setNames(c("Y", "X"))
+}
+
+# ampute the data
+amputation <- function(dat, mech, type, prop){
+  amp <- dat %>% 
+    mice::ampute(
+      prop = prop,
+      mech = mech,
+      type = type
+    )}
+
+# impute the data
+imputation <- function(amp, meth, m, it){
+  imp <- amp$amp %>%
+    mice::mice(
+      m = m,
+      method = meth,
+      maxit = it,
+      print = FALSE
+    )}
+
+# evaluate the imputations
+evaluation <- function(imp, amp, dat, r){
+  # fit analysis model
+  fit_imps <- with(imp, lm(Y ~ X))
+  fits <- list(full = lm(Y~X, dat),
+               cca = lm(Y~X, amp$amp, na.action = na.omit),
+               imp = suppressWarnings(mice::pool(fit_imps)))
+  # performance
+  out <- purrr::map_dfr(fits, ~{
+    broom::tidy(.x, conf.int = TRUE) %>% 
+      .[2, c("estimate", "conf.low", "conf.high")]}) %>% 
+    mutate(
[…]

### 2022-05-02 R/utils.R

@@ -1,0 +1,1 @@
+rmse <- function(obs, pred = 0){sqrt(mean((obs-pred)^2))}

### 2022-05-03 R/simulation.R
Add simulation pipeline (for all conditions)

@@ -1,126 +1,0 @@
-## simulation parameters ##
-
-# conditions
-n_obs = c(50, 500, 5000) 
-corr = c(0, .4, .8)
-mis_mech = c("MCAR", "MAR", "MNAR")
-mis_type = c("LEFT", "RIGHT", "MID", "TAIL")
-mis_prop = c(.1, .25, .5)
-n_imp = c(1, 5)
-imp_meth = c("mean", "norm.predict", "norm") 
-n_it = c(1, 5, 10)
-
-# combine into lists for mapping
-dat_par <- expand.grid(n_obs = n_obs, corr = corr)
-amp_par <- expand.grid(dat_nr = 1:nrow(dat_par), mis_mech = mis_mech, mis_type = mis_type, mis_prop = mis_prop, stringsAsFactors = FALSE)
-imp_par <- expand.grid(amp_nr = 1:nrow(amp_par), imp_meth = imp_meth, n_imp = n_imp, n_it = n_it, stringsAsFactors = FALSE)
-
-
-## simulation functions ##
-
-# generate data
-generation <- function(n, r){
-  dat <- data.frame(mvtnorm::rmvnorm(
-    n = n,
-    mean = c(10, 10),
-    sigma = matrix(c(1, r, r, 1), 2, 2)
-  )) %>% setNames(c("Y", "X"))
-}
-
-# ampute the data
-amputation <- function(dat, mech, type, prop){
-  amp <- dat %>% 
-    mice::ampute(
-      prop = prop,
-      mech = mech,
-      type = type
-    )}
-
-# impute the data
-imputation <- function(amp, meth, m, it){
-  imp <- amp$amp %>%
-    mice::mice(
-      m = m,
-      method = meth,
-      maxit = it,
-      print = FALSE
-    )}
-
-# evaluate the imputations
-evaluation <- function(imp, amp, dat, r){
-  # fit analysis model
-  fit_imps <- with(imp, lm(Y ~ X))
-  fits <- list(full = lm(Y~X, dat),
-               cca = lm(Y~X, amp$amp, na.action = na.omit),
-               imp = suppressWarnings(mice::pool(fit_imps)))
-  # performance
-  out <- purrr::map_dfr(fits, ~{
-    broom::tidy(.x, conf.int = TRUE) %>% 
-      .[2, c("estimate", "conf.low", "conf.high")]}) %>% 
-    mutate(
[…]

### 2022-05-03 simulation.R

@@ -1,0 +1,26 @@
+## simulation parameters ##
+
+# setup
+set.seed(123)
+n_sim = 2
+
+# data generation parameters
+n_obs = 500 # c(50, 500, 5000) 
+corr = 0.4 # c(0, 0.4, 0.8)
+dat_par <- expand.grid(n_obs = n_obs, corr = corr)
+# amputation parameters
+mis_mech = c("MCAR", "MAR") # c("MCAR", "MAR", "MNAR")
+mis_type = "RIGHT"  # c("LEFT", "RIGHT", "MID", "TAIL")
+mis_prop = c(0.1, 0.25, 0.5)
+amp_par <- expand.grid(dat_nr = 1:nrow(dat_par), mis_mech = mis_mech, mis_type = mis_type, mis_prop = mis_prop, stringsAsFactors = FALSE)
+# imputation parameters
+imp_meth = c("none", "norm") # c("none", "mean", "norm.predict", "norm") 
+n_imp = 5 # c(1, 5, 10)
+n_it = 7 # c(1, 5, 10)
+imp_par <- expand.grid(amp_nr = 1:nrow(amp_par), imp_meth = imp_meth, n_imp = n_imp, n_it = n_it, stringsAsFactors = FALSE)
+
+# simulate
+all_dat <- purrr::pmap(dat_par, ~{generation(n = ..1, r = ..2)})
+all_amp <- purrr::pmap(amp_par, ~{amputation(all_dat[[..1]], mech = ..2, type = ..3, prop = ..4)})
+all_imp2 <- purrr::pmap_dfr(imp_par, ~{imputation(all_amp[[..1]], meth = ..2, m = ..3, it = ..4) %>% 
+    evaluation()})

### 2022-05-03 simulation.R
Replicate simulation function

@@ -19,8 +19,17 @@ n_imp = 5 # c(1, 5, 10)
 n_it = 7 # c(1, 5, 10)
 imp_par <- expand.grid(amp_nr = 1:nrow(amp_par), imp_meth = imp_meth, n_imp = n_imp, n_it = n_it, stringsAsFactors = FALSE)
 
-# simulate
-all_dat <- purrr::pmap(dat_par, ~{generation(n = ..1, r = ..2)})
-all_amp <- purrr::pmap(amp_par, ~{amputation(all_dat[[..1]], mech = ..2, type = ..3, prop = ..4)})
-all_imp2 <- purrr::pmap_dfr(imp_par, ~{imputation(all_amp[[..1]], meth = ..2, m = ..3, it = ..4) %>% 
-    evaluation()})
+# simulation function
+simulate_once <- function(...) {
+  all_dat <- purrr::pmap(dat_par, ~{generation(n = ..1, r = ..2)})
+  all_amp <- purrr::pmap(amp_par, ~{amputation(all_dat[[..1]], mech = ..2, type = ..3, prop = ..4)})
+  all_imp <- purrr::pmap(imp_par, ~{imputation(all_amp[[..1]], meth = ..2, m = ..3, it = ..4)})
+  out <- purrr::map_dfr(all_imp, ~evaluation(.x))
+  return(out)
+}
+
+# test function
+simulate_once()
+
+# testrun simulation 
+a <- replicate(n_sim, simulate_once(), simplify = FALSE)

### 2022-06-13 Manuscript/Manuscript.Rmd
Visualize simulation results

--- 2022-05-02 Manuscript.Rmd
@@
 knitr::opts_chunk$set(echo = TRUE)
-library(dplyr)
-library(purrr)
-library(mice)
-miceadds::source.all("R")
+library(tidyverse)
+
+# filters
+coeff <- "X4"
+maxit <- 20
+```
+
+```{r data}
+performance <- readRDS("./Results/performance.RDS")
+summary <- performance %>% 
+  group_by(method, mech, prop, .it, term) %>% 
+  summarise(across(c(bias, cov, ciw, ac, psrf), mean, na.rm = TRUE))
+SEs <- performance %>% 
+  group_by(method, mech, prop, .it, term) %>% 
+  summarise(across(c(bias, ciw, ac, psrf), var, na.rm = TRUE)) %>% 
+  mutate(across(c(bias, ciw, ac, psrf), sqrt))
+results <- left_join(summary, SEs, by = c("method", "mech", "prop", ".it", "term"), suffix = c("", "_SE")) %>% 
+  mutate(cov = cov,
+         cov_SE = sqrt((cov * (1 - cov)) / 2000))
 ```
@@
 
-```{r eval=FALSE, include=FALSE}
-dat <- generation()
-amp <- amputation(dat)
-imp <- imputation(amp)
-est <- evaluation(imp)
-```
-
 # Results
@@
 Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+```{r}
+# results %>% 
+#   filter(term == "X2", .it < 26) %>% 
+#   ggplot(aes(x = .it, y = bias)) +
+#   geom_hline(yintercept = 0, color = "grey", size = 2) +
+#   geom_line(color = "blue") +
+#   geom_line(aes(x = .it, y = bias + bias_SE), color = "lightblue", linetype = "dashed") +
+#   geom_line(aes(x = .it, y = bias - bias_SE), color = "lightblue", linetype = "dashed") +
+#   facet_grid(as.factor(prop) ~ mech) +
+#   theme_classic()
+
+results %>% 
+  filter(term == coeff, .it > 0, .it < maxit + 1) %>% 
+  ggplot(aes(x = .it, y = bias)) +
+  geom_hline(yintercept = 0, color = "black") +
+  geom_errorbar(aes(ymin = bias - bias_SE, ymax = bias + bias_SE), alpha = 0.1) +
+  geom_line(color = "blue") +
+  geom_point(color = "blue") +
+  facet_grid(as.factor(prop) ~ mech) +
+  theme_classic()
+
+results %>% 
+  filter(term == coeff, .it > 0, .it < maxit + 1) %>% 
+  ggplot(aes(x = .it, y = ac)) +
+  geom_hline(yintercept = 0, color = "black") +
+  geom_errorbar(aes(ymin = ac - ac_SE, ymax = ac + ac_SE), alpha = 0.1) +
+  geom_line(color = "blue") +
+  geom_point(color = "blue") +
+  facet_grid(as.factor(prop) ~ mech) +
+  theme_classic()
+
+results %>% 
+  filter(term == coeff, .it > 0, .it < maxit + 1) %>% 
+  ggplot(aes(x = .it, y = psrf)) +
+  geom_hline(yintercept = 1, color = "black") +
+  geom_errorbar(aes(ymin = psrf - psrf_SE, ymax = psrf + psrf_SE), alpha = 0.1) +
+  geom_line(color = "blue") +
+  geom_point(color = "blue") +
+  facet_grid(as.factor(prop) ~ mech) +
+  theme_classic()
+```
 

### 2022-06-13 Functions/generate_datasets.R

@@ -3,6 +3,7 @@
 # generate complete data
 generate_complete <- function(
     n_obs = 200,
+    corr = 0.3,
     betas = c(-0.5,-0.1, 0.1, 0.5)
 ) {
   # create variance-covariance matrix with moderate correlations
@@ -45,12 +46,13 @@ induce_missingness <- function(
 # combine into one function
 create_data <- function(
     sample_size = 200, 
+    correlations = 0.3,
     effects = c(-0.5,-0.1, 0.1, 0.5),
     mechanisms = c("MCAR", "MAR"),
     proportions = c(0.1, 0.25, 0.5)
 ) {
   # create a single complete dataset
-  dat <- generate_complete(n_obs = sample_size, betas = effects)
+  dat <- generate_complete(n_obs = sample_size, corr = correlations, betas = effects)
   # ampute the data with different missingness
   amps <- induce_missingness(dat, mis_mech = mechanisms, mis_prop = proportions)
   # output

### 2022-06-13 simulation.R

@@ -17,17 +17,18 @@ set.seed(11)
 # parameters
 n_sim <- 2000
 n_obs <- 200
-n_it <- 100
+corr <- 0.3
 betas <- c(-0.5, -0.1, 0.1, 0.5)
 mis_mech = c("MCAR", "MAR")
 mis_prop = c(0.1, 0.25, 0.5)
+n_it <- 100
 
 # #################################
 # ### TEST LOWER LEVEL FUCTIONS ###
 # #################################
 # 
 # # generate data
-# dat <- generate_complete(n_obs, betas)
+# dat <- generate_complete(n_obs, corr, betas)
 # 
 # # ampute data
 # amps <- induce_missingness(dat, mis_mech = "MAR", mis_prop = 0.5)
@@ -82,8 +83,8 @@ results_raw <- replicate(
   simulate_once(n_obs, betas, mis_mech, mis_prop),
   simplify = FALSE
   )
-# save raw results
-saveRDS(results_raw, "./Results/raw.RDS")
+# # save raw results
+# saveRDS(results_raw, "./Results/raw.RDS")
 
 ########################
 ### EVALUATE RESULTS ###
@@ -91,7 +92,7 @@ saveRDS(results_raw, "./Results/raw.RDS")
 
 # calculate bias, coverage rate and CI width
 performance <- evaluate_est(results_raw)
-saveRDS(performance, "./Results/performance.RDS")
+# saveRDS(performance, "./Results/performance.RDS")
 
 # simulation results across all conditions
 performance %>% 

### 2022-06-13 testrun.R

@@ -1,0 +1,37 @@
+# nonconv
+
+# packages
+library(dplyr)
+library(mvtnorm)
+library(mice)
+library(miceadds)
+
+# functions
+miceadds::source.all("./functions")
+
+# parameters
+n_sim <- 2000
+n_obs <- 200
+corr <- 0.1
+betas <- c(-0.5, -0.1, 0.1, 0.5)
+mis_mech = c("MCAR", "MAR")
+mis_prop = c(0.1, 0.25, 0.5)
+n_it <- 100
+
+
+# generate data
+dat <- generate_complete(n_obs, corr, betas)
+dat$X5 <- (2 * dat$X3) + dat$X4
+
+# ampute data
+amps <- induce_missingness(dat, mis_mech = "MAR", mis_prop = 0.5)
+
+# impute data with MICE
+imp <- mice(amps[[1]]$amp, maxit = 10)
+plot(imp)
+ggmice::plot_trace(imp)
+
+# convergence
+conv <- convergence(imp)
+ggplot(conv, aes(x = .it, y = ac, color = vrb)) + geom_hline(yintercept = 0) + geom_line() + theme_classic()
+ggplot(conv, aes(x = .it, y = psrf, color = vrb)) + geom_hline(yintercept = 1) + geom_line() + theme_classic()

### 2022-06-16 Functions/generate_datasets.R
Edit missing data pattern (for multivariate missingness)

@@ -22,9 +22,19 @@ generate_complete <- function(
   return(dat)
 }
 
+# generate multivariate missingness patterns patterns
+create_patterns <- function() {
+  mis_pat <-
+    expand.grid(c(0, 1), c(0, 1), c(0, 1), c(0, 1), c(0, 1)) %>%
+    setNames(., c("Y", "X1", "X2", "X3", "X4")) %>%
+    .[rowSums(.) > 1 & rowSums(.) < 5, ]
+  return(mis_pat)
+}
+
 # ampute the complete data
 induce_missingness <- function(
     dat,
+    mis_pat = NULL, 
     mis_mech = c("MCAR", "MAR"),
     mis_prop = c(0.1, 0.25, 0.5)
 ) {
@@ -33,7 +43,7 @@ induce_missingness <- function(
   amps <- purrr::map(mis_mech, function(mm) {
     purrr::map(mis_prop, function(mp) {
       # ampute the data
-      mice::ampute(dat, mech = mm, prop = mp) 
+      mice::ampute(dat, patterns = mis_pat, mech = mm, prop = mp) 
     }) 
   }) %>% 
     # unlist just make sure the proportions 
@@ -48,13 +58,14 @@ create_data <- function(
     sample_size = 200, 
     correlations = 0.3,
     effects = c(-0.5,-0.1, 0.1, 0.5),
+    patterns = NULL, 
     mechanisms = c("MCAR", "MAR"),
     proportions = c(0.1, 0.25, 0.5)
 ) {
   # create a single complete dataset
   dat <- generate_complete(n_obs = sample_size, corr = correlations, betas = effects)
   # ampute the data with different missingness
-  amps <- induce_missingness(dat, mis_mech = mechanisms, mis_prop = proportions)
+  amps <- induce_missingness(dat, mis_pat, mis_mech = mechanisms, mis_prop = proportions)
   # output
   return(amps)
 }

### 2022-06-16 simulation.R

@@ -15,13 +15,14 @@ miceadds::source.all("./functions")
 set.seed(11)
 
 # parameters
-n_sim <- 2000
+n_sim <- 10
 n_obs <- 200
-corr <- 0.3
-betas <- c(-0.5, -0.1, 0.1, 0.5)
+corr <- 0.5
+betas <- c(1, 1, -1, -1)
+mis_pat <- create_patterns()
 mis_mech = c("MCAR", "MAR")
-mis_prop = c(0.1, 0.25, 0.5)
-n_it <- 100
+mis_prop = c(0.25, 0.5, 0.75)
+n_it <- 10
 
 # #################################
 # ### TEST LOWER LEVEL FUCTIONS ###
@@ -31,16 +32,7 @@ n_it <- 100
 # dat <- generate_complete(n_obs, corr, betas)
 # 
 # # ampute data
-# amps <- induce_missingness(dat, mis_mech = "MAR", mis_prop = 0.5)
-# 
-# # apply complete case analysis
-# CCA <- apply_CCA(amps[[1]])
-# 
-# # impute data with MICE
-# MICE <- apply_MICE(amps[[1]])
-# 
-# # impute data with python
-# ### [YOUR FUNCTION HERE] ###
+# amps <- induce_missingness(dat, mis_pat, mis_mech = "MAR", mis_prop = 0.5)
 # 
 # ##################################
 # ### TEST HIGHER LEVEL FUCTIONS ###
@@ -53,11 +45,12 @@ n_it <- 100
 ### COMBINE INTO ONE FUCTION ###
 ################################
 
-simulate_once <- function(n_obs, betas, mis_mech, mis_prop) {
+simulate_once <- function(n_obs, betas, mis_pat, mis_mech, mis_prop) {
   # generate incomplete data
   amps <- create_data(
     sample_size = n_obs,
     effects = betas,
+    patterns = mis_pat,
     mechanisms = mis_mech,
     proportions = mis_prop
   )
@@ -80,7 +73,7 @@ simulate_once <- function(n_obs, betas, mis_mech, mis_prop) {
 # repeat the simulation function n_sim times
 results_raw <- replicate(
   n_sim, 
-  simulate_once(n_obs, betas, mis_mech, mis_prop),
+  simulate_once(n_obs, betas, mis_pat, mis_mech, mis_prop),
   simplify = FALSE
   )
 # # save raw results

### 2022-06-16 testrun.R

@@ -1,30 +1,12 @@
 # nonconv
-
-# packages
-library(dplyr)
-library(mvtnorm)
-library(mice)
-library(miceadds)
-
-# functions
-miceadds::source.all("./functions")
-
-# parameters
-n_sim <- 2000
-n_obs <- 200
-corr <- 0.1
-betas <- c(-0.5, -0.1, 0.1, 0.5)
-mis_mech = c("MCAR", "MAR")
-mis_prop = c(0.1, 0.25, 0.5)
-n_it <- 100
-
+library(ggplot2)
 
 # generate data
 dat <- generate_complete(n_obs, corr, betas)
-dat$X5 <- (2 * dat$X3) + dat$X4
+# dat$X5 <- (2 * dat$X3) + dat$X4
 
 # ampute data
-amps <- induce_missingness(dat, mis_mech = "MAR", mis_prop = 0.5)
+amps <- induce_missingness(dat, mis_pat = NULL, mis_mech = "MAR", mis_prop = 0.5)
 
 # impute data with MICE
 imp <- mice(amps[[1]]$amp, maxit = 10)

### 2023-01-17 Manuscript/manuscript.qmd
setup manuscript with motivating example

---
title: "Non-convergence in iterative imputation"
author: "H. I. Oberman"
abstract: "Iterative imputation has become the de facto standard to accommodate for the ubiquitous problem of missing data. While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence. Our study provides insight into identifying non-convergence in iterative imputation algorithms. We show that these algorithms can yield correct outcomes even when a converged state has not yet formally been reached. In the cases considered, inferential validity is achieved after five to ten iterations, much earlier than indicated by diagnostic methods. We conclude that it never hurts to iterate longer, but such calculations hardly bring added value."
# keywords: "missing data, iterative imputation, non-convergence, MICE"
# bibliography: references.bib
bibliography: references.bib
---

```{r setup, echo=FALSE}
library(mice)
library(ggmice)
library(ggplot2)
```

# Introduction

Iterative imputation has become the de facto standard to accommodate for missing data. The aim is usually to draw valid inferences, i.e. to get unbiased, confidence-valid estimates that incorporate the effects of the missingness. Such estimates are obtained with iterative imputation by separating the missing data problem from the scientific problem. The missing values are imputed (i.e., filled in) using some sort of algorithm. And subsequently, the scientific model of interest is performed on the completed data. To obtain valid scientific estimates, both the missing data problem and the scientific problem should be appropriately considered. The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputations.

All inferences rely on the convergence of the imputation algorithm, yet determining whether an algorithm has converged is not trivial. There has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms. A widely accepted practice is visual inspection of the algorithm, although diagnosing convergence through visual inspection may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18]. Therefore, a quantitative diagnostic method to assess convergence would be preferred.

It is challenging to arrive upon a single point at which convergence has been reached. Since the aim is to converge to a distribution and not to a single point, the algorithm may produce some fluctuations even after it has converged. Because of this property, it may be more desirable to focus on *non*-convergence. Fortunately, there are non-convergence identifiers for other iterative algorithms, but the validity of these identifiers has not been systematically evaluated on imputation algorithms.

In this study, we explore different methods for identifying non-convergence in iterative imputation algorithms. We evaluate whether these methods are able to cover the extent of the non-convergence, and we also investigate the relation between non-convergence and the validity of the inferences. We translate the results of our simulation study into guidelines for practice, which we demonstrate by means of a motivating example.

# Motivating Example

We use empirical incomplete data: the `boys` dataset from the `mice` package, which contains data on some health-related variables for 748 Dutch boys [@mice]. Say we're interested in the relation between children's heights and their respective ages, we could use a linear regression model to predict `hgt` from `age`. However, as figure XYZ shows, the variable `hgt` is not completely observed. To be able to analyze these data, we need to solve the missing data problem first.

```{r boys, echo=FALSE}
ggmice::ggmice(mice::boys, ggplot2::aes(age, hgt)) + ggplot2::geom_point()
```

The incomplete height variable can be imputed based on auxiliary variables, such as weight. After imputation with `mice`, one would conventionally inspect the trace plots for signs of non-convergence.

```{r trace, echo=FALSE}
imp <- mice::mice(mice::boys[, c("age", "hgt", "wgt")], maxit = 20, printFlag = FALSE)
plot_trace(imp)
```

-   Conventional inspection of algorithmic

-   We are interested in predicting age from the other variables, in particular in the regression coefficient of `hgt`

-   We compare non-convergence identified using visual inspection versus rhat in the chain variances, scientific estimate and lambda.

-   The figures show results of a `mice` run with 20 iterations but otherwise default settings.

```{r case, fig.cap = "Case study.", echo=FALSE, eval=FALSE}
trace <- param %>% 
  ggplot() + 
  geom_line(aes(x = it, y = mu.hgt, color = as.factor(m))) + 
  theme_classic() + 
  labs(x = "Iteration number",
       y = "Chain mean variable 'hgt'",
       color = "Imputation")
means <- casestudy_plot(v="r.hat.max.mu.hgt", name = "Rhat chain means") + 
  list(geom_hline(yintercept = 1.2, color = "grey"))
qhats <-
  casestudy_plot(v = "r.hat.max.qhat", name = "Rhat regression estimate") +
  list(geom_hline(yintercept = 1.2, color = "grey"))
lambdas <- casestudy_plot(v = "r.hat.max.lambda", name = "Rhat lambda") +
  list(geom_hline(yintercept = 1.2, color = "grey"))
(trace + means) / (qhats + lambdas) + plot_layout(guides="collect") + plot_annotation(tag_levels = 'A')

```

From the traceplot of the chain means (see \@ref(fig:case)A) it seems that mixing improves up-to 10 iterations, while trending is only apparent in the first three iterations.

This figure (\@ref(fig:case)B) shows that 7 iterations are required before the $\widehat{R}$-values of the chain means drop below the threshold for non-convergence.

The $\widehat{R}$-values for the scientific estimate reaches the threshold much sooner, when $n_{it}=14$ (see \@ref(fig:case)C).

According to the $\widehat{R}$-values with $\lambda$ as parameter, at least 15 iterations are required (see \@ref(fig:case)D).

```{r}

```

### 2025-04-07 Functions/apply_methods.R
make sim parallel

@@ -3,15 +3,14 @@
 # complete case analysis
 apply_CCA <- function(amp) {
   # list-wise deletion
-  est <- na.omit(amp$amp) %>% 
-    # fit regression 
-    lm(Y ~ X1 + X2 + X3 + X4, .) %>% 
+  est <- # fit regression 
+    lm(Y ~ X1 + X2 + X3 + X4, data = na.omit(amp$amp)) |> 
     # clean results
-    broom::tidy(conf.int = TRUE) %>% 
+    broom::tidy(conf.int = TRUE) |> 
     # choose estimates
-    select(term, estimate, conf.low, conf.high) %>% 
-    # add method name and missingness
-    cbind(method = "CCA", mech = amp$mech, prop = amp$prop, .it = 0, ., ac_mean = NA, psrf_mean = NA, ac_sd = NA, psrf_sd = NA) 
+    select(term, estimate, conf.low, conf.high)
+  # add method name and missingness
+  est <- cbind(method = "CCA", mech = amp$mech, prop = amp$prop, .it = 0, est, ac_mean = NA, psrf_mean = NA, ac_sd = NA, psrf_sd = NA)
   # rename "(Intercept)" to "Y" for easier processing
   est[est$term == "(Intercept)", "term"] <- "Y"
   # output
@@ -51,11 +50,11 @@ apply_MICE <- function(amp, n_it) {
 # internal function to calculate pooled regression estimates
 estimate_param <- function(imp) {
   # run analysis on all imputations
-  est <- with(imp, lm(Y ~ X1 + X2 + X3 + X4)) %>% 
+  est <- with(imp, lm(Y ~ X1 + X2 + X3 + X4)) |> 
     # pool results
-    mice::pool() %>% 
+    mice::pool() |> 
     # clean results
-    broom::tidy(conf.int = TRUE) %>% 
+    broom::tidy(conf.int = TRUE) |> 
     # select estimates
     select(term, estimate, conf.low, conf.high)
   # rename "(Intercept)" to "Y" for easier processing
@@ -77,12 +76,13 @@ add_iteration <- function(implist) {
 # combine into one function
 apply_methods <- function(amps, betas, n_it) {
   # apply CCA to each incomplete dataset
-  CCA <- purrr::map_dfr(amps, ~{apply_CCA(.)})
+  CCA <- purrr::map_dfr(amps, ~{apply_CCA(.x)})
[…]

### 2025-04-07 Functions/generate_datasets.R

@@ -23,11 +23,13 @@ generate_complete <- function(
 }
 
 # generate multivariate missingness patterns patterns
-create_patterns <- function() {
-  mis_pat <-
-    expand.grid(c(0, 1), c(0, 1), c(0, 1), c(0, 1), c(0, 1)) %>%
-    setNames(., c("Y", "X1", "X2", "X3", "X4")) %>%
-    .[rowSums(.) > 1 & rowSums(.) < 5, ]
+create_patterns <- function(n_col) {
+  pat_list <- vector(mode = "list", length = n_col + 1)
+  pat_list <- lapply(pat_list, \(x) {x <- c(0, 1)})
+  mis_pat <- expand.grid(pat_list) 
+  names(mis_pat) <- c("Y", paste0("X", 1:n_col))
+  # omit patterns with all/none missing
+  mis_pat <- mis_pat[rowSums(mis_pat) > 1 & rowSums(mis_pat) < n_col + 1, ]
   return(mis_pat)
 }
 
@@ -45,7 +47,7 @@ induce_missingness <- function(
       # ampute the data
       mice::ampute(dat, patterns = mis_pat, mech = mm, prop = mp) 
     }) 
-  }) %>% 
+  }) |> 
     # unlist just make sure the proportions 
     # are not nested within the meachnisms
     unlist(recursive = FALSE)

### 2025-04-07 simulation.R

@@ -17,8 +17,10 @@ set.seed(11)
 # parameters
 n_sim <- 10
 n_obs <- 200
-corr <- 0.5
-betas <- c(1, 1, -1, -1)
+n_col <- 4
+corr <- 0.3
+beta <- 0.6
+betas <- rep(beta, n_col)
 mis_pat <- create_patterns()
 mis_mech = c("MCAR", "MAR")
 mis_prop = c(0.25, 0.5, 0.75)
@@ -79,6 +81,40 @@ results_raw <- replicate(
 # # save raw results
 # saveRDS(results_raw, "./Results/raw.RDS")
 
+####################
+### RUN PARALLEL ###
+####################
+
+library(pbapply)
+
+cl <- parallel::makeCluster(4)
+parallel::clusterExport(
+  cl,
+  c("generate_complete",
+    "induce_missingness",
+    "create_data",
+    "apply_methods",
+    "apply_CCA",
+    "apply_MICE",
+    "evaluate_est",
+    "estimate_param",
+    "create_patterns",
+    "n_obs", "betas", "mis_pat", "mis_mech", "mis_prop",
+    "simulate_once",
+    "add_iteration"
+    
+    
+  )
+)
+
+out <- pbreplicate(n_sim, 
+                   simulate_once(n_obs, betas, mis_pat, mis_mech, mis_prop),
+                   cl = cl)
+
+parallel::stopCluster(cl)
+
+save(out, file = "results/out.RData")
+
 ########################
 ### EVALUATE RESULTS ###
 ########################
@@ -102,4 +138,28 @@ performance %>%
   group_by(method, mech, prop, .it, term) %>% 
   summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), mean, na.rm = TRUE))
 
+# plot results
+performance |>
+  filter(term == "X1", mech == "MCAR") |>
+  ggplot(aes(x = .it, y = bias)) +
+  geom_smooth() +
+  geom_point() +
+  labs(
+    title = "Bias of regression coefficient estimates",
+    x = "Iteration number",
+    y = "Bias"
+  ) +
+  theme_classic()
 
+# plot results
+performance |>
+  filter(term == "X1", mech == "MCAR") |>
+  ggplot(aes(x = .it, y = psrf_sd)) +
+  geom_smooth() +
+  geom_point() +
+  labs(
+    title = "Bias of regression coefficient estimates",
[…]

### 2025-04-08 Functions/apply_methods.R
run parallel

@@ -8,7 +8,7 @@ apply_CCA <- function(amp) {
     # clean results
     broom::tidy(conf.int = TRUE) |> 
     # choose estimates
-    select(term, estimate, conf.low, conf.high)
+    dplyr::select(term, estimate, conf.low, conf.high)
   # add method name and missingness
   est <- cbind(method = "CCA", mech = amp$mech, prop = amp$prop, .it = 0, est, ac_mean = NA, psrf_mean = NA, ac_sd = NA, psrf_sd = NA)
   # rename "(Intercept)" to "Y" for easier processing
@@ -56,7 +56,7 @@ estimate_param <- function(imp) {
     # clean results
     broom::tidy(conf.int = TRUE) |> 
     # select estimates
-    select(term, estimate, conf.low, conf.high)
+    dplyr::select(term, estimate, conf.low, conf.high)
   # rename "(Intercept)" to "Y" for easier processing
   est[est$term == "(Intercept)", "term"] <- "Y"
   # output

### 2025-04-08 Functions/evaluate_estimates.R

@@ -3,7 +3,7 @@
 
 evaluate_est <- function(results) {
   performance <- purrr::map_dfr(results, ~{
-    mutate(.x,
+    dplyr::mutate(.x,
       bias = truth - estimate,
       cov = conf.low <= truth & conf.high >= truth,
       ciw = conf.high - conf.low,

### 2025-04-08 simulation.R

@@ -21,7 +21,7 @@ n_col <- 4
 corr <- 0.3
 beta <- 0.6
 betas <- rep(beta, n_col)
-mis_pat <- create_patterns()
+mis_pat <- create_patterns(n_col)
 mis_mech = c("MCAR", "MAR")
 mis_prop = c(0.25, 0.5, 0.75)
 n_it <- 10
@@ -72,12 +72,12 @@ simulate_once <- function(n_obs, betas, mis_pat, mis_mech, mis_prop) {
 ### RUN SIMULATION ###
 ######################
 
-# repeat the simulation function n_sim times
-results_raw <- replicate(
-  n_sim, 
-  simulate_once(n_obs, betas, mis_pat, mis_mech, mis_prop),
-  simplify = FALSE
-  )
+# # repeat the simulation function n_sim times
+# results_raw <- replicate(
+#   n_sim, 
+#   simulate_once(n_obs, betas, mis_pat, mis_mech, mis_prop),
+#   simplify = FALSE
+#   )
 # # save raw results
 # saveRDS(results_raw, "./Results/raw.RDS")
 
@@ -101,41 +101,43 @@ parallel::clusterExport(
     "create_patterns",
     "n_obs", "betas", "mis_pat", "mis_mech", "mis_prop",
     "simulate_once",
-    "add_iteration"
-    
+    "add_iteration",
+    "n_it", "n_sim", "n_col", "corr", "beta", "betas", "mis_pat", "mis_mech", "mis_prop"
     
   )
 )
 
 out <- pbreplicate(n_sim, 
                    simulate_once(n_obs, betas, mis_pat, mis_mech, mis_prop),
-                   cl = cl)
+                   cl = cl, 
+                   simplify = FALSE)
 
 parallel::stopCluster(cl)
 
-save(out, file = "results/out.RData")
+save(out, file = "out.RData")
 
 ########################
 ### EVALUATE RESULTS ###
 ########################
 
 # calculate bias, coverage rate and CI width
-performance <- evaluate_est(results_raw)
-# saveRDS(performance, "./Results/performance.RDS")
+# performance <- evaluate_est(results_raw)
+# # saveRDS(performance, "./Results/performance.RDS")
+performance <- evaluate_est(out)
 
 # simulation results across all conditions
-performance %>% 
-  group_by(method, .it) %>% 
+performance |> 
+  group_by(method, .it) |> 
[…]

### 2025-04-08 Manuscript/manuscript.qmd

@@
 ```
+
+## Discussion
+
+The potential scale reduction factor metric assumes over-dispersed starting values of the iterative algorithm.
+The MICE algorithm does not start in an over-dispersed state.
+MICE does not rely on staring values for parameters at all: instead, the algorithm is initialized based on a 'zeroth' iteration, where missing values are filled in using a random draw from observed values.
+Under MCAR, this would typically not yield an over-dispersed state at all, depending on the parameter of interest: means and variances are not affected, regression coefficients and other multivariate parameters start biased downwards (because sampling starting values distorts multivariate structures in the data).
+Whether this accrues to over-dispersion is questionable.
+Under MAR (or MNAR), some over-dispersion might occur when the observed data and imputed data differ impactfully.
+The algorithm should 'escape' the initial state (based on the staring values drawn from observed data alone).
+Over iterations, all parameters of interest (e.g.
+means, variances, multivariate estimands) should converge towards a stable state.
+Moreover, the MICE algorithm is initialized with more information than a typical MCMC algorithm: parameter estimates depend not only on imputed data, but also on observed data that does not change over iterations.
+Therefore, the initial state of the algorithm is 'too good' to observe a relative decrease in $\widehat{R}$.

### 2025-04-10 Manuscript/manuscript.qmd
add complete data to sim

@@
 
-```{r setup, echo=FALSE}
+```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
 library(mice)
@@
 
-All inferences rely on the convergence of the imputation algorithm, yet determining whether an algorithm has converged is not trivial.
+All inferences with imputed data rely on the convergence of the imputation algorithm.
+Yet, determining whether an algorithm has converged is not trivial.
 There has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms.
-A widely accepted practice is visual inspection of the algorithm, although diagnosing convergence through visual inspection may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18].
+A widely accepted practice is visual inspection of the algorithm.
+Diagnosing convergence through visual inspection, however, may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18].
 Therefore, a quantitative diagnostic method to assess convergence would be preferred.
@@
 
-We use empirical incomplete data: the `boys` dataset from the `mice` package, which contains data on some health-related variables for 748 Dutch boys [@mice].
-Say we're interested in the relation between children's heights and their respective ages, we could use a linear regression model to predict `hgt` from `age`.
+We use empirical incomplete data: the `boys` dataset from the `mice` package, which contains health-related data for 748 Dutch boys [@mice].
+Say, we're interested in the relation between children's heights and their respective ages, we could use a linear regression model to predict `hgt` from `age`.
 However, as figure XYZ shows, the variable `hgt` is not completely observed.
@@
 ```{r boys, echo=FALSE}
-ggmice::ggmice(mice::boys, ggplot2::aes(age, hgt)) + ggplot2::geom_point()
+ggmice::ggmice(mice::boys, ggplot2::aes(age, hgt, shape = .where)) + 
+  ggplot2::geom_point() + 
+  ggplot2::scale_shape_manual(values = c(1, 4)) +
+  ggplot2::labs(x = "Age (years)", y = "Height (cm)", shape = "")
 ```
@@
 imp <- mice::mice(mice::boys[, c("age", "hgt", "wgt")], maxit = 20, printFlag = FALSE)
-plot_trace(imp)
+plot_trace(imp, "hgt") 
 ```
 
--   Conventional inspection of algorithmic
+```{r}
+# impute missingness with mis-specified imputation model
+meth <- make.method(boys) #define imputation model
+meth["bmi"] <- "~I(wgt / (hgt / 100)^2)" #mis-specify model
+nonconv <- mice( #impute missingness
+  boys,
+  meth = meth,
+  maxit = 20,
+  print = FALSE,
+  seed = 60109
+)
 
--   We are interested in predicting age from the other variables, in particular in the regression coefficient of `hgt`
+# # save results
+# save(nonconv, file = "Results/example_nonconv.Rdata")
 
--   We compare non-convergence identified using visual inspection versus rhat in the chain variances, scientific estimate and lambda.
+#####################
+# Regular convergence
+#####################
 
--   The figures show results of a `mice` run with 20 iterations but otherwise default settings.
+# impute missingness with correctly specified imputation model
+pred <- make.predictorMatrix(boys) #mend imputation method
+pred[c("hgt", "wgt"), "bmi"] <- 0 #remove dependency
+conv <- mice( #impute missingness
+  boys,
+  meth = meth,
+  pred = pred,
+  maxit = 20,
+  print = FALSE,
+  seed = 60109
+)
 
-```{r case, fig.cap = "Case study.", echo=FALSE, eval=FALSE}
-trace <- param %>% 
-  ggplot() + 
-  geom_line(aes(x = it, y = mu.hgt, color = as.factor(m))) + 
-  theme_classic() + 
-  labs(x = "Iteration number",
-       y = "Chain mean variable 'hgt'",
-       color = "Imputation")
-means <- casestudy_plot(v="r.hat.max.mu.hgt", name = "Rhat chain means") + 
-  list(geom_hline(yintercept = 1.2, color = "grey"))
-qhats <-
-  casestudy_plot(v = "r.hat.max.qhat", name = "Rhat regression estimate") +
-  list(geom_hline(yintercept = 1.2, color = "grey"))
-lambdas <- casestudy_plot(v = "r.hat.max.lambda", name = "Rhat lambda") +
-  list(geom_hline(yintercept = 1.2, color = "grey"))
-(trace + means) / (qhats + lambdas) + plot_layout(guides="collect") + plot_annotation(tag_levels = 'A')
-
+plot_trace(nonconv, "hgt") +
+  ggplot2::labs(title = "Pathological non-convergence")
+plot_trace(conv, "hgt") +
+  ggplot2::labs(title = "Typical convergence")
 ```
@@
 
-This figure (\@ref(fig:case)B) shows that 7 iterations are required before the $\widehat{R}$-values of the chain means drop below the threshold for non-convergence.
+## Aims
 
-The $\widehat{R}$-values for the scientific estimate reaches the threshold much sooner, when $n_{it}=14$ (see \@ref(fig:case)C).
-
-According to the $\widehat{R}$-values with $\lambda$ as parameter, at least 15 iterations are required (see \@ref(fig:case)D).
-
-```{r}
-
-```
+- can we quantify non-convergence quantitatively?
+- which parameter should we track?
+- how many iterations are needed before convergence is reached?
+- how many iterations are needed before valid inferences are reached?
 
@@
 The MICE algorithm does not start in an over-dispersed state.
-MICE does not rely on staring values for parameters at all: instead, the algorithm is initialized based on a 'zeroth' iteration, where missing values are filled in using a random draw from observed values.
+MICE does not rely on starting values for parameters at all: instead, the algorithm is initialized based on a 'zeroth' iteration, where missing values are filled in using a random draw from observed values.
 Under MCAR, this would typically not yield an over-dispersed state at all, depending on the parameter of interest: means and variances are not affected, regression coefficients and other multivariate parameters start biased downwards (because sampling starting values distorts multivariate structures in the data).
@@
 Under MAR (or MNAR), some over-dispersion might occur when the observed data and imputed data differ impactfully.
-The algorithm should 'escape' the initial state (based on the staring values drawn from observed data alone).
+The algorithm should 'escape' the initial state (based on the starting values drawn from observed data alone).
 Over iterations, all parameters of interest (e.g.

### 2025-04-10 Manuscript/template.Rmd
add complete data to sim

--- 2022-05-02 Manuscript.Rmd
@@
 knitr::opts_chunk$set(echo = TRUE)
-library(dplyr)
-library(purrr)
-library(mice)
-miceadds::source.all("R")
+library(tidyverse)
+
+# filters
+coeff <- "X1"
+maxit <- 20
+```
+
+```{r data}
+performance <- readRDS("./Results/performance.RDS")
+summary <- performance %>% 
+  group_by(method, mech, prop, .it, term) %>% 
+  summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), mean, na.rm = TRUE))
+SEs <- performance %>% 
+  group_by(method, mech, prop, .it, term) %>% 
+  summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), var, na.rm = TRUE)) %>% 
+  mutate(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), sqrt))
+results <- left_join(summary, SEs, by = c("method", "mech", "prop", ".it", "term"), suffix = c("", "_SE")) %>% 
+  mutate(cov = cov,
+         cov_SE = sqrt((cov * (1 - cov)) / 2000))
 ```
@@
 Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+```{r}
+# results %>% 
+#   filter(term == "X2", .it < 26) %>% 
+#   ggplot(aes(x = .it, y = bias)) +
+#   geom_hline(yintercept = 0, color = "grey", size = 2) +
+#   geom_line(color = "blue") +
+#   geom_line(aes(x = .it, y = bias + bias_SE), color = "lightblue", linetype = "dashed") +
+#   geom_line(aes(x = .it, y = bias - bias_SE), color = "lightblue", linetype = "dashed") +
+#   facet_grid(as.factor(prop) ~ mech) +
+#   theme_classic()
+
+results %>% 
+  filter(term == coeff, .it > 0, .it < maxit + 1) %>% 
+  ggplot(aes(x = .it, y = bias)) +
+  geom_hline(yintercept = 0, color = "black") +
+  geom_errorbar(aes(ymin = bias - bias_SE, ymax = bias + bias_SE), alpha = 0.1) +
+  geom_line(color = "blue") +
+  geom_point(color = "blue") +
+  facet_grid(as.factor(prop) ~ mech) +
+  theme_classic()
+
+results %>% 
+  filter(term == coeff, .it > 0, .it < maxit + 1) %>% 
+  ggplot(aes(x = .it, y = ac_mean)) +
+  geom_hline(yintercept = 0, color = "black") +
+  geom_errorbar(aes(ymin = ac_mean - ac_mean_SE, ymax = ac_mean + ac_mean_SE), alpha = 0.1) +
+  geom_line(color = "blue") +
+  geom_point(color = "blue") +
+  facet_grid(as.factor(prop) ~ mech) +
+  theme_classic()
+
+results %>% 
+  filter(term == coeff, .it > 0, .it < maxit + 1) %>% 
+  ggplot(aes(x = .it, y = psrf_mean)) +
+  geom_hline(yintercept = 1, color = "black") +
+  geom_errorbar(aes(ymin = psrf_mean - psrf_mean_SE, ymax = psrf_mean + psrf_mean_SE), alpha = 0.1) +
+  geom_line(color = "blue") +
+  geom_point(color = "blue") +
+  facet_grid(as.factor(prop) ~ mech) +
+  theme_classic()
+
+results %>% 
+  filter(term == coeff, .it > 0, .it < maxit + 1) %>% 
+  ggplot(aes(x = .it, y = ac_sd)) +
+  geom_hline(yintercept = 0, color = "black") +
+  geom_errorbar(aes(ymin = ac_sd - ac_sd_SE, ymax = ac_sd + ac_sd_SE), alpha = 0.1) +
+  geom_line(color = "blue") +
+  geom_point(color = "blue") +
+  facet_grid(as.factor(prop) ~ mech) +
+  theme_classic()
+
+results %>% 
+  filter(term == coeff, .it > 0, .it < maxit + 1) %>% 
+  ggplot(aes(x = .it, y = psrf_sd)) +
+  geom_hline(yintercept = 1, color = "black") +
+  geom_errorbar(aes(ymin = psrf_sd - psrf_sd_SE, ymax = psrf_sd + psrf_sd_SE), alpha = 0.1) +
+  geom_line(color = "blue") +
+  geom_point(color = "blue") +
+  facet_grid(as.factor(prop) ~ mech) +
+  theme_classic()
+```
 

### 2025-04-10 Functions/apply_methods.R

@@ -1,5 +1,22 @@
 # functions to apply missing data methods
 
+# comparative truth, fully observed data
+apply_full <- function(amp) {
+  # list-wise deletion
+  est <- # fit regression 
+    lm(Y ~ X1 + X2 + X3 + X4, data = amp$data) |> 
+    # clean results
+    broom::tidy(conf.int = TRUE) |> 
+    # choose estimates
+    dplyr::select(term, estimate, conf.low, conf.high)
+  # add method name and missingness
+  est <- cbind(method = "full", mech = amp$mech, prop = amp$prop, .it = 0, est, ac_mean = NA, psrf_mean = NA, ac_sd = NA, psrf_sd = NA)
+  # rename "(Intercept)" to "Y" for easier processing
+  est[est$term == "(Intercept)", "term"] <- "Y"
+  # output
+  return(est)
+}
+
 # complete case analysis
 apply_CCA <- function(amp) {
   # list-wise deletion
@@ -75,12 +92,14 @@ add_iteration <- function(implist) {
 
 # combine into one function
 apply_methods <- function(amps, betas, n_it) {
+  # comparative truth on complete dataset
+  full <- purrr::map_dfr(amps, ~{apply_full(.x)})
   # apply CCA to each incomplete dataset
   CCA <- purrr::map_dfr(amps, ~{apply_CCA(.x)})
   # impute with MICE and estimate effects
   MICE <- purrr::map_dfr(amps, ~{apply_MICE(.x, n_it)})
   # combine estimates 
-  ests <- rbind(CCA, MICE) |> 
+  ests <- rbind(full, CCA, MICE) |> 
     cbind(truth = c(0, betas))
   row.names(ests) <- NULL
   # output
@@ -103,4 +122,27 @@ apply_methods <- function(amps, betas, n_it) {
 # collect_theta <- function(imp) {
 #  a <-  mice::complete(imp, "all")
 #  purrr::map_dbl(a, ~{princomp(., cor = TRUE) %>% .$sdev %>% .[1] %>% . ^ 2})
+# }
+
+# multivar_metrics <- function(implist){
+# # compute AC
+# ac <-
+#   purrr::map(vrbs, function(.vrb) {
+#     c(NA, dplyr::cummean(dplyr::coalesce(
+#       purrr::map_dbl(2:t, function(.itr) {
+#         suppressWarnings(stats::cor(
[…]

### 2025-04-10 simulation.R

@@ -15,7 +15,7 @@ miceadds::source.all("./functions")
 set.seed(11)
 
 # parameters
-n_sim <- 10
+n_sim <- 100
 n_obs <- 200
 n_col <- 4
 corr <- 0.3
@@ -66,7 +66,7 @@ simulate_once <- function(n_obs, betas, mis_pat, mis_mech, mis_prop) {
 # ### TEST SIMULATION FUNCTION ###
 # ################################
 # 
-# ests <- simulate_once(n_obs, betas, mis_mech, mis_prop)
+# ests <- simulate_once(n_obs, betas, mis_pat, mis_mech, mis_prop)
 
 ######################
 ### RUN SIMULATION ###
@@ -94,6 +94,7 @@ parallel::clusterExport(
     "induce_missingness",
     "create_data",
     "apply_methods",
+    "apply_full",
     "apply_CCA",
     "apply_MICE",
     "evaluate_est",

### 2025-04-10 Manuscript/manuscript.qmd
clean up repo

@@
 bibliography: references.bib
+format: pdf
 ---

### 2025-04-10 simulation.R

@@ -9,13 +9,13 @@ library(mice)
 library(miceadds)
 
 # functions
-miceadds::source.all("./functions")
+miceadds::source.all("./R")
 
 # randomness
 set.seed(11)
 
 # parameters
-n_sim <- 100
+n_sim <- 2
 n_obs <- 200
 n_col <- 4
 corr <- 0.3
@@ -24,7 +24,7 @@ betas <- rep(beta, n_col)
 mis_pat <- create_patterns(n_col)
 mis_mech = c("MCAR", "MAR")
 mis_prop = c(0.25, 0.5, 0.75)
-n_it <- 10
+n_it <- 50
 
 # #################################
 # ### TEST LOWER LEVEL FUCTIONS ###
@@ -115,7 +115,7 @@ out <- pbreplicate(n_sim,
 
 parallel::stopCluster(cl)
 
-save(out, file = "out.RData")
+save(out, file = "results/out.RData")
 
 ########################
 ### EVALUATE RESULTS ###

### 2025-04-10 Manuscript/manuscript.qmd
add contents thesis

@@
 
-# Introduction
+# Abstract (old)
+
+# Introduction (new-ish)
 
@@
 
-```{r}
+A mis-specified imputation model can lead to non-convergence in the imputation algorithm.
+
+```{r, echo=FALSE}
 # impute missingness with mis-specified imputation model
@@
 
-plot_trace(nonconv, "hgt") +
-  ggplot2::labs(title = "Pathological non-convergence")
-plot_trace(conv, "hgt") +
-  ggplot2::labs(title = "Typical convergence")
+plot_trace(nonconv, "hgt") 
+# plot_trace(conv, "hgt") +
+#   ggplot2::labs(title = "Typical convergence")
 ```
-
-From the traceplot of the chain means (see \@ref(fig:case)A) it seems that mixing improves up-to 10 iterations, while trending is only apparent in the first three iterations.
 
@@
 
-## Discussion
+# Introduction (old)
+
+Anyone who analyzes person-data may run into a missing data problem.
+Missing data is not only ubiquitous across science, but treating it can also be a tedious task.
+If a dataset contains just one incomplete observation, calculations are not defined, and statistical models cannot be fitted to the data.
+To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations).
+Unfortunately, this *ad hoc* solution may yield invalid results [@buur18].
+An alternative is to apply imputation.
+With imputation, we 'fill in' the missing values in an incomplete dataset.
+Subsequently, the model of scientific interest can be fitted to the completed dataset.
+By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness.
+This technique is known as 'multiple imputation' [MI; @rubin76].
+MI has proven to be a powerful tool to draw valid inferences from incomplete data under many circumstances [@buur18].
+
+Figure \ref{fig:diagram} provides an overview of the steps involved with MI.
+The missing part $y_{\mathrm mis}$ of an incomplete dataset is imputed $m$ times.
+This creates $m$ sets of imputed data $\dot y_{\mathrm imp, \ell}$, where $\ell = 1, 2, ..., m$.
+The imputed data is then combined with the observed data $y_{\mathrm obs}$ to create $m$ completed datasets.
+On each of these datasets the analysis of scientific interest is performed to estimate $Q$: the quantity of scientific interest (e.g., a regression coefficient).
+Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}_\ell$-values are obtained.
+Finally, the $\hat{Q}_\ell$-values are combined into a single pooled estimate $\bar{Q}$.
+The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence-valid estimate of the true---but unobserved---scientific estimand $Q$ [@rubin96].
+<!-- "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->
+
+```{r diagram, echo=FALSE, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\\hat{Q}$, to a single pooled estimate $\\bar{Q}$."}
+
+knitr::include_graphics("../figures/diagram.pdf") #adapted from [@buur18] section 1.4.1.
+```
+
+A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice].
+With MICE, imputed values are drawn from the posterior predictive distribution of the missing values.
+The algorithm is named after its iterative nature: a multivariate distribution is obtained by iterating over a sequence of univariate imputations.
+Iteration, however, also introduces a potential threat to the validity of the imputations: What if the algorithm has not converged?
+Are the imputations then to be trusted?
+And can we rely on the inference obtained on the completed data?
+
+These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18].
+There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@zhu15; @takahashi2017].
+Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murray2018].
+Therefore, algorithmic convergence should be monitored carefully---although this is not straightforward.
+Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods.
+In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another.
+The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13].
+Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09].
+Several non-convergence diagnostics exist, but it is not known whether these are appropriate within the imputation framework.
+
+<!-- https://www-tandfonline-com.proxy.library.uu.nl/doi/pdf/10.1207/s15327906mbr3304_5?needAccess=true -->
+
+<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->
+
+In this paper we study different methods for assessing non-convergence in iterative imputation algorithms.
+We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications.
+We then address the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in `R` [@R].
+For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package in `R` [@mice].
+The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged.
+And additionally, to evaluate the behavior and performance of several diagnostic methods to identify non-convergence.
+With that, we formulate an informed advice on when it is safe to conclude that the algorithm is converged *enough* for valid inferences.
+We translate the results of the study into guidelines for applied researchers, which may facilitate drawing valid inferences from incomplete data.
+
+<!-- The default number of iterations in common MI software: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100??
+[(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p.
+139), Amelia = NA, because re-sampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf).
+-->
+
+## Some notation
+
+Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample.
+The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing.
+The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\mathrm inc}$ in dataset $y$.
+The collection of observed data values in $y$ is denoted by $y_{\mathrm obs}$; the missing part of $y$ is referred to as $y_{\mathrm mis}$.
+For each datapoint in $y_{\mathrm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations in the imputation algorithm ($t = 1, 2, \dots, T$).
+The state-space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values).
+The collection of $\theta$-values between $t=1$ (summarizing the state-space of the algorithm at initialization) and $t=T$ (summarizing the state-space for the imputed values) will be referred to as an 'imputation chain'.
+
+# Algorithmic non-convergence
+
+There are two requirements for convergence of iterative algorithms: mixing and stationarity [@gelm13].
+In iterative imputation algorithms, mixing implies that imputation chains intermingle nicely, and stationarity is characterized by the absence of trending across iterations.
+If one of the two requirements is not met, we speak of non-convergence.
+Without mixing, chains may be 'stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values.
+The distribution of imputed values then differs across imputations.
+This may cause under-estimation of the variance between chains, which results in spurious, invalid inferences.
+Without stationarity, there is trending within imputation chains.
+Trending implies that further iterations would yield a systematically lower or higher set of imputations.
+Iterative imputation algorithms that have not (yet) reached stationarity, may thus yield biased estimates.
+
+To illustrate what non-mixing and non-stationarity look like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2].
+Figure \ref{fig:non-conv} displays two scenarios from the example.
+The panel on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm.
+The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
+Each line portrays one imputation chain (i.e., the values of a scalar summary $\theta$ across iterations).
+The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in the $m$ sets of imputations $\dot y_{\mathrm imp, \ell}$.
+
+In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending.
+In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.
+Importantly, the chain means at the last iteration (the imputed values per imputation $\ell$) are very different between the two scenarios.
+The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
+It is obvious that non-convergence in this example impacts the distribution of the imputed values per imputation $\dot y_{\mathrm imp, \ell}$.
+This effect may translate into the distribution of the $m$ sets of completed data {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$, and finally the pooled estimate $\bar{Q}$.
+$\bar{Q}$ is then a  biased, invalid estimate of $Q$.
+Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.
+
+```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence (reproduced from van Buuren; 2018).
+Depicted are chain means of variable $j$ in each imputation ($m=5$; $T=10$).
+Note that any difference between the two scenarios (see e.g., the magnitude of the chain means on the y-axes) stems from the imputation algorithms---the two scenarios have equal incomplete data $y$."}
+
+# par(mar = c(2, 2, 0.1, 0.1))
+# con_plot
+# non_plot
+
+```
+
+## Identifying non-convergence
+
+Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection.
+After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots.
+In a traceplot, a scalar summary of the state-space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}.
+The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances.
+Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18, $\S$ 6.5.2].
+<!-- Recommendation @buur18: "the different streams should be freely intermingled with one another, without showing any definite trends". -->
+
+Identifying non-convergence by inspecting traceplots may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18, $\S$ 6.5.2].
+Moreover, traceplots are typically only used to inspect  univariate $\theta$s.
+Monitoring univariate summaries of the state-space of the algorithm may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of the imputations.
+Ideally, we would monitor a multivariate $\theta$.
+
+A suggestion by @buur18 for a multivariate $\theta$ to monitor is the estimated quantity of scientific interest $\hat{Q}$, which usually is multivariate in nature.
+This $\theta$ is computed as the estimate in each imputation, $\hat{Q}_\ell$, across iterations.
+Implementing this, however, might be somewhat too technical for empirical researchers.
+Besides, this scalar summary is not model-independent.
+That is, $\theta=\hat{Q}$ is not universal to all complete data problems, while one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently.
+Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest.
+On these grounds, such a $\theta$ can be considered insufficient too.
+<!-- Researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2].  -->
+
+As alternative, van Buuren [-@buur18, $\S$ 4.5.2] proposed multivariate evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of @mack03.
+This technique may yield a model-independent, multivariate scalar summary to monitor, but it is not implemented in the  `mice` package [@mice].
+
+## A novel scalar summary
+
+To monitor non-convergence in iterative imputation algorithms, we may  summarize the state-space of the algorithm with several $\theta$s.
+The downside to the current $\theta$s is that they either focus on the univariate state-space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
+Ideally, one would like to evaluate a model-independent $\theta$ that summarizes the multivariate nature of the data.
+
+We propose $\lambda_{1}$ as such a scalar summary.
+We define $\lambda_{1}$ as the first eigenvalue of the variance-covariance matrix in the $m$ completed datasets.
+The eigenvalues of a variance-covariance matrix $S$ summarize the total set of covariances in the data.
+Let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ...
+\geq \lambda_{j, \ell}$ be the eigenvalues of $S_\ell$ in each imputation.
+The first eigenvalue, $\lambda_{1,\ell}$, then summarizes the largest possible amount of covariance in each completed dataset {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$}.
+By definition, $\lambda_{1}$-values are equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on the completed data.
+As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the model of scientific interest, yet still summarizes the multivariate state-space of the algorithm.
+
+In this study, we consider each of the four $\theta$s that we discussed.
+Namely, two univariate scalar summaries (chain mean and chain variance), a model-dependent multivariate summary $\hat{Q}$, and the novel model-independent multivariate summary $\lambda_1$.
+
+# Non-convergence diagnostics
+
+There are many diagnostic tools to identify non-convergence in iterative (MCMC) algorithms [@broo98; @elad06].
+We consider only two of them that may be appropriate for imputation algorithms---one to monitor signs of non-mixing, and one for non-stationarity.
+As recommended by e.g.
+@cowl96 we will use the potential scale reduction factor $\widehat{R}$ to evaluate mixing ['Gelman-Rubin statistic'; @gelm92], and autocorrelation to diagnose trending [$AC$; @scha97; @gelm13].
+With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@vehtari2021].
+Therefore, we will evaluate the appropriateness of both $\widehat{R}$ versions, in addition to $AC$.
+Other methods are outside the scope of this study because they e.g.
+assume that values within chains represent independent samples, whereas the MICE algorithm only uses the final iteration to produce imputations.
+
+<!-- Non-stationarity within chains may be diagnosed with e.g., autocorrelation [$AC$; @scha97; @gelm13], numeric standard error ['MC error'; @gewe92], or Raftery and Lewis's [-@raft91] procedure to determine the effect of trending on the precision of estimates.
+-->
+
+<!-- "We can use effective sample size ^neff to give us a sense of the precision obtained from our simulations.
+As we have discussed in Section 10.5, for many purposes it should suffice to have 100 or even 10 independent simulation draws.
+(If neff = 10, the simulation standard error is increased by p 1 + 1/10 = 1.05).
+As a default rule, we suggest running the simulation until ^neff is at least 5m, that is, until there are the equivalent of at least 10 independent draws per sequence (recall that m is twice the number of sequences, as we have split each sequence into two parts so that Rb can assess stationarity as well as mixing).
+Having an effective sample size of 10 per sequence should typically correspond to stability of all the simulated sequences."  @gelm13, p.
+287 -->
+
+<!-- We don't use effective sample size and MC error because they assume independent samples *within* chains.
+In MI, we only use the final estimate.
+Also, computing n_eff assumes infinitely many iterations, which is fine in Bayesian analyses where T is often at least one thousand.
+But the default in iterative imp is often substantially lower.
+-->
+
+## Potential scale reduction factor
+
+In 2019, Vehtari et al.
+proposed an updated version of the potential scale reduction factor $\widehat{R}$, originally coined in 1992.
+The adapted version would be better suited to detect non-mixing in the tails of distributions, and even identify non-stationarity.
+This version uses three transformations on the scalar summary $\theta$, before computing $\widehat{R}$-values.
+Namely, rank-normalization, folding, and localization.
+The definition of the diagnostic itself is equal to the original $\widehat{R}$.
+Therefore, we may follow Vehtari et al.'s formulation (2019, p.
+5) to define both $\widehat{R}$ versions.
+Let $m$ be the total number of chains, $T$ the number of iterations per chain (where $T\geq2$), and $\theta$ the scalar summary of interest.
+For each chain ($\ell = 1, 2, \dots, m$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.
+
+\begin{align*}
+W&=\frac{1}{m} \sum_{\ell=1}^{m} s_{\ell}^{2}, \text { where } s_{\ell}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t \ell)}-\bar{\theta}^{(\cdot \ell)}\right)^{2}. 
+\end{align*}
+
+\noindent We then estimate between-chain variance $B$ (note that we diverge from the typical notation in MI, where $B$ denotes the variance between the estimated quantities of scientific interest $\hat{Q}_{\ell}$).
+$B$ is defined as the variance of the collection of average $\theta$s per chain:
+
+\begin{align*}
+B&=\frac{T}{m-1} \sum_{\ell=1}^{m}\left(\bar{\theta}^{(\cdot \ell)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot \ell)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t \ell)} \text{, } \bar{\theta}^{(\cdot \cdot)}=\frac{1}{m} \sum_{\ell=1}^{m} \bar{\theta}^{(\cdot \ell)}.
+\end{align*}
+
+\noindent From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$.
+$\widehat{R}$ is then obtained as a ratio between this total variance and the within-chain variance:
+
+\begin{equation*}
+\widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
+\text{ where } \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{T-1}{T} W+\frac{1}{T} B.
+\end{equation*}
+
+We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run [@gelm92].
+The assumption underlying this interpretation is that chains are 'over-dispersed' at $t=1$, and reach convergence as $T \to \infty$.
+Over-dispersion implies that the initial values of the chains are 'far away' from the target distribution and each other.
+When the sampled values in each chain are independent of the chain's initial value, the mixing component of convergence is satisfied.
+The variance between chains, $B$, is then equivalent to the variance within chains, $W$, and $\widehat{R}$-values will be close to one.
+High $\widehat{R}$-values thus indicate non-convergence.
+
+## Autocorrelation
+
+Autocorrelation is defined as the correlation between two subsequent $\theta$-values within the same chain [@lync07, p.
+147].
+In this study, we only consider $AC$ at lag 1, i.e., the correlation between the $t^{th}$ and $(t+1)^{th}$ iteration of the same chain.
+Following the same notation as for $\widehat{R}$,
+
+\begin{equation*}
+AC = \left( \frac{T}{T-1} \right) \frac{\sum_{t=1}^{T-1}(\theta_t - \bar{\theta}^{(\cdot m)})(\theta_{t+1} - \bar{\theta}^{(\cdot m)})}{\sum_{t=1}^{T}(\theta_t - \bar{\theta}^{(\cdot m)})^2}.
+\end{equation*}
+
+We can interpret $AC$-values as a measure of non-stationarity.
+If there is dependence between subsequent $\theta$-values in imputation chains, $AC$-values are non-zero.
+Positive $AC$-values occur when $\theta$-values are recurring (i.e., high $\theta$-values are followed by high $\theta$-values, and low $\theta$-values are followed by low $\theta$-values).
+Recurrence within imputation chains may lead to trending.
+Negative $AC$-values occur when $\theta$-values of subsequent iterations are less similar, or diverge from one another.
+Divergence within imputation poses no threat to the convergence of the algorithm---it may even speed up convergence.
+Complete stationarity is reached when $AC=0$.
+As non-convergence diagnostic, our interest is in positive $AC$-values.
+
+## Thresholds
+
+It is unlikely that iterative algorithms such as MICE will achieve the ideal values of $\widehat{R}=1$ and $AC=0$.
+Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state.
+The aim, therefore, is to reach approximate convergence [@gelm13].
+Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} > 1$), and there is little dependency between subsequent iterations of imputation chains ($AC > 0$).
+In practice, we diagnose non-convergence when approximate convergence is violated, i.e., when $\widehat{R}$ and $AC$ exceed a certain threshold.
+The conventional thresholds to diagnose non-mixing are $\widehat{R} > 1.2$ [@gelm92] or $\widehat{R} > 1.1$ [@gelm13].
+@vehtari2021 proposed a much more stringent threshold of $\widehat{R} > 1.01$.
+The magnitude of $AC$-values may be evaluated statistically, using a Wald test with $AC=0$ as null hypothesis [@box15].
+$AC$-values that are significantly higher than zero indicate non-stationarity.
+<!-- Threshold Rhat > 1.1 @gelm13, p. 288 -->
+
+## In practice
+
+Before we evaluate the performance of the non-convergence diagnostics $\widehat{R}$ and $AC$ quantitatively through simulation, we first assess their appropriateness qualitatively.
+We do this by applying them to the example of pathological non-convergence that we reproduced form @buur18.
+Ideally, the diagnostics are as informative as, or better than visual inspection of the traceplots.
+The methods should at least indicate worse performance (higher $\widehat{R}$- and $AC$-values) for the scenario with pathological non-convergence, compared to the typically converged algorithm.
+
+Each panel in Figure \ref{fig:diagnostics} depicts one method to identify non-convergence, applied to the two scenarios from Figure \ref{fig:non-conv}.
+Panel A consists of the two traceplots that may be evaluated through visual inspection.
+Panel B shows two versions of the $AC$: the default calculation with `R` function `stats::acf()` [@R], and manual calculation as the correlation between $\theta$ in iteration $t$ and $\theta$ in iteration $t+1$.
+Panel C displays the traditional computation of $\widehat{R}$ conform Gelman and Rubin (original), and in panel D we see $\widehat{R}$ as computed by implementing Vehtari et al.'s recommendations (adapted).
+
+```{r diagnostics, fig.cap="Convergence diagnostics applied on the two scenarios from Figure 2 ($\\theta$ = chain mean of variable $j$; $m=5$; $T=10$).
+Depicted are the imputation chains in panel A, two versions of the autocorrelation in panel B, and two versions of $\\widehat{R}$ in panels C and D."}
+
+# (theta + plot_layout(guides = "collect", ncol = 2)) + (ac_both + plot_layout(guides = "keep")) + old_rhat + new_rhat +  plot_annotation(tag_levels = "A", tag_suffix = ".") 
+```
+
+By visually inspecting the imputation chains in panel A, we conclude that the two scenarios show very different convergence.
+The typically converged algorithm initially portrays some signs of non-mixing (around $t=2$), but intermingles nicely overall.
+Additionally, there is very little trending in this scenario.
+The algorithm with pathological non-convergence shows severe non-mixing, although this gradually improves (beyond $t=7$).
+In this scenario, we see a lot of trending initially (up-to $t=6$), after which the chains reached a somewhat more stationary state.
+
+When we look at panel B, we conclude something weird.
+The $AC$-values calculated with the default function indicate equal performance for the typical convergence and the pathological non-convergence scenarios (up-to $t=5$), while there is obvious trending in the $\theta$s of the latter.
+Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$ for both scenarios.
+However, if we look at the chain means of the non-convergence scenario, there should be signs of trending  up-to iteration number seven.
+After consulting the documentation on `stats::acf()`, we conclude that this $AC$ function is not suited for iterative imputation algorithms.
+The function is optimized for performance when $t\geq50$ [@box15], while the default number of iterations in iterative imputation is often much lower.
+Therefore, we compute $AC$ manually (presented with solid lines in panel B).
+These $AC$-values behave as expected from visual inspection.
+We will, therefore, only consider manually calculated $AC$-values as non-convergence diagnostic.
+
+In panel C, we do not see a lot of difference between the two scenarios either.
+The $\widehat{R}$-values are even somewhat lower for the non-convergence scenario than for the typical convergence.
+And apparently, the large variance causes an increase in $\widehat{R}$-values.
+Convergence seems to worsen with every additional iteration.
+This paints the wrong picture.
+An explanation for this behavior may be that under severe trending, the assumption of over-dispersion is violated.
+Taken together, this version of $\widehat{R}$ is sub-optimal for assessing non-convergence in MI.
+
+The $\widehat{R}$-values in panel D are conform our expectations.
+The adapted version of $\widehat{R}$ depicted here does indicate less signs of non-convergence as the number of iterations goes up.
+An exception to this general trend is the 'dip' around $t=3$.
+As diagnostic tool, this may lead to the incorrect conclusion that the non-convergence scenario has superior convergence compared to the typical scenario.
+We can explain this as a downside of the low number of iterations in iterative imputation: the adapted version of $\widehat{R}$ can only be completely employed if the number of iterations is at least four.
+Otherwise, it is not possible to perform all three transformations to the chains of $\theta$-values.
+For low values of $t$, the $\widehat{R}$-values conform Vehtari et al.
+are, therefore, more similar to the original $\widehat{R}$ by Gelman and Rubin.
+
+In short, both non-convergence diagnostics have a version that may be appropriate for iterative imputation algorithms.
+We will compute $\widehat{R}$ conform @vehtari2021, and calculate $AC$ manually.
+The diagnostics will be applied to the four types of scalar summaries $\theta$ described earlier: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix.
+We evaluate the performance of these eight sets of diagnostic methods (two diagnostics; four $\theta$s) through simulation.
+
+# Simulation study
+
+The simulation study aims to determine to what extend non-convergence affects iterative imputation algorithms.
+Specifically, we are interested in the validity of inferences drawn from incomplete data using the MICE algorithm.
+And, whether invalid inferences due to non-convergence may be detected using eight sets of diagnostic methods.
+If the estimated quantities of scientific interest $\bar{Q}$ are unbiased and confidence-valid estimates of ${Q}$, we will conclude that the algorithm is sufficiently converged for practical purposes.
+In contrast to this, we defined *approximate* convergence as the most converged state that the algorithm can obtain.
+Continued iteration after reaching approximate convergence does not yield better mixing or stationarity (indicated by non-improving $\widehat{R}$- and $AC$-values).
+Continued iteration after obtaining valid inferences may lead to a more converged state of the algorithm, but not better estimates.
+
+To induce non-convergence, we consider two sets of simulation conditions: 'missingness' and 'early stopping'.
+Missingness refers to the severity of the missing data problem, which we determine by the proportion of incomplete cases in dataset $y$, $p_{\mathrm inc}$.
+Early stopping is defined by the number of iterations in the imputation algorithm, $T$.
+
+The missingness conditions are chosen to reflect the difficulty of the missingness problem.
+The underlying assumption is that a low proportion of incomplete cases $p_{\mathrm inc}$ leads to quick algorithmic convergence, since there is a lot of information in the observed data.
+A higher $p_{\mathrm inc}$ yields slower convergence---unless there is so little information in $y_{\mathrm obs}$ that this is outweighed by the random component in the imputation algorithm.
+Then, convergence to a stable but highly variable state may be reached instantly.
+Still, we expect that missingness conditions with higher $p_{\mathrm inc}$ will result in more signs of non-convergence.
+
+The assumption inherent to the early stopping conditions is that terminating the imputation algorithm too early causes non-convergence.
+Generally, the algorithm will not reach convergence if $T=1$, because the imputed values in the first iteration (at $t=1$) depend on the starting values [which are sampled randomly from the set of observed datapoints; @buur18].
+As the number of iterations increases, the imputation chains should become independent of the initial values, until approximate convergence is reached (i.e., when an additional iteration does not lead to a more converged state).
+We expect that we can induce non-convergence by early stopping in at least those conditions where $T$ is smaller than the default number of iterations in `mice`, $T=5$ [@mice].
+
+## Hypotheses
+
+1.
+We expect that simulation conditions with a high proportion of incomplete cases $p_{\mathrm inc}$ and a low number of iterations $T$ will more often result in biased, invalid estimates of the quantities of scientific interest $Q$.
+
+2.
+We hypothesize that $\widehat{R}$ and $AC$ will correctly identify signs of non-convergence in those simulation conditions where any $\bar{Q}$ is *not* an unbiased and confidence-valid estimate of $Q$.
+
+3.
+We hypothesize that the recommended thresholds to diagnose non-convergence with $\widehat{R}$ ($\widehat{R} > 1.2$, $\widehat{R} > 1.1$, and $\widehat{R} > 1.01$) may be too stringent for iterative imputation applications.
+In an empirical study, where $\widehat{R}$ was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold $\widehat{R}>1.2$.
+Yet, scientific estimates were insensitive to continued iteration from $t=5$ onward [@lace07].
+We, therefore, suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates.
+In contrast to this, signs of non-convergence can be under-estimated by $\widehat{R}$, in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
+437].
+In  `mice`, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values.
+In practice, we do not expect this to cause problems for identifying non-convergence with $\widehat{R}$.
+
+4.
+We expect that high $AC$-values are implausible in iterative imputation algorithms with typical convergence.
+After only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
+
+5. We further hypothesize that multivariate $\theta$s are better at detecting non-convergence than univariate $\theta$s.
+
+## Set-up
+
+We investigate non-convergence of the MICE algorithm through model-based simulation in `R` [version 3.6.3; @R].
+The number of simulation repetitions is 1000, and the simulation set-up is summarized in the pseudo-code below.
+The complete `R` script of the simulation study is available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).
+
+```
+# pseudo-code of simulation 
+1. simulate data 
+for (number of simulation runs from 1 to 1000)
+ for (proportion of missing cases between 5% and 95%)
+  2. create missingness
+  for (number of iterations from 1 to 100)
+   3. impute missingness
+   4. perform analysis of scientific interest
+   5. compute non-convergence diagnostics 
+   6. pool results across imputations
+   7. compute performance measures
+ 8. combine outcomes of all conditions
+9. aggregate outcomes across simulation runs
+
+```
+### Data-generating mechanism.
+
+In this study, sampling variance is not of interest.
+Therefore, a single complete dataset may serve as comparative truth in all simulation repetitions [@vink14].
+The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem.
+Let the predictor space be defined as
+$$
+\begin{aligned}
+\begin{pmatrix}X_1\\
+X_2\\
+X_3
+\end{pmatrix} \sim \mathcal{N}
+\begin{bmatrix}
+\begin{pmatrix}
+12\\
+3\\
+0.5
+\end{pmatrix}\!\!,
+\begin{pmatrix}
+4 & 4 & 1.8 \\
+4 & 16 & 4.8\\
+1.8 & 4.8 & 9
+\end{pmatrix}
+\end{bmatrix}\!\!\text{.}\\[2\jot]
+\end{aligned}
+$$
+A finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm]. Subsequently, a fourth variable is constructed as outcome variable $Y$. For each unit $i = 1, 2,..., N$, let 
+$$
+Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
+$$
+where $\epsilon \sim \mathcal{N}(0, 100)$.
+This results in a dataset $y$, where $y_{\mathrm mis} = \emptyset$.
+In other words, $y$ contains the complete set of observations for all units $i$ on all variables $j$ (where $j =  Y, X_1, X_2, X_3$).
+From the complete dataset we obtain the true values of the scientific estimands $Q$.
+
+### Scientific estimands.
+
+We consider four types of $Q$s that are often of interest in empirical research.
+Namely, two descriptive statistics: the mean $\mu_j$ and standard deviation $\sigma_j$ of each variable $j$.
+We also consider the regression coefficients of the predictors: $\beta_1$, $\beta_2$, and $\beta_3$ in regression equation
+$$Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,$$
+where $Y'$ is the expected value of the outcome.
+The final estimand is the proportion of variance explained by the set of predictors: coefficient of determination $r^2$ (note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$).
+The scientific estimands are estimated by $\bar{Q}$ in each simulation condition.
+
+### Simulation conditions.
+
+In each simulation repetition, we *ampute* the complete dataset conform the missingness conditions.
+The proportion of incomplete cases $p_{\mathrm inc}$ varies between 5% and 95%, in incremental steps ($p_{\mathrm inc}$ = .05, .25, .50, .75, .95).
+We consider all possible univariate and multivariate patterns of missingness, conform a 'missing completely at random' missingness mechanism [MCAR; @rubin87].
+I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on $p_{\mathrm inc}$.
+We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\mathrm obs}, y_{\mathrm mis}$} for each of the five missingness conditions.
+
+We impute the missing values in {$y_{\mathrm obs}, \dot y_{\mathrm mis}$} according to the early stopping conditions.
+That is, we vary the number of iterations in the imputation algorithm between one and one hundred ($T = 1,2,.,100$).
+All imputation procedures are performed using `mice` [function `mice()`; @mice], with Bayesian linear regression imputation, and five imputation chains ($m=5$).
+This results in $m=5$ sets of imputations for each of the one hundred early stopping conditions, $\dot y_{\mathrm imp, \ell}$.
+
+Subsequently, we obtain the completed data per imputation, {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice].
+We estimate univariate $Q$s directly from the completed data, by averaging across imputations to obtain $\bar{Q}$.
+Estimates of multivariate $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R].
+We pool the resulting estimates, $\hat{Q}_\ell$, into a single $\bar{Q}$ conform @vink14.
+
+### Methods.
+
+We consider eight sets of diagnostic methods to identify non-convergence, by applying $\widehat{R}$ and $AC$ on four different $\theta$s.
+Univariate $\theta$s are obtained from $\dot y_{\mathrm imp, \ell}$.
+The model-independent multivariate $\theta$, $\lambda_{1}$, is obtained from {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$} as the variance in the first PCA component [function `stats::princomp`; R].
+The model-dependent multivariate $\theta$ is one of the quantities of scientific interest: the estimated regression coefficients in {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$}.
+We apply the two non-convergence diagnostics on the $\theta$s by implementing the adapted version of $\widehat{R}$ and by manually programming the $AC$ function.
+
+### Performance measures.
+
+As recommended by @buur18, we evaluate the performance of the imputation algorithm with bias, coverage rate (CR), and confidence interval width (CIW).
+Bias is calculated for each $Q$, whereas CR and CIW are only obtained for $Q = \beta$.
+We calculate bias as $\bar{Q} - Q$.
+CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$.
+Let
+$$\text{CI} = \bar{Q} \pm t_{(m-1)} \times \mathrm SE_{\bar{Q}},$$
+where $t_{(M-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $\mathrm SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
+If we obtain nominal coverage (CR = 95%), we can conclude that $\bar{Q}$ is a confidence-valid estimate of $Q$.
+Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$.
+CIW is of interest because it is a measure of efficiency.
+Under nominal coverage, short CIs are preferred, since wider CIs indicate lower statistical power.
+
+We evaluate the diagnostic methods to identify non-convergence against these performance measures.
+$\widehat{R}$ and $AC$ should identify non-convergence in simulation conditions, where $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
+<!-- Performance measures @buur18, $\S$ 2.5.2 -->
+
+<!-- We could also look at distributional characteristics, and plausibility of imputed values, see @vinknd. For now, this is outside of the scope of this study. -->
+
+# Results
+
+Our results show the performance of the MICE algorithm under different conditions of missingness and early stopping.
+For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias.
+For univariate scientific estimands ($Q=\mu$ and $Q=\sigma$) we observe the largest bias in the outcome variable $Y$.
+And for the regression coefficients ($Q=\beta$) the bias is most pronounced in $\beta_1$ (the effect of $X_1$ on $Y$).
+Since there is just one estimate for $Q=r^2$ to evaluate, we consider $Q=\mu_Y, \sigma_Y,r^2, \beta_1$.
+In the figures, we present all missingness conditions, but only early stopping conditions where $1 \leq T\leq50$.
+That is, the results are more or less stable for conditions where $T \geq 30$.
+Full results are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).
+
+## Quantities of scientific interest
+
+```{r Qs, fig.height = 6.8, fig.cap="Impact of non-convergence on statistical inferences.
+Depicted are the bias, coverage rate (CR) and confidence interval width (CIW) of the estimates $\\bar{Q}$.
+We present the subset of $Q$s with the worst performance in terms of bias: the mean and standard deviation of $Y$, and the regression coefficient of $X_1$.
+The solid gray line depicts the targets (unbiased estimates and nominal coverage rate)."}
+
+# mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_annotation(tag_levels = "A", tag_suffix = ".") + plot_layout(ncol = 2, guides = "collect") 
+```
+
+Figure \ref{fig:Qs} displays the influence of missingness and early stopping on the estimated quantities of scientific interest, $Q$s.
+In the first row of the figure, we see the bias in the estimated descriptive statistics ($Q=\mu_Y$ and $Q=\sigma_Y$; panels A and B).
+The second row shows the bias in the multivariate scientific estimands ($Q=r^2$ and $Q=\beta_1$; C and D).
+The third row consists of the coverage rate (CR) and confidence interval width (CIW) of the estimated regression coefficient ($Q=\beta_1$; E and F).
+
+### $Q=\mu$.
+The estimated univariate means seem unaffected by early stopping.
+This implies that approximately unbiased estimates may be obtained with as little as one iteration ($T \geq 1$).
+The bias in $\bar{Q}$ depends solely on the proportion of incomplete cases, with $p_{\mathrm inc}\geq.75$ leading to more extreme biases than other conditions.
+Completely unbiased estimates are only obtained in conditions where $p_{\mathrm inc}\leq.50$.
+This is curious since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.
+Note however, that the magnitude of the bias in these conditions is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$).
+We, therefore, conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.
+
+### $Q=\sigma$.
+The estimated standard deviations are not substantially impacted by early stopping either.
+While the estimates for $T=1$ are not equal to other conditions, the bias is actually *less* severe than for $T\geq2$.
+Therefore, no iteration at all is needed to obtain approximately unbiased estimates for $Q=\sigma$.
+Completely unbiased estimates are only obtained in conditions where $p_{\mathrm inc}=.05$.
+The other missingness conditions are impacted by non-convergence in the order of increasing $p_{\mathrm inc}$.
+Similar to $Q=\mu$, the magnitude of the bias in $Q=\sigma$ is negligible.
+We, therefore, conclude that neither one of the univariate $Q$s is affected substantially by non-convergence.
+
+### $Q=r^2$.
+In contrast to the univariate estimates, the estimated coefficient of determination is clearly affected by early stopping, see Figure \ref{fig:Qs}C.
+The bias in $Q$ appears to decrease with each additional iteration, while the magnitude of the bias still depends on the proportion of incomplete cases.
+Therefore, we see that the number of iterations that is necessary to reach stable, non-improving estimates also differs across missingness conditions.
+In conditions where $p_{\mathrm inc}=.05$, estimates are unbiased after one iteration.
+The highest number of iterations necessary to reach approximate unbiasedness is seven ($p_{\mathrm inc}=.95$; $T\geq7$).
+Completely unbiased estimates are only obtained in conditions where $p_{\mathrm inc}\leq.25$.
+This implies that even for biased estimates, $T\geq7$ would suffice to reach a stable solution.
+
+### $Q=\beta$.
+For the estimated regression coefficients, we first consider bias, before discussing CR and CIW.
+As Figure \ref{fig:Qs}D shows, the bias in $\bar{Q}$ is affected by both the proportion of incomplete cases and the number of iterations.
+Similar to $Q=r^2$, we observe approximately unbiased estimates after at least one, and at most seven iterations, depending on $p_{\mathrm inc}$.
+Completely unbiased estimates are only obtained in conditions where $T\geq3$ and $p_{\mathrm inc}\leq .50$.
+Despite persistent bias in conditions where $p_{\mathrm inc}\geq.75$, the coverage rates are non-improving after just three iterations, irrespective of $p_{\mathrm inc}$.
+We conclude that any condition where $T\geq3$ yields approximately nominal coverage rates, and thus confidence-valid estimates.
+The nominal coverages can be explained by the CIWs.
+The CIWs depicted in Figure \ref{fig:Qs}F are conform the theoretical foundation of MI: CIs are wider in conditions with higher proportions of incomplete cases (i.e., there is less information in the data, and thus more uncertainty due to missingness).
+Since conditions with higher $p_{\mathrm inc}$ result in both wider CIs and more severe bias, the true value of $Q$ may be included in the CI, despite of the bias in $\bar{Q}$.
+We conclude that approximately unbiased estimates may be obtained in conditions where $T\geq7$, whereas confidence-valid estimates require at most three iterations.
+
+### Summary. 
+These results demonstrate that the estimates of univariate scientific estimands $Q$ are not impacted by early stopping of the MICE algorithm or the proportion of incomplete cases in $y$.
+Unbiased estimates may be obtained after just one iteration.
+Multivariate estimates, by contrast, are affected by both the number of iterations and the proportion of missing cases.
+Completely unbiased estimates are only obtained under low to moderate missingness ($p_{\mathrm inc}\leq.50$), after at most three iterations.
+We observe approximately unbiased estimates after at most seven iterations (for any $p_{\mathrm inc}$ considered).
+This implies that the algorithm produces stable, non-improving estimates when $T\geq7$.
+
+## Non-convergence diagnostics
+
+Figure \ref{fig:rhats-acs} displays results of the eight sets of diagnostic methods.
+The two columns in the figure represent the  non-convergence diagnostics $\widehat{R}$ and $AC$; the rows depict the four scalar summaries $\theta$ under consideration.
+The first row in Figure \ref{fig:rhats-acs} thus shows $\widehat{R}$ and $AC$ applied on the chain means (panels A and B).
+In the second row, we see the diagnostics applied on the chain variances (C and D).
+The third row depicts the same for $\theta=\hat{Q}$ (E and F).
+And in the last row, we see the diagnostics applied on the novel $\theta$ to consider: $\lambda_1$ (G and H).
+We evaluate the performance of the methods by establishing whether they correctly identify conditions in which $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
+As we concluded in the last section, substantially biased estimates were only obtained in conditions where $T\leq6$, and we observed non-nominal coverage only when $T\leq3$.
+However, there was some persistent bias in conditions where $p_{\mathrm inc}\geq.75$, irrespective of the number of iterations.
+If the diagnostic methods are appropriate for MI, we will detect non-convergence in these conditions.
+
+```{r rhats-acs, fig.height = 6.8, fig.cap="Non-convergence diagnostics $\\widehat{R}$ and $AC$ applied on several $\\theta$s.
+The left-hand side of the figure contains $\\widehat{R}$-values, and the right-hand side contains $AC$-values.
+Depicted in the rows are the scalar summaries $\\theta$: chain mean, chain variance, the quantity of scientific interest $\\hat{Q}$, and the first eigenvalue of the variance-covariance matrix $\\lambda_1$.
+The dashed gray lines depict the different thresholds to diagnose non-convergence."}
+
+# mean_Rh + mean_AC + var_Rh + var_AC + est_Rh + est_AC + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 2) + plot_annotation(tag_levels = "A", tag_suffix = ".")
+```
+
+### $\theta=$ chain mean. 
+As expected, $\widehat{R}$ lowers with the number of iterations.
+After an initial 'dip' in conditions where $T=3$, we observe a gradual decrease in $\widehat{R}$-values with every additional iteration.
+This implies improving convergence, until the decrease tapers off after 30 to 40 iterations.
+Using the threshold $\widehat{R}>1.2$, we diagnose non-convergence in conditions where $T\leq7$, whereas the threshold 1.1 is exceeded when $T\leq13$.
+The very strict threshold 1.01 is surpassed in all conditions $T\leq100$ (not shown), and therefore not informative under the current specifications.
+With this set of methods, we fail to identify non-convergence in conditions with persistent bias due to missingness.
+
+Similarly, $AC$ decreases with higher $T$s.
+$AC$ indicates improving stationarity in conditions where $T\leq6$.
+The $AC$-values do not exceed the threshold defined by statistical significance of the $AC$-values.
+Based on this threshold, we fail to diagnose non-convergence in any condition.
+
+### $\theta=$ chain variance.
+As panel C and D in Figure \ref{fig:rhats-acs} show, the results of this set of methods are highly similar to those with chain means as $\theta$.
+This holds for both $\widehat{R}$ and $AC$.
+The only apparent difference is in the $\widehat{R}$ threshold 1.1.
+Instead of diagnosing non-convergence in conditions where $T\leq13$, we now diagnose this when $T\leq11$.
+All other observations are equivalent.
+
+### $\theta=\hat{Q}$.
+The $\widehat{R}$-values in Figure \ref{fig:rhats-acs}E show a similar trend across iterations as we observed for chain means and chain variances.
+Only conditions with a very high  proportion of missing cases ($p_{\mathrm inc}\geq.75$) diverge from the earlier observations.
+The $\widehat{R}$-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations.
+The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are 11 for $\widehat{R}>1.2$, 23 for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).
+
+The $AC$-values also follow earlier observations, with the same exception as for $\widehat{R}$ (when $p_{\mathrm inc}\geq.75$).
+To reach maximum stationarity, these conditions require at most seven iterations instead of five.
+Moreover, the $AC$-values in these conditions exceed the threshold to diagnose non-convergence.
+Interestingly, this only occurs after ten or even thirty iterations, while we would expect more signs of trending early in the iterations.
+Since the $AC$-values are not improving or worsening when $T\geq7$, we conclude that non-convergence is incorrectly diagnosed based on the threshold.
+
+### $\theta=\lambda_{1}$.
+Once again, we observe $\widehat{R}$-values with a similar trend across iterations.
+For this $\theta$, the results differ only in the most extreme missingness condition.
+When $p_{\mathrm inc}=.95$, we diagnose non-convergence at $T\leq9$ according to $\widehat{R}>1.2$, $T\leq19$ for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).
+The 'dip' in $\widehat{R}$-values at $T=3$ is even more pronounced than with other $\theta$s.
+Some $\widehat{R}$-values even overcome the threshold of 1.2.
+If we would terminate the algorithm at $T=3$, we would incorrectly conclude that the algorithm reached approximate convergence, according to this threshold.
+
+$AC$-values are systematically higher for this $\theta$ than for other scalar summaries.
+The $AC$-values also taper off at a later point in the iterations.
+This suggests improving stationarity up-to ten, or even thirty iterations, depending on the missingness condition.
+According to the threshold, we should diagnose non-convergence at some point in any missingness condition.
+However, this occurs only after reasonably decreasing $AC$-values are obtained---again suggesting that non-convergence is incorrectly diagnosed.
+
+### Summary. 
+Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
+Under low to moderate missingness, it does not seem to matter on which $\theta$ the non-convergence diagnostics are applied.
+If we look at complete unbiasedness, however, we notice that univariate $\theta$s fail to diagnose the persistent bias in conditions where $p_{\mathrm inc}\geq.75$.
+
+The number of iterations necessary to obtain approximately unbiased, confidence-valid estimates corresponds to the $\widehat{R}$ threshold 1.2.
+The thresholds 1.1 and 1.01 seem too strict compared to the validity of the inferences.
+The threshold to diagnose significant $AC$-values does not appear to be appropriate in at least the current set-up, and perhaps in iterative imputation in general.
+A better heuristic to diagnose non-stationarity with $AC$ may be through evaluation of the $AC$-values across iterations: If the values do not substantially decrease with $T$, approximate stationarity may be concluded.
+<!-- [similar to the elbow heuristic in PCA]. -->
+
+# Discussion
+
+Before drawing any further conclusions, let us first return to the hypotheses that we formulated for the simulation study.
+
+1.
+As hypothesized, biased, invalid estimates of quantities of scientific interest occurred more often in simulation conditions with a high proportion of incomplete cases $p_{\mathrm inc}$ and a low number of iterations $T$.
+However, this only holds for multivariate scientific estimands.
+Univariate $Q$s are not substantially impacted by early stopping.
+
+2.
+Our results also support the hypothesis that $\widehat{R}$ and $AC$ would correctly identify signs of non-convergence.
+We observed higher $\widehat{R}$ and $AC$-values in conditions where $\bar{Q}$s were biased, invalid estimates of $Q$s.
+
+3.
+We hypothesized that the recommended thresholds to diagnose non-convergence with $\widehat{R}$ would be too stringent for iterative imputation applications, compared to the number of iterations necessary for statistically valid inferences.
+This hypothesis is partially supported: On the one hand, the 1.01 and 1.1 thresholds indeed seem to over-estimate the signs of non-convergence.
+On the other hand, however, we observed that the traditional threshold of 1.2 roughly corresponds to the number of iterations required to obtain approximately unbiased, confidence-valid estimates.
+
+4.
+The results of this simulation study do not support the hypothesis that high $AC$-values are implausible in iterative imputation algorithms with typical convergence.
+We expected that the randomness induced by the algorithm would effectively mitigate the risk of dependency within chains after a few iterations.
+In this study, however, we observed $AC$-values as high as theoretically possible ($AC=1$).
+The second part of the hypothesis is not refuted: the $AC$-values did decrease quickly as a function of $T$.
+
+5.
+Our final hypothesis was that multivariate $\theta$s would be better at detecting non-convergence than univariate $\theta$s.
+Multivariate $\theta$s did indeed show superior performance under certain conditions (e.g., multivariate $Q$s and extreme missingness).
+
+## Conclusion
+
+We have shown that non-convergence in iterative imputation algorithms goes hand in hand with biased, invalid estimates---at least for the multivariate quantities of scientific interest considered in this study.
+Identifying non-convergence may be a crucial aspect of drawing valid statistical inferences from incomplete data.
+We concluded that the current practice of visually inspecting non-convergence through traceplots of univariate scalar summaries of the state-space of the algorithm does not suffice.
+We, therefore, considered eight sets of diagnostic methods to identify non-convergence: two non-convergence diagnostics ($\widehat{R}$ and $AC$), two univariate $\theta$s, a model-dependent multivariate $\theta$, and a novel, model-independent multivariate $\theta$.
+Signs of non-convergence due to early stopping were identified with each of the methods.
+Bias due to high proportions of incomplete cases was only identified with multivariate $\theta$s.
+
+Since we obtained approximately unbiased, confidence-valid estimates after at most seven iterations, we conclude that the $\widehat{R}$ threshold 1.2 is the most appropriate diagnostic cut-off.
+The threshold to diagnose non-stationarity with $AC$ does not seem to apply.
+However, some signs of non-convergence were detected in simulation conditions many more iterations.
+Under moderate missingness ($p_{\mathrm inc}\leq.50$) $\widehat{R}$-values decreased substantially until 30 to 40 iterations, which implies that mixing in the algorithm still improved with each additional iteration.
+$AC$-values only improved until about six iterations, suggesting minimal improvement in trending was obtained beyond $T=6$ in the current simulation set-up.
+
+The main finding of this study is that valid inferences may be obtained much quicker than approximate algorithmic convergence is reached.
+Under the current specifications, univariate $Q$s did not require algorithmic convergence at all.
+They are unbiased almost instantly after the algorithm is initiated.
+Approximately unbiased, confidence-valid estimates of multivariate $Q$s were obtained after a maximum of seven iterations.
+Continued iterations beyond $T = 7$ did not yield better estimates.
+
+## Implications
+
+Iterative imputation algorithms such as MICE are known to yield valid results under severe missingness.
+With this study, we have shown that valid inferences can also be obtained under a combination of severe missingness and early stopping.
+Based on our results, the traditional threshold of $\widehat{R}> 1.2$ would be appropriate for diagnosing non-convergence in iterative imputation algorithms.
+This is, however, not in line with the recent recommendations by @vehtari2021 to lower the threshold to 1.01.
+The discrepancy between our conclusion and Vehtari et al.
+may be explained by the nature of iterative imputation algorithms.
+In the limit, iterative imputation algorithms have the same characteristics as other MCMC algorithms.
+But, in imputation procedures, a part of the distribution is already determined by the observed data, whereas the entire distribution is unknown in many other MCMC applications.
+Since we combine a known distribution with an unknown distribution, valid estimates may be reached much sooner.
+Convergence may, therefore, be diagnosed at a less stringent threshold.
+
+The threshold to diagnose non-convergence with $AC$ does not seem appropriate at all in iterative imputation algorithms.
+A better diagnostic cut-off would be the number of iterations at which an additional iteration does not substantially decrease the $AC$-values.
+In practice, this implies that the default number of iterations in MICE is not sufficient.
+$AC$ can only be computed if $T\geq3$, while the current default in the `mice` package is five [@mice].
+Identifying a reasonable decrease in the $AC$-values across iterations (an 'elbow') requires more than three observations.
+We, therefore, suggest to increase the default number of iterations in MICE to ten.
+The computational cost of five additional iterations has become less of a burden since MICE was introduced [-@mice].
+Moreover, an increased number of iterations would grant the opportunity to exclude the first iterations from evaluation with $\widehat{R}$.
+Excluding $T\leq3$ provides empirical researchers with a less ambiguous heuristic to diagnose non-convergence, because we expect the initial 'dip' in $\widehat{R}$-values to disappear.
 
@@
 Therefore, the initial state of the algorithm is 'too good' to observe a relative decrease in $\widehat{R}$.
+
+## Recommendations for empirical researchers
+
+Based on our results, we formulate several recommendations for empirical researchers who employ iterative imputation to draw inferences from incomplete data.
+We suggest that non-convergence may still be evaluated visually, but in addition to inspecting univariate summaries of the state-space of the algorithm ($\theta$s, e.g., chain means), multivariate $\theta$s should be considered.
+We propose the following steps:
+
+1. First, check traceplots of the default $\theta$s (e.g., chain means and chain variances) for signs of pathological non-convergence. Adjust the imputation model if necessary.
+
+2.
+Subsequently, decide which multivariate $\theta$s to track across iterations.
+The novel $\theta$ that we propose in this study, $\lambda_1$ is scientific model-independent and may thus always be employed.
+Alternatively, specify your own scalar summary of interest [see e.g., @buur18].
+Monitor these $\theta$s through visual inspection, or using a non-convergence diagnostic.
+
+3.
+Compute non-convergence diagnostics $\widehat{R}$ and $AC$.
+Do *not* use the original implementation to calculate $\widehat{R}$ [@gelm92], or the `R` function `stats::acf()` to compute $AC$ [@R].
+Instead, calculate $\widehat{R}$ conform @vehtari2021 and compute autocorrelations manually (see e.g., [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint)).
+
+4.
+And finally, use the threshold $\widehat{R}>1.2$ to diagnose non-mixing, and assess stationarity by plotting the autocorrelation over iterations.
+Keep iterating until the threshold for mixing is overcome, and until reasonably decreasing asymptotic $AC$-values are obtained.
+At that point, inferences are unlikely to improve by continued iteration.
+
+## Limitations
+
+Much remains unknown about non-convergence in iterative imputation algorithms.
+Even though we have demonstrated the appropriateness of $\widehat{R}$ and $AC$ as non-convergence diagnostics in this study, results may not extrapolate to other situations.
+In this study, we only considered the iterative imputation algorithm implemented in the `mice` package.
+The performance of the non-convergence diagnostics may depend on the characteristics of this specific algorithm.
+A potential threat to the ability to detect non-mixing with $\widehat{R}$, for example, is the assumption of over-dispersed initial states.
+In `mice`, the algorithm is initiated by sampling starting values at random from the observed data, whereas other imputation software may use the average of the observed cases.
+The latter could lead to a systematical under-estimation of the variability in the first few iterations and violate the assumption of over-dispersion.
+In those cases, it would be more difficult to detect non-convergence with $\widehat{R}$.
+Therefore, the diagnostic might not be  appropriate outside of `mice`.
+
+Moreover, the current simulation conditions were restricted to a single missingness mechanism.
+Proper performance under a 'missing completely at random' (MCAR) mechanism is a necessary condition for any missing data method.
+It does not, however, guarantee equal performance under different missingness mechanisms.
+Future research should determine the performance of $\widehat{R}$ and $AC$ under 'missing at random' and 'missing not at random' mechanisms.
+Another parameter to consider in future research is the choice of the imputation method.
+As @murray2018 concluded, the behavior of algorithms such as MICE under certain default imputation models is still an open research question.
+We have investigated the behavior of MICE under only one type of model (Bayesian linear regression).
+Different imputations models might converge more poorly.
+
+The imputation model may even be mis-specified on purpose, to induce non-convergence of different severity levels.
+Qualitatively, we have shown that $\widehat{R}$ and $AC$ are appropriate under clear violation of convergence.
+Quantitatively, we have only demonstrated that they can identify signs of non-convergence under MCAR, when the imputation model is a correctly specified Bayesian linear regression model.
+\newline \newline
+
+\noindent In short, we have shown that iterative imputation algorithms can yield correct outcomes, even when a converged state has not yet formally been reached.
+Any further iterations would then burn computational resources without improving the statistical inferences.
+Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the $\widehat{R}$ and AC diagnostics.
+Of course, it never hurts to iterate longer, but such calculations hardly bring added value.

### 2025-04-10 figures/diagram.tex
add contents thesis

\documentclass{standalone}\usepackage[]{graphicx}\usepackage[]{xcolor}
% maxwidth is the original width if it is less than linewidth
% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlsng}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hldef}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
%% recommended packages

%% more packages 
\usepackage{amsmath, tikz} % for table, equations, and flow chart figure (respectively)
\usetikzlibrary{fit, positioning, shapes} % for flow chart figure
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

\begin{tikzpicture}
\tikzstyle{main}=[rectangle, minimum width = 22mm, minimum height = 9mm, thick, draw =black!80, node distance = 14mm]
\tikzstyle{connect}=[draw, -latex, thick, out=180,in=270, looseness=0]
\node[main, fill = white!100] (data) [label=below:Incomplete data] {$\{y_{\rm obs}, y_{\rm mis}\}$};
\node[main] (mids) [right=of data] { $\{y_{\rm obs}, \dot y_{\rm imp, \ell}\}$};
\node[main] (mids2) [above=10mm of mids] {$\{y_{\rm obs}, \dot y_{\rm imp, \ell}\}$ };
\node[main] (mids3) [below=10mm of mids,label=below: $m$ completed datasets] { $\{y_{\rm obs}, \dot y_{\rm imp, \ell}\}$};
\node[main] (mira) [right=of mids] {$\hat{Q}_\ell$};
\node[main] (mira2) [above=10mm of mira] { $\hat{Q}_\ell$ };
\node[main] (mira3) [below=10mm of mira,label=below: $m$ results] { $\hat{Q}_\ell$ };
\node[main, fill = white!100] (mipo) [right=of mira,label=below:Pooled result] {$\bar{Q}$ };
\path      
    (data.east) edge [connect] node [right] {} (mids.west)
(data.east) edge [connect] node [right] {} (mids2.west)
(data.east) edge [connect] node [right] {} (mids3.west)
      (mids.east) edge [connect] node [right] {} (mira.west)
      (mids2.east) edge [connect] node [right] {} (mira2.west)
      (mids3.east) edge [connect] node [right] {} (mira3.west)
		(mira.east) edge [connect] node [right] {} (mipo.west)
		(mira2.east) edge [connect] node [right] {} (mipo.west)
		(mira3.east) edge [connect] node [right] {} (mipo.west)
;
\end{tikzpicture}
%\caption{Scheme of the main steps in multiple imputation \citep[$m = 3$; adapted from][\S~1.4.1]{buur18}. {\footnotesize Missing data in dataset $Y$ is `imputed' (i.e., filled in) $m$ times. The imputed data is combined with the observed data ($Y_{obs}$) to create $m$ completed datasets. On each completed dataset, the analysis of scientific interest (or `substantive model') is performed. The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$. Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}$-values are obtained. These $m$ values are combined into a single pooled estimate $\bar{Q}$. $\{y_{obs}, y_{mis}\}$ }}

\end{document}

### 2025-04-10 figures/non-conv.R

@@ -1,0 +1,83 @@
+# reproduce example of pathological non-convergence traceplots
+# by van Buuren (2018) and apply non-convergence diagnostics
+# requires the packages 'mice', 'purrr', and 'dplyr'/'magrittr'
+# and the non-convergence diagnostics function
+
+library(mice, warn.conflicts = FALSE)
+
+# non-conv
+# impute missingness with mis-specified imputation model
+meth <- make.method(boys) #define imputation model
+meth["bmi"] <- "~I(wgt / (hgt / 100)^2)" #mis-specify model
+imp_nonconv <- mice( #impute missingness
+  boys,
+  meth = meth,
+  maxit = 10,
+  print = FALSE,
+  seed = 60109
+)
+
+
+# impute missingness with correctly specified imputation model
+pred <- make.predictorMatrix(boys) #mend imputation method
+pred[c("hgt", "wgt"), "bmi"] <- 0 #remove dependency
+imp_conv <- conv <- mice( #impute missingness
+  boys,
+  meth = meth,
+  pred = pred,
+  maxit = 10,
+  print = FALSE,
+  seed = 60109
+)
+
+plot_wgt <- function(imp) {
+# extract parameters for plotting
+mn <- imp$chainMean
+sm <- sqrt(imp$chainVar)
+vrb <-  "wgt" #names(imp$data)
+p <- length(vrb)
+m <- imp$m
+it <- imp$iteration
+long <- cbind(expand.grid(.it = seq_len(it), .m = seq_len(m)),
+              data.frame(
+                .ms = rep(c("mean", "sd"), each = m * it *
+                            p),
+                vrb = rep(vrb, each = m * it, times = 2),
+                val = c(matrix(aperm(mn[vrb, , , drop = FALSE], c(
+                  2, 3, 1
+                )), nrow = m * it * p), matrix(aperm(sm[vrb, , , drop = FALSE], c(
+                  2, 3, 1
+                )), nrow = m * it * p))
+              ))
+ggplot2::ggplot(long[long$.ms == "mean", ],
+                ggplot2::aes(
+                  x = .data$.it,
+                  y = .data$val,
+                  color = as.factor(.data$.m)
+                )) + ggplot2::geom_line(linewidth = 0.6) +
[…]

### 2025-04-16 Manuscript/manuscript.qmd
update sim design

@@
 library(ggplot2)
+library(dplyr)
+load("../results/out.RData")
+source("../R/evaluate_estimates.R")
+performance <- evaluate_est(out)
+performance[performance$term == "Y", "term"] <- "b0"
+results <- # simulation results split by condition and regression coefficient
+performance |> 
+  group_by(method, mech, prop, .it, term) |> 
+  summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), \(x) mean(x, na.rm = TRUE))) |>
+  mutate(mech = factor(mech, levels = c("MCAR", "MAR"), ordered = TRUE),
+         prop = factor(prop),
+         term = factor(term))
 ```
 
-# Abstract (old)
+```{r, echo=FALSE}
+# # plot results per sim rep
+# performance |>
+#   filter(term == "X1", mech == "MAR", method == "MICE", prop == 0.75) |>
+#   ggplot(aes(x = .it, y = bias)) + #, color = as.factor(prop)
+#   geom_line(aes(group = .sim)) +
+#   geom_smooth(method = "loess") +
+#   # labs(
+#   #   title = "Bias regression coefficient",
+#   #   x = "Iteration number",
+#   #   y = "Bias"
+#   # ) +
+#   theme_classic()
+```
 
@@
 - how many iterations are needed before valid inferences are reached?
+
+```{r}
+results |>
+  filter(method == "MICE") |> #, prop == 0.75 mech == "MAR",
+  ggplot(aes(x = .it, y = bias, color = prop)) +
+  geom_hline(yintercept = 0, alpha = 0.1) + 
+  geom_line() +
+  facet_grid(term ~ mech) +
+  theme_classic()
+
+results |>
+  filter(method == "MICE") |> #, prop == 0.75 mech == "MAR",
+  ggplot(aes(x = .it, y = psrf_mean, color = prop)) +
+  geom_hline(yintercept = 1, alpha = 0.1) + 
+  geom_line() +
+  facet_grid(term ~ mech) +
+  theme_classic()
+
+```
 
@@
 
-We investigate non-convergence of the MICE algorithm through model-based simulation in `R` [version 3.6.3; @R].
-The number of simulation repetitions is 1000, and the simulation set-up is summarized in the pseudo-code below.
-The complete `R` script of the simulation study is available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).
+We investigate non-convergence of the MICE algorithm through model-based simulation in `R` [version 4.4.2; @R].
+The number of simulation repetitions is 2000, and the simulation set-up is summarized in the pseudo-code below.
+The complete `R` script of the simulation study is available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint) [TODO: make Zenodo DOI].
 
@@
 # pseudo-code of simulation 
-1. simulate data 
-for (number of simulation runs from 1 to 1000)
- for (proportion of missing cases between 5% and 95%)
+for (number of simulation runs) {
+  1. simulate complete data 
+  for (missingness conditions) {
     2. create missingness
-  for (number of iterations from 1 to 100)
+    for (number of iterations) {
       3. impute missingness
-   4. perform analysis of scientific interest
-   5. compute non-convergence diagnostics 
-   6. pool results across imputations
-   7. compute performance measures
- 8. combine outcomes of all conditions
-9. aggregate outcomes across simulation runs
+      4. estimate quantities of scientific interest
+      5. apply performance measures to the estimates
+      6. compute non-convergence diagnostics
+    }
+  }
+7. aggregate outcomes across simulation runs
+```
 
-```
 ### Data-generating mechanism.
 
-In this study, sampling variance is not of interest.
-Therefore, a single complete dataset may serve as comparative truth in all simulation repetitions [@vink14].
 The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem.
 Let the predictor space be defined as
+
 $$
@@
 \begin{pmatrix}
-12\\
-3\\
-0.5
+0\\
+0\\
+0
 \end{pmatrix}\!\!,
 \begin{pmatrix}
-4 & 4 & 1.8 \\
-4 & 16 & 4.8\\
-1.8 & 4.8 & 9
+1 & 0.5 & 0.5 \\
+0.5 & 1 & 0.5 \\
+0.5 & 0.5 & 1
 \end{pmatrix}
@@
 $$
-A finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm]. Subsequently, a fourth variable is constructed as outcome variable $Y$. For each unit $i = 1, 2,..., N$, let 
+
+In each simulation repetition, a finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm].
+Subsequently, a fourth variable is constructed as outcome variable $Y$.
+For each unit $i = 1, 2,..., N$, let
+
 $$
-Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
+Y_i = X_{1i} + X_{2i} + X_{3i} + \epsilon_i ,
 $$
-where $\epsilon \sim \mathcal{N}(0, 100)$.
+where $\epsilon \sim \mathcal{N}(0, 1)$.
 This results in a dataset $y$, where $y_{\mathrm mis} = \emptyset$.
 In other words, $y$ contains the complete set of observations for all units $i$ on all variables $j$ (where $j =  Y, X_1, X_2, X_3$).
-From the complete dataset we obtain the true values of the scientific estimands $Q$.
 
@@
 
-We consider four types of $Q$s that are often of interest in empirical research.
-Namely, two descriptive statistics: the mean $\mu_j$ and standard deviation $\sigma_j$ of each variable $j$.
-We also consider the regression coefficients of the predictors: $\beta_1$, $\beta_2$, and $\beta_3$ in regression equation
+We consider four multivariate quantities of scientific interest ($Qs$). Namely, the regression coefficients of the analysis model,
+
 $$Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,$$
 where $Y'$ is the expected value of the outcome.
-The final estimand is the proportion of variance explained by the set of predictors: coefficient of determination $r^2$ (note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$).
+From the data-generating mechanism, we obtain the true values of the scientific estimands: $\beta_0 = 0$, $\beta_1 = \beta_2 = \beta_3 = 1$.
 The scientific estimands are estimated by $\bar{Q}$ in each simulation condition.
+
+<!-- The final estimand is the proportion of variance explained by the set of predictors: coefficient of determination $r^2$ (note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$).
+-->
 
@@
 
-In each simulation repetition, we *ampute* the complete dataset conform the missingness conditions.
-The proportion of incomplete cases $p_{\mathrm inc}$ varies between 5% and 95%, in incremental steps ($p_{\mathrm inc}$ = .05, .25, .50, .75, .95).
-We consider all possible univariate and multivariate patterns of missingness, conform a 'missing completely at random' missingness mechanism [MCAR; @rubin87].
-I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on $p_{\mathrm inc}$.
-We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\mathrm obs}, y_{\mathrm mis}$} for each of the five missingness conditions.
+We consider two sets of simulation conditions, 'missingness conditions' and 'early stopping' conditions. The simulation study is a fully factorial design, as visualized in Table XYZ.
+
+In each simulation repetition, we *ampute* the complete dataset conform six missingness conditions: two missing data mechanisms and three proportions of incomplete cases.
+The missing data mechanisms are 'missing completely at random' [MCAR; @rubin87], where the probability to be missing is the same for all $N \times k$ cells in $y$, and a right-tailed 'missing at random' (MAR) mechanism, where the probability to be missing is a function of the observed data and higher values are more likely to be missing.
+The proportion of incomplete cases $p_{\mathrm inc}$ is set to 25%, 50%, and 75%.
+
+<!-- Table -->
+<!-- mechanisms: MCAR, MAR -->
+<!-- proportion of incomplete cases: 25%, 50%, 75% -->
+
+<!-- \begin{table}[] -->
+<!-- \begin{tabular}{lllll} -->
+<!-- \hline -->
+<!-- Missingness &  &                 &  & Early stopping            \\ \cline{1-3} \cline{5-5}  -->
+<!-- Mechanism   &  & Proportion      &  & Iteration number          \\ \cline{1-1} \cline{3-3} \cline{5-5}  -->
+<!-- MCAR, MAR   &  & 0.25, 0.5, 0.75 &  & 1, 2, 3, ..., 98, 99, 100 \\ \hline -->
+<!-- \end{tabular} -->
+<!-- \end{table} -->
+
+$$
+\begin{aligned}
+\hline
+&\mathrm{Missingness \ conditions} & & & &\mathrm{Early \ stopping \ conditions} \\
+\hline
+&\begin{matrix}
+\mathrm{MCAR} \\
+\mathrm{MAR \ } 
+\end{matrix}
+\quad
+&\times
+\quad
+&\begin{matrix}
+25\% \\
+50\% \\
+75\%
+\end{matrix}
+\quad
+&\times
+\quad
+&\begin{matrix}
+1 \\
+2 \\
+... \\
+99 \\
+100 \\
+\end{matrix} \\
+&\mathrm{mechansim} &&\mathrm{incomplete \ cases} &&\mathrm{iteration(s)} \\
+\hline
+\end{aligned}
+$$
+
+We consider all possible multivariate patterns of missingness, as visualized in Figure \ref{fig:patterns}.
+
+```{r patterns, echo=FALSE, eval=FALSE}
+source("../R/generate_datasets.R")
+pat <- create_patterns(3)
+d <- dat[1:10,]
+is.na(d) <- pat == 1
+d
+ggmice::plot_pattern(d)
+```
+
+We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\mathrm obs}, y_{\mathrm mis}$} for each of the six missingness conditions.
 
@@
 Subsequently, we obtain the completed data per imputation, {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice].
-We estimate univariate $Q$s directly from the completed data, by averaging across imputations to obtain $\bar{Q}$.
-Estimates of multivariate $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R].
-We pool the resulting estimates, $\hat{Q}_\ell$, into a single $\bar{Q}$ conform @vink14.
+Estimates of scientific quantities $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R].
+We pool the resulting $m$ estimates, $\hat{Q}_\ell$, into a single $\bar{Q}$ conform Rubin's rules [function `mice::pool()`; @mice].
 

### 2025-04-16 simulation.R
run sim with different param

@@ -15,23 +15,23 @@ miceadds::source.all("./R")
 set.seed(11)
 
 # parameters
-n_sim <- 2
+n_sim <- 100
 n_obs <- 200
-n_col <- 4
-corr <- 0.3
-beta <- 0.6
+n_col <- 3
+corr <- 0.5
+beta <- 1
 betas <- rep(beta, n_col)
 mis_pat <- create_patterns(n_col)
 mis_mech = c("MCAR", "MAR")
 mis_prop = c(0.25, 0.5, 0.75)
-n_it <- 50
+n_it <- 100
 
 # #################################
 # ### TEST LOWER LEVEL FUCTIONS ###
 # #################################
 # 
 # # generate data
-# dat <- generate_complete(n_obs, corr, betas)
+# dat <- generate_complete(n_obs, n_col, corr, betas)
 # 
 # # ampute data
 # amps <- induce_missingness(dat, mis_pat, mis_mech = "MAR", mis_prop = 0.5)
@@ -115,54 +115,6 @@ out <- pbreplicate(n_sim,
 
 parallel::stopCluster(cl)
 
+out <- purrr::map(1:length(out), ~{cbind(.sim = .x, out[[.x]])}) 
+save(out, file = paste0("results/out", length(out), "sim", max(out[[1]]$.it), "it", ".RData"))
 save(out, file = "results/out.RData")
-
-########################
-### EVALUATE RESULTS ###
-########################
-
-# calculate bias, coverage rate and CI width
-# performance <- evaluate_est(results_raw)
-# # saveRDS(performance, "./Results/performance.RDS")
-performance <- evaluate_est(out)
-
-# simulation results across all conditions
-performance |> 
-  group_by(method, .it) |> 
-  summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), mean, na.rm = TRUE))
-
-# simulation results split by condition
-performance |> 
-  group_by(method, mech, prop, .it) |> 
-  summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), mean, na.rm = TRUE))
-
-# simulation results split by condition and regression coefficient
-performance |> 
-  group_by(method, mech, prop, .it, term) |> 
-  summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), mean, na.rm = TRUE))
-
-# plot results
-performance |>
[…]

### 2025-04-16 R/apply_methods.R
make analysis functions data-independent

@@ -4,7 +4,7 @@
 apply_full <- function(amp) {
   # list-wise deletion
   est <- # fit regression 
-    lm(Y ~ X1 + X2 + X3 + X4, data = amp$data) |> 
+    lm(Y ~ ., data = amp$data) |> 
     # clean results
     broom::tidy(conf.int = TRUE) |> 
     # choose estimates
@@ -12,7 +12,7 @@ apply_full <- function(amp) {
   # add method name and missingness
   est <- cbind(method = "full", mech = amp$mech, prop = amp$prop, .it = 0, est, ac_mean = NA, psrf_mean = NA, ac_sd = NA, psrf_sd = NA)
   # rename "(Intercept)" to "Y" for easier processing
-  est[est$term == "(Intercept)", "term"] <- "Y"
+  est[est$term == "(Intercept)", "term"] <- "b0"
   # output
   return(est)
 }
@@ -21,7 +21,7 @@ apply_full <- function(amp) {
 apply_CCA <- function(amp) {
   # list-wise deletion
   est <- # fit regression 
-    lm(Y ~ X1 + X2 + X3 + X4, data = na.omit(amp$amp)) |> 
+    lm(Y ~ ., data = na.omit(amp$amp)) |> 
     # clean results
     broom::tidy(conf.int = TRUE) |> 
     # choose estimates
@@ -29,7 +29,7 @@ apply_CCA <- function(amp) {
   # add method name and missingness
   est <- cbind(method = "CCA", mech = amp$mech, prop = amp$prop, .it = 0, est, ac_mean = NA, psrf_mean = NA, ac_sd = NA, psrf_sd = NA)
   # rename "(Intercept)" to "Y" for easier processing
-  est[est$term == "(Intercept)", "term"] <- "Y"
+  est[est$term == "(Intercept)", "term"] <- "b0"
   # output
   return(est)
 }
@@ -67,7 +67,7 @@ apply_MICE <- function(amp, n_it) {
 # internal function to calculate pooled regression estimates
 estimate_param <- function(imp) {
   # run analysis on all imputations
-  est <- with(imp, lm(Y ~ X1 + X2 + X3 + X4)) |> 
+  est <- purrr::map(mice::complete(imp, "all"), ~{lm(Y ~ ., data = .x)}) |> 
     # pool results
     mice::pool() |> 
     # clean results
@@ -75,7 +75,7 @@ estimate_param <- function(imp) {
     # select estimates
[…]

### 2025-04-16 R/generate_datasets.R
make data-generation function independent of n_col

@@ -1,20 +1,16 @@
 # functions to create incomplete datasets
 
 # generate complete data
-generate_complete <- function(
-    n_obs = 200,
-    corr = 0.3,
-    betas = c(-0.5,-0.1, 0.1, 0.5)
-) {
+generate_complete <- function(n_obs, n_col, corr, betas) {
   # create variance-covariance matrix with moderate correlations
-  vcov <- matrix(0.3, nrow = 4, ncol = 4)
+  vcov <- matrix(corr, nrow = n_col, ncol = n_col)
   diag(vcov) <- 1
   # create predictor space data
   X <- mvtnorm::rmvnorm(n = n_obs, sigma = vcov)
   # multiply each predictor observation by the corresponding beta
   Y <- X %*% betas
   # generate residual error for each observation
-  e <- rnorm(n_obs)
+  e <- rnorm(n_obs, mean = 0, sd = 1)
   # combine predictors and outcome plus residual
   dat <- data.frame(Y = Y + e,
                     X)
@@ -59,13 +55,13 @@ induce_missingness <- function(
 create_data <- function(
     sample_size = 200, 
     correlations = 0.3,
-    effects = c(-0.5,-0.1, 0.1, 0.5),
+    effects = c(0.5, 0.5, 0.5),
     patterns = NULL, 
     mechanisms = c("MCAR", "MAR"),
     proportions = c(0.1, 0.25, 0.5)
 ) {
   # create a single complete dataset
-  dat <- generate_complete(n_obs = sample_size, corr = correlations, betas = effects)
+  dat <- generate_complete(n_obs = sample_size, n_col = length(effects), corr = correlations, betas = effects)
   # ampute the data with different missingness
   amps <- induce_missingness(dat, mis_pat, mis_mech = mechanisms, mis_prop = proportions)
   # output

### 2025-04-16 figures/non-conv.R
put results in seperate folder

@@ -56,7 +56,7 @@ ggplot2::ggplot(long[long$.ms == "mean", ],
                   color = as.factor(.data$.m)
                 )) + ggplot2::geom_line(linewidth = 0.6) +
   ggplot2::geom_hline(yintercept = -Inf) + 
-  ggplot2::labs(x = "Iteration", y = "Imputation parameter", color = "") + 
+  ggplot2::labs(x = "Iteration number", y = "Average imputed value", color = "") + 
   ggplot2::theme_classic() + 
   ggplot2::theme(legend.position = "none")
 }

### 2025-04-17 Manuscript/manuscript.qmd
add contents MissingThePoint manuscript

@@
 
+# Introduction (newest)
+
+Iterative imputation has become the de facto standard to accommodate missing data in social scientific research [ref: e.g., rubin after 18 years paper?
+murray 2018?].
+The technique enables researchers to draw valid inferences when faced with incomplete observations, without resorting to ad hoc solutions such as list-wise deletion, which may bias results.
+With iterative imputation, every missing cell in an incomplete dataset is imputed (i.e., filled in) algorithmically.
+And once the dataset is completed, the analysis of scientific interest can be performed.
+This analysis will yield unbiased and confidence-valid estimates if--and only if--both the missing data problem and the scientific problem are appropriately considered.
+The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputed values.
+Yet, there has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms.
+
+<!-- Any scientific inferences obtained after iterative imputation rely on the convergence of the imputation algorithm,  -->
+
+Determining whether an algorithm has converged is not trivial, especially in the context of iterative imputation.
+Since the aim of iterative imputation is to converge to a distribution and not to a single point, the algorithm may produce some fluctuations even after it has converged.
+Because of this property, it may be more desirable to focus on *non*-convergence.
+A widely accepted practice is visual inspection of the algorithm [@ragh07; @buur18].
+Diagnosing non-convergence through visual inspection may, however, be undesirable: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18].
+[add argument: "Inspection of such plots is a notoriously unreliable method of assessing convergence and in addition is unwieldy when monitoring a large number of quantities of interest, such as can arise in complicated hierarchical models" (Gelman et al., 2013, p.
+285).] Therefore, a quantitative diagnostic method to assess convergence would be preferred.
+Fortunately, there are non-convergence identifiers for *other* iterative algorithms, but the validity of these identifiers has not been systematically evaluated on imputation algorithms.
+
+<!-- It is challenging to arrive upon a single point at which convergence has been reached. -->
+
+In this study, we explore different methods for identifying non-convergence in iterative imputation algorithms.
+We evaluate whether these methods are able to cover the extent of the non-convergence, and we also investigate the relation between non-convergence and the validity of the inferences.
+We translate the results of our simulation study into guidelines for practice, which we demonstrate by means of a case study.
+
+# Simulation set-up
+
+We investigate non-convergence in iterative imputation through model-based simulation in R (version 4.4.2; R Core Team 2024).
+We provide a summary of the simulation set-up in Algorithm 1; the complete script and technical details are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint) [TODO: make this a Zenodo DOI].
+
+**Algorithm 1: simulation set-up (pseudo-code)**
+
+    for each simulation repetition 
+      1. simulate complete data
+      for each missingness condition 
+        2. create missingness
+        for each iteration 
+          3. impute missingness
+          4. perform analysis of scientific interest
+          5. apply non-convergence identifiers
+          6. pool results across imputations
+          7. compute performance measures
+        8. combine outcomes from all iterations
+      9. combine outcomes from all missingness conditions
+    10. aggregate outcomes from all simulation repetitions
+
+## Aims
+
+With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using the imputation package `{mice}` [@mice].
+Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$).
+To assess non-convergence, we terminate the imputation algorithm after a varying number of iterations ($n_{it} = 1, 2, ..., 50$).
+We differentiate between nine different missingness scenarios that are defined in the data generating mechanism.
+
+## Data generating mechanism
+
+We generate one complete set of $n_{obs} = 1000$ cases in each simulation repetition. This set consists of three multivariately normal random variables, such that
+
+$$
+\begin{pmatrix}
+Y\\
+X_1\\
+X_2
+\end{pmatrix}
+\sim \mathcal{N}
+\begin{bmatrix}
+\begin{pmatrix}
+0\\
+0\\
+0
+\end{pmatrix},
+\begin{pmatrix}
+2     &       &  \\
+0.5   & 1     &  \\
+-0.5  & 0.5   & 1
+\end{pmatrix}
+\end{bmatrix}.
+$$
+
+The complete set is 'amputed' (i.e., made incomplete) according to nine missingness conditions.
+We use a $3 \times 3$ factorial design with three missingness mechanisms and three proportions of incomplete cases, see Figure XYZ.
+Missingness mechanisms refer to the probability of being missing for any given entry in a dataset.
+There are three distinct types of mechanisms, as defined by Rubin (1976).
+Roughly translated, the probability of being missing may be equal for all entries (MCAR; Missing Completely At Random), the probability of being missing depends on observed information (MAR; Missing At Random), or the probability depends on *un*observed information, making the missingness non-ignorable (MNAR; Missing Not At Random).
+[add left-tailed M(N)AR mechanism, md pattern, and maybe refer to ampute paper or "Dance of..." by Schouten 2018].
+
+## Estimands
+
+We impute the missing data five times ($m = 5$) using Bayesian linear regression imputation with `mice` [@mice].
+On each imputed dataset, we perform multiple linear regression as our analysis of scientific interest.
+We predict outcome variable $Y$ from the other two variables,
+
+$$
+\hat{Y} = \beta_0 + \beta_1 X_1 + \beta_2 X_2,
+$$
+
+where $\hat{Y}$ is the predicted value of the outcome.
+Our estimands are the regression coefficient $\beta_1$ and coefficient of determination $\rho^2$ that we obtain after pooling the regression results across the imputations.
+
+## Diagnostic methods
+
+- non-convergence in iterative algorithms is diagnosed using an identifier and a parameter
+
+- the parameter can be any statistic that we track across iterations, for example the average imputed value per imputation (i.e., chain means)
+
+- the identifier is a calculation of some sort that quantifies non-convergence
+
+- identifiers are historically focused on either non-mixing between chains or non-stationarity within chains
+
+- the popular non-mixing identifier rhat has recently been updated by Vehtari et al, and should now work for non-stationarity as well
+
+- just to be sure, we also use autocorrelation to quantify trending within chains
+
+- we apply these identifiers to four different univariate and multivariate parameters
+
+- the univariate parameters are those commonly used for visual inspection: chain means and chain variances
+
+- one multivariate parameter that is of immediate interest for our estimand is the estimated value itself: regression coefficient beta_1
+
+- we also propose a new multivariate parameter that is not dependent on the model of scientific interest: the first eigenvalue of the variance-covariance matrix of the completed data [explain!]
+
+We use eight different methods to diagnose non-convergence: a combination of two non-convergence identifiers---autocorrelation and $\widehat{R}$---and four parameters---chain means, chain variances, $\beta_1$, and $\lambda$.
+
+## Performance measures
+
+As recommended by van Buuren, our performance measures are bias, confidence interval width, and coverage rate of the estimands (-@buur18 $\S$ 2.5.2).
+These measures are averaged across the simulation repetitions.
+<!-- [or just of the regression estimate?? otherwise add the ciw and cov of r2 too!!]. -->
+
+# Simulation Results
+
+The following figures display the simulation results for the eight diagnostic methods and four performance measures, contrasted to the number of iterations in the imputation algorithm.
+Within the figures, we split the results according to the missingness conditions [missingness mechanisms as line types, and proportion of incomplete cases as colors].
+Note that these results are averages of the $n_{sim} = 1000$ simulation repetitions.
+
+## Diagnostic Methods
+
+Figure \@ref(fig:ac) shows the autocorrelations in the chain means (panel A), chain variances (panel B), regression estimates (panel C), and eigenvalues (panel D).
+
+Autocorrelation in the chain means decreases rapidly in the first few iterations (see \@ref(fig:ac)A).
+The decrease is substantive until $n_{it} \geq 6$.
+This means that there is some initial trending within chains, but the average imputed value quickly reaches stationarity.
+These results hold irrespective of the missingness condition.
+
+Autocorrelation in the chain variances show us something similar (see \@ref(fig:ac)B).
+The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous than for chain means, but generally around $n_{it} \geq 10$.
+We do not observe a systematic difference between missingness conditions here either.
+
+There is more autocorrelation in the scientific estimates than in the chain means and chain variances (see \@ref(fig:ac)C).
+We observe the highest autocorrelations in conditions where 75% of cases are incomplete.
+Overall, the autocorrelations reach a plateau when $n_{it} \geq 20$ to $30$.
+There is no clear effect of the missingness mechanisms.
+
+The autocorrelation in the eigenvalues exhibits a similar trend to the autocorrelation in the scientific estimates (see \@ref(fig:ac)D). Trending in this parameter diminishes when $n_{it} \geq 20$.
+
+Figure \@ref(fig:rh) shows the potential scale reduction factor in the chain means (panel A), chain variances (panel B), regression estimates (panel C), and eigenvalues (panel D).
+
+We observe that $\widehat{R}$-values of the chain means generally decreases as a function of the number of iterations (see \@ref(fig:rh)A).
+An exception to this observation is the initial increase when $3 \leq n_{it} \leq 5$ [interpret??
+due to initialization or is there really more mixing initially??].
+After the first couple of iterations, the mixing between chain means generally improves until $n_{it} \geq 30$ to $40$.
+There is no apparent differentiation between the missingness conditions.
+
+The mixing between chain variances mimics the mixing between chain means almost perfectly (see \@ref(fig:rh)B).
+Irrespective of the missingness condition, the $\widehat{R}$-values taper off around $n_{it} \geq 30$.
+
+With the regression estimate as parameter we observe very similar $\widehat{R}$-values than with chain means and chain variances (see \@ref(fig:rh)C).
+We do, however, see some differences between missingness conditions.
+Conditions where 75% of the cases are incomplete show more extreme non-mixing.
+The overall trend remains the same: about 30 iterations are required before mixing stops improving substantially.
+
+$\widehat{R}$-values of the eigenvalues show a trend similar to the chain means and chain variances (see \@ref(fig:rh)D). [add something about conditions??]
+
+[These rhat plots all show some initialization before the fifth iteration: is rhat useful before that??]
+
+## Performance Measures
+
+In Figure \@ref(fig:perf) we show the performance measures: bias in the regression estimate (panel A), bias in the coefficient of determination (panel B), the empirical coverage rate of the regression estimate (panel C) and the average confidence interval width of this estimate (panel D).
+
+We see that within a few iterations the bias in the regression estimate approaches zero (see \@ref(fig:perf)A).
+When $n_{it} \geq 6$, even the worst-performing conditions (e.g., with a proportion of incomplete cases of 75%) produce stable, non-improving estimates [regression coefficient is underestimated because there is less info to estimate the relation??].
+
+Equivalent to the bias in the regression estimate, the bias in the coefficient of determination tapers off within a couple of iterations (see \@ref(fig:perf)B).
+We observe stable estimates in all conditions when $n_{it} \geq 6$ [interpret the over-estimation in MNAR+75% condition?].
+
+Nominal coverage is quickly reached (see \@ref(fig:perf)C).
+After just three iterations, the coverage rates are non-improving in every missingness condition [but MNAR with 5% incomplete cases does not reach nominal coverage --\> due to bias in the estimate in combination with very narrow CI (see CIW!)].
+
+The average confidence interval width decreases quickly with every added iteration until a stable plateau is reached (see \@ref(fig:perf)D).
+Depending on the proportion of incomplete cases this takes up-to $n_{it} \geq 9$.
+
+# Discussion
+
+Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the non-convergence identifiers.
+Of course, it never hurts to iterate longer, but such calculations hardly bring added value.
+
+- Convergence diagnostics keep improving substantially until n_it = 20-30
+
+- Performance measures do not improve after n_it = 9
+
+- [methodological explanation is that rhat and ac have a lag (few it to inform your statistic) --\> will always indicate convergence slower than inferential validity is reached]
+
+- Univariate thetas may under-estimate non-convergence.
+
+- Determining non-stationarity with lambda is more difficult than with qhat :(
+
+In short, the validity of iterative imputation stands or falls with algorithmic convergence---or so it's thought.
+We have shown that iterative imputation algorithms can yield correct outcomes even when a converged state has not yet formally been reached.
+Any further iterations just burn computational resources without improving the statistical inferences.
+
+# Case Study
+
+- We use real data: the `boys` dataset from the `mice` package
+
+- We are interested in predicting age from the other variables, in particular in the regression coefficient of `hgt`
+
+- We compare non-convergence identified using visual inspection versus rhat in the chain variances, scientific estimate and lambda.
+
+- The figures show results of a `mice` run with 20 iterations but otherwise default settings.
+
+From the traceplot of the chain means (see \@ref(fig:case)A) it seems that mixing improves up-to 10 iterations, while trending is only apparent in the first three iterations.
+
+This figure (\@ref(fig:case)B) shows that 7 iterations are required before the $\widehat{R}$-values of the chain means drop below the threshold for non-convergence.
+
+The $\widehat{R}$-values for the scientific estimate reaches the threshold much sooner, when $n_{it}=14$ (see \@ref(fig:case)C).
+
+According to the $\widehat{R}$-values with $\lambda$ as parameter, at least 15 iterations are required (see \@ref(fig:case)D).
+
 # Introduction (new-ish)
@@
 
-- can we quantify non-convergence quantitatively?
-- which parameter should we track?
+- can we diagnose non-convergence quantitatively?
+- ~~which parameter should we track?~~
 - how many iterations are needed before convergence is reached?

### 2025-04-17 Manuscript/manuscript.qmd
tables in methods section

@@
 # bibliography: references.bib
+execute:
+  echo: false
+  message: false
+  warning: false
+format: pdf
 bibliography: references.bib
-format: pdf
 ---
@@
 performance <- evaluate_est(out)
-performance[performance$term == "Y", "term"] <- "b0"
+# performance[performance$term == "b0", "term"] <- "Y"
 results <- # simulation results split by condition and regression coefficient
@@
   mutate(mech = factor(mech, levels = c("MCAR", "MAR"), ordered = TRUE),
-         prop = factor(prop),
-         term = factor(term))
+         prop = factor(prop, levels = c(0.25, 0.5, 0.75), ordered = TRUE),
+         term = factor(term, levels = c("Y", "X1", "X2", "X3"), ordered = TRUE))
 ```
@@
 With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using the imputation package `{mice}` [@mice].
-Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$).
-To assess non-convergence, we terminate the imputation algorithm after a varying number of iterations ($n_{it} = 1, 2, ..., 50$).
+Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 2000$).
+To assess non-convergence, we terminate the imputation algorithm after a varying number of iterations ($n_{it} = 1, 2, ..., 100$).
 We differentiate between nine different missingness scenarios that are defined in the data generating mechanism.
@@
 
-We generate one complete set of $n_{obs} = 1000$ cases in each simulation repetition. This set consists of three multivariately normal random variables, such that
+We generate a complete set of $n_{\mathrm{obs}} = 200$ cases in each simulation repetition. This set consists of three multivariately normal random variables, such that
 
@@
 \begin{pmatrix}
-Y\\
 X_1\\
-X_2
+X_2\\
+X_3
 \end{pmatrix}
@@
 \begin{pmatrix}
-2     &       &  \\
+1     &       &  \\
 0.5   & 1     &  \\
--0.5  & 0.5   & 1
+0.5  & 0.5   & 1
 \end{pmatrix}
-\end{bmatrix}.
-$$
-
-The complete set is 'amputed' (i.e., made incomplete) according to nine missingness conditions.
-We use a $3 \times 3$ factorial design with three missingness mechanisms and three proportions of incomplete cases, see Figure XYZ.
+\end{bmatrix},
+$$ 
+and an outcome variable $Y$ as a linear combination of the three predictors. For each unit $i = 1, 2,..., n_\mathrm{obs}$,
+
+$$
+Y_i = X_{1i} + X_{2i} + X_{3i} + \epsilon_i ,
+$$
+where $\epsilon \sim \mathcal{N}(0, 1)$.
+This results in a complete dataset all units $i$ on all variables $j$ (where $j =  Y, X_1, X_2, X_3$) in each simulation repetition.
+Multivariate normal data are generated using the `mvtnorm` package [@mvtnorm].
+
+Subsequently, the complete set is 'amputed' (i.e., made incomplete) according to six missingness conditions.
+We use a $2 \times 3$ factorial design with three missingness mechanisms and three proportions of incomplete cases, see Table XYZ.
 Missingness mechanisms refer to the probability of being missing for any given entry in a dataset.
 There are three distinct types of mechanisms, as defined by Rubin (1976).
-Roughly translated, the probability of being missing may be equal for all entries (MCAR; Missing Completely At Random), the probability of being missing depends on observed information (MAR; Missing At Random), or the probability depends on *un*observed information, making the missingness non-ignorable (MNAR; Missing Not At Random).
-[add left-tailed M(N)AR mechanism, md pattern, and maybe refer to ampute paper or "Dance of..." by Schouten 2018].
+Roughly translated, the probability of being missing may be equal for all entries (MCAR; Missing Completely At Random), may depend on observed information (MAR; Missing At Random), or may depend on *un*observed information, making the missingness non-ignorable (MNAR; Missing Not At Random).
+[TODO: add left-tailed M(N)AR mechanism, md pattern, and maybe refer to ampute paper or "Dance of..." by Schouten 2018; TODO: add `ampute()`].
+
+\begin{table}[]
+\begin{tabular}{lcr}
+\hline
+\multicolumn{3}{l}{Missingness conditions} \\ \hline
+missing data mechanism &  & proportion of incomplete cases  \\
+                       & $\times$  &                         \\
+MCAR, MAR              &   & 25\%, 50\%, 75\%                \\ \hline
+\end{tabular}
+\end{table}
 
@@
 On each imputed dataset, we perform multiple linear regression as our analysis of scientific interest.
-We predict outcome variable $Y$ from the other two variables,
-
-$$
-\hat{Y} = \beta_0 + \beta_1 X_1 + \beta_2 X_2,
-$$
-
-where $\hat{Y}$ is the predicted value of the outcome.
-Our estimands are the regression coefficient $\beta_1$ and coefficient of determination $\rho^2$ that we obtain after pooling the regression results across the imputations.
+We predict outcome variable $Y$ from the other variables,
+
+$$
+\hat{Y} = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,
+$$
+
+where $\hat{Y}$ is the predicted value of the outcome. Our estimands are the regression coefficients $\boldsymbol{\beta}$, including the intercept $\beta_0$.
 
@@
 
+\begin{table}[]
+\begin{tabular}{lcccr}
+\hline
+\multicolumn{5}{l}{Non-convergence diagnostics}   \\ \hline
+identifier                  &          & parameter                    &          & variable                 \\
+                            & $\times$ &                              & $\times$ &                          \\
+AC, $\widehat{R}$           &          & chain means, chain variances &          & $Y$, $X_1$, $X_2$, $X_3$ \\ \hline
+\end{tabular}
+\end{table}
+
 - non-convergence in iterative algorithms is diagnosed using an identifier and a parameter
@@
 
-- we apply these identifiers to four different univariate and multivariate parameters
-
-- the univariate parameters are those commonly used for visual inspection: chain means and chain variances
-
-- one multivariate parameter that is of immediate interest for our estimand is the estimated value itself: regression coefficient beta_1
-
-- we also propose a new multivariate parameter that is not dependent on the model of scientific interest: the first eigenvalue of the variance-covariance matrix of the completed data [explain!]
-
-We use eight different methods to diagnose non-convergence: a combination of two non-convergence identifiers---autocorrelation and $\widehat{R}$---and four parameters---chain means, chain variances, $\beta_1$, and $\lambda$.
+- we apply these identifiers to the two parameters that we typically evaluate after imputation using visual inspection: chain means and chain variances
+
+We use two non-convergence identifiers---autocorrelation and $\widehat{R}$---to diagnose non-convergence in the imputation models of the four incomplete variables.
+For each variable we apply the two identifiers on two parameters---chain means and chain variances---and four variables, resulting in 16 sets of identifier-parameter-variable pairs.
+
+$$
+\begin{aligned}
+\hline
+&\mathrm{Non-convergence \ diagnostics} & & & & \\
+\hline
+&\begin{matrix}
+\widehat{R} \\
+\mathrm{AC } 
+\end{matrix}
+\quad
+&\times
+\quad
+&\begin{matrix}
+\mathrm{chain \ means} \\
+\mathrm{chain \ variances} 
+\end{matrix}
+\quad
+&\times
+\quad
+&\begin{matrix}
+Y\\
+X_1\\
+X_2\\
+X_3
+\end{matrix} \\
+&\mathrm{identifier} &&\mathrm{parameter} &&\mathrm{variable} \\
+\hline
+\end{aligned}
+$$
 
@@
 
-As recommended by van Buuren, our performance measures are bias, confidence interval width, and coverage rate of the estimands (-@buur18 $\S$ 2.5.2).
+As recommended by van Buuren (-@buur18 $\S$ 2.5.2), our performance measures are bias, confidence interval width, and coverage rate of the estimands.
 These measures are averaged across the simulation repetitions.
@@
 
-The following figures display the simulation results for the eight diagnostic methods and four performance measures, contrasted to the number of iterations in the imputation algorithm.
-Within the figures, we split the results according to the missingness conditions [missingness mechanisms as line types, and proportion of incomplete cases as colors].
-Note that these results are averages of the $n_{sim} = 1000$ simulation repetitions.
+The following figures display the simulation results for the sixteen diagnostic identifier-parameter-variable pairs, and sixteen performance measure-variable pairs, contrasted to the number of iterations in the imputation algorithm.
+Within the figures, we split the results according to the missingness conditions [TODO: missingness mechanisms as line types, and proportion of incomplete cases as colors?].
+Note that these results are averages of the $n_{sim} = 2000$ simulation repetitions.
 
@@
 
-Figure \@ref(fig:ac) shows the autocorrelations in the chain means (panel A), chain variances (panel B), regression estimates (panel C), and eigenvalues (panel D).
+Figure \@ref(fig:ac) shows the autocorrelations in the chain means (panel A) and chain variances (panel B).
+
+```{r ac, fig.ncol=2}
+results |>
+  filter(method == "MICE") |> 
+  ggplot(aes(x = .it, y = ac_mean, color = mech, linetype = mech)) +
+  geom_hline(yintercept = 0, alpha = 0.1) + 
+  geom_line() +
+  facet_grid(term ~ prop) +
+  theme_classic()
+
+results |>
+  filter(method == "MICE") |> 
+  ggplot(aes(x = .it, y = ac_sd, color = mech, linetype = mech)) +
+  geom_hline(yintercept = 0, alpha = 0.1) + 
+  geom_line() +
+  facet_grid(term ~ prop) +
+  theme_classic()
+```
 
@@
 These results hold irrespective of the missingness condition.
-
 Autocorrelation in the chain variances show us something similar (see \@ref(fig:ac)B).
@@
 
-There is more autocorrelation in the scientific estimates than in the chain means and chain variances (see \@ref(fig:ac)C).
-We observe the highest autocorrelations in conditions where 75% of cases are incomplete.
-Overall, the autocorrelations reach a plateau when $n_{it} \geq 20$ to $30$.
-There is no clear effect of the missingness mechanisms.
-
-The autocorrelation in the eigenvalues exhibits a similar trend to the autocorrelation in the scientific estimates (see \@ref(fig:ac)D). Trending in this parameter diminishes when $n_{it} \geq 20$.
-
-Figure \@ref(fig:rh) shows the potential scale reduction factor in the chain means (panel A), chain variances (panel B), regression estimates (panel C), and eigenvalues (panel D).
-
-We observe that $\widehat{R}$-values of the chain means generally decreases as a function of the number of iterations (see \@ref(fig:rh)A).
-An exception to this observation is the initial increase when $3 \leq n_{it} \leq 5$ [interpret??
+```{r psrf, fig.ncol=2}
+results |>
+  filter(method == "MICE") |> 
+  ggplot(aes(x = .it, y = psrf_mean, color = mech, linetype = mech)) +
+  geom_hline(yintercept = 1, alpha = 0.1) + 
+  geom_line() +
+  facet_grid(term ~ prop) +
+  theme_classic()
+
+results |>
+  filter(method == "MICE") |> 
+  ggplot(aes(x = .it, y = psrf_sd, color = mech, linetype = mech)) +
+  geom_hline(yintercept = 1, alpha = 0.1) + 
+  geom_line() +
+  facet_grid(term ~ prop) +
+  theme_classic()
+```
+
+Figure \@ref(fig:rh) shows the potential scale reduction factor in the chain means (panel A) and chain variances (panel B).
+
+We observe that $\widehat{R}$-values of the chain means generally decrease as a function of the number of iterations (see \@ref(fig:rh)A).
+An exception to this observation is a steep increase in iterations $3 \leq n_{it} \leq 5$ [TODO: interpret??
 due to initialization or is there really more mixing initially??].
@@
 There is no apparent differentiation between the missingness conditions.
-
 The mixing between chain variances mimics the mixing between chain means almost perfectly (see \@ref(fig:rh)B).
 Irrespective of the missingness condition, the $\widehat{R}$-values taper off around $n_{it} \geq 30$.
-
-With the regression estimate as parameter we observe very similar $\widehat{R}$-values than with chain means and chain variances (see \@ref(fig:rh)C).
-We do, however, see some differences between missingness conditions.
-Conditions where 75% of the cases are incomplete show more extreme non-mixing.
-The overall trend remains the same: about 30 iterations are required before mixing stops improving substantially.
-
-$\widehat{R}$-values of the eigenvalues show a trend similar to the chain means and chain variances (see \@ref(fig:rh)D). [add something about conditions??]
-
-[These rhat plots all show some initialization before the fifth iteration: is rhat useful before that??]
+[The rhat plots all show some initialization before the fifth iteration: is rhat useful before that??]
 
@@
 
-In Figure \@ref(fig:perf) we show the performance measures: bias in the regression estimate (panel A), bias in the coefficient of determination (panel B), the empirical coverage rate of the regression estimate (panel C) and the average confidence interval width of this estimate (panel D).
+```{r bias}
+results |>
+  filter(method == "MICE") |> 
+  ggplot(aes(x = .it, y = bias, color = mech, linetype = mech)) +
+  geom_hline(yintercept = 0, alpha = 0.1) + 
+  geom_line() +
+  facet_grid(term ~ prop) +
+  theme_classic()
+```
+
+- more missingness = more extreme bias, but also steeper decrease over iterations
+
+```{r cov}
+results |>
+  filter(method == "MICE") |> 
+  ggplot(aes(x = .it, y = cov, color = mech, linetype = mech)) +
+  geom_hline(yintercept = .95, alpha = 0.1) + 
+  geom_line() +
+  facet_grid(term ~ prop) +
+  theme_classic()
+```
+
+- no clear trend
+
+```{r ciw}
+results |>
+  filter(method == "MICE") |> 
+  ggplot(aes(x = .it, y = ciw, color = mech, linetype = mech)) +
+  geom_hline(yintercept = .95, alpha = 0.1) + 
+  geom_line() +
+  facet_grid(term ~ prop) +
+  theme_classic()
+```
+
+- more missingness = wider CIs, with steeper decrease in CIW, but not stababiling at same value between missingness conditions.
+
+In Figure \@ref(fig:perf) we show the performance measures: bias in the regression estimate (panel A), the empirical coverage rate of the regression estimate (panel C) and the average confidence interval width of this estimate (panel D).
 
@@
 
-Equivalent to the bias in the regression estimate, the bias in the coefficient of determination tapers off within a couple of iterations (see \@ref(fig:perf)B).
-We observe stable estimates in all conditions when $n_{it} \geq 6$ [interpret the over-estimation in MNAR+75% condition?].
-
 Nominal coverage is quickly reached (see \@ref(fig:perf)C).
@@
 - [methodological explanation is that rhat and ac have a lag (few it to inform your statistic) --\> will always indicate convergence slower than inferential validity is reached]
-
-- Univariate thetas may under-estimate non-convergence.
-
-- Determining non-stationarity with lambda is more difficult than with qhat :(
 
@@
 - how many iterations are needed before valid inferences are reached?
-
-```{r}
-results |>
-  filter(method == "MICE") |> #, prop == 0.75 mech == "MAR",
-  ggplot(aes(x = .it, y = bias, color = prop)) +
-  geom_hline(yintercept = 0, alpha = 0.1) + 
-  geom_line() +
-  facet_grid(term ~ mech) +
-  theme_classic()
-
-results |>
-  filter(method == "MICE") |> #, prop == 0.75 mech == "MAR",
-  ggplot(aes(x = .it, y = psrf_mean, color = prop)) +
-  geom_hline(yintercept = 1, alpha = 0.1) + 
-  geom_line() +
-  facet_grid(term ~ mech) +
-  theme_classic()
-
-```
 

### 2025-04-17 R/apply_methods.R

@@ -12,7 +12,7 @@ apply_full <- function(amp) {
   # add method name and missingness
   est <- cbind(method = "full", mech = amp$mech, prop = amp$prop, .it = 0, est, ac_mean = NA, psrf_mean = NA, ac_sd = NA, psrf_sd = NA)
   # rename "(Intercept)" to "Y" for easier processing
-  est[est$term == "(Intercept)", "term"] <- "b0"
+  est[est$term == "(Intercept)", "term"] <- "Y"
   # output
   return(est)
 }
@@ -29,7 +29,7 @@ apply_CCA <- function(amp) {
   # add method name and missingness
   est <- cbind(method = "CCA", mech = amp$mech, prop = amp$prop, .it = 0, est, ac_mean = NA, psrf_mean = NA, ac_sd = NA, psrf_sd = NA)
   # rename "(Intercept)" to "Y" for easier processing
-  est[est$term == "(Intercept)", "term"] <- "b0"
+  est[est$term == "(Intercept)", "term"] <- "Y"
   # output
   return(est)
 }
@@ -75,7 +75,7 @@ estimate_param <- function(imp) {
     # select estimates
     dplyr::select(term, estimate, conf.low, conf.high)
   # rename "(Intercept)" to "Y" for easier processing
-  est[est$term == "(Intercept)", "term"] <- "b0"
+  est[est$term == "(Intercept)", "term"] <- "Y"
   # output
   return(est)
 }

### 2025-04-17 Manuscript/manuscript.qmd
merge methods sections

@@
   warning: false
+  fig_width: 6
 format: pdf
@@
 performance <- evaluate_est(out)
-# performance[performance$term == "b0", "term"] <- "Y"
+performance[performance$term == "b0", "term"] <- "Y"
 results <- # simulation results split by condition and regression coefficient
@@
 
+## Aims
+
+- can we diagnose non-convergence quantitatively?
+- ~~which parameter should we track?~~
+- how many iterations are needed before convergence is reached?
+- how many iterations are needed before valid inferences are reached?
+
 # Introduction (newest)
@@
 
-We investigate non-convergence in iterative imputation through model-based simulation in R (version 4.4.2; R Core Team 2024).
+We investigate non-convergence in iterative imputation through model-based simulation in R (version 4.4.2; @R).
 We provide a summary of the simulation set-up in Algorithm 1; the complete script and technical details are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint) [TODO: make this a Zenodo DOI].
+The number of simulation repetitions $n_{\mathrm{sim}} = 2000$.
 
@@
 
+```
+# pseudo-code of simulation 
+for (number of simulation runs) {
+  1. simulate complete data 
+  for (missingness conditions) {
+    2. create missingness
+    for (number of iterations) {
+      3. impute missingness
+      4. estimate quantities of scientific interest
+      5. apply performance measures to the estimates
+      6. compute non-convergence diagnostics
+    }
+  }
+7. aggregate outcomes across simulation runs
+```
+
 ## Aims
@@
 With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using the imputation package `{mice}` [@mice].
-Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 2000$).
-To assess non-convergence, we terminate the imputation algorithm after a varying number of iterations ($n_{it} = 1, 2, ..., 100$).
-We differentiate between nine different missingness scenarios that are defined in the data generating mechanism.
+Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{\mathrm{sim}} = 2000$).
+To evaluate convergence, we terminate the imputation algorithm after a varying number of iterations ($n_{\mathrm{it}} = 1, 2, ..., 100$).
+We differentiate between six different missingness scenarios that are defined in the data generating mechanism.
 
@@
 
-We generate a complete set of $n_{\mathrm{obs}} = 200$ cases in each simulation repetition. This set consists of three multivariately normal random variables, such that
+In each simulation repetition, we first generate a complete set of $n_{\mathrm{obs}} = 200$ cases, representing person-data in a multiple linear regression problem.
+The predictor space consists of three multivariately normal random variables,
 
@@
 \end{pmatrix}
-\end{bmatrix},
+\end{bmatrix}.
 $$ 
-and an outcome variable $Y$ as a linear combination of the three predictors. For each unit $i = 1, 2,..., n_\mathrm{obs}$,
+The outcome variable $Y$ is a linear combination of the three predictors, such that for each unit $i = 1, 2,..., n_\mathrm{obs}$,
 
@@
 where $\epsilon \sim \mathcal{N}(0, 1)$.
-This results in a complete dataset all units $i$ on all variables $j$ (where $j =  Y, X_1, X_2, X_3$) in each simulation repetition.
+This results in a complete dataset of size $n_{\mathrm{obs}} \times p$, with units $i = 1, 2,..., n_\mathrm{obs}$ and variables $j =  Y, X_1, X_2, X_3$).
 Multivariate normal data are generated using the `mvtnorm` package [@mvtnorm].
 
-Subsequently, the complete set is 'amputed' (i.e., made incomplete) according to six missingness conditions.
-We use a $2 \times 3$ factorial design with three missingness mechanisms and three proportions of incomplete cases, see Table XYZ.
-Missingness mechanisms refer to the probability of being missing for any given entry in a dataset.
+Subsequently, the complete are 'amputed' (i.e., made incomplete) according to six missingness conditions.
+We use a $2 \times 3$ factorial design with two missing data mechanisms and three proportions of incomplete cases, see Table XYZ.
+
+[TODO: make this a box] Missingness mechanisms refer to the probability of being missing for any given entry in a dataset.
 There are three distinct types of mechanisms, as defined by Rubin (1976).
 Roughly translated, the probability of being missing may be equal for all entries (MCAR; Missing Completely At Random), may depend on observed information (MAR; Missing At Random), or may depend on *un*observed information, making the missingness non-ignorable (MNAR; Missing Not At Random).
-[TODO: add left-tailed M(N)AR mechanism, md pattern, and maybe refer to ampute paper or "Dance of..." by Schouten 2018; TODO: add `ampute()`].
+
+The missing data mechanisms under consideration are 'missing completely at random' [MCAR; @rubin87], where the probability to be missing is the same for all $n_\mathrm{obs} \times p$ cells in $y$, and a right-tailed 'missing at random' (MAR) mechanism, where the probability to be missing is a function of the observed data, and higher values are more likely to be missing.
+The proportion of incomplete cases $\pi_{\mathrm{inc}}$ is set to 25%, 50%, and 75%.
+We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\mathrm{obs}}, y_{\mathrm{mis}}$} for every missingness condition.
+The missing data patterns are all possible combinations of multivariate missingness.
+
+[TODO: add left-tailed M(N)AR mechanism, md pattern, and maybe refer to ampute paper or "Dance of..." by Schouten 2018?].
 
@@
 On each imputed dataset, we perform multiple linear regression as our analysis of scientific interest.
-We predict outcome variable $Y$ from the other variables,
+Our estimands are the regression coefficients $\boldsymbol{\beta}$,
 
@@
 
-where $\hat{Y}$ is the predicted value of the outcome. Our estimands are the regression coefficients $\boldsymbol{\beta}$, including the intercept $\beta_0$.
+where $\hat{Y}$ is the predicted value of the outcome.
+We estimate these coefficients using the `lm()` function [@R] in each imputed dataset, and subsequently pool the estimates across the five imputations using Rubin's rules [function `mice::pool()`; @mice].
+
+<!-- All imputation procedures are performed using `mice` [function `mice()`; @mice], with Bayesian linear regression imputation, and five imputation chains ($m=5$).
+This results in $m=5$ sets of imputations for each of the conditions, $\dot y_{\mathrm{imp}, \ell}$.
+-->
+
+<!-- Subsequently, we obtain the completed data per imputation, {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice].
+Estimates of scientific quantities $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R].
+We pool the resulting $m$ estimates, $\hat{Q}_\ell$, into a single $\bar{Q}$ conform Rubin's rules [function `mice::pool()`; @mice].
+-->
+
+## Performance measures
+
+We evaluate the pooled estimates against their true population values using the performance measures bias, confidence interval width, and coverage rate, as recommended by van Buuren (-@buur18 $\S$ 2.5.2).
+We calculate bias as $\bar{Q} - Q$.
+CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$.
+Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$.
+CIW is of interest because it is a measure of efficiency.
+Under nominal coverage, short CIs are preferred, since wider CIs indicate lower statistical power.
+
+<!-- We evaluate the diagnostic methods to identify non-convergence against these performance measures.
+$\widehat{R}$ and $AC$ should identify non-convergence in simulation conditions, where $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
+-->
 
@@
 
-$$
-\begin{aligned}
-\hline
-&\mathrm{Non-convergence \ diagnostics} & & & & \\
-\hline
-&\begin{matrix}
-\widehat{R} \\
-\mathrm{AC } 
-\end{matrix}
-\quad
-&\times
-\quad
-&\begin{matrix}
-\mathrm{chain \ means} \\
-\mathrm{chain \ variances} 
-\end{matrix}
-\quad
-&\times
-\quad
-&\begin{matrix}
-Y\\
-X_1\\
-X_2\\
-X_3
-\end{matrix} \\
-&\mathrm{identifier} &&\mathrm{parameter} &&\mathrm{variable} \\
-\hline
-\end{aligned}
-$$
-
-## Performance measures
-
-As recommended by van Buuren (-@buur18 $\S$ 2.5.2), our performance measures are bias, confidence interval width, and coverage rate of the estimands.
-These measures are averaged across the simulation repetitions.
-<!-- [or just of the regression estimate?? otherwise add the ciw and cov of r2 too!!]. -->
+<!-- We consider eight sets of diagnostic methods to identify non-convergence, by applying $\widehat{R}$ and $AC$ on four different $\theta$s.
+Univariate $\theta$s are obtained from $\dot y_{\mathrm{imp}, \ell}$.
+The model-independent multivariate $\theta$, $\lambda_{1}$, is obtained from {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$} as the variance in the first PCA component [function `stats::princomp`; R].
+The model-dependent multivariate $\theta$ is one of the quantities of scientific interest: the estimated regression coefficients in {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}.
+We apply the two non-convergence diagnostics on the $\theta$s by implementing the adapted version of $\widehat{R}$ and by manually programming the $AC$ function.
+-->
 
@@
 
-```{r ac, fig.ncol=2}
+```{r ac_mean}
 results |>
@@
   theme_classic()
-
+```
+
+```{r ac_sd}
 results |>
@@
 
-```{r psrf, fig.ncol=2}
+```{r psrf_mean}
 results |>
@@
   theme_classic()
-
+```
+
+```{r psrf_sd}
 results |>
@@
 
-## Aims
-
-- can we diagnose non-convergence quantitatively?
-- ~~which parameter should we track?~~
-- how many iterations are needed before convergence is reached?
-- how many iterations are needed before valid inferences are reached?
-
 # Introduction (old)
@@
 Figure \ref{fig:diagram} provides an overview of the steps involved with MI.
-The missing part $y_{\mathrm mis}$ of an incomplete dataset is imputed $m$ times.
-This creates $m$ sets of imputed data $\dot y_{\mathrm imp, \ell}$, where $\ell = 1, 2, ..., m$.
-The imputed data is then combined with the observed data $y_{\mathrm obs}$ to create $m$ completed datasets.
+The missing part $y_{\mathrm{mis}}$ of an incomplete dataset is imputed $m$ times.
+This creates $m$ sets of imputed data $\dot y_{\mathrm{imp}, \ell}$, where $\ell = 1, 2, ..., m$.
+The imputed data is then combined with the observed data $y_{\mathrm{obs}}$ to create $m$ completed datasets.
 On each of these datasets the analysis of scientific interest is performed to estimate $Q$: the quantity of scientific interest (e.g., a regression coefficient).
@@
 The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing.
-The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\mathrm inc}$ in dataset $y$.
-The collection of observed data values in $y$ is denoted by $y_{\mathrm obs}$; the missing part of $y$ is referred to as $y_{\mathrm mis}$.
-For each datapoint in $y_{\mathrm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations in the imputation algorithm ($t = 1, 2, \dots, T$).
+The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\mathrm{inc}}$ in dataset $y$.
+The collection of observed data values in $y$ is denoted by $y_{\mathrm{obs}}$; the missing part of $y$ is referred to as $y_{\mathrm{mis}}$.
+For each datapoint in $y_{\mathrm{mis}}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations in the imputation algorithm ($t = 1, 2, \dots, T$).
 The state-space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values).
@@
 Each line portrays one imputation chain (i.e., the values of a scalar summary $\theta$ across iterations).
-The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in the $m$ sets of imputations $\dot y_{\mathrm imp, \ell}$.
+The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in the $m$ sets of imputations $\dot y_{\mathrm{imp}, \ell}$.
 
@@
 The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
-It is obvious that non-convergence in this example impacts the distribution of the imputed values per imputation $\dot y_{\mathrm imp, \ell}$.
-This effect may translate into the distribution of the $m$ sets of completed data {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$, and finally the pooled estimate $\bar{Q}$.
+It is obvious that non-convergence in this example impacts the distribution of the imputed values per imputation $\dot y_{\mathrm{imp}, \ell}$.
+This effect may translate into the distribution of the $m$ sets of completed data {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$, and finally the pooled estimate $\bar{Q}$.
 $\bar{Q}$ is then a  biased, invalid estimate of $Q$.
@@
 \geq \lambda_{j, \ell}$ be the eigenvalues of $S_\ell$ in each imputation.
-The first eigenvalue, $\lambda_{1,\ell}$, then summarizes the largest possible amount of covariance in each completed dataset {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$}.
+The first eigenvalue, $\lambda_{1,\ell}$, then summarizes the largest possible amount of covariance in each completed dataset {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}.
 By definition, $\lambda_{1}$-values are equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on the completed data.
@@
 To induce non-convergence, we consider two sets of simulation conditions: 'missingness' and 'early stopping'.
-Missingness refers to the severity of the missing data problem, which we determine by the proportion of incomplete cases in dataset $y$, $p_{\mathrm inc}$.
+Missingness refers to the severity of the missing data problem, which we determine by the proportion of incomplete cases in dataset $y$, $p_{\mathrm{inc}}$.
 Early stopping is defined by the number of iterations in the imputation algorithm, $T$.
@@
 The missingness conditions are chosen to reflect the difficulty of the missingness problem.
-The underlying assumption is that a low proportion of incomplete cases $p_{\mathrm inc}$ leads to quick algorithmic convergence, since there is a lot of information in the observed data.
-A higher $p_{\mathrm inc}$ yields slower convergence---unless there is so little information in $y_{\mathrm obs}$ that this is outweighed by the random component in the imputation algorithm.
+The underlying assumption is that a low proportion of incomplete cases $p_{\mathrm{inc}}$ leads to quick algorithmic convergence, since there is a lot of information in the observed data.
+A higher $p_{\mathrm{inc}}$ yields slower convergence---unless there is so little information in $y_{\mathrm{obs}}$ that this is outweighed by the random component in the imputation algorithm.
 Then, convergence to a stable but highly variable state may be reached instantly.
-Still, we expect that missingness conditions with higher $p_{\mathrm inc}$ will result in more signs of non-convergence.
+Still, we expect that missingness conditions with higher $p_{\mathrm{inc}}$ will result in more signs of non-convergence.
 
@@
 1.
-We expect that simulation conditions with a high proportion of incomplete cases $p_{\mathrm inc}$ and a low number of iterations $T$ will more often result in biased, invalid estimates of the quantities of scientific interest $Q$.
+We expect that simulation conditions with a high proportion of incomplete cases $p_{\mathrm{inc}}$ and a low number of iterations $T$ will more often result in biased, invalid estimates of the quantities of scientific interest $Q$.
 
@@
 where $\epsilon \sim \mathcal{N}(0, 1)$.
-This results in a dataset $y$, where $y_{\mathrm mis} = \emptyset$.
+This results in a dataset $y$, where $y_{\mathrm{mis}} = \emptyset$.
 In other words, $y$ contains the complete set of observations for all units $i$ on all variables $j$ (where $j =  Y, X_1, X_2, X_3$).
@@
 The missing data mechanisms are 'missing completely at random' [MCAR; @rubin87], where the probability to be missing is the same for all $N \times k$ cells in $y$, and a right-tailed 'missing at random' (MAR) mechanism, where the probability to be missing is a function of the observed data and higher values are more likely to be missing.
-The proportion of incomplete cases $p_{\mathrm inc}$ is set to 25%, 50%, and 75%.
+The proportion of incomplete cases $p_{\mathrm{inc}}$ is set to 25%, 50%, and 75%.
 
@@
 <!-- proportion of incomplete cases: 25%, 50%, 75% -->
-
-<!-- \begin{table}[] -->
-<!-- \begin{tabular}{lllll} -->
-<!-- \hline -->
-<!-- Missingness &  &                 &  & Early stopping            \\ \cline{1-3} \cline{5-5}  -->
-<!-- Mechanism   &  & Proportion      &  & Iteration number          \\ \cline{1-1} \cline{3-3} \cline{5-5}  -->
-<!-- MCAR, MAR   &  & 0.25, 0.5, 0.75 &  & 1, 2, 3, ..., 98, 99, 100 \\ \hline -->
-<!-- \end{tabular} -->
-<!-- \end{table} -->
-
-$$
-\begin{aligned}
-\hline
-&\mathrm{Missingness \ conditions} & & & &\mathrm{Early \ stopping \ conditions} \\
-\hline
-&\begin{matrix}
-\mathrm{MCAR} \\
-\mathrm{MAR \ } 
-\end{matrix}
-\quad
-&\times
-\quad
-&\begin{matrix}
-25\% \\
-50\% \\
-75\%
-\end{matrix}
-\quad
-&\times
-\quad
-&\begin{matrix}
-1 \\
-2 \\
-... \\
-99 \\
-100 \\
-\end{matrix} \\
-&\mathrm{mechansim} &&\mathrm{incomplete \ cases} &&\mathrm{iteration(s)} \\
-\hline
-\end{aligned}
-$$
 
@@
 
-We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\mathrm obs}, y_{\mathrm mis}$} for each of the six missingness conditions.
-
-We impute the missing values in {$y_{\mathrm obs}, \dot y_{\mathrm mis}$} according to the early stopping conditions.
+We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\mathrm{obs}}, y_{\mathrm{mis}}$} for each of the six missingness conditions.
+
+We impute the missing values in {$y_{\mathrm{obs}}, \dot y_{\mathrm{mis}}$} according to the early stopping conditions.
 That is, we vary the number of iterations in the imputation algorithm between one and one hundred ($T = 1,2,.,100$).
 All imputation procedures are performed using `mice` [function `mice()`; @mice], with Bayesian linear regression imputation, and five imputation chains ($m=5$).
-This results in $m=5$ sets of imputations for each of the one hundred early stopping conditions, $\dot y_{\mathrm imp, \ell}$.
-
-Subsequently, we obtain the completed data per imputation, {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice].
+This results in $m=5$ sets of imputations for each of the one hundred early stopping conditions, $\dot y_{\mathrm{imp}, \ell}$.
+
+Subsequently, we obtain the completed data per imputation, {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice].
 Estimates of scientific quantities $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R].
@@
 We consider eight sets of diagnostic methods to identify non-convergence, by applying $\widehat{R}$ and $AC$ on four different $\theta$s.
-Univariate $\theta$s are obtained from $\dot y_{\mathrm imp, \ell}$.
-The model-independent multivariate $\theta$, $\lambda_{1}$, is obtained from {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$} as the variance in the first PCA component [function `stats::princomp`; R].
-The model-dependent multivariate $\theta$ is one of the quantities of scientific interest: the estimated regression coefficients in {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$}.
+Univariate $\theta$s are obtained from $\dot y_{\mathrm{imp}, \ell}$.
+The model-independent multivariate $\theta$, $\lambda_{1}$, is obtained from {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$} as the variance in the first PCA component [function `stats::princomp`; R].
+The model-dependent multivariate $\theta$ is one of the quantities of scientific interest: the estimated regression coefficients in {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}.
 We apply the two non-convergence diagnostics on the $\theta$s by implementing the adapted version of $\widehat{R}$ and by manually programming the $AC$ function.
@@
 Let
-$$\text{CI} = \bar{Q} \pm t_{(m-1)} \times \mathrm SE_{\bar{Q}},$$
-where $t_{(M-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $\mathrm SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
+$$\text{CI} = \bar{Q} \pm t_{(m-1)} \times \mathrm{SE}_{\bar{Q}},$$
+where $t_{(M-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $\mathrm{SE}_{\bar{Q}}$ is the square root of the pooled variance estimate.
 If we obtain nominal coverage (CR = 95%), we can conclude that $\bar{Q}$ is a confidence-valid estimate of $Q$.
@@
 This implies that approximately unbiased estimates may be obtained with as little as one iteration ($T \geq 1$).
-The bias in $\bar{Q}$ depends solely on the proportion of incomplete cases, with $p_{\mathrm inc}\geq.75$ leading to more extreme biases than other conditions.
-Completely unbiased estimates are only obtained in conditions where $p_{\mathrm inc}\leq.50$.
+The bias in $\bar{Q}$ depends solely on the proportion of incomplete cases, with $p_{\mathrm{inc}}\geq.75$ leading to more extreme biases than other conditions.
+Completely unbiased estimates are only obtained in conditions where $p_{\mathrm{inc}}\leq.50$.
 This is curious since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.
@@
 Therefore, no iteration at all is needed to obtain approximately unbiased estimates for $Q=\sigma$.
-Completely unbiased estimates are only obtained in conditions where $p_{\mathrm inc}=.05$.
-The other missingness conditions are impacted by non-convergence in the order of increasing $p_{\mathrm inc}$.
+Completely unbiased estimates are only obtained in conditions where $p_{\mathrm{inc}}=.05$.
+The other missingness conditions are impacted by non-convergence in the order of increasing $p_{\mathrm{inc}}$.
 Similar to $Q=\mu$, the magnitude of the bias in $Q=\sigma$ is negligible.
@@
 Therefore, we see that the number of iterations that is necessary to reach stable, non-improving estimates also differs across missingness conditions.
-In conditions where $p_{\mathrm inc}=.05$, estimates are unbiased after one iteration.
-The highest number of iterations necessary to reach approximate unbiasedness is seven ($p_{\mathrm inc}=.95$; $T\geq7$).
-Completely unbiased estimates are only obtained in conditions where $p_{\mathrm inc}\leq.25$.
+In conditions where $p_{\mathrm{inc}}=.05$, estimates are unbiased after one iteration.
+The highest number of iterations necessary to reach approximate unbiasedness is seven ($p_{\mathrm{inc}}=.95$; $T\geq7$).
+Completely unbiased estimates are only obtained in conditions where $p_{\mathrm{inc}}\leq.25$.
 This implies that even for biased estimates, $T\geq7$ would suffice to reach a stable solution.
@@
 As Figure \ref{fig:Qs}D shows, the bias in $\bar{Q}$ is affected by both the proportion of incomplete cases and the number of iterations.
-Similar to $Q=r^2$, we observe approximately unbiased estimates after at least one, and at most seven iterations, depending on $p_{\mathrm inc}$.
-Completely unbiased estimates are only obtained in conditions where $T\geq3$ and $p_{\mathrm inc}\leq .50$.
-Despite persistent bias in conditions where $p_{\mathrm inc}\geq.75$, the coverage rates are non-improving after just three iterations, irrespective of $p_{\mathrm inc}$.
+Similar to $Q=r^2$, we observe approximately unbiased estimates after at least one, and at most seven iterations, depending on $p_{\mathrm{inc}}$.
+Completely unbiased estimates are only obtained in conditions where $T\geq3$ and $p_{\mathrm{inc}}\leq .50$.
+Despite persistent bias in conditions where $p_{\mathrm{inc}}\geq.75$, the coverage rates are non-improving after just three iterations, irrespective of $p_{\mathrm{inc}}$.
 We conclude that any condition where $T\geq3$ yields approximately nominal coverage rates, and thus confidence-valid estimates.
@@
 The CIWs depicted in Figure \ref{fig:Qs}F are conform the theoretical foundation of MI: CIs are wider in conditions with higher proportions of incomplete cases (i.e., there is less information in the data, and thus more uncertainty due to missingness).
-Since conditions with higher $p_{\mathrm inc}$ result in both wider CIs and more severe bias, the true value of $Q$ may be included in the CI, despite of the bias in $\bar{Q}$.
+Since conditions with higher $p_{\mathrm{inc}}$ result in both wider CIs and more severe bias, the true value of $Q$ may be included in the CI, despite of the bias in $\bar{Q}$.
 We conclude that approximately unbiased estimates may be obtained in conditions where $T\geq7$, whereas confidence-valid estimates require at most three iterations.
@@
 Multivariate estimates, by contrast, are affected by both the number of iterations and the proportion of missing cases.
-Completely unbiased estimates are only obtained under low to moderate missingness ($p_{\mathrm inc}\leq.50$), after at most three iterations.
-We observe approximately unbiased estimates after at most seven iterations (for any $p_{\mathrm inc}$ considered).
+Completely unbiased estimates are only obtained under low to moderate missingness ($p_{\mathrm{inc}}\leq.50$), after at most three iterations.
+We observe approximately unbiased estimates after at most seven iterations (for any $p_{\mathrm{inc}}$ considered).
 This implies that the algorithm produces stable, non-improving estimates when $T\geq7$.
@@
 As we concluded in the last section, substantially biased estimates were only obtained in conditions where $T\leq6$, and we observed non-nominal coverage only when $T\leq3$.
-However, there was some persistent bias in conditions where $p_{\mathrm inc}\geq.75$, irrespective of the number of iterations.
+However, there was some persistent bias in conditions where $p_{\mathrm{inc}}\geq.75$, irrespective of the number of iterations.
 If the diagnostic methods are appropriate for MI, we will detect non-convergence in these conditions.
@@
 The $\widehat{R}$-values in Figure \ref{fig:rhats-acs}E show a similar trend across iterations as we observed for chain means and chain variances.
-Only conditions with a very high  proportion of missing cases ($p_{\mathrm inc}\geq.75$) diverge from the earlier observations.
+Only conditions with a very high  proportion of missing cases ($p_{\mathrm{inc}}\geq.75$) diverge from the earlier observations.
 The $\widehat{R}$-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations.
@@
 
-The $AC$-values also follow earlier observations, with the same exception as for $\widehat{R}$ (when $p_{\mathrm inc}\geq.75$).
+The $AC$-values also follow earlier observations, with the same exception as for $\widehat{R}$ (when $p_{\mathrm{inc}}\geq.75$).
 To reach maximum stationarity, these conditions require at most seven iterations instead of five.
@@
 For this $\theta$, the results differ only in the most extreme missingness condition.
-When $p_{\mathrm inc}=.95$, we diagnose non-convergence at $T\leq9$ according to $\widehat{R}>1.2$, $T\leq19$ for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).
+When $p_{\mathrm{inc}}=.95$, we diagnose non-convergence at $T\leq9$ according to $\widehat{R}>1.2$, $T\leq19$ for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).
 The 'dip' in $\widehat{R}$-values at $T=3$ is even more pronounced than with other $\theta$s.
@@
 Under low to moderate missingness, it does not seem to matter on which $\theta$ the non-convergence diagnostics are applied.
-If we look at complete unbiasedness, however, we notice that univariate $\theta$s fail to diagnose the persistent bias in conditions where $p_{\mathrm inc}\geq.75$.
+If we look at complete unbiasedness, however, we notice that univariate $\theta$s fail to diagnose the persistent bias in conditions where $p_{\mathrm{inc}}\geq.75$.
 
@@
 1.
-As hypothesized, biased, invalid estimates of quantities of scientific interest occurred more often in simulation conditions with a high proportion of incomplete cases $p_{\mathrm inc}$ and a low number of iterations $T$.
+As hypothesized, biased, invalid estimates of quantities of scientific interest occurred more often in simulation conditions with a high proportion of incomplete cases $p_{\mathrm{inc}}$ and a low number of iterations $T$.
 However, this only holds for multivariate scientific estimands.
@@
 However, some signs of non-convergence were detected in simulation conditions many more iterations.
-Under moderate missingness ($p_{\mathrm inc}\leq.50$) $\widehat{R}$-values decreased substantially until 30 to 40 iterations, which implies that mixing in the algorithm still improved with each additional iteration.
+Under moderate missingness ($p_{\mathrm{inc}}\leq.50$) $\widehat{R}$-values decreased substantially until 30 to 40 iterations, which implies that mixing in the algorithm still improved with each additional iteration.
 $AC$-values only improved until about six iterations, suggesting minimal improvement in trending was obtained beyond $T=6$ in the current simulation set-up.

### 2025-04-17 figures/patterns.R

@@ -1,0 +1,5 @@
+source("../R/generate_datasets.R")
+pat <- create_patterns(3)
+d <- dat[1:10,]
+is.na(d) <- pat == 1
+ggmice::plot_pattern(d)

### 2025-04-22 Manuscript/manuscript.qmd
add miss pat

@@
 
-```{r patterns, echo=FALSE, eval=FALSE}
+```{r patterns, echo=FALSE}
 source("../R/generate_datasets.R")
 pat <- create_patterns(3)
-d <- dat[1:10,]
-is.na(d) <- pat == 1
-d
-ggmice::plot_pattern(d)
+is.na(pat) <- pat == 1
+pat
+ggmice::plot_pattern(pat, caption = FALSE) +
+  theme(axis.text.x=element_blank(), 
+      axis.ticks.x=element_blank(), 
+      axis.text.y=element_blank(), 
+      axis.ticks.y=element_blank(),
+      axis.text = element_blank()) +
+  theme_void()
 ```

### 2025-04-22 Manuscript/manuscript.qmd
merge intro sections

--- 2025-04-17 Manuscript/manuscript.qmd
@@
 performance[performance$term == "b0", "term"] <- "Y"
-results <- # simulation results split by condition and regression coefficient
-performance |> 
-  group_by(method, mech, prop, .it, term) |> 
-  summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), \(x) mean(x, na.rm = TRUE))) |>
-  mutate(mech = factor(mech, levels = c("MCAR", "MAR"), ordered = TRUE),
+performance <- mutate(performance,
+  mech = factor(mech, levels = c("MCAR", "MAR"), ordered = TRUE),
   prop = factor(prop, levels = c(0.25, 0.5, 0.75), ordered = TRUE),
   term = factor(term, levels = c("Y", "X1", "X2", "X3"), ordered = TRUE))
+results <- performance |> 
+  group_by(method, mech, prop, .it, term) |> 
+  summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), \(x) mean(x, na.rm = TRUE))) 
 ```
@@
 
-# Introduction (newest)
+# Introduction
 
@@
 This analysis will yield unbiased and confidence-valid estimates if--and only if--both the missing data problem and the scientific problem are appropriately considered.
-The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputed values.
+The validity of this whole process naturally depends on the convergence of the algorithm that was used to generate the imputed values.
 Yet, there has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms.
@@
 
+[TODO: intrgrate this section] In an empirical study, where $\widehat{R}$ was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold $\widehat{R}>1.2$.
+Yet, scientific estimates were insensitive to continued iteration from $t=5$ onward [@lace07].
+We, therefore, suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates.
+In contrast to this, signs of non-convergence can be under-estimated by $\widehat{R}$, in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
+437].
+In  `mice`, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values.
+In practice, we do not expect this to cause problems for identifying non-convergence with $\widehat{R}$.
+
+[TODO: intrgrate this section] If the estimated quantities of scientific interest $\bar{Q}$ are unbiased and confidence-valid estimates of ${Q}$, we will conclude that the algorithm is sufficiently converged for practical purposes.
+In contrast to this, we defined *approximate* convergence as the most converged state that the algorithm can obtain.
+Continued iteration after reaching approximate convergence does not yield better mixing or stationarity (indicated by non-improving $\widehat{R}$- and $AC$-values).
+Continued iteration after obtaining valid inferences may lead to a more converged state of the algorithm, but not better estimates.
+
+<!-- Iterative imputation has become the de facto standard to accommodate for missing data.
+The aim is usually to draw valid inferences, i.e.
+to get unbiased, confidence-valid estimates that incorporate the effects of the missingness.
+Such estimates are obtained with iterative imputation by separating the missing data problem from the scientific problem.
+The missing values are imputed (i.e., filled in) using some sort of algorithm.
+And subsequently, the scientific model of interest is performed on the completed data.
+To obtain valid scientific estimates, both the missing data problem and the scientific problem should be appropriately considered.
+The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputations.
+-->
+
+<!-- All inferences with imputed data rely on the convergence of the imputation algorithm.
+Yet, determining whether an algorithm has converged is not trivial.
+There has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms.
+A widely accepted practice is visual inspection of the algorithm.
+Diagnosing convergence through visual inspection, however, may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18].
+Therefore, a quantitative diagnostic method to assess convergence would be preferred.
+-->
+
+<!-- It is challenging to arrive upon a single point at which convergence has been reached.
+Since the aim is to converge to a distribution and not to a single point, the algorithm may produce some fluctuations even after it has converged.
+Because of this property, it may be more desirable to focus on *non*-convergence.
+Fortunately, there are non-convergence identifiers for other iterative algorithms, but the validity of these identifiers has not been systematically evaluated on imputation algorithms.
+-->
+
+<!-- In this study, we explore different methods for identifying non-convergence in iterative imputation algorithms.
+We evaluate whether these methods are able to cover the extent of the non-convergence, and we also investigate the relation between non-convergence and the validity of the inferences.
+We translate the results of our simulation study into guidelines for practice, which we demonstrate by means of a motivating example.
+-->
+
+In this paper we study different methods for assessing non-convergence in iterative imputation algorithms.
+We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications.
+We then address the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in `R` [@R].
+For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package in `R` [@mice].
+The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged.
+And additionally, to evaluate the behavior and performance of several diagnostic methods to identify non-convergence.
+With that, we formulate an informed advice on when it is safe to conclude that the algorithm is converged *enough* for valid inferences.
+We translate the results of the study into guidelines for applied researchers, which may facilitate drawing valid inferences from incomplete data.
+
+# Background
+
+## Some notation
+
+Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample.
+The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing.
+The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\mathrm{inc}}$ in dataset $y$.
+The collection of observed data values in $y$ is denoted by $y_{\mathrm{obs}}$; the missing part of $y$ is referred to as $y_{\mathrm{mis}}$.
+For each datapoint in $y_{\mathrm{mis}}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations in the imputation algorithm ($t = 1, 2, \dots, T$).
+The state-space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values).
+The collection of $\theta$-values between $t=1$ (summarizing the state-space of the algorithm at initialization) and $t=T$ (summarizing the state-space for the imputed values) will be referred to as an 'imputation chain'.
+
+## Iterative imputation
+
+Anyone who analyzes person-data may run into a missing data problem.
+Missing data is not only ubiquitous across science, but treating it can also be a tedious task.
+If a dataset contains just one incomplete observation, calculations are not defined, and statistical models cannot be fitted to the data.
+To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations).
+Unfortunately, this *ad hoc* solution may yield invalid results [@buur18].
+An alternative is to apply imputation.
+With imputation, we 'fill in' the missing values in an incomplete dataset.
+Subsequently, the model of scientific interest can be fitted to the completed dataset.
+By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness.
+This technique is known as 'multiple imputation' [MI; @rubin76].
+MI has proven to be a powerful tool to draw valid inferences from incomplete data under many circumstances [@buur18].
+
+Figure \ref{fig:diagram} provides an overview of the steps involved with MI.
+The missing part $y_{\mathrm{mis}}$ of an incomplete dataset is imputed $m$ times.
+This creates $m$ sets of imputed data $\dot y_{\mathrm{imp}, \ell}$, where $\ell = 1, 2, ..., m$.
+The imputed data is then combined with the observed data $y_{\mathrm{obs}}$ to create $m$ completed datasets.
+On each of these datasets the analysis of scientific interest is performed to estimate $Q$: the quantity of scientific interest (e.g., a regression coefficient).
+Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}_\ell$-values are obtained.
+Finally, the $\hat{Q}_\ell$-values are combined into a single pooled estimate $\bar{Q}$.
+The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence-valid estimate of the true---but unobserved---scientific estimand $Q$ [@rubin96].
+<!-- "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->
+
+```{r diagram, echo=FALSE, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\\hat{Q}$, to a single pooled estimate $\\bar{Q}$."}
+
+knitr::include_graphics("../figures/diagram.pdf") #adapted from [@buur18] section 1.4.1.
+```
+
+A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice].
+With MICE, imputed values are drawn from the posterior predictive distribution of the missing values.
+The algorithm is named after its iterative nature: a multivariate distribution is obtained by iterating over a sequence of univariate imputations.
+Iteration, however, also introduces a potential threat to the validity of the imputations: What if the algorithm has not converged?
+Are the imputations then to be trusted?
+And can we rely on the inference obtained on the completed data?
+
+These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18].
+There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@zhu15; @takahashi2017].
+Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murray2018].
+Therefore, algorithmic convergence should be monitored carefully---although this is not straightforward.
+Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods.
+In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another.
+The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13].
+Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09].
+Several non-convergence diagnostics exist, but it is not known whether these are appropriate within the imputation framework.
+
+<!-- https://www-tandfonline-com.proxy.library.uu.nl/doi/pdf/10.1207/s15327906mbr3304_5?needAccess=true -->
+
+<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->
+
+<!-- The default number of iterations in common MI software: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100??
+[(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p.
+139), Amelia = NA, because re-sampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf).
+-->
+
+## Algorithmic non-convergence
+
+There are two requirements for convergence of iterative algorithms: mixing and stationarity [@gelm13].
+In iterative imputation algorithms, mixing implies that imputation chains intermingle nicely, and stationarity is characterized by the absence of trending across iterations.
+If one of the two requirements is not met, we speak of non-convergence.
+Without mixing, chains may be 'stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values.
+The distribution of imputed values then differs across imputations.
+This may cause under-estimation of the variance between chains, which results in spurious, invalid inferences.
+Without stationarity, there is trending within imputation chains.
+Trending implies that further iterations would yield a systematically lower or higher set of imputations.
+Iterative imputation algorithms that have not (yet) reached stationarity, may thus yield biased estimates.
+
+To illustrate what non-mixing and non-stationarity look like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2].
+Figure \ref{fig:non-conv} displays two scenarios from the example.
+The panel on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm.
+The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
+Each line portrays one imputation chain (i.e., the values of a scalar summary $\theta$ across iterations).
+The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in the $m$ sets of imputations $\dot y_{\mathrm{imp}, \ell}$.
+
+In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending.
+In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.
+Importantly, the chain means at the last iteration (the imputed values per imputation $\ell$) are very different between the two scenarios.
+The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
+It is obvious that non-convergence in this example impacts the distribution of the imputed values per imputation $\dot y_{\mathrm{imp}, \ell}$.
+This effect may translate into the distribution of the $m$ sets of completed data {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$, and finally the pooled estimate $\bar{Q}$.
+$\bar{Q}$ is then a  biased, invalid estimate of $Q$.
+Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.
+
+```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence (reproduced from van Buuren; 2018).
+Depicted are chain means of variable $j$ in each imputation ($m=5$; $T=10$).
+Note that any difference between the two scenarios (see e.g., the magnitude of the chain means on the y-axes) stems from the imputation algorithms---the two scenarios have equal incomplete data $y$."}
+
+# par(mar = c(2, 2, 0.1, 0.1))
+# con_plot
+# non_plot
+
+```
+
+## Identifying non-convergence
+
+Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection.
+After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots.
+In a traceplot, a scalar summary of the state-space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}.
+The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances.
+Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18, $\S$ 6.5.2].
+<!-- Recommendation @buur18: "the different streams should be freely intermingled with one another, without showing any definite trends". -->
+
+<!-- Identifying non-convergence by inspecting traceplots may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18, $\S$ 6.5.2].
+Moreover, traceplots are typically only used to inspect  univariate $\theta$s.
+Monitoring univariate summaries of the state-space of the algorithm may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of the imputations.
+Ideally, we would monitor a multivariate $\theta$.
+-->
+
+<!-- A suggestion by @buur18 for a multivariate $\theta$ to monitor is the estimated quantity of scientific interest $\hat{Q}$, which usually is multivariate in nature.
+This $\theta$ is computed as the estimate in each imputation, $\hat{Q}_\ell$, across iterations.
+Implementing this, however, might be somewhat too technical for empirical researchers.
+Besides, this scalar summary is not model-independent.
+That is, $\theta=\hat{Q}$ is not universal to all complete data problems, while one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently.
+Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest.
+On these grounds, such a $\theta$ can be considered insufficient too.
+-->
+<!-- Researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2].  -->
+
+<!-- As alternative, van Buuren [-@buur18, $\S$ 4.5.2] proposed multivariate evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of @mack03.
+This technique may yield a model-independent, multivariate scalar summary to monitor, but it is not implemented in the  `mice` package [@mice].
+-->
+
+<!-- ## A novel scalar summary -->
+
+<!-- To monitor non-convergence in iterative imputation algorithms, we may  summarize the state-space of the algorithm with several $\theta$s.
+The downside to the current $\theta$s is that they either focus on the univariate state-space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
+Ideally, one would like to evaluate a model-independent $\theta$ that summarizes the multivariate nature of the data.
+-->
+
+<!-- We propose $\lambda_{1}$ as such a scalar summary.
+We define $\lambda_{1}$ as the first eigenvalue of the variance-covariance matrix in the $m$ completed datasets.
+The eigenvalues of a variance-covariance matrix $S$ summarize the total set of covariances in the data.
+Let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ...
+\geq \lambda_{j, \ell}$ be the eigenvalues of $S_\ell$ in each imputation.
+The first eigenvalue, $\lambda_{1,\ell}$, then summarizes the largest possible amount of covariance in each completed dataset {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}.
+By definition, $\lambda_{1}$-values are equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on the completed data.
+As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the model of scientific interest, yet still summarizes the multivariate state-space of the algorithm.
+-->
+
+<!-- In this study, we consider each of the four $\theta$s that we discussed.
+Namely, two univariate scalar summaries (chain mean and chain variance), a model-dependent multivariate summary $\hat{Q}$, and the novel model-independent multivariate summary $\lambda_1$.
+-->
+
+## Non-convergence diagnostics
+
+There are many diagnostic tools to identify non-convergence in iterative (MCMC) algorithms [@broo98; @elad06].
+We consider only two of them that may be appropriate for imputation algorithms---one to monitor signs of non-mixing, and one for non-stationarity.
+As recommended by e.g.
+@cowl96 we will use the potential scale reduction factor $\widehat{R}$ to evaluate mixing ['Gelman-Rubin statistic'; @gelm92], and autocorrelation to diagnose trending [$AC$; @scha97; @gelm13].
+With a recently proposed adaptation, $\widehat{R}$ may also serve to diagnose non-stationarity [@vehtari2021].
+We will evaluate the appropriateness of both $\widehat{R}$ and $AC$.
+Other methods are outside the scope of this study because they e.g.
+assume that values within chains represent independent samples, whereas the MICE algorithm only uses the final iteration to produce imputations.
+
+<!-- Non-stationarity within chains may be diagnosed with e.g., autocorrelation [$AC$; @scha97; @gelm13], numeric standard error ['MC error'; @gewe92], or Raftery and Lewis's [-@raft91] procedure to determine the effect of trending on the precision of estimates.
+-->
+
+<!-- "We can use effective sample size ^neff to give us a sense of the precision obtained from our simulations.
+As we have discussed in Section 10.5, for many purposes it should suffice to have 100 or even 10 independent simulation draws.
+(If neff = 10, the simulation standard error is increased by p 1 + 1/10 = 1.05).
+As a default rule, we suggest running the simulation until ^neff is at least 5m, that is, until there are the equivalent of at least 10 independent draws per sequence (recall that m is twice the number of sequences, as we have split each sequence into two parts so that Rb can assess stationarity as well as mixing).
+Having an effective sample size of 10 per sequence should typically correspond to stability of all the simulated sequences."  @gelm13, p.
+287 -->
+
+<!-- We don't use effective sample size and MC error because they assume independent samples *within* chains.
+In MI, we only use the final estimate.
+Also, computing n_eff assumes infinitely many iterations, which is fine in Bayesian analyses where T is often at least one thousand.
+But the default in iterative imp is often substantially lower.
+-->
+
+## Potential scale reduction factor
+
+In 2019, Vehtari et al.
+proposed an updated version of the potential scale reduction factor $\widehat{R}$, originally coined in 1992.
+The adapted version would be better suited to detect non-mixing in the tails of distributions, and even identify non-stationarity.
+This version uses three transformations on the scalar summary $\theta$, before computing $\widehat{R}$-values.
+Namely, rank-normalization, folding, and localization.
+We follow Vehtari et al.'s formulation (2019, p.
+5) to define $\widehat{R}$.
+Let $m$ be the total number of chains, $T$ the number of iterations per chain (where $T\geq2$), and $\theta$ the scalar summary of interest.
+For each chain ($\ell = 1, 2, \dots, m$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.
+
+\begin{align*}
+W&=\frac{1}{m} \sum_{\ell=1}^{m} s_{\ell}^{2}, \text { where } s_{\ell}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t \ell)}-\bar{\theta}^{(\cdot \ell)}\right)^{2}. 
+\end{align*}
+
+\noindent We then estimate between-chain variance $B$ (note that we diverge from the typical notation in MI, where $B$ denotes the variance between the estimated quantities of scientific interest $\hat{Q}_{\ell}$).
+$B$ is defined as the variance of the collection of average $\theta$s per chain:
+
+\begin{align*}
+B&=\frac{T}{m-1} \sum_{\ell=1}^{m}\left(\bar{\theta}^{(\cdot \ell)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot \ell)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t \ell)} \text{, } \bar{\theta}^{(\cdot \cdot)}=\frac{1}{m} \sum_{\ell=1}^{m} \bar{\theta}^{(\cdot \ell)}.
+\end{align*}
+
+\noindent From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$.
+$\widehat{R}$ is then obtained as a ratio between this total variance and the within-chain variance:
+
+\begin{equation*}
+\widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
+\text{ where } \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{T-1}{T} W+\frac{1}{T} B.
+\end{equation*}
+
+We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run [@gelm92].
+The assumption underlying this interpretation is that chains are 'over-dispersed' at $t=1$, and reach convergence as $T \to \infty$.
+Over-dispersion implies that the initial values of the chains are 'far away' from the target distribution and each other.
+When the sampled values in each chain are independent of the chain's initial value, the mixing component of convergence is satisfied.
+The variance between chains, $B$, is then equivalent to the variance within chains, $W$, and $\widehat{R}$-values will be close to one.
+High $\widehat{R}$-values thus indicate non-convergence.
+
+## Autocorrelation
+
+Autocorrelation is defined as the correlation between two subsequent $\theta$-values within the same chain [@lync07, p.
+147].
+In this study, we only consider $AC$ at lag 1, i.e., the correlation between the $t^{th}$ and $(t+1)^{th}$ iteration of the same chain.
+Following the same notation as for $\widehat{R}$,
+
+\begin{equation*}
+AC = \left( \frac{T}{T-1} \right) \frac{\sum_{t=1}^{T-1}(\theta_t - \bar{\theta}^{(\cdot m)})(\theta_{t+1} - \bar{\theta}^{(\cdot m)})}{\sum_{t=1}^{T}(\theta_t - \bar{\theta}^{(\cdot m)})^2}.
+\end{equation*}
+
+We can interpret $AC$-values as a measure of non-stationarity.
+If there is dependence between subsequent $\theta$-values in imputation chains, $AC$-values are non-zero.
+Positive $AC$-values occur when $\theta$-values are recurring (i.e., high $\theta$-values are followed by high $\theta$-values, and low $\theta$-values are followed by low $\theta$-values).
+Recurrence within imputation chains may lead to trending.
+Negative $AC$-values occur when $\theta$-values of subsequent iterations are less similar, or diverge from one another.
+Divergence within imputation poses no threat to the convergence of the algorithm---it may even speed up convergence.
+Complete stationarity is reached when $AC=0$.
+As non-convergence diagnostic, our interest is in positive $AC$-values.
+
+## Thresholds
+
+It is unlikely that iterative algorithms such as MICE will achieve the ideal values of $\widehat{R}=1$ and $AC=0$.
+[TODO: check this!
+isn't it because of under-dispersion and reliance on observed data?] Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state.
+The aim, therefore, is to reach approximate convergence [@gelm13].
+Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} \gtrapprox 1$), and there is little dependency between subsequent iterations of imputation chains ($AC \gtrapprox 0$).
+In practice, we diagnose non-convergence when approximate convergence is violated, i.e., when $\widehat{R}$ and $AC$ exceed a certain threshold.
+The conventional thresholds to diagnose non-mixing are $\widehat{R} > 1.2$ [@gelm92] or $\widehat{R} > 1.1$ [@gelm13].
+@vehtari2021 proposed a much more stringent threshold of $\widehat{R} > 1.01$.
+The magnitude of $AC$-values may be evaluated statistically, using a Wald test with $AC=0$ as null hypothesis [@box15].
+$AC$-values that are significantly higher than zero indicate non-stationarity.
+<!-- Threshold Rhat > 1.1 @gelm13, p. 288 -->
+
 # Simulation set-up
@@
 **Algorithm 1: simulation set-up (pseudo-code)**
-
-    for each simulation repetition 
-      1. simulate complete data
-      for each missingness condition 
-        2. create missingness
-        for each iteration 
-          3. impute missingness
-          4. perform analysis of scientific interest
-          5. apply non-convergence identifiers
-          6. pool results across imputations
-          7. compute performance measures
-        8. combine outcomes from all iterations
-      9. combine outcomes from all missingness conditions
-    10. aggregate outcomes from all simulation repetitions
-
-```
-# pseudo-code of simulation 
-for (number of simulation runs) {
-  1. simulate complete data 
-  for (missingness conditions) {
-    2. create missingness
-    for (number of iterations) {
-      3. impute missingness
-      4. estimate quantities of scientific interest
-      5. apply performance measures to the estimates
-      6. compute non-convergence diagnostics
-    }
-  }
-7. aggregate outcomes across simulation runs
+```
+for (each simulation run) {
+  simulate complete data;
+  for (each missingness condition) {
+    create missingness;
+    for (each imputation model iteration) {
+      impute missingness;
+      estimate quantities of scientific interest;
+      apply performance measures to the estimates;
+      compute non-convergence diagnostics
+} } }
 ```
@@
 To evaluate convergence, we terminate the imputation algorithm after a varying number of iterations ($n_{\mathrm{it}} = 1, 2, ..., 100$).
-We differentiate between six different missingness scenarios that are defined in the data generating mechanism.
+We differentiate between six different missingness scenarios that are defined in the data generating mechanism (see Table XYZ).
 
 ## Data generating mechanism
+
+\begin{table}[]
+\begin{tabular}{lcr}
+\hline
+\multicolumn{3}{l}{Missingness conditions} \\ \hline
+missing data mechanism &  & proportion of incomplete cases  \\
+                       & $\times$  &                         \\
+MCAR, MAR              &   & 25\%, 50\%, 75\%                \\ \hline
+\end{tabular}
+\end{table}
 
@@
 We use a $2 \times 3$ factorial design with two missing data mechanisms and three proportions of incomplete cases, see Table XYZ.
-
-[TODO: make this a box] Missingness mechanisms refer to the probability of being missing for any given entry in a dataset.
+We consider all possible multivariate patterns of missingness, as visualized in Figure \ref{fig:patterns}.
+
+```{r patterns, echo=FALSE}
+source("../R/generate_datasets.R")
+pat <- create_patterns(3)
+is.na(pat) <- pat == 1
+ggmice::plot_pattern(pat, caption = FALSE) +
+  theme(axis.text.x=element_blank(), 
+      axis.ticks.x=element_blank(), 
+      axis.text.y=element_blank(), 
+      axis.ticks.y=element_blank(),
+      axis.text = element_blank()) +
+  theme_void()
+```
+
+::: {.callout-note}
+# Missing Data Mechanism
+
+Missingness mechanisms refer to the probability of being missing for any given entry in a dataset.
 There are three distinct types of mechanisms, as defined by Rubin (1976).
 Roughly translated, the probability of being missing may be equal for all entries (MCAR; Missing Completely At Random), may depend on observed information (MAR; Missing At Random), or may depend on *un*observed information, making the missingness non-ignorable (MNAR; Missing Not At Random).
+:::
 
@@
 We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\mathrm{obs}}, y_{\mathrm{mis}}$} for every missingness condition.
-The missing data patterns are all possible combinations of multivariate missingness.
 
 [TODO: add left-tailed M(N)AR mechanism, md pattern, and maybe refer to ampute paper or "Dance of..." by Schouten 2018?].
+
+## Estimands
+
+We impute the missing data five times ($m = 5$) using Bayesian linear regression imputation with `mice` [@mice].
+On each imputed dataset, we perform multiple linear regression as our analysis of scientific interest.
+Our estimands are the regression coefficients $\boldsymbol{\beta}$,
+
+$$
+\hat{Y} = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,
+$$
+
+where $\hat{Y}$ is the predicted value of the outcome.
+We estimate these coefficients using the `lm()` function [@R] in each imputed dataset, and subsequently pool the estimates across the five imputations using Rubin's rules [function `mice::pool()`; @mice].
+
+<!-- All imputation procedures are performed using `mice` [function `mice()`; @mice], with Bayesian linear regression imputation, and five imputation chains ($m=5$).
+This results in $m=5$ sets of imputations for each of the conditions, $\dot y_{\mathrm{imp}, \ell}$.
+-->
+
+<!-- Subsequently, we obtain the completed data per imputation, {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice].
+Estimates of scientific quantities $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R].
+We pool the resulting $m$ estimates, $\hat{Q}_\ell$, into a single $\bar{Q}$ conform Rubin's rules [function `mice::pool()`; @mice].
+-->
+
+## Performance measures
 
@@
 \hline
-\multicolumn{3}{l}{Missingness conditions} \\ \hline
-missing data mechanism &  & proportion of incomplete cases  \\
+\multicolumn{3}{l}{Performance measures} \\ \hline
+estimand &  & metric  \\
                        & $\times$  &                         \\
-MCAR, MAR              &   & 25\%, 50\%, 75\%                \\ \hline
+$\beta_0$, $\beta_1$, $\beta_2$, $\beta_3$   &   &   bias, coverage rate, CI width  \\ \hline
 \end{tabular}
 \end{table}
-
-## Estimands
-
-We impute the missing data five times ($m = 5$) using Bayesian linear regression imputation with `mice` [@mice].
-On each imputed dataset, we perform multiple linear regression as our analysis of scientific interest.
-Our estimands are the regression coefficients $\boldsymbol{\beta}$,
-
-$$
-\hat{Y} = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,
-$$
-
-where $\hat{Y}$ is the predicted value of the outcome.
-We estimate these coefficients using the `lm()` function [@R] in each imputed dataset, and subsequently pool the estimates across the five imputations using Rubin's rules [function `mice::pool()`; @mice].
-
-<!-- All imputation procedures are performed using `mice` [function `mice()`; @mice], with Bayesian linear regression imputation, and five imputation chains ($m=5$).
-This results in $m=5$ sets of imputations for each of the conditions, $\dot y_{\mathrm{imp}, \ell}$.
--->
-
-<!-- Subsequently, we obtain the completed data per imputation, {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice].
-Estimates of scientific quantities $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R].
-We pool the resulting $m$ estimates, $\hat{Q}_\ell$, into a single $\bar{Q}$ conform Rubin's rules [function `mice::pool()`; @mice].
--->
-
-## Performance measures
 
@@
 
-# Simulation Results
+# Simulation results
 
@@
 
+```{r, cache=TRUE, fig.height=9}
+performance |>
+  filter(method == "MICE") |>
+ggplot(aes(x = .it, y = psrf_mean)) +
+  geom_point(aes(group = .sim), shape = 1, alpha = 0.05) +
+  geom_smooth(method = "loess") +
+  facet_grid(term ~ mech + prop) +
+  labs(
+    title = "Potential Scale reduction Factor(PSRF) applied to chain means",
+    x = "Iteration number",
+    y = "PSRF"
+  ) +
+  theme_classic() 
+```
+
 ## Performance Measures
@@
 
+<!-- Our results show the performance of the MICE algorithm under different conditions of missingness and early stopping.
+For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias.
+For univariate scientific estimands ($Q=\mu$ and $Q=\sigma$) we observe the largest bias in the outcome variable $Y$.
+And for the regression coefficients ($Q=\beta$) the bias is most pronounced in $\beta_1$ (the effect of $X_1$ on $Y$).
+Since there is just one estimate for $Q=r^2$ to evaluate, we consider $Q=\mu_Y, \sigma_Y,r^2, \beta_1$.
+In the figures, we present all missingness conditions, but only early stopping conditions where $1 \leq T\leq50$.
+That is, the results are more or less stable for conditions where $T \geq 30$.
+Full results are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).
+-->
+
+<!-- ## Quantities of scientific interest -->
+
+<!-- ```{r Qs, fig.height = 6.8, fig.cap="Impact of non-convergence on statistical inferences.
+Depicted are the bias, coverage rate (CR) and confidence interval width (CIW) of the estimates $\\bar{Q}$.
+We present the subset of $Q$s with the worst performance in terms of bias: the mean and standard deviation of $Y$, and the regression coefficient of $X_1$.
+The solid gray line depicts the targets (unbiased estimates and nominal coverage rate)."} -->
+
+<!-- # mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_annotation(tag_levels = "A", tag_suffix = ".") + plot_layout(ncol = 2, guides = "collect")  -->
+<!-- ``` -->
+
+<!-- Figure \ref{fig:Qs} displays the influence of missingness and early stopping on the estimated quantities of scientific interest, $Q$s.
+In the first row of the figure, we see the bias in the estimated descriptive statistics ($Q=\mu_Y$ and $Q=\sigma_Y$; panels A and B).
+The second row shows the bias in the multivariate scientific estimands ($Q=r^2$ and $Q=\beta_1$; C and D).
+The third row consists of the coverage rate (CR) and confidence interval width (CIW) of the estimated regression coefficient ($Q=\beta_1$; E and F).
+-->
+
+<!-- ### $Q=\mu$. -->
+<!-- The estimated univariate means seem unaffected by early stopping.
+This implies that approximately unbiased estimates may be obtained with as little as one iteration ($T \geq 1$).
+The bias in $\bar{Q}$ depends solely on the proportion of incomplete cases, with $p_{\mathrm{inc}}\geq.75$ leading to more extreme biases than other conditions.
+Completely unbiased estimates are only obtained in conditions where $p_{\mathrm{inc}}\leq.50$.
+This is curious since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.
+Note however, that the magnitude of the bias in these conditions is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$).
+We, therefore, conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.
+-->
+
+<!-- ### $Q=\sigma$. -->
+<!-- The estimated standard deviations are not substantially impacted by early stopping either.
+While the estimates for $T=1$ are not equal to other conditions, the bias is actually *less* severe than for $T\geq2$.
+Therefore, no iteration at all is needed to obtain approximately unbiased estimates for $Q=\sigma$.
+Completely unbiased estimates are only obtained in conditions where $p_{\mathrm{inc}}=.05$.
+The other missingness conditions are impacted by non-convergence in the order of increasing $p_{\mathrm{inc}}$.
+Similar to $Q=\mu$, the magnitude of the bias in $Q=\sigma$ is negligible.
+We, therefore, conclude that neither one of the univariate $Q$s is affected substantially by non-convergence.
+-->
+
+<!-- ### $Q=r^2$. -->
+<!-- In contrast to the univariate estimates, the estimated coefficient of determination is clearly affected by early stopping, see Figure \ref{fig:Qs}C.
+The bias in $Q$ appears to decrease with each additional iteration, while the magnitude of the bias still depends on the proportion of incomplete cases.
+Therefore, we see that the number of iterations that is necessary to reach stable, non-improving estimates also differs across missingness conditions.
+In conditions where $p_{\mathrm{inc}}=.05$, estimates are unbiased after one iteration.
+The highest number of iterations necessary to reach approximate unbiasedness is seven ($p_{\mathrm{inc}}=.95$; $T\geq7$).
+Completely unbiased estimates are only obtained in conditions where $p_{\mathrm{inc}}\leq.25$.
+This implies that even for biased estimates, $T\geq7$ would suffice to reach a stable solution.
+-->
+
+<!-- ### $Q=\beta$. -->
+<!-- For the estimated regression coefficients, we first consider bias, before discussing CR and CIW.
+As Figure \ref{fig:Qs}D shows, the bias in $\bar{Q}$ is affected by both the proportion of incomplete cases and the number of iterations.
+Similar to $Q=r^2$, we observe approximately unbiased estimates after at least one, and at most seven iterations, depending on $p_{\mathrm{inc}}$.
+Completely unbiased estimates are only obtained in conditions where $T\geq3$ and $p_{\mathrm{inc}}\leq .50$.
+Despite persistent bias in conditions where $p_{\mathrm{inc}}\geq.75$, the coverage rates are non-improving after just three iterations, irrespective of $p_{\mathrm{inc}}$.
+We conclude that any condition where $T\geq3$ yields approximately nominal coverage rates, and thus confidence-valid estimates.
+-->
+<!-- The nominal coverages can be explained by the CIWs.
+The CIWs depicted in Figure \ref{fig:Qs}F are conform the theoretical foundation of MI: CIs are wider in conditions with higher proportions of incomplete cases (i.e., there is less information in the data, and thus more uncertainty due to missingness).
+Since conditions with higher $p_{\mathrm{inc}}$ result in both wider CIs and more severe bias, the true value of $Q$ may be included in the CI, despite of the bias in $\bar{Q}$.
+-->
+<!-- We conclude that approximately unbiased estimates may be obtained in conditions where $T\geq7$, whereas confidence-valid estimates require at most three iterations.  -->
+
+<!-- ### Summary.  -->
+<!-- These results demonstrate that the estimates of univariate scientific estimands $Q$ are not impacted by early stopping of the MICE algorithm or the proportion of incomplete cases in $y$.
+Unbiased estimates may be obtained after just one iteration.
+Multivariate estimates, by contrast, are affected by both the number of iterations and the proportion of missing cases.
+Completely unbiased estimates are only obtained under low to moderate missingness ($p_{\mathrm{inc}}\leq.50$), after at most three iterations.
+We observe approximately unbiased estimates after at most seven iterations (for any $p_{\mathrm{inc}}$ considered).
+This implies that the algorithm produces stable, non-improving estimates when $T\geq7$.
+-->
+
+<!-- ## Non-convergence diagnostics -->
+
+<!-- Figure \ref{fig:rhats-acs} displays results of the eight sets of diagnostic methods.
+The two columns in the figure represent the  non-convergence diagnostics $\widehat{R}$ and $AC$; the rows depict the four scalar summaries $\theta$ under consideration.
+The first row in Figure \ref{fig:rhats-acs} thus shows $\widehat{R}$ and $AC$ applied on the chain means (panels A and B).
+In the second row, we see the diagnostics applied on the chain variances (C and D).
+The third row depicts the same for $\theta=\hat{Q}$ (E and F).
+And in the last row, we see the diagnostics applied on the novel $\theta$ to consider: $\lambda_1$ (G and H).
+We evaluate the performance of the methods by establishing whether they correctly identify conditions in which $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
+As we concluded in the last section, substantially biased estimates were only obtained in conditions where $T\leq6$, and we observed non-nominal coverage only when $T\leq3$.
+However, there was some persistent bias in conditions where $p_{\mathrm{inc}}\geq.75$, irrespective of the number of iterations.
+If the diagnostic methods are appropriate for MI, we will detect non-convergence in these conditions.
+-->
+
+<!-- ```{r rhats-acs, fig.height = 6.8, fig.cap="Non-convergence diagnostics $\\widehat{R}$ and $AC$ applied on several $\\theta$s.
+The left-hand side of the figure contains $\\widehat{R}$-values, and the right-hand side contains $AC$-values.
+Depicted in the rows are the scalar summaries $\\theta$: chain mean, chain variance, the quantity of scientific interest $\\hat{Q}$, and the first eigenvalue of the variance-covariance matrix $\\lambda_1$.
+The dashed gray lines depict the different thresholds to diagnose non-convergence."} -->
+
+<!-- # mean_Rh + mean_AC + var_Rh + var_AC + est_Rh + est_AC + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 2) + plot_annotation(tag_levels = "A", tag_suffix = ".") -->
+<!-- ``` -->
+
+<!-- ### $\theta=$ chain mean.  -->
+<!-- As expected, $\widehat{R}$ lowers with the number of iterations.
+After an initial 'dip' in conditions where $T=3$, we observe a gradual decrease in $\widehat{R}$-values with every additional iteration.
+This implies improving convergence, until the decrease tapers off after 30 to 40 iterations.
+Using the threshold $\widehat{R}>1.2$, we diagnose non-convergence in conditions where $T\leq7$, whereas the threshold 1.1 is exceeded when $T\leq13$.
+The very strict threshold 1.01 is surpassed in all conditions $T\leq100$ (not shown), and therefore not informative under the current specifications.
+With this set of methods, we fail to identify non-convergence in conditions with persistent bias due to missingness.
+-->
+
+<!-- Similarly, $AC$ decreases with higher $T$s.
+$AC$ indicates improving stationarity in conditions where $T\leq6$.
+The $AC$-values do not exceed the threshold defined by statistical significance of the $AC$-values.
+Based on this threshold, we fail to diagnose non-convergence in any condition.
+-->
+
+<!-- ### $\theta=$ chain variance. -->
+<!-- As panel C and D in Figure \ref{fig:rhats-acs} show, the results of this set of methods are highly similar to those with chain means as $\theta$.
+This holds for both $\widehat{R}$ and $AC$.
+The only apparent difference is in the $\widehat{R}$ threshold 1.1.
+Instead of diagnosing non-convergence in conditions where $T\leq13$, we now diagnose this when $T\leq11$.
+All other observations are equivalent.
+-->
+
+<!-- ### $\theta=\hat{Q}$. -->
+<!-- The $\widehat{R}$-values in Figure \ref{fig:rhats-acs}E show a similar trend across iterations as we observed for chain means and chain variances.
+Only conditions with a very high  proportion of missing cases ($p_{\mathrm{inc}}\geq.75$) diverge from the earlier observations.
+The $\widehat{R}$-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations.
+The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are 11 for $\widehat{R}>1.2$, 23 for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).
+-->
+
+<!-- The $AC$-values also follow earlier observations, with the same exception as for $\widehat{R}$ (when $p_{\mathrm{inc}}\geq.75$).
+To reach maximum stationarity, these conditions require at most seven iterations instead of five.
+Moreover, the $AC$-values in these conditions exceed the threshold to diagnose non-convergence.
+Interestingly, this only occurs after ten or even thirty iterations, while we would expect more signs of trending early in the iterations.
+Since the $AC$-values are not improving or worsening when $T\geq7$, we conclude that non-convergence is incorrectly diagnosed based on the threshold.
+-->
+
+<!-- ### $\theta=\lambda_{1}$. -->
+<!-- Once again, we observe $\widehat{R}$-values with a similar trend across iterations.
+For this $\theta$, the results differ only in the most extreme missingness condition.
+When $p_{\mathrm{inc}}=.95$, we diagnose non-convergence at $T\leq9$ according to $\widehat{R}>1.2$, $T\leq19$ for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).
+The 'dip' in $\widehat{R}$-values at $T=3$ is even more pronounced than with other $\theta$s.
+Some $\widehat{R}$-values even overcome the threshold of 1.2.
+If we would terminate the algorithm at $T=3$, we would incorrectly conclude that the algorithm reached approximate convergence, according to this threshold.
+-->
+
+<!-- $AC$-values are systematically higher for this $\theta$ than for other scalar summaries.
+The $AC$-values also taper off at a later point in the iterations.
+This suggests improving stationarity up-to ten, or even thirty iterations, depending on the missingness condition.
+According to the threshold, we should diagnose non-convergence at some point in any missingness condition.
+However, this occurs only after reasonably decreasing $AC$-values are obtained---again suggesting that non-convergence is incorrectly diagnosed.
+-->
+
+<!-- ### Summary.  -->
+<!-- Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
+Under low to moderate missingness, it does not seem to matter on which $\theta$ the non-convergence diagnostics are applied.
+If we look at complete unbiasedness, however, we notice that univariate $\theta$s fail to diagnose the persistent bias in conditions where $p_{\mathrm{inc}}\geq.75$.
+-->
+
+<!-- The number of iterations necessary to obtain approximately unbiased, confidence-valid estimates corresponds to the $\widehat{R}$ threshold 1.2.
+The thresholds 1.1 and 1.01 seem too strict compared to the validity of the inferences.
+The threshold to diagnose significant $AC$-values does not appear to be appropriate in at least the current set-up, and perhaps in iterative imputation in general.
+A better heuristic to diagnose non-stationarity with $AC$ may be through evaluation of the $AC$-values across iterations: If the values do not substantially decrease with $T$, approximate stationarity may be concluded.
+-->
+
 # Discussion
@@
 
+<!-- Before drawing any further conclusions, let us first return to the hypotheses that we formulated for the simulation study.  -->
+
+<!-- 1.
+As hypothesized, biased, invalid estimates of quantities of scientific interest occurred more often in simulation conditions with a high proportion of incomplete cases $p_{\mathrm{inc}}$ and a low number of iterations $T$.
+However, this only holds for multivariate scientific estimands.
+Univariate $Q$s are not substantially impacted by early stopping.
+-->
+
+<!-- 2.
+Our results also support the hypothesis that $\widehat{R}$ and $AC$ would correctly identify signs of non-convergence.
+We observed higher $\widehat{R}$ and $AC$-values in conditions where $\bar{Q}$s were biased, invalid estimates of $Q$s.
+-->
+
+<!-- 3.
+We hypothesized that the recommended thresholds to diagnose non-convergence with $\widehat{R}$ would be too stringent for iterative imputation applications, compared to the number of iterations necessary for statistically valid inferences.
+This hypothesis is partially supported: On the one hand, the 1.01 and 1.1 thresholds indeed seem to over-estimate the signs of non-convergence.
+On the other hand, however, we observed that the traditional threshold of 1.2 roughly corresponds to the number of iterations required to obtain approximately unbiased, confidence-valid estimates.
+-->
+
+<!-- 4.
+The results of this simulation study do not support the hypothesis that high $AC$-values are implausible in iterative imputation algorithms with typical convergence.
+We expected that the randomness induced by the algorithm would effectively mitigate the risk of dependency within chains after a few iterations.
+In this study, however, we observed $AC$-values as high as theoretically possible ($AC=1$).
+The second part of the hypothesis is not refuted: the $AC$-values did decrease quickly as a function of $T$.
+-->
+
+<!-- 5.
+Our final hypothesis was that multivariate $\theta$s would be better at detecting non-convergence than univariate $\theta$s.
+Multivariate $\theta$s did indeed show superior performance under certain conditions (e.g., multivariate $Q$s and extreme missingness).
+-->
+
+<!-- ## Conclusion -->
+
+<!-- We have shown that non-convergence in iterative imputation algorithms goes hand in hand with biased, invalid estimates---at least for the multivariate quantities of scientific interest considered in this study.
+Identifying non-convergence may be a crucial aspect of drawing valid statistical inferences from incomplete data.
+We concluded that the current practice of visually inspecting non-convergence through traceplots of univariate scalar summaries of the state-space of the algorithm does not suffice.
+We, therefore, considered eight sets of diagnostic methods to identify non-convergence: two non-convergence diagnostics ($\widehat{R}$ and $AC$), two univariate $\theta$s, a model-dependent multivariate $\theta$, and a novel, model-independent multivariate $\theta$.
+Signs of non-convergence due to early stopping were identified with each of the methods.
+Bias due to high proportions of incomplete cases was only identified with multivariate $\theta$s.
+-->
+
+<!-- Since we obtained approximately unbiased, confidence-valid estimates after at most seven iterations, we conclude that the $\widehat{R}$ threshold 1.2 is the most appropriate diagnostic cut-off.
+The threshold to diagnose non-stationarity with $AC$ does not seem to apply.
+However, some signs of non-convergence were detected in simulation conditions many more iterations.
+Under moderate missingness ($p_{\mathrm{inc}}\leq.50$) $\widehat{R}$-values decreased substantially until 30 to 40 iterations, which implies that mixing in the algorithm still improved with each additional iteration.
+$AC$-values only improved until about six iterations, suggesting minimal improvement in trending was obtained beyond $T=6$ in the current simulation set-up.
+-->
+
+<!-- The main finding of this study is that valid inferences may be obtained much quicker than approximate algorithmic convergence is reached.
+Under the current specifications, univariate $Q$s did not require algorithmic convergence at all.
+They are unbiased almost instantly after the algorithm is initiated.
+Approximately unbiased, confidence-valid estimates of multivariate $Q$s were obtained after a maximum of seven iterations.
+Continued iterations beyond $T = 7$ did not yield better estimates.
+-->
+
+<!-- ## Implications -->
+
+<!-- Iterative imputation algorithms such as MICE are known to yield valid results under severe missingness.
+With this study, we have shown that valid inferences can also be obtained under a combination of severe missingness and early stopping.
+Based on our results, the traditional threshold of $\widehat{R}> 1.2$ would be appropriate for diagnosing non-convergence in iterative imputation algorithms.
+This is, however, not in line with the recent recommendations by @vehtari2021 to lower the threshold to 1.01.
+The discrepancy between our conclusion and Vehtari et al.
+may be explained by the nature of iterative imputation algorithms.
+In the limit, iterative imputation algorithms have the same characteristics as other MCMC algorithms.
+But, in imputation procedures, a part of the distribution is already determined by the observed data, whereas the entire distribution is unknown in many other MCMC applications.
+Since we combine a known distribution with an unknown distribution, valid estimates may be reached much sooner.
+Convergence may, therefore, be diagnosed at a less stringent threshold.
+-->
+
+<!-- The threshold to diagnose non-convergence with $AC$ does not seem appropriate at all in iterative imputation algorithms.
+A better diagnostic cut-off would be the number of iterations at which an additional iteration does not substantially decrease the $AC$-values.
+In practice, this implies that the default number of iterations in MICE is not sufficient.
+$AC$ can only be computed if $T\geq3$, while the current default in the `mice` package is five [@mice].
+Identifying a reasonable decrease in the $AC$-values across iterations (an 'elbow') requires more than three observations.
+We, therefore, suggest to increase the default number of iterations in MICE to ten.
+The computational cost of five additional iterations has become less of a burden since MICE was introduced [-@mice].
+Moreover, an increased number of iterations would grant the opportunity to exclude the first iterations from evaluation with $\widehat{R}$.
+Excluding $T\leq3$ provides empirical researchers with a less ambiguous heuristic to diagnose non-convergence, because we expect the initial 'dip' in $\widehat{R}$-values to disappear.
+-->
+
+<!-- The potential scale reduction factor metric assumes over-dispersed starting values of the iterative algorithm.
+The MICE algorithm does not start in an over-dispersed state.
+MICE does not rely on starting values for parameters at all: instead, the algorithm is initialized based on a 'zeroth' iteration, where missing values are filled in using a random draw from observed values.
+Under MCAR, this would typically not yield an over-dispersed state at all, depending on the parameter of interest: means and variances are not affected, regression coefficients and other multivariate parameters start biased downwards (because sampling starting values distorts multivariate structures in the data).
+Whether this accrues to over-dispersion is questionable.
+Under MAR (or MNAR), some over-dispersion might occur when the observed data and imputed data differ impactfully.
+The algorithm should 'escape' the initial state (based on the starting values drawn from observed data alone).
+Over iterations, all parameters of interest (e.g.
+means, variances, multivariate estimands) should converge towards a stable state.
+Moreover, the MICE algorithm is initialized with more information than a typical MCMC algorithm: parameter estimates depend not only on imputed data, but also on observed data that does not change over iterations.
+Therefore, the initial state of the algorithm is 'too good' to observe a relative decrease in $\widehat{R}$.
+-->
+
+<!-- ## Recommendations for empirical researchers -->
+
+<!-- Based on our results, we formulate several recommendations for empirical researchers who employ iterative imputation to draw inferences from incomplete data.
+We suggest that non-convergence may still be evaluated visually, but in addition to inspecting univariate summaries of the state-space of the algorithm ($\theta$s, e.g., chain means), multivariate $\theta$s should be considered.
+We propose the following steps: -->
+
+<!-- 1. First, check traceplots of the default $\theta$s (e.g., chain means and chain variances) for signs of pathological non-convergence. Adjust the imputation model if necessary.  -->
+
+<!-- 2.
+Subsequently, decide which multivariate $\theta$s to track across iterations.
+The novel $\theta$ that we propose in this study, $\lambda_1$ is scientific model-independent and may thus always be employed.
+Alternatively, specify your own scalar summary of interest [see e.g., @buur18].
+Monitor these $\theta$s through visual inspection, or using a non-convergence diagnostic.
+-->
+
+<!-- 3.
+Compute non-convergence diagnostics $\widehat{R}$ and $AC$.
+Do *not* use the original implementation to calculate $\widehat{R}$ [@gelm92], or the `R` function `stats::acf()` to compute $AC$ [@R].
+Instead, calculate $\widehat{R}$ conform @vehtari2021 and compute autocorrelations manually (see e.g., [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint)).
+-->
+
+<!-- 4.
+And finally, use the threshold $\widehat{R}>1.2$ to diagnose non-mixing, and assess stationarity by plotting the autocorrelation over iterations.
+Keep iterating until the threshold for mixing is overcome, and until reasonably decreasing asymptotic $AC$-values are obtained.
+At that point, inferences are unlikely to improve by continued iteration.
+-->
+
+<!-- ## Limitations -->
+
+<!-- Much remains unknown about non-convergence in iterative imputation algorithms.
+Even though we have demonstrated the appropriateness of $\widehat{R}$ and $AC$ as non-convergence diagnostics in this study, results may not extrapolate to other situations.
+In this study, we only considered the iterative imputation algorithm implemented in the `mice` package.
+The performance of the non-convergence diagnostics may depend on the characteristics of this specific algorithm.
+A potential threat to the ability to detect non-mixing with $\widehat{R}$, for example, is the assumption of over-dispersed initial states.
+In `mice`, the algorithm is initiated by sampling starting values at random from the observed data, whereas other imputation software may use the average of the observed cases.
+The latter could lead to a systematical under-estimation of the variability in the first few iterations and violate the assumption of over-dispersion.
+In those cases, it would be more difficult to detect non-convergence with $\widehat{R}$.
+Therefore, the diagnostic might not be  appropriate outside of `mice`.
+-->
+
+<!-- Moreover, the current simulation conditions were restricted to a single missingness mechanism.
+Proper performance under a 'missing completely at random' (MCAR) mechanism is a necessary condition for any missing data method.
+It does not, however, guarantee equal performance under different missingness mechanisms.
+Future research should determine the performance of $\widehat{R}$ and $AC$ under 'missing at random' and 'missing not at random' mechanisms.
+Another parameter to consider in future research is the choice of the imputation method.
+As @murray2018 concluded, the behavior of algorithms such as MICE under certain default imputation models is still an open research question.
+We have investigated the behavior of MICE under only one type of model (Bayesian linear regression).
+Different imputations models might converge more poorly.
+-->
+
+<!-- The imputation model may even be mis-specified on purpose, to induce non-convergence of different severity levels.
+Qualitatively, we have shown that $\widehat{R}$ and $AC$ are appropriate under clear violation of convergence.
+Quantitatively, we have only demonstrated that they can identify signs of non-convergence under MCAR, when the imputation model is a correctly specified Bayesian linear regression model.
+\newline \newline -->
+
+<!-- \noindent In short, we have shown that iterative imputation algorithms can yield correct outcomes, even when a converged state has not yet formally been reached.
+Any further iterations would then burn computational resources without improving the statistical inferences.
+Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the $\widehat{R}$ and AC diagnostics.
+Of course, it never hurts to iterate longer, but such calculations hardly bring added value.
+-->
+
 # Case Study
-
-- We use real data: the `boys` dataset from the `mice` package
-
-- We are interested in predicting age from the other variables, in particular in the regression coefficient of `hgt`
-
-- We compare non-convergence identified using visual inspection versus rhat in the chain variances, scientific estimate and lambda.
-
-- The figures show results of a `mice` run with 20 iterations but otherwise default settings.
-
-From the traceplot of the chain means (see \@ref(fig:case)A) it seems that mixing improves up-to 10 iterations, while trending is only apparent in the first three iterations.
-
-This figure (\@ref(fig:case)B) shows that 7 iterations are required before the $\widehat{R}$-values of the chain means drop below the threshold for non-convergence.
-
-The $\widehat{R}$-values for the scientific estimate reaches the threshold much sooner, when $n_{it}=14$ (see \@ref(fig:case)C).
-
-According to the $\widehat{R}$-values with $\lambda$ as parameter, at least 15 iterations are required (see \@ref(fig:case)D).
-
-# Introduction (new-ish)
-
-Iterative imputation has become the de facto standard to accommodate for missing data.
-The aim is usually to draw valid inferences, i.e.
-to get unbiased, confidence-valid estimates that incorporate the effects of the missingness.
-Such estimates are obtained with iterative imputation by separating the missing data problem from the scientific problem.
-The missing values are imputed (i.e., filled in) using some sort of algorithm.
-And subsequently, the scientific model of interest is performed on the completed data.
-To obtain valid scientific estimates, both the missing data problem and the scientific problem should be appropriately considered.
-The validity of this whole process naturally depends on the convergence of the algorithm that was used to to generate the imputations.
-
-All inferences with imputed data rely on the convergence of the imputation algorithm.
-Yet, determining whether an algorithm has converged is not trivial.
-There has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms.
-A widely accepted practice is visual inspection of the algorithm.
-Diagnosing convergence through visual inspection, however, may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18].
-Therefore, a quantitative diagnostic method to assess convergence would be preferred.
-
-It is challenging to arrive upon a single point at which convergence has been reached.
-Since the aim is to converge to a distribution and not to a single point, the algorithm may produce some fluctuations even after it has converged.
-Because of this property, it may be more desirable to focus on *non*-convergence.
-Fortunately, there are non-convergence identifiers for other iterative algorithms, but the validity of these identifiers has not been systematically evaluated on imputation algorithms.
-
-In this study, we explore different methods for identifying non-convergence in iterative imputation algorithms.
-We evaluate whether these methods are able to cover the extent of the non-convergence, and we also investigate the relation between non-convergence and the validity of the inferences.
-We translate the results of our simulation study into guidelines for practice, which we demonstrate by means of a motivating example.
-
-# Motivating Example
 
@@
 
-# Introduction (old)
-
-Anyone who analyzes person-data may run into a missing data problem.
-Missing data is not only ubiquitous across science, but treating it can also be a tedious task.
-If a dataset contains just one incomplete observation, calculations are not defined, and statistical models cannot be fitted to the data.
-To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations).
-Unfortunately, this *ad hoc* solution may yield invalid results [@buur18].
-An alternative is to apply imputation.
-With imputation, we 'fill in' the missing values in an incomplete dataset.
-Subsequently, the model of scientific interest can be fitted to the completed dataset.
-By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness.
-This technique is known as 'multiple imputation' [MI; @rubin76].
-MI has proven to be a powerful tool to draw valid inferences from incomplete data under many circumstances [@buur18].
-
-Figure \ref{fig:diagram} provides an overview of the steps involved with MI.
-The missing part $y_{\mathrm{mis}}$ of an incomplete dataset is imputed $m$ times.
-This creates $m$ sets of imputed data $\dot y_{\mathrm{imp}, \ell}$, where $\ell = 1, 2, ..., m$.
-The imputed data is then combined with the observed data $y_{\mathrm{obs}}$ to create $m$ completed datasets.
-On each of these datasets the analysis of scientific interest is performed to estimate $Q$: the quantity of scientific interest (e.g., a regression coefficient).
-Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}_\ell$-values are obtained.
-Finally, the $\hat{Q}_\ell$-values are combined into a single pooled estimate $\bar{Q}$.
-The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence-valid estimate of the true---but unobserved---scientific estimand $Q$ [@rubin96].
-<!-- "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->
-
-```{r diagram, echo=FALSE, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\\hat{Q}$, to a single pooled estimate $\\bar{Q}$."}
-
-knitr::include_graphics("../figures/diagram.pdf") #adapted from [@buur18] section 1.4.1.
-```
-
-A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice].
-With MICE, imputed values are drawn from the posterior predictive distribution of the missing values.
-The algorithm is named after its iterative nature: a multivariate distribution is obtained by iterating over a sequence of univariate imputations.
-Iteration, however, also introduces a potential threat to the validity of the imputations: What if the algorithm has not converged?
-Are the imputations then to be trusted?
-And can we rely on the inference obtained on the completed data?
-
-These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18].
-There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@zhu15; @takahashi2017].
-Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murray2018].
-Therefore, algorithmic convergence should be monitored carefully---although this is not straightforward.
-Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods.
-In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another.
-The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13].
-Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09].
-Several non-convergence diagnostics exist, but it is not known whether these are appropriate within the imputation framework.
-
-<!-- https://www-tandfonline-com.proxy.library.uu.nl/doi/pdf/10.1207/s15327906mbr3304_5?needAccess=true -->
-
-<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->
-
-In this paper we study different methods for assessing non-convergence in iterative imputation algorithms.
-We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications.
-We then address the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in `R` [@R].
-For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package in `R` [@mice].
-The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged.
-And additionally, to evaluate the behavior and performance of several diagnostic methods to identify non-convergence.
-With that, we formulate an informed advice on when it is safe to conclude that the algorithm is converged *enough* for valid inferences.
-We translate the results of the study into guidelines for applied researchers, which may facilitate drawing valid inferences from incomplete data.
-
-<!-- The default number of iterations in common MI software: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100??
-[(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p.
-139), Amelia = NA, because re-sampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf).
--->
-
-## Some notation
-
-Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample.
-The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing.
-The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\mathrm{inc}}$ in dataset $y$.
-The collection of observed data values in $y$ is denoted by $y_{\mathrm{obs}}$; the missing part of $y$ is referred to as $y_{\mathrm{mis}}$.
-For each datapoint in $y_{\mathrm{mis}}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations in the imputation algorithm ($t = 1, 2, \dots, T$).
-The state-space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values).
-The collection of $\theta$-values between $t=1$ (summarizing the state-space of the algorithm at initialization) and $t=T$ (summarizing the state-space for the imputed values) will be referred to as an 'imputation chain'.
-
-# Algorithmic non-convergence
-
-There are two requirements for convergence of iterative algorithms: mixing and stationarity [@gelm13].
-In iterative imputation algorithms, mixing implies that imputation chains intermingle nicely, and stationarity is characterized by the absence of trending across iterations.
-If one of the two requirements is not met, we speak of non-convergence.
-Without mixing, chains may be 'stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values.
-The distribution of imputed values then differs across imputations.
-This may cause under-estimation of the variance between chains, which results in spurious, invalid inferences.
-Without stationarity, there is trending within imputation chains.
-Trending implies that further iterations would yield a systematically lower or higher set of imputations.
-Iterative imputation algorithms that have not (yet) reached stationarity, may thus yield biased estimates.
-
-To illustrate what non-mixing and non-stationarity look like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2].
-Figure \ref{fig:non-conv} displays two scenarios from the example.
-The panel on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm.
-The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
-Each line portrays one imputation chain (i.e., the values of a scalar summary $\theta$ across iterations).
-The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in the $m$ sets of imputations $\dot y_{\mathrm{imp}, \ell}$.
-
-In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending.
-In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.
-Importantly, the chain means at the last iteration (the imputed values per imputation $\ell$) are very different between the two scenarios.
-The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
-It is obvious that non-convergence in this example impacts the distribution of the imputed values per imputation $\dot y_{\mathrm{imp}, \ell}$.
-This effect may translate into the distribution of the $m$ sets of completed data {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$, and finally the pooled estimate $\bar{Q}$.
-$\bar{Q}$ is then a  biased, invalid estimate of $Q$.
-Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.
-
-```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence (reproduced from van Buuren; 2018).
-Depicted are chain means of variable $j$ in each imputation ($m=5$; $T=10$).
-Note that any difference between the two scenarios (see e.g., the magnitude of the chain means on the y-axes) stems from the imputation algorithms---the two scenarios have equal incomplete data $y$."}
-
-# par(mar = c(2, 2, 0.1, 0.1))
-# con_plot
-# non_plot
-
-```
-
-## Identifying non-convergence
-
-Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection.
-After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots.
-In a traceplot, a scalar summary of the state-space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}.
-The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances.
-Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18, $\S$ 6.5.2].
-<!-- Recommendation @buur18: "the different streams should be freely intermingled with one another, without showing any definite trends". -->
-
-Identifying non-convergence by inspecting traceplots may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18, $\S$ 6.5.2].
-Moreover, traceplots are typically only used to inspect  univariate $\theta$s.
-Monitoring univariate summaries of the state-space of the algorithm may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of the imputations.
-Ideally, we would monitor a multivariate $\theta$.
-
-A suggestion by @buur18 for a multivariate $\theta$ to monitor is the estimated quantity of scientific interest $\hat{Q}$, which usually is multivariate in nature.
-This $\theta$ is computed as the estimate in each imputation, $\hat{Q}_\ell$, across iterations.
-Implementing this, however, might be somewhat too technical for empirical researchers.
-Besides, this scalar summary is not model-independent.
-That is, $\theta=\hat{Q}$ is not universal to all complete data problems, while one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently.
-Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest.
-On these grounds, such a $\theta$ can be considered insufficient too.
-<!-- Researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2].  -->
-
-As alternative, van Buuren [-@buur18, $\S$ 4.5.2] proposed multivariate evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of @mack03.
-This technique may yield a model-independent, multivariate scalar summary to monitor, but it is not implemented in the  `mice` package [@mice].
-
-## A novel scalar summary
-
-To monitor non-convergence in iterative imputation algorithms, we may  summarize the state-space of the algorithm with several $\theta$s.
-The downside to the current $\theta$s is that they either focus on the univariate state-space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
-Ideally, one would like to evaluate a model-independent $\theta$ that summarizes the multivariate nature of the data.
-
-We propose $\lambda_{1}$ as such a scalar summary.
-We define $\lambda_{1}$ as the first eigenvalue of the variance-covariance matrix in the $m$ completed datasets.
-The eigenvalues of a variance-covariance matrix $S$ summarize the total set of covariances in the data.
-Let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ...
-\geq \lambda_{j, \ell}$ be the eigenvalues of $S_\ell$ in each imputation.
-The first eigenvalue, $\lambda_{1,\ell}$, then summarizes the largest possible amount of covariance in each completed dataset {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}.
-By definition, $\lambda_{1}$-values are equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on the completed data.
-As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the model of scientific interest, yet still summarizes the multivariate state-space of the algorithm.
-
-In this study, we consider each of the four $\theta$s that we discussed.
-Namely, two univariate scalar summaries (chain mean and chain variance), a model-dependent multivariate summary $\hat{Q}$, and the novel model-independent multivariate summary $\lambda_1$.
-
-# Non-convergence diagnostics
-
-There are many diagnostic tools to identify non-convergence in iterative (MCMC) algorithms [@broo98; @elad06].
-We consider only two of them that may be appropriate for imputation algorithms---one to monitor signs of non-mixing, and one for non-stationarity.
-As recommended by e.g.
-@cowl96 we will use the potential scale reduction factor $\widehat{R}$ to evaluate mixing ['Gelman-Rubin statistic'; @gelm92], and autocorrelation to diagnose trending [$AC$; @scha97; @gelm13].
-With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@vehtari2021].
-Therefore, we will evaluate the appropriateness of both $\widehat{R}$ versions, in addition to $AC$.
-Other methods are outside the scope of this study because they e.g.
-assume that values within chains represent independent samples, whereas the MICE algorithm only uses the final iteration to produce imputations.
-
-<!-- Non-stationarity within chains may be diagnosed with e.g., autocorrelation [$AC$; @scha97; @gelm13], numeric standard error ['MC error'; @gewe92], or Raftery and Lewis's [-@raft91] procedure to determine the effect of trending on the precision of estimates.
--->
-
-<!-- "We can use effective sample size ^neff to give us a sense of the precision obtained from our simulations.
-As we have discussed in Section 10.5, for many purposes it should suffice to have 100 or even 10 independent simulation draws.
-(If neff = 10, the simulation standard error is increased by p 1 + 1/10 = 1.05).
-As a default rule, we suggest running the simulation until ^neff is at least 5m, that is, until there are the equivalent of at least 10 independent draws per sequence (recall that m is twice the number of sequences, as we have split each sequence into two parts so that Rb can assess stationarity as well as mixing).
-Having an effective sample size of 10 per sequence should typically correspond to stability of all the simulated sequences."  @gelm13, p.
-287 -->
-
-<!-- We don't use effective sample size and MC error because they assume independent samples *within* chains.
-In MI, we only use the final estimate.
-Also, computing n_eff assumes infinitely many iterations, which is fine in Bayesian analyses where T is often at least one thousand.
-But the default in iterative imp is often substantially lower.
--->
-
-## Potential scale reduction factor
-
-In 2019, Vehtari et al.
-proposed an updated version of the potential scale reduction factor $\widehat{R}$, originally coined in 1992.
-The adapted version would be better suited to detect non-mixing in the tails of distributions, and even identify non-stationarity.
-This version uses three transformations on the scalar summary $\theta$, before computing $\widehat{R}$-values.
-Namely, rank-normalization, folding, and localization.
-The definition of the diagnostic itself is equal to the original $\widehat{R}$.
-Therefore, we may follow Vehtari et al.'s formulation (2019, p.
-5) to define both $\widehat{R}$ versions.
-Let $m$ be the total number of chains, $T$ the number of iterations per chain (where $T\geq2$), and $\theta$ the scalar summary of interest.
-For each chain ($\ell = 1, 2, \dots, m$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.
-
-\begin{align*}
-W&=\frac{1}{m} \sum_{\ell=1}^{m} s_{\ell}^{2}, \text { where } s_{\ell}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t \ell)}-\bar{\theta}^{(\cdot \ell)}\right)^{2}. 
-\end{align*}
-
-\noindent We then estimate between-chain variance $B$ (note that we diverge from the typical notation in MI, where $B$ denotes the variance between the estimated quantities of scientific interest $\hat{Q}_{\ell}$).
-$B$ is defined as the variance of the collection of average $\theta$s per chain:
-
-\begin{align*}
-B&=\frac{T}{m-1} \sum_{\ell=1}^{m}\left(\bar{\theta}^{(\cdot \ell)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot \ell)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t \ell)} \text{, } \bar{\theta}^{(\cdot \cdot)}=\frac{1}{m} \sum_{\ell=1}^{m} \bar{\theta}^{(\cdot \ell)}.
-\end{align*}
-
-\noindent From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$.
-$\widehat{R}$ is then obtained as a ratio between this total variance and the within-chain variance:
-
-\begin{equation*}
-\widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
-\text{ where } \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{T-1}{T} W+\frac{1}{T} B.
-\end{equation*}
-
-We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run [@gelm92].
-The assumption underlying this interpretation is that chains are 'over-dispersed' at $t=1$, and reach convergence as $T \to \infty$.
-Over-dispersion implies that the initial values of the chains are 'far away' from the target distribution and each other.
-When the sampled values in each chain are independent of the chain's initial value, the mixing component of convergence is satisfied.
-The variance between chains, $B$, is then equivalent to the variance within chains, $W$, and $\widehat{R}$-values will be close to one.
-High $\widehat{R}$-values thus indicate non-convergence.
-
-## Autocorrelation
-
-Autocorrelation is defined as the correlation between two subsequent $\theta$-values within the same chain [@lync07, p.
-147].
-In this study, we only consider $AC$ at lag 1, i.e., the correlation between the $t^{th}$ and $(t+1)^{th}$ iteration of the same chain.
-Following the same notation as for $\widehat{R}$,
-
-\begin{equation*}
-AC = \left( \frac{T}{T-1} \right) \frac{\sum_{t=1}^{T-1}(\theta_t - \bar{\theta}^{(\cdot m)})(\theta_{t+1} - \bar{\theta}^{(\cdot m)})}{\sum_{t=1}^{T}(\theta_t - \bar{\theta}^{(\cdot m)})^2}.
-\end{equation*}
-
-We can interpret $AC$-values as a measure of non-stationarity.
-If there is dependence between subsequent $\theta$-values in imputation chains, $AC$-values are non-zero.
-Positive $AC$-values occur when $\theta$-values are recurring (i.e., high $\theta$-values are followed by high $\theta$-values, and low $\theta$-values are followed by low $\theta$-values).
-Recurrence within imputation chains may lead to trending.
-Negative $AC$-values occur when $\theta$-values of subsequent iterations are less similar, or diverge from one another.
-Divergence within imputation poses no threat to the convergence of the algorithm---it may even speed up convergence.
-Complete stationarity is reached when $AC=0$.
-As non-convergence diagnostic, our interest is in positive $AC$-values.
-
-## Thresholds
-
-It is unlikely that iterative algorithms such as MICE will achieve the ideal values of $\widehat{R}=1$ and $AC=0$.
-Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state.
-The aim, therefore, is to reach approximate convergence [@gelm13].
-Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} > 1$), and there is little dependency between subsequent iterations of imputation chains ($AC > 0$).
-In practice, we diagnose non-convergence when approximate convergence is violated, i.e., when $\widehat{R}$ and $AC$ exceed a certain threshold.
-The conventional thresholds to diagnose non-mixing are $\widehat{R} > 1.2$ [@gelm92] or $\widehat{R} > 1.1$ [@gelm13].
-@vehtari2021 proposed a much more stringent threshold of $\widehat{R} > 1.01$.
-The magnitude of $AC$-values may be evaluated statistically, using a Wald test with $AC=0$ as null hypothesis [@box15].
-$AC$-values that are significantly higher than zero indicate non-stationarity.
-<!-- Threshold Rhat > 1.1 @gelm13, p. 288 -->
-
-## In practice
+- We use real data: the `boys` dataset from the `mice` package
+
+- We are interested in predicting age from the other variables, in particular in the regression coefficient of `hgt`
+
+- We compare non-convergence identified using visual inspection versus rhat in the chain variances, scientific estimate and lambda.
+
+- The figures show results of a `mice` run with 20 iterations but otherwise default settings.
+
+From the traceplot of the chain means (see \@ref(fig:case)A) it seems that mixing improves up-to 10 iterations, while trending is only apparent in the first three iterations.
+
+This figure (\@ref(fig:case)B) shows that 7 iterations are required before the $\widehat{R}$-values of the chain means drop below the threshold for non-convergence.
+
+The $\widehat{R}$-values for the scientific estimate reaches the threshold much sooner, when $n_{it}=14$ (see \@ref(fig:case)C).
+
+According to the $\widehat{R}$-values with $\lambda$ as parameter, at least 15 iterations are required (see \@ref(fig:case)D).
 
@@
 
-By visually inspecting the imputation chains in panel A, we conclude that the two scenarios show very different convergence.
+<!-- By visually inspecting the imputation chains in panel A, we conclude that the two scenarios show very different convergence.
 The typically converged algorithm initially portrays some signs of non-mixing (around $t=2$), but intermingles nicely overall.
@@
 In this scenario, we see a lot of trending initially (up-to $t=6$), after which the chains reached a somewhat more stationary state.
-
-When we look at panel B, we conclude something weird.
+-->
+
+<!-- When we look at panel B, we conclude something weird.
 The $AC$-values calculated with the default function indicate equal performance for the typical convergence and the pathological non-convergence scenarios (up-to $t=5$), while there is obvious trending in the $\theta$s of the latter.
@@
 We will, therefore, only consider manually calculated $AC$-values as non-convergence diagnostic.
-
-In panel C, we do not see a lot of difference between the two scenarios either.
+-->
+
+<!-- In panel C, we do not see a lot of difference between the two scenarios either.
 The $\widehat{R}$-values are even somewhat lower for the non-convergence scenario than for the typical convergence.
@@
 Taken together, this version of $\widehat{R}$ is sub-optimal for assessing non-convergence in MI.
-
-The $\widehat{R}$-values in panel D are conform our expectations.
+-->
+
+<!-- The $\widehat{R}$-values in panel D are conform our expectations.
 The adapted version of $\widehat{R}$ depicted here does indicate less signs of non-convergence as the number of iterations goes up.
@@
 are, therefore, more similar to the original $\widehat{R}$ by Gelman and Rubin.
-
-In short, both non-convergence diagnostics have a version that may be appropriate for iterative imputation algorithms.
+-->
+
+<!-- In short, both non-convergence diagnostics have a version that may be appropriate for iterative imputation algorithms.
 We will compute $\widehat{R}$ conform @vehtari2021, and calculate $AC$ manually.
@@
 We evaluate the performance of these eight sets of diagnostic methods (two diagnostics; four $\theta$s) through simulation.
-
-# Simulation study
-
-The simulation study aims to determine to what extend non-convergence affects iterative imputation algorithms.
-Specifically, we are interested in the validity of inferences drawn from incomplete data using the MICE algorithm.
-And, whether invalid inferences due to non-convergence may be detected using eight sets of diagnostic methods.
-If the estimated quantities of scientific interest $\bar{Q}$ are unbiased and confidence-valid estimates of ${Q}$, we will conclude that the algorithm is sufficiently converged for practical purposes.
-In contrast to this, we defined *approximate* convergence as the most converged state that the algorithm can obtain.
-Continued iteration after reaching approximate convergence does not yield better mixing or stationarity (indicated by non-improving $\widehat{R}$- and $AC$-values).
-Continued iteration after obtaining valid inferences may lead to a more converged state of the algorithm, but not better estimates.
-
-To induce non-convergence, we consider two sets of simulation conditions: 'missingness' and 'early stopping'.
-Missingness refers to the severity of the missing data problem, which we determine by the proportion of incomplete cases in dataset $y$, $p_{\mathrm{inc}}$.
-Early stopping is defined by the number of iterations in the imputation algorithm, $T$.
-
-The missingness conditions are chosen to reflect the difficulty of the missingness problem.
-The underlying assumption is that a low proportion of incomplete cases $p_{\mathrm{inc}}$ leads to quick algorithmic convergence, since there is a lot of information in the observed data.
-A higher $p_{\mathrm{inc}}$ yields slower convergence---unless there is so little information in $y_{\mathrm{obs}}$ that this is outweighed by the random component in the imputation algorithm.
-Then, convergence to a stable but highly variable state may be reached instantly.
-Still, we expect that missingness conditions with higher $p_{\mathrm{inc}}$ will result in more signs of non-convergence.
-
-The assumption inherent to the early stopping conditions is that terminating the imputation algorithm too early causes non-convergence.
-Generally, the algorithm will not reach convergence if $T=1$, because the imputed values in the first iteration (at $t=1$) depend on the starting values [which are sampled randomly from the set of observed datapoints; @buur18].
-As the number of iterations increases, the imputation chains should become independent of the initial values, until approximate convergence is reached (i.e., when an additional iteration does not lead to a more converged state).
-We expect that we can induce non-convergence by early stopping in at least those conditions where $T$ is smaller than the default number of iterations in `mice`, $T=5$ [@mice].
-
-## Hypotheses
-
-1.
-We expect that simulation conditions with a high proportion of incomplete cases $p_{\mathrm{inc}}$ and a low number of iterations $T$ will more often result in biased, invalid estimates of the quantities of scientific interest $Q$.
-
-2.
-We hypothesize that $\widehat{R}$ and $AC$ will correctly identify signs of non-convergence in those simulation conditions where any $\bar{Q}$ is *not* an unbiased and confidence-valid estimate of $Q$.
-
-3.
-We hypothesize that the recommended thresholds to diagnose non-convergence with $\widehat{R}$ ($\widehat{R} > 1.2$, $\widehat{R} > 1.1$, and $\widehat{R} > 1.01$) may be too stringent for iterative imputation applications.
-In an empirical study, where $\widehat{R}$ was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold $\widehat{R}>1.2$.
-Yet, scientific estimates were insensitive to continued iteration from $t=5$ onward [@lace07].
-We, therefore, suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates.
-In contrast to this, signs of non-convergence can be under-estimated by $\widehat{R}$, in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
-437].
-In  `mice`, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values.
-In practice, we do not expect this to cause problems for identifying non-convergence with $\widehat{R}$.
-
-4.
-We expect that high $AC$-values are implausible in iterative imputation algorithms with typical convergence.
-After only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
-
-5. We further hypothesize that multivariate $\theta$s are better at detecting non-convergence than univariate $\theta$s.
-
-## Set-up
-
-We investigate non-convergence of the MICE algorithm through model-based simulation in `R` [version 4.4.2; @R].
-The number of simulation repetitions is 2000, and the simulation set-up is summarized in the pseudo-code below.
-The complete `R` script of the simulation study is available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint) [TODO: make Zenodo DOI].
-
-```
-# pseudo-code of simulation 
-for (number of simulation runs) {
-  1. simulate complete data 
-  for (missingness conditions) {
-    2. create missingness
-    for (number of iterations) {
-      3. impute missingness
-      4. estimate quantities of scientific interest
-      5. apply performance measures to the estimates
-      6. compute non-convergence diagnostics
-    }
-  }
-7. aggregate outcomes across simulation runs
-```
-
-### Data-generating mechanism.
-
-The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem.
-Let the predictor space be defined as
-
-$$
-\begin{aligned}
-\begin{pmatrix}X_1\\
-X_2\\
-X_3
-\end{pmatrix} \sim \mathcal{N}
-\begin{bmatrix}
-\begin{pmatrix}
-0\\
-0\\
-0
-\end{pmatrix}\!\!,
-\begin{pmatrix}
-1 & 0.5 & 0.5 \\
-0.5 & 1 & 0.5 \\
-0.5 & 0.5 & 1
-\end{pmatrix}
-\end{bmatrix}\!\!\text{.}\\[2\jot]
-\end{aligned}
-$$
-
-In each simulation repetition, a finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm].
-Subsequently, a fourth variable is constructed as outcome variable $Y$.
-For each unit $i = 1, 2,..., N$, let
-
-$$
-Y_i = X_{1i} + X_{2i} + X_{3i} + \epsilon_i ,
-$$
-where $\epsilon \sim \mathcal{N}(0, 1)$.
-This results in a dataset $y$, where $y_{\mathrm{mis}} = \emptyset$.
-In other words, $y$ contains the complete set of observations for all units $i$ on all variables $j$ (where $j =  Y, X_1, X_2, X_3$).
-
-### Scientific estimands.
-
-We consider four multivariate quantities of scientific interest ($Qs$). Namely, the regression coefficients of the analysis model,
-
-$$Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,$$
-where $Y'$ is the expected value of the outcome.
-From the data-generating mechanism, we obtain the true values of the scientific estimands: $\beta_0 = 0$, $\beta_1 = \beta_2 = \beta_3 = 1$.
-The scientific estimands are estimated by $\bar{Q}$ in each simulation condition.
-
-<!-- The final estimand is the proportion of variance explained by the set of predictors: coefficient of determination $r^2$ (note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$).
--->
-
-### Simulation conditions.
-
-We consider two sets of simulation conditions, 'missingness conditions' and 'early stopping' conditions. The simulation study is a fully factorial design, as visualized in Table XYZ.
-
-In each simulation repetition, we *ampute* the complete dataset conform six missingness conditions: two missing data mechanisms and three proportions of incomplete cases.
-The missing data mechanisms are 'missing completely at random' [MCAR; @rubin87], where the probability to be missing is the same for all $N \times k$ cells in $y$, and a right-tailed 'missing at random' (MAR) mechanism, where the probability to be missing is a function of the observed data and higher values are more likely to be missing.
-The proportion of incomplete cases $p_{\mathrm{inc}}$ is set to 25%, 50%, and 75%.
-
-<!-- Table -->
-<!-- mechanisms: MCAR, MAR -->
-<!-- proportion of incomplete cases: 25%, 50%, 75% -->
-
-We consider all possible multivariate patterns of missingness, as visualized in Figure \ref{fig:patterns}.
-
-```{r patterns, echo=FALSE, eval=FALSE}
-source("../R/generate_datasets.R")
-pat <- create_patterns(3)
-d <- dat[1:10,]
-is.na(d) <- pat == 1
-d
-ggmice::plot_pattern(d)
-```
-
-We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\mathrm{obs}}, y_{\mathrm{mis}}$} for each of the six missingness conditions.
-
-We impute the missing values in {$y_{\mathrm{obs}}, \dot y_{\mathrm{mis}}$} according to the early stopping conditions.
-That is, we vary the number of iterations in the imputation algorithm between one and one hundred ($T = 1,2,.,100$).
-All imputation procedures are performed using `mice` [function `mice()`; @mice], with Bayesian linear regression imputation, and five imputation chains ($m=5$).
-This results in $m=5$ sets of imputations for each of the one hundred early stopping conditions, $\dot y_{\mathrm{imp}, \ell}$.
-
-Subsequently, we obtain the completed data per imputation, {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice].
-Estimates of scientific quantities $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R].
-We pool the resulting $m$ estimates, $\hat{Q}_\ell$, into a single $\bar{Q}$ conform Rubin's rules [function `mice::pool()`; @mice].
-
-### Methods.
-
-We consider eight sets of diagnostic methods to identify non-convergence, by applying $\widehat{R}$ and $AC$ on four different $\theta$s.
-Univariate $\theta$s are obtained from $\dot y_{\mathrm{imp}, \ell}$.
-The model-independent multivariate $\theta$, $\lambda_{1}$, is obtained from {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$} as the variance in the first PCA component [function `stats::princomp`; R].
-The model-dependent multivariate $\theta$ is one of the quantities of scientific interest: the estimated regression coefficients in {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}.
-We apply the two non-convergence diagnostics on the $\theta$s by implementing the adapted version of $\widehat{R}$ and by manually programming the $AC$ function.
-
-### Performance measures.
-
-As recommended by @buur18, we evaluate the performance of the imputation algorithm with bias, coverage rate (CR), and confidence interval width (CIW).
-Bias is calculated for each $Q$, whereas CR and CIW are only obtained for $Q = \beta$.
-We calculate bias as $\bar{Q} - Q$.
-CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$.
-Let
-$$\text{CI} = \bar{Q} \pm t_{(m-1)} \times \mathrm{SE}_{\bar{Q}},$$
-where $t_{(M-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $\mathrm{SE}_{\bar{Q}}$ is the square root of the pooled variance estimate.
-If we obtain nominal coverage (CR = 95%), we can conclude that $\bar{Q}$ is a confidence-valid estimate of $Q$.
-Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$.
-CIW is of interest because it is a measure of efficiency.
-Under nominal coverage, short CIs are preferred, since wider CIs indicate lower statistical power.
-
-We evaluate the diagnostic methods to identify non-convergence against these performance measures.
-$\widehat{R}$ and $AC$ should identify non-convergence in simulation conditions, where $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
-<!-- Performance measures @buur18, $\S$ 2.5.2 -->
-
-<!-- We could also look at distributional characteristics, and plausibility of imputed values, see @vinknd. For now, this is outside of the scope of this study. -->
-
-# Results
-
-Our results show the performance of the MICE algorithm under different conditions of missingness and early stopping.
-For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias.
-For univariate scientific estimands ($Q=\mu$ and $Q=\sigma$) we observe the largest bias in the outcome variable $Y$.
-And for the regression coefficients ($Q=\beta$) the bias is most pronounced in $\beta_1$ (the effect of $X_1$ on $Y$).
-Since there is just one estimate for $Q=r^2$ to evaluate, we consider $Q=\mu_Y, \sigma_Y,r^2, \beta_1$.
-In the figures, we present all missingness conditions, but only early stopping conditions where $1 \leq T\leq50$.
-That is, the results are more or less stable for conditions where $T \geq 30$.
-Full results are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).
-
-## Quantities of scientific interest
-
-```{r Qs, fig.height = 6.8, fig.cap="Impact of non-convergence on statistical inferences.
-Depicted are the bias, coverage rate (CR) and confidence interval width (CIW) of the estimates $\\bar{Q}$.
-We present the subset of $Q$s with the worst performance in terms of bias: the mean and standard deviation of $Y$, and the regression coefficient of $X_1$.
-The solid gray line depicts the targets (unbiased estimates and nominal coverage rate)."}
-
-# mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_annotation(tag_levels = "A", tag_suffix = ".") + plot_layout(ncol = 2, guides = "collect") 
-```
-
-Figure \ref{fig:Qs} displays the influence of missingness and early stopping on the estimated quantities of scientific interest, $Q$s.
-In the first row of the figure, we see the bias in the estimated descriptive statistics ($Q=\mu_Y$ and $Q=\sigma_Y$; panels A and B).
-The second row shows the bias in the multivariate scientific estimands ($Q=r^2$ and $Q=\beta_1$; C and D).
-The third row consists of the coverage rate (CR) and confidence interval width (CIW) of the estimated regression coefficient ($Q=\beta_1$; E and F).
-
-### $Q=\mu$.
-The estimated univariate means seem unaffected by early stopping.
-This implies that approximately unbiased estimates may be obtained with as little as one iteration ($T \geq 1$).
-The bias in $\bar{Q}$ depends solely on the proportion of incomplete cases, with $p_{\mathrm{inc}}\geq.75$ leading to more extreme biases than other conditions.
-Completely unbiased estimates are only obtained in conditions where $p_{\mathrm{inc}}\leq.50$.
-This is curious since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.
-Note however, that the magnitude of the bias in these conditions is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$).
-We, therefore, conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.
-
-### $Q=\sigma$.
-The estimated standard deviations are not substantially impacted by early stopping either.
-While the estimates for $T=1$ are not equal to other conditions, the bias is actually *less* severe than for $T\geq2$.
-Therefore, no iteration at all is needed to obtain approximately unbiased estimates for $Q=\sigma$.
-Completely unbiased estimates are only obtained in conditions where $p_{\mathrm{inc}}=.05$.
-The other missingness conditions are impacted by non-convergence in the order of increasing $p_{\mathrm{inc}}$.
-Similar to $Q=\mu$, the magnitude of the bias in $Q=\sigma$ is negligible.
-We, therefore, conclude that neither one of the univariate $Q$s is affected substantially by non-convergence.
-
-### $Q=r^2$.
-In contrast to the univariate estimates, the estimated coefficient of determination is clearly affected by early stopping, see Figure \ref{fig:Qs}C.
-The bias in $Q$ appears to decrease with each additional iteration, while the magnitude of the bias still depends on the proportion of incomplete cases.
-Therefore, we see that the number of iterations that is necessary to reach stable, non-improving estimates also differs across missingness conditions.
-In conditions where $p_{\mathrm{inc}}=.05$, estimates are unbiased after one iteration.
-The highest number of iterations necessary to reach approximate unbiasedness is seven ($p_{\mathrm{inc}}=.95$; $T\geq7$).
-Completely unbiased estimates are only obtained in conditions where $p_{\mathrm{inc}}\leq.25$.
-This implies that even for biased estimates, $T\geq7$ would suffice to reach a stable solution.
-
-### $Q=\beta$.
-For the estimated regression coefficients, we first consider bias, before discussing CR and CIW.
-As Figure \ref{fig:Qs}D shows, the bias in $\bar{Q}$ is affected by both the proportion of incomplete cases and the number of iterations.
-Similar to $Q=r^2$, we observe approximately unbiased estimates after at least one, and at most seven iterations, depending on $p_{\mathrm{inc}}$.
-Completely unbiased estimates are only obtained in conditions where $T\geq3$ and $p_{\mathrm{inc}}\leq .50$.
-Despite persistent bias in conditions where $p_{\mathrm{inc}}\geq.75$, the coverage rates are non-improving after just three iterations, irrespective of $p_{\mathrm{inc}}$.
-We conclude that any condition where $T\geq3$ yields approximately nominal coverage rates, and thus confidence-valid estimates.
-The nominal coverages can be explained by the CIWs.
-The CIWs depicted in Figure \ref{fig:Qs}F are conform the theoretical foundation of MI: CIs are wider in conditions with higher proportions of incomplete cases (i.e., there is less information in the data, and thus more uncertainty due to missingness).
-Since conditions with higher $p_{\mathrm{inc}}$ result in both wider CIs and more severe bias, the true value of $Q$ may be included in the CI, despite of the bias in $\bar{Q}$.
-We conclude that approximately unbiased estimates may be obtained in conditions where $T\geq7$, whereas confidence-valid estimates require at most three iterations.
-
-### Summary. 
-These results demonstrate that the estimates of univariate scientific estimands $Q$ are not impacted by early stopping of the MICE algorithm or the proportion of incomplete cases in $y$.
-Unbiased estimates may be obtained after just one iteration.
-Multivariate estimates, by contrast, are affected by both the number of iterations and the proportion of missing cases.
-Completely unbiased estimates are only obtained under low to moderate missingness ($p_{\mathrm{inc}}\leq.50$), after at most three iterations.
-We observe approximately unbiased estimates after at most seven iterations (for any $p_{\mathrm{inc}}$ considered).
-This implies that the algorithm produces stable, non-improving estimates when $T\geq7$.
-
-## Non-convergence diagnostics
-
-Figure \ref{fig:rhats-acs} displays results of the eight sets of diagnostic methods.
-The two columns in the figure represent the  non-convergence diagnostics $\widehat{R}$ and $AC$; the rows depict the four scalar summaries $\theta$ under consideration.
-The first row in Figure \ref{fig:rhats-acs} thus shows $\widehat{R}$ and $AC$ applied on the chain means (panels A and B).
-In the second row, we see the diagnostics applied on the chain variances (C and D).
-The third row depicts the same for $\theta=\hat{Q}$ (E and F).
-And in the last row, we see the diagnostics applied on the novel $\theta$ to consider: $\lambda_1$ (G and H).
-We evaluate the performance of the methods by establishing whether they correctly identify conditions in which $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
-As we concluded in the last section, substantially biased estimates were only obtained in conditions where $T\leq6$, and we observed non-nominal coverage only when $T\leq3$.
-However, there was some persistent bias in conditions where $p_{\mathrm{inc}}\geq.75$, irrespective of the number of iterations.
-If the diagnostic methods are appropriate for MI, we will detect non-convergence in these conditions.
-
-```{r rhats-acs, fig.height = 6.8, fig.cap="Non-convergence diagnostics $\\widehat{R}$ and $AC$ applied on several $\\theta$s.
-The left-hand side of the figure contains $\\widehat{R}$-values, and the right-hand side contains $AC$-values.
-Depicted in the rows are the scalar summaries $\\theta$: chain mean, chain variance, the quantity of scientific interest $\\hat{Q}$, and the first eigenvalue of the variance-covariance matrix $\\lambda_1$.
-The dashed gray lines depict the different thresholds to diagnose non-convergence."}
-
-# mean_Rh + mean_AC + var_Rh + var_AC + est_Rh + est_AC + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 2) + plot_annotation(tag_levels = "A", tag_suffix = ".")
-```
-
-### $\theta=$ chain mean. 
-As expected, $\widehat{R}$ lowers with the number of iterations.
-After an initial 'dip' in conditions where $T=3$, we observe a gradual decrease in $\widehat{R}$-values with every additional iteration.
-This implies improving convergence, until the decrease tapers off after 30 to 40 iterations.
-Using the threshold $\widehat{R}>1.2$, we diagnose non-convergence in conditions where $T\leq7$, whereas the threshold 1.1 is exceeded when $T\leq13$.
-The very strict threshold 1.01 is surpassed in all conditions $T\leq100$ (not shown), and therefore not informative under the current specifications.
-With this set of methods, we fail to identify non-convergence in conditions with persistent bias due to missingness.
-
-Similarly, $AC$ decreases with higher $T$s.
-$AC$ indicates improving stationarity in conditions where $T\leq6$.
-The $AC$-values do not exceed the threshold defined by statistical significance of the $AC$-values.
-Based on this threshold, we fail to diagnose non-convergence in any condition.
-
-### $\theta=$ chain variance.
-As panel C and D in Figure \ref{fig:rhats-acs} show, the results of this set of methods are highly similar to those with chain means as $\theta$.
-This holds for both $\widehat{R}$ and $AC$.
-The only apparent difference is in the $\widehat{R}$ threshold 1.1.
-Instead of diagnosing non-convergence in conditions where $T\leq13$, we now diagnose this when $T\leq11$.
-All other observations are equivalent.
-
-### $\theta=\hat{Q}$.
-The $\widehat{R}$-values in Figure \ref{fig:rhats-acs}E show a similar trend across iterations as we observed for chain means and chain variances.
-Only conditions with a very high  proportion of missing cases ($p_{\mathrm{inc}}\geq.75$) diverge from the earlier observations.
-The $\widehat{R}$-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations.
-The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are 11 for $\widehat{R}>1.2$, 23 for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).
-
-The $AC$-values also follow earlier observations, with the same exception as for $\widehat{R}$ (when $p_{\mathrm{inc}}\geq.75$).
-To reach maximum stationarity, these conditions require at most seven iterations instead of five.
-Moreover, the $AC$-values in these conditions exceed the threshold to diagnose non-convergence.
-Interestingly, this only occurs after ten or even thirty iterations, while we would expect more signs of trending early in the iterations.
-Since the $AC$-values are not improving or worsening when $T\geq7$, we conclude that non-convergence is incorrectly diagnosed based on the threshold.
-
-### $\theta=\lambda_{1}$.
-Once again, we observe $\widehat{R}$-values with a similar trend across iterations.
-For this $\theta$, the results differ only in the most extreme missingness condition.
-When $p_{\mathrm{inc}}=.95$, we diagnose non-convergence at $T\leq9$ according to $\widehat{R}>1.2$, $T\leq19$ for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).
-The 'dip' in $\widehat{R}$-values at $T=3$ is even more pronounced than with other $\theta$s.
-Some $\widehat{R}$-values even overcome the threshold of 1.2.
-If we would terminate the algorithm at $T=3$, we would incorrectly conclude that the algorithm reached approximate convergence, according to this threshold.
-
-$AC$-values are systematically higher for this $\theta$ than for other scalar summaries.
-The $AC$-values also taper off at a later point in the iterations.
-This suggests improving stationarity up-to ten, or even thirty iterations, depending on the missingness condition.
-According to the threshold, we should diagnose non-convergence at some point in any missingness condition.
-However, this occurs only after reasonably decreasing $AC$-values are obtained---again suggesting that non-convergence is incorrectly diagnosed.
-
-### Summary. 
-Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
-Under low to moderate missingness, it does not seem to matter on which $\theta$ the non-convergence diagnostics are applied.
-If we look at complete unbiasedness, however, we notice that univariate $\theta$s fail to diagnose the persistent bias in conditions where $p_{\mathrm{inc}}\geq.75$.
-
-The number of iterations necessary to obtain approximately unbiased, confidence-valid estimates corresponds to the $\widehat{R}$ threshold 1.2.
-The thresholds 1.1 and 1.01 seem too strict compared to the validity of the inferences.
-The threshold to diagnose significant $AC$-values does not appear to be appropriate in at least the current set-up, and perhaps in iterative imputation in general.
-A better heuristic to diagnose non-stationarity with $AC$ may be through evaluation of the $AC$-values across iterations: If the values do not substantially decrease with $T$, approximate stationarity may be concluded.
-<!-- [similar to the elbow heuristic in PCA]. -->
-
-# Discussion
-
-Before drawing any further conclusions, let us first return to the hypotheses that we formulated for the simulation study.
-
-1.
-As hypothesized, biased, invalid estimates of quantities of scientific interest occurred more often in simulation conditions with a high proportion of incomplete cases $p_{\mathrm{inc}}$ and a low number of iterations $T$.
-However, this only holds for multivariate scientific estimands.
-Univariate $Q$s are not substantially impacted by early stopping.
-
-2.
-Our results also support the hypothesis that $\widehat{R}$ and $AC$ would correctly identify signs of non-convergence.
-We observed higher $\widehat{R}$ and $AC$-values in conditions where $\bar{Q}$s were biased, invalid estimates of $Q$s.
-
-3.
-We hypothesized that the recommended thresholds to diagnose non-convergence with $\widehat{R}$ would be too stringent for iterative imputation applications, compared to the number of iterations necessary for statistically valid inferences.
-This hypothesis is partially supported: On the one hand, the 1.01 and 1.1 thresholds indeed seem to over-estimate the signs of non-convergence.
-On the other hand, however, we observed that the traditional threshold of 1.2 roughly corresponds to the number of iterations required to obtain approximately unbiased, confidence-valid estimates.
-
-4.
-The results of this simulation study do not support the hypothesis that high $AC$-values are implausible in iterative imputation algorithms with typical convergence.
-We expected that the randomness induced by the algorithm would effectively mitigate the risk of dependency within chains after a few iterations.
-In this study, however, we observed $AC$-values as high as theoretically possible ($AC=1$).
-The second part of the hypothesis is not refuted: the $AC$-values did decrease quickly as a function of $T$.
-
-5.
-Our final hypothesis was that multivariate $\theta$s would be better at detecting non-convergence than univariate $\theta$s.
-Multivariate $\theta$s did indeed show superior performance under certain conditions (e.g., multivariate $Q$s and extreme missingness).
-
-## Conclusion
-
-We have shown that non-convergence in iterative imputation algorithms goes hand in hand with biased, invalid estimates---at least for the multivariate quantities of scientific interest considered in this study.
-Identifying non-convergence may be a crucial aspect of drawing valid statistical inferences from incomplete data.
-We concluded that the current practice of visually inspecting non-convergence through traceplots of univariate scalar summaries of the state-space of the algorithm does not suffice.
-We, therefore, considered eight sets of diagnostic methods to identify non-convergence: two non-convergence diagnostics ($\widehat{R}$ and $AC$), two univariate $\theta$s, a model-dependent multivariate $\theta$, and a novel, model-independent multivariate $\theta$.
-Signs of non-convergence due to early stopping were identified with each of the methods.
-Bias due to high proportions of incomplete cases was only identified with multivariate $\theta$s.
-
-Since we obtained approximately unbiased, confidence-valid estimates after at most seven iterations, we conclude that the $\widehat{R}$ threshold 1.2 is the most appropriate diagnostic cut-off.
-The threshold to diagnose non-stationarity with $AC$ does not seem to apply.
-However, some signs of non-convergence were detected in simulation conditions many more iterations.
-Under moderate missingness ($p_{\mathrm{inc}}\leq.50$) $\widehat{R}$-values decreased substantially until 30 to 40 iterations, which implies that mixing in the algorithm still improved with each additional iteration.
-$AC$-values only improved until about six iterations, suggesting minimal improvement in trending was obtained beyond $T=6$ in the current simulation set-up.
-
-The main finding of this study is that valid inferences may be obtained much quicker than approximate algorithmic convergence is reached.
-Under the current specifications, univariate $Q$s did not require algorithmic convergence at all.
-They are unbiased almost instantly after the algorithm is initiated.
-Approximately unbiased, confidence-valid estimates of multivariate $Q$s were obtained after a maximum of seven iterations.
-Continued iterations beyond $T = 7$ did not yield better estimates.
-
-## Implications
-
-Iterative imputation algorithms such as MICE are known to yield valid results under severe missingness.
-With this study, we have shown that valid inferences can also be obtained under a combination of severe missingness and early stopping.
-Based on our results, the traditional threshold of $\widehat{R}> 1.2$ would be appropriate for diagnosing non-convergence in iterative imputation algorithms.
-This is, however, not in line with the recent recommendations by @vehtari2021 to lower the threshold to 1.01.
-The discrepancy between our conclusion and Vehtari et al.
-may be explained by the nature of iterative imputation algorithms.
-In the limit, iterative imputation algorithms have the same characteristics as other MCMC algorithms.
-But, in imputation procedures, a part of the distribution is already determined by the observed data, whereas the entire distribution is unknown in many other MCMC applications.
-Since we combine a known distribution with an unknown distribution, valid estimates may be reached much sooner.
-Convergence may, therefore, be diagnosed at a less stringent threshold.
-
-The threshold to diagnose non-convergence with $AC$ does not seem appropriate at all in iterative imputation algorithms.
-A better diagnostic cut-off would be the number of iterations at which an additional iteration does not substantially decrease the $AC$-values.
-In practice, this implies that the default number of iterations in MICE is not sufficient.
-$AC$ can only be computed if $T\geq3$, while the current default in the `mice` package is five [@mice].
-Identifying a reasonable decrease in the $AC$-values across iterations (an 'elbow') requires more than three observations.
-We, therefore, suggest to increase the default number of iterations in MICE to ten.
-The computational cost of five additional iterations has become less of a burden since MICE was introduced [-@mice].
-Moreover, an increased number of iterations would grant the opportunity to exclude the first iterations from evaluation with $\widehat{R}$.
-Excluding $T\leq3$ provides empirical researchers with a less ambiguous heuristic to diagnose non-convergence, because we expect the initial 'dip' in $\widehat{R}$-values to disappear.
-
-The potential scale reduction factor metric assumes over-dispersed starting values of the iterative algorithm.
-The MICE algorithm does not start in an over-dispersed state.
-MICE does not rely on starting values for parameters at all: instead, the algorithm is initialized based on a 'zeroth' iteration, where missing values are filled in using a random draw from observed values.
-Under MCAR, this would typically not yield an over-dispersed state at all, depending on the parameter of interest: means and variances are not affected, regression coefficients and other multivariate parameters start biased downwards (because sampling starting values distorts multivariate structures in the data).
-Whether this accrues to over-dispersion is questionable.
-Under MAR (or MNAR), some over-dispersion might occur when the observed data and imputed data differ impactfully.
-The algorithm should 'escape' the initial state (based on the starting values drawn from observed data alone).
-Over iterations, all parameters of interest (e.g.
-means, variances, multivariate estimands) should converge towards a stable state.
-Moreover, the MICE algorithm is initialized with more information than a typical MCMC algorithm: parameter estimates depend not only on imputed data, but also on observed data that does not change over iterations.
-Therefore, the initial state of the algorithm is 'too good' to observe a relative decrease in $\widehat{R}$.
-
-## Recommendations for empirical researchers
-
-Based on our results, we formulate several recommendations for empirical researchers who employ iterative imputation to draw inferences from incomplete data.
-We suggest that non-convergence may still be evaluated visually, but in addition to inspecting univariate summaries of the state-space of the algorithm ($\theta$s, e.g., chain means), multivariate $\theta$s should be considered.
-We propose the following steps:
-
-1. First, check traceplots of the default $\theta$s (e.g., chain means and chain variances) for signs of pathological non-convergence. Adjust the imputation model if necessary.
-
-2.
-Subsequently, decide which multivariate $\theta$s to track across iterations.
-The novel $\theta$ that we propose in this study, $\lambda_1$ is scientific model-independent and may thus always be employed.
-Alternatively, specify your own scalar summary of interest [see e.g., @buur18].
-Monitor these $\theta$s through visual inspection, or using a non-convergence diagnostic.
-
-3.
-Compute non-convergence diagnostics $\widehat{R}$ and $AC$.
-Do *not* use the original implementation to calculate $\widehat{R}$ [@gelm92], or the `R` function `stats::acf()` to compute $AC$ [@R].
-Instead, calculate $\widehat{R}$ conform @vehtari2021 and compute autocorrelations manually (see e.g., [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint)).
-
-4.
-And finally, use the threshold $\widehat{R}>1.2$ to diagnose non-mixing, and assess stationarity by plotting the autocorrelation over iterations.
-Keep iterating until the threshold for mixing is overcome, and until reasonably decreasing asymptotic $AC$-values are obtained.
-At that point, inferences are unlikely to improve by continued iteration.
-
-## Limitations
-
-Much remains unknown about non-convergence in iterative imputation algorithms.
-Even though we have demonstrated the appropriateness of $\widehat{R}$ and $AC$ as non-convergence diagnostics in this study, results may not extrapolate to other situations.
-In this study, we only considered the iterative imputation algorithm implemented in the `mice` package.
-The performance of the non-convergence diagnostics may depend on the characteristics of this specific algorithm.
-A potential threat to the ability to detect non-mixing with $\widehat{R}$, for example, is the assumption of over-dispersed initial states.
-In `mice`, the algorithm is initiated by sampling starting values at random from the observed data, whereas other imputation software may use the average of the observed cases.
-The latter could lead to a systematical under-estimation of the variability in the first few iterations and violate the assumption of over-dispersion.
-In those cases, it would be more difficult to detect non-convergence with $\widehat{R}$.
-Therefore, the diagnostic might not be  appropriate outside of `mice`.
-
-Moreover, the current simulation conditions were restricted to a single missingness mechanism.
-Proper performance under a 'missing completely at random' (MCAR) mechanism is a necessary condition for any missing data method.
-It does not, however, guarantee equal performance under different missingness mechanisms.
-Future research should determine the performance of $\widehat{R}$ and $AC$ under 'missing at random' and 'missing not at random' mechanisms.
-Another parameter to consider in future research is the choice of the imputation method.
-As @murray2018 concluded, the behavior of algorithms such as MICE under certain default imputation models is still an open research question.
-We have investigated the behavior of MICE under only one type of model (Bayesian linear regression).
-Different imputations models might converge more poorly.
-
-The imputation model may even be mis-specified on purpose, to induce non-convergence of different severity levels.
-Qualitatively, we have shown that $\widehat{R}$ and $AC$ are appropriate under clear violation of convergence.
-Quantitatively, we have only demonstrated that they can identify signs of non-convergence under MCAR, when the imputation model is a correctly specified Bayesian linear regression model.
-\newline \newline
-
-\noindent In short, we have shown that iterative imputation algorithms can yield correct outcomes, even when a converged state has not yet formally been reached.
-Any further iterations would then burn computational resources without improving the statistical inferences.
-Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the $\widehat{R}$ and AC diagnostics.
-Of course, it never hurts to iterate longer, but such calculations hardly bring added value.
+-->

### 2025-04-22 simulation.R

@@ -15,7 +15,7 @@ miceadds::source.all("./R")
 set.seed(11)
 
 # parameters
-n_sim <- 100
+n_sim <- 2000
 n_obs <- 200
 n_col <- 3
 corr <- 0.5

### 2025-04-22 Manuscript/manuscript.qmd
sim 3 mech

@@
 
-[TODO: intrgrate this section] In an empirical study, where $\widehat{R}$ was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold $\widehat{R}>1.2$.
+[TODO: integrate this section] In an empirical study with STATA, where $\widehat{R}$ was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold $\widehat{R}>1.2$.
 Yet, scientific estimates were insensitive to continued iteration from $t=5$ onward [@lace07].

### 2025-04-22 simulation.R

@@ -15,14 +15,14 @@ miceadds::source.all("./R")
 set.seed(11)
 
 # parameters
-n_sim <- 2000
+n_sim <- 200
 n_obs <- 200
 n_col <- 3
 corr <- 0.5
 beta <- 1
 betas <- rep(beta, n_col)
 mis_pat <- create_patterns(n_col)
-mis_mech = c("MCAR", "MAR")
+mis_mech = c("MCAR", "MAR", "MNAR")
 mis_prop = c(0.25, 0.5, 0.75)
 n_it <- 100
 

### 2025-04-23 Manuscript/manuscript.qmd
plot results with CI

@@
 performance <- mutate(performance,
-  mech = factor(mech, levels = c("MCAR", "MAR"), ordered = TRUE),
+  mech = factor(mech, levels = c("MCAR", "MAR", "MNAR"), ordered = TRUE),
   prop = factor(prop, levels = c(0.25, 0.5, 0.75), ordered = TRUE),
@@
   summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), \(x) mean(x, na.rm = TRUE)))
+results_ll <- performance |> 
+  group_by(method, mech, prop, .it, term) |> 
+  summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), \(x) quantile(x, probs = c(0.025), na.rm = TRUE))) 
+results_ul <- performance |> 
+  group_by(method, mech, prop, .it, term) |> 
+  summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), \(x) quantile(x, probs = c(0.975), na.rm = TRUE))) 
+results <- left_join(results, results_ll, by = c("method", "mech", "prop", ".it", "term"), suffix = c("", "_ll"))
+results <- left_join(results, results_ul, by = c("method", "mech", "prop", ".it", "term"), suffix = c("", "_ul"))
 ```
@@
 ```{r, echo=FALSE}
-# # plot results per sim rep
-# performance |>
-#   filter(term == "X1", mech == "MAR", method == "MICE", prop == 0.75) |>
-#   ggplot(aes(x = .it, y = bias)) + #, color = as.factor(prop)
-#   geom_line(aes(group = .sim)) +
-#   geom_smooth(method = "loess") +
-#   # labs(
-#   #   title = "Bias regression coefficient",
-#   #   x = "Iteration number",
-#   #   y = "Bias"
-#   # ) +
-#   theme_classic()
+results |>
+  filter(method == "MICE") |> 
+  ggplot(aes(x = .it, y = psrf_mean)) +
+  geom_hline(yintercept = 1, alpha = 0.1) + 
+  geom_line() +
+  geom_line(aes(y = psrf_mean_ll), linetype = "dashed") +
+  geom_line(aes(y = psrf_mean_ul), linetype = "dashed") +
+  facet_grid(term ~ prop + mech) +
+  theme_classic()
 ```
@@
   filter(method == "MICE") |> 
-  ggplot(aes(x = .it, y = psrf_mean, color = mech, linetype = mech)) +
+  ggplot(aes(x = .it, y = psrf_mean)) +
   geom_hline(yintercept = 1, alpha = 0.1) + 
   geom_line() +
-  facet_grid(term ~ prop) +
+  facet_grid(term ~ prop + mech) +
   theme_classic()

### 2025-04-23 figures/plot_results.R

@@ -1,0 +1,19 @@
+# plot results 
+plot_results <- function(summarized_results, metric, hline, mis = "MAR") {
+  # Check if the metric is valid
+  if (!metric %in% c("bias", "cov", "ciw", "ac_mean", "ac_sd", "psrf_mean", "psrf_sd")) {
+    stop("Invalid metric. Choose from 'bias', 'cov', 'ciw', 'ac_mean', 'ac_sd', 'psrf_mean', or 'psrf_sd'.")
+  }
+  metric_ll <- paste0(metric, "_ll")
+  metric_ul <- paste0(metric, "_ul")
+  summarized_results |>
+    filter(method == "MICE") |> 
+    filter(mis %in% mech) |>
+    ggplot(aes(x = .it, y = .data[[metric]])) +
+    geom_hline(yintercept = hline, alpha = 0.1) + 
+    geom_line() +
+    geom_line(aes(y = .data[[metric_ll]]), linetype = "dashed", alpha = 0.5) +
+    geom_line(aes(y = .data[[metric_ul]]), linetype = "dashed", alpha = 0.5) +
+    facet_grid(term ~ prop + mech) +
+    theme_classic()
+}

### 2025-04-29 Manuscript/manuscript.qmd
rerun simulation without defaults, update intro section

--- 2025-04-22 Manuscript/manuscript.qmd
@@
 library(dplyr)
+
 load("../results/out.RData")
-source("../R/evaluate_estimates.R")
+source("../R/evaluate_performance.R")
 performance <- evaluate_est(out) 
-performance[performance$term == "b0", "term"] <- "Y"
-performance <- mutate(performance,
-  mech = factor(mech, levels = c("MCAR", "MAR"), ordered = TRUE),
-  prop = factor(prop, levels = c(0.25, 0.5, 0.75), ordered = TRUE),
-  term = factor(term, levels = c("Y", "X1", "X2", "X3"), ordered = TRUE))
-results <- performance |> 
-  group_by(method, mech, prop, .it, term) |> 
-  summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), \(x) mean(x, na.rm = TRUE))) 
-```
-
-```{r, echo=FALSE}
-# # plot results per sim rep
-# performance |>
-#   filter(term == "X1", mech == "MAR", method == "MICE", prop == 0.75) |>
-#   ggplot(aes(x = .it, y = bias)) + #, color = as.factor(prop)
-#   geom_line(aes(group = .sim)) +
-#   geom_smooth(method = "loess") +
-#   # labs(
-#   #   title = "Bias regression coefficient",
-#   #   x = "Iteration number",
-#   #   y = "Bias"
-#   # ) +
-#   theme_classic()
+results <- summarize_res(performance)
+source("../figures/plot_results.R")
+```
+
+```{r, echo=FALSE, eval=FALSE}
+results |>
+  filter(method == "MICE") |> 
+  ggplot(aes(x = .it, y = psrf_mean)) +
+  geom_hline(yintercept = 1, alpha = 0.1) + 
+  geom_line() +
+  geom_line(aes(y = psrf_mean_ll), linetype = "dashed", alpha = 0.3) +
+  geom_line(aes(y = psrf_mean_ul), linetype = "dashed", alpha = 0.3) +
+  facet_grid(term ~ mech + prop) +
+  theme_classic()
+
+plot_results(results, "bias", 0, "MCAR")
+plot_results(results, "cov", 0.95, "MCAR")
+plot_results(results, "ciw", NULL, "MCAR")
+plot_results(results, "ac_mean", 0, "MCAR")
+plot_results(results, "ac_sd", 0, "MCAR")
+plot_results(results, "ac_mean", 0, "MCAR")
+plot_results(results, "psrf_mean", 1, "MCAR")
+plot_results(results, "psrf_sd", 1, "MCAR")
 ```
@@
 - ~~which parameter should we track?~~
-- how many iterations are needed before convergence is reached?
+- how many iterations are needed before converge to a stationary distribution is reached?
 - how many iterations are needed before valid inferences are reached?
@@
 
-Iterative imputation has become the de facto standard to accommodate missing data in social scientific research [ref: e.g., rubin after 18 years paper?
-murray 2018?].
-The technique enables researchers to draw valid inferences when faced with incomplete observations, without resorting to ad hoc solutions such as list-wise deletion, which may bias results.
-With iterative imputation, every missing cell in an incomplete dataset is imputed (i.e., filled in) algorithmically.
-And once the dataset is completed, the analysis of scientific interest can be performed.
-This analysis will yield unbiased and confidence-valid estimates if--and only if--both the missing data problem and the scientific problem are appropriately considered.
-The validity of this whole process naturally depends on the convergence of the algorithm that was used to generate the imputed values.
+Multiple imputation has become the de facto standard to accommodate missing data in social scientific research.
+The technique enables researchers to draw valid inferences from incomplete data [@rubin76].
+[TODO: remove the next sentence?] This is in contrast to ad hoc solutions such as list-wise deletion, which may bias scientific estimates.
+Multiple imputation separates the missing data problem from the scientific problem.
+First, the missing values are imputed (i.e., filled in) using an imputation algorithm.
+Once the data are completed, the analysis of scientific interest can be performed on the completed data.
+This process is repeated several times to obtain a distribution of plausible values, which reflects the uncertainty in the data due to missingness.
+Finally, the scientific estimates are pooled across the multiple imputations.
+This pooled estimate is unbiased and confidence-valid if--and only if--both the missing data problem and the scientific problem are appropriately considered.
+
+[TODO: add refs? e.g., rubin after 18 years paper? murray 2018?]
+
+To obtain valid scientific estimates from incomplete data, both the missing data problem and the scientific problem should be appropriately considered.
+The popular imputation algorithm 'MICE' has proven to be a powerful tool to draw valid inferences from incomplete data under many circumstances [@buuren2011; @buur18].
+The algorithm is named after its iterative nature: missing data are imputed on a variable-by-variable basis, using a sequence of univariate imputation models.
+The validity of this iterative process naturally depends on the convergence of the imputation algorithm.
 Yet, there has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms.
@@
 Determining whether an algorithm has converged is not trivial, especially in the context of iterative imputation.
-Since the aim of iterative imputation is to converge to a distribution and not to a single point, the algorithm may produce some fluctuations even after it has converged.
+Since the aim of iterative imputation is to converge to a distribution and not to a single point, the algorithm produces some fluctuations even after it has converged.
 Because of this property, it may be more desirable to focus on *non*-convergence.
@@
 285).] Therefore, a quantitative diagnostic method to assess convergence would be preferred.
-Fortunately, there are non-convergence identifiers for *other* iterative algorithms, but the validity of these identifiers has not been systematically evaluated on imputation algorithms.
-
+Fortunately, there are non-convergence identifiers for *other* iterative algorithms, but the validity of these identifiers has not been evaluated on imputation algorithms.
+
+[TODO: add *approximate* convergence as the most converged state that the algorithm can obtain?]
 <!-- It is challenging to arrive upon a single point at which convergence has been reached. -->
 
-In this study, we explore different methods for identifying non-convergence in iterative imputation algorithms.
+In this study, we explore different methods for identifying non-convergence in the iterative imputation algorithm MICE.
 We evaluate whether these methods are able to cover the extent of the non-convergence, and we also investigate the relation between non-convergence and the validity of the inferences.
 We translate the results of our simulation study into guidelines for practice, which we demonstrate by means of a case study.
-
-[TODO: intrgrate this section] In an empirical study, where $\widehat{R}$ was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold $\widehat{R}>1.2$.
-Yet, scientific estimates were insensitive to continued iteration from $t=5$ onward [@lace07].
-We, therefore, suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates.
-In contrast to this, signs of non-convergence can be under-estimated by $\widehat{R}$, in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
-437].
-In  `mice`, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values.
-In practice, we do not expect this to cause problems for identifying non-convergence with $\widehat{R}$.
-
-[TODO: intrgrate this section] If the estimated quantities of scientific interest $\bar{Q}$ are unbiased and confidence-valid estimates of ${Q}$, we will conclude that the algorithm is sufficiently converged for practical purposes.
-In contrast to this, we defined *approximate* convergence as the most converged state that the algorithm can obtain.
-Continued iteration after reaching approximate convergence does not yield better mixing or stationarity (indicated by non-improving $\widehat{R}$- and $AC$-values).
-Continued iteration after obtaining valid inferences may lead to a more converged state of the algorithm, but not better estimates.
 
@@
 
-In this paper we study different methods for assessing non-convergence in iterative imputation algorithms.
+[TODO: merge with prev section] In this paper we study different methods for assessing non-convergence in iterative imputation algorithms.
 We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications.
@@
 
+<!-- [TODO: integrate this section] In an empirical study with STATA, where $\widehat{R}$ was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold $\widehat{R}>1.2$.
+Yet, scientific estimates were insensitive to continued iteration from $t=5$ onward [@lace07].
+We, therefore, suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates.
+In contrast to this, signs of non-convergence can be under-estimated by $\widehat{R}$, in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
+437].
+In  `mice`, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values.
+In practice, we do not expect this to cause problems for identifying non-convergence with $\widehat{R}$.
+-->
+
+<!-- [TODO: intrgrate this section] If the estimated quantities of scientific interest $\bar{Q}$ are unbiased and confidence-valid estimates of ${Q}$, we will conclude that the algorithm is sufficiently converged for practical purposes.
+In contrast to this, we defined *approximate* convergence as the most converged state that the algorithm can obtain.
+Continued iteration after reaching approximate convergence does not yield better mixing or stationarity (indicated by non-improving $\widehat{R}$- and $AC$-values).
+Continued iteration after obtaining valid inferences may lead to a more converged state of the algorithm, but not better estimates.
+-->
+
 # Background
 
-## Some notation
+## Notation
 
@@
 The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing.
-The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\mathrm{inc}}$ in dataset $y$.
 The collection of observed data values in $y$ is denoted by $y_{\mathrm{obs}}$; the missing part of $y$ is referred to as $y_{\mathrm{mis}}$.
+We define the proportion of incomplete cases, $p_{\mathrm{inc}}$, as the number of units in dataset $y$ with at least one missing value divided by the total number of units $n$.
+
 For each datapoint in $y_{\mathrm{mis}}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations in the imputation algorithm ($t = 1, 2, \dots, T$).
@@
 
-## Iterative imputation
+## The MICE algorithm
 
@@
 An alternative is to apply imputation.
-With imputation, we 'fill in' the missing values in an incomplete dataset.
-Subsequently, the model of scientific interest can be fitted to the completed dataset.
-By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness.
+With imputation, we 'fill in' the missing values in an incomplete dataset several times.
 This technique is known as 'multiple imputation' [MI; @rubin76].
-MI has proven to be a powerful tool to draw valid inferences from incomplete data under many circumstances [@buur18].
-
-Figure \ref{fig:diagram} provides an overview of the steps involved with MI.
+
+Figure \ref{diagram} provides an overview of the steps involved with MI.
 The missing part $y_{\mathrm{mis}}$ of an incomplete dataset is imputed $m$ times.
@@
 
-A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice].
-With MICE, imputed values are drawn from the posterior predictive distribution of the missing values.
-The algorithm is named after its iterative nature: a multivariate distribution is obtained by iterating over a sequence of univariate imputations.
-Iteration, however, also introduces a potential threat to the validity of the imputations: What if the algorithm has not converged?
-Are the imputations then to be trusted?
-And can we rely on the inference obtained on the completed data?
-
-These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18].
-There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@zhu15; @takahashi2017].
-Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murray2018].
-Therefore, algorithmic convergence should be monitored carefully---although this is not straightforward.
-Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods.
-In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another.
-The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13].
-Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09].
-Several non-convergence diagnostics exist, but it is not known whether these are appropriate within the imputation framework.
+A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@buuren2011].
+With MICE, imputed values are drawn from the posterior predictive distribution of the missing values on a variable-by-variable basis.
+The algorithm is initialized by filling in $m$ sets of starting values, randomly drawn from the observed data.
+The algorithm then iterates over the $k$ variables, generating $m$ sets of imputed values for each variable $j$.
+After one full run through the data ($t=1$), the imputed values will be 'updated' in the second iteration ($t=2$).
+This process should be repeated until convergence is reached.
 
@@
 
-## Algorithmic non-convergence
-
-There are two requirements for convergence of iterative algorithms: mixing and stationarity [@gelm13].
-In iterative imputation algorithms, mixing implies that imputation chains intermingle nicely, and stationarity is characterized by the absence of trending across iterations.
+## Convergence of MICE
+
+Iterative imputation algorithms such as MICE often are special cases of Markov chain Monte Carlo (MCMC) methods.
+In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another.
+The values generated by the algorithm will vary even after convergence [@gelm13].
+Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09].
+
+Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection [@ragh07; @buur18].
+After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots.
+In a traceplot, a scalar summary of the state-space of the algorithm $\theta$ is plotted against the iteration number.
+Non-convergence is diagnosed if the chains show definite trends, or if the imputation chains are not freely intermingled with one another [@buur18, $\S$ 6.5.2].
+These guidelines are in line with the two requirements for convergence in MCMC algorithms: stationarity and mixing [@gelm13].
+Stationarity is characterized by the absence of trending across iterations.
+Mixing implies that the chains intermingle nicely.
 If one of the two requirements is not met, we speak of non-convergence.
-Without mixing, chains may be 'stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values.
+Without mixing, imputation chains may be 'stuck' at a local optimum instead of sampling imputed values from the entire predictive posterior distribution of the missing values.
 The distribution of imputed values then differs across imputations.
@@
 
-To illustrate what non-mixing and non-stationarity look like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2].
-Figure \ref{fig:non-conv} displays two scenarios from the example.
-The panel on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm.
-The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
-Each line portrays one imputation chain (i.e., the values of a scalar summary $\theta$ across iterations).
-The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in the $m$ sets of imputations $\dot y_{\mathrm{imp}, \ell}$.
-
-In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending.
-In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.
-Importantly, the chain means at the last iteration (the imputed values per imputation $\ell$) are very different between the two scenarios.
-The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
-It is obvious that non-convergence in this example impacts the distribution of the imputed values per imputation $\dot y_{\mathrm{imp}, \ell}$.
-This effect may translate into the distribution of the $m$ sets of completed data {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$, and finally the pooled estimate $\bar{Q}$.
-$\bar{Q}$ is then a  biased, invalid estimate of $Q$.
-Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.
-
-```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence (reproduced from van Buuren; 2018).
-Depicted are chain means of variable $j$ in each imputation ($m=5$; $T=10$).
-Note that any difference between the two scenarios (see e.g., the magnitude of the chain means on the y-axes) stems from the imputation algorithms---the two scenarios have equal incomplete data $y$."}
-
-# par(mar = c(2, 2, 0.1, 0.1))
-# con_plot
-# non_plot
-
-```
-
-## Identifying non-convergence
-
-Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection.
-After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots.
-In a traceplot, a scalar summary of the state-space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}.
-The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances.
-Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18, $\S$ 6.5.2].
+<!-- The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances. -->
+
+<!-- Several non-convergence diagnostics exist, but it is not known whether these are appropriate within the imputation framework.  -->
+
+<!-- *approximate* convergence as the most converged state that the algorithm can obtain. Continued iteration after reaching approximate convergence does not yield better mixing or stationarity -->
+
+[TODO: move/integrate this section?] There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@zhu15; @takahashi2017].
+Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murray2018].
+Therefore, algorithmic convergence should be monitored carefully---although this is not straightforward.
+
 <!-- Recommendation @buur18: "the different streams should be freely intermingled with one another, without showing any definite trends". -->
@@
 With a recently proposed adaptation, $\widehat{R}$ may also serve to diagnose non-stationarity [@vehtari2021].
-We will evaluate the appropriateness of both $\widehat{R}$ and $AC$.
-Other methods are outside the scope of this study because they e.g.
-assume that values within chains represent independent samples, whereas the MICE algorithm only uses the final iteration to produce imputations.
+We will evaluate the appropriateness of $\widehat{R}$ and $AC$.
+Other methods are outside the scope of this study, e.g.
+because they assume that values within chains represent independent samples, whereas the MICE algorithm only uses the final iteration to produce imputations.
 
@@
 The number of simulation repetitions $n_{\mathrm{sim}} = 2000$.
+
+[TODO: add guaranteed convergence because compatible imputation models, equivalent to JM and true Gibbs sampler, @vanbuuren2007]
+
+[TODO: add different correlations? add different imputation models (incompatible/PMM)?]
 
@@
 ```{r patterns, echo=FALSE}
-source("../R/generate_datasets.R")
-pat <- create_patterns(3)
-is.na(pat) <- pat == 1
-ggmice::plot_pattern(pat, caption = FALSE) +
-  theme(axis.text.x=element_blank(), 
-      axis.ticks.x=element_blank(), 
-      axis.text.y=element_blank(), 
-      axis.ticks.y=element_blank(),
-      axis.text = element_blank()) +
-  theme_void()
+source("../figures/plot_patterns.R")
+plot_patterns()
 ```
@@
 ::: {.callout-note}
-# Missing Data Mechanism
+# Missing data mechanism
 
@@
 
-## Diagnostic Methods
+[TODO: add tables, with MC SEs]
+
+## Performance measures
+
+```{r bias}
+plot_results(results, "bias", 0, "MCAR")
+```
+
+- more missingness = more extreme bias, but also steeper decrease over iterations
+
+```{r cov}
+plot_results(results, "cov", 0.95, "MCAR")
+```
+
+- no clear trend
+
+```{r ciw}
+plot_results(results, "ciw", NULL, "MCAR")
+```
+
+- more missingness = wider CIs, with steeper decrease in CIW, but not stababiling at same value between missingness conditions.
+
+In Figure \@ref(fig:perf) we show the performance measures: bias in the regression estimate (panel A), the empirical coverage rate of the regression estimate (panel C) and the average confidence interval width of this estimate (panel D).
+
+We see that within a few iterations the bias in the regression estimate approaches zero (see \@ref(fig:perf)A).
+When $n_{it} \geq 6$, even the worst-performing conditions (e.g., with a proportion of incomplete cases of 75%) produce stable, non-improving estimates [regression coefficient is underestimated because there is less info to estimate the relation??].
+
+Nominal coverage is quickly reached (see \@ref(fig:perf)C).
+After just three iterations, the coverage rates are non-improving in every missingness condition [but MNAR with 5% incomplete cases does not reach nominal coverage --\> due to bias in the estimate in combination with very narrow CI (see CIW!)].
+
+The average confidence interval width decreases quickly with every added iteration until a stable plateau is reached (see \@ref(fig:perf)D).
+Depending on the proportion of incomplete cases this takes up-to $n_{it} \geq 9$.
+
+## Diagnostic methods
 
@@
 ```{r ac_mean}
-results |>
-  filter(method == "MICE") |> 
-  ggplot(aes(x = .it, y = ac_mean, color = mech, linetype = mech)) +
-  geom_hline(yintercept = 0, alpha = 0.1) + 
-  geom_line() +
-  facet_grid(term ~ prop) +
-  theme_classic()
+plot_results(results, "ac_mean", 0, "MCAR")
 ```
@@
 ```{r ac_sd}
-results |>
-  filter(method == "MICE") |> 
-  ggplot(aes(x = .it, y = ac_sd, color = mech, linetype = mech)) +
-  geom_hline(yintercept = 0, alpha = 0.1) + 
-  geom_line() +
-  facet_grid(term ~ prop) +
-  theme_classic()
+plot_results(results, "ac_sd", 0, "MCAR")
 ```
@@
 ```{r psrf_mean}
-results |>
-  filter(method == "MICE") |> 
-  ggplot(aes(x = .it, y = psrf_mean, color = mech, linetype = mech)) +
-  geom_hline(yintercept = 1, alpha = 0.1) + 
-  geom_line() +
-  facet_grid(term ~ prop) +
-  theme_classic()
+plot_results(results, "psrf_mean", 1, "MCAR")
 ```
@@
 ```{r psrf_sd}
-results |>
-  filter(method == "MICE") |> 
-  ggplot(aes(x = .it, y = psrf_sd, color = mech, linetype = mech)) +
-  geom_hline(yintercept = 1, alpha = 0.1) + 
-  geom_line() +
-  facet_grid(term ~ prop) +
-  theme_classic()
+plot_results(results, "psrf_sd", 1, "MCAR")
 ```
@@
 
-```{r, cache=TRUE, fig.height=9}
+```{r, cache=TRUE, eval=FALSE, fig.height=9}
 performance |>
@@
 
-## Performance Measures
-
-```{r bias}
-results |>
-  filter(method == "MICE") |> 
-  ggplot(aes(x = .it, y = bias, color = mech, linetype = mech)) +
-  geom_hline(yintercept = 0, alpha = 0.1) + 
-  geom_line() +
-  facet_grid(term ~ prop) +
-  theme_classic()
-```
-
-- more missingness = more extreme bias, but also steeper decrease over iterations
-
-```{r cov}
-results |>
-  filter(method == "MICE") |> 
-  ggplot(aes(x = .it, y = cov, color = mech, linetype = mech)) +
-  geom_hline(yintercept = .95, alpha = 0.1) + 
-  geom_line() +
-  facet_grid(term ~ prop) +
-  theme_classic()
-```
-
-- no clear trend
-
-```{r ciw}
-results |>
-  filter(method == "MICE") |> 
-  ggplot(aes(x = .it, y = ciw, color = mech, linetype = mech)) +
-  geom_hline(yintercept = .95, alpha = 0.1) + 
-  geom_line() +
-  facet_grid(term ~ prop) +
-  theme_classic()
-```
-
-- more missingness = wider CIs, with steeper decrease in CIW, but not stababiling at same value between missingness conditions.
-
-In Figure \@ref(fig:perf) we show the performance measures: bias in the regression estimate (panel A), the empirical coverage rate of the regression estimate (panel C) and the average confidence interval width of this estimate (panel D).
-
-We see that within a few iterations the bias in the regression estimate approaches zero (see \@ref(fig:perf)A).
-When $n_{it} \geq 6$, even the worst-performing conditions (e.g., with a proportion of incomplete cases of 75%) produce stable, non-improving estimates [regression coefficient is underestimated because there is less info to estimate the relation??].
-
-Nominal coverage is quickly reached (see \@ref(fig:perf)C).
-After just three iterations, the coverage rates are non-improving in every missingness condition [but MNAR with 5% incomplete cases does not reach nominal coverage --\> due to bias in the estimate in combination with very narrow CI (see CIW!)].
-
-The average confidence interval width decreases quickly with every added iteration until a stable plateau is reached (see \@ref(fig:perf)D).
-Depending on the proportion of incomplete cases this takes up-to $n_{it} \geq 9$.
-
 <!-- Our results show the performance of the MICE algorithm under different conditions of missingness and early stopping.
@@
 
+To illustrate what non-mixing and non-stationarity look like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2].
+Figure \ref{fig:non-conv} displays two scenarios from the example.
+The panel on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm.
+The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
+Each line portrays one imputation chain (i.e., the values of a scalar summary $\theta$ across iterations).
+The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in the $m$ sets of imputations $\dot y_{\mathrm{imp}, \ell}$.
+
+In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending.
+In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.
+Importantly, the chain means at the last iteration (the imputed values per imputation $\ell$) are very different between the two scenarios.
+The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
+It is obvious that non-convergence in this example impacts the distribution of the imputed values per imputation $\dot y_{\mathrm{imp}, \ell}$.
+This effect may translate into the distribution of the $m$ sets of completed data {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$, and finally the pooled estimate $\bar{Q}$.
+$\bar{Q}$ is then a  biased, invalid estimate of $Q$.
+Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.
+
+```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence (reproduced from van Buuren; 2018).
+Depicted are chain means of variable $j$ in each imputation ($m=5$; $T=10$).
+Note that any difference between the two scenarios (see e.g., the magnitude of the chain means on the y-axes) stems from the imputation algorithms---the two scenarios have equal incomplete data $y$."}
+
+# par(mar = c(2, 2, 0.1, 0.1))
+# con_plot
+# non_plot
+
+```
+
 Before we evaluate the performance of the non-convergence diagnostics $\widehat{R}$ and $AC$ quantitatively through simulation, we first assess their appropriateness qualitatively.

### 2025-04-29 R/evaluate_estimates.R

@@ -1,14 +1,0 @@
-# function(s) to evaluate the estimates
-### [YOUR FUNCTION HERE] ###
-
-evaluate_est <- function(results) {
-  performance <- purrr::map_dfr(results, ~{
-    dplyr::mutate(.x,
-      bias = truth - estimate,
-      cov = conf.low <= truth & conf.high >= truth,
-      ciw = conf.high - conf.low,
-      .keep = "unused"
-        )
-  })
-  return(performance)
-}

### 2025-04-29 R/evaluate_performance.R

@@ -1,0 +1,36 @@
+# function(s) to evaluate the estimates
+evaluate_est <- function(results) {
+  performance <- purrr::map_dfr(results, ~{
+    dplyr::mutate(.x,
+      bias = truth - estimate,
+      cov = conf.low <= truth & conf.high >= truth,
+      ciw = conf.high - conf.low,
+      .keep = "unused"
+        )
+  })
+  return(performance)
+}
+
+
+summarize_res <- function(performance) {
+  performance <- mutate(performance,
+                        mech = factor(mech, levels = c("MCAR", "MAR", "MNAR"), ordered = TRUE),
+                        prop = factor(prop, ordered = TRUE),
+                        term = factor(term, levels = c("Y", "X1", "X2", "X3"), ordered = TRUE))
+  results <- performance |> 
+    group_by(method, mech, prop, .it, term) |> 
+    summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), \(x) mean(x, na.rm = TRUE)))
+  results_ll <- performance |> 
+    group_by(method, mech, prop, .it, term) |> 
+    summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), \(x) quantile(x, probs = c(0.025), na.rm = TRUE))) 
+  results <- left_join(results, results_ll, by = c("method", "mech", "prop", ".it", "term"), suffix = c("", "_ll"))
+  results_ul <- performance |> 
+    group_by(method, mech, prop, .it, term) |> 
+    summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), \(x) quantile(x, probs = c(0.975), na.rm = TRUE))) 
+  results <- left_join(results, results_ul, by = c("method", "mech", "prop", ".it", "term"), suffix = c("", "_ul"))
+  cov_SE <- sqrt((results$cov * (1 - results$cov)) / max(performance$.sim))
+  results$cov_ll <- results$cov - qnorm(0.975) * cov_SE
+  results$cov_ul <- results$cov + qnorm(0.975) * cov_SE
+  return(results)
+}
+# performance[performance$term == "b0", "term"] <- "Y"

### 2025-04-29 R/generate_datasets.R

@@ -25,17 +25,13 @@ create_patterns <- function(n_col) {
   mis_pat <- expand.grid(pat_list) 
   names(mis_pat) <- c("Y", paste0("X", 1:n_col))
   # omit patterns with all/none missing
-  mis_pat <- mis_pat[rowSums(mis_pat) > 1 & rowSums(mis_pat) < n_col + 1, ]
+  mis_pat <- mis_pat[rowSums(mis_pat) > 0 & rowSums(mis_pat) < n_col, ]
   return(mis_pat)
 }
 
 # ampute the complete data
 induce_missingness <- function(
-    dat,
-    mis_pat = NULL, 
-    mis_mech = c("MCAR", "MAR"),
-    mis_prop = c(0.1, 0.25, 0.5)
-) {
+    dat, mis_pat, mis_mech, mis_prop) {
   # create a list of amputed data objects 
   # for each of the mechanisms and proportions
   amps <- purrr::map(mis_mech, function(mm) {
@@ -52,18 +48,24 @@ induce_missingness <- function(
 }
 
 # combine into one function
-create_data <- function(
-    sample_size = 200, 
-    correlations = 0.3,
-    effects = c(0.5, 0.5, 0.5),
-    patterns = NULL, 
-    mechanisms = c("MCAR", "MAR"),
-    proportions = c(0.1, 0.25, 0.5)
-) {
+create_data <- function(sample_size,
+                        correlations,
+                        effects,
+                        patterns,
+                        mechanisms,
+                        proportions) {
   # create a single complete dataset
-  dat <- generate_complete(n_obs = sample_size, n_col = length(effects), corr = correlations, betas = effects)
+  dat <- generate_complete(
+    n_obs = sample_size,
+    n_col = length(effects),
+    corr = correlations,
+    betas = effects
+  )
   # ampute the data with different missingness
-  amps <- induce_missingness(dat, mis_pat, mis_mech = mechanisms, mis_prop = proportions)
+  amps <- induce_missingness(dat,
+                             mis_pat = patterns,
+                             mis_mech = mechanisms,
+                             mis_prop = proportions)
   # output
   return(amps)
 }

### 2025-04-29 figures/patterns.R

@@ -1,5 +1,0 @@
-source("../R/generate_datasets.R")
-pat <- create_patterns(3)
-d <- dat[1:10,]
-is.na(d) <- pat == 1
-ggmice::plot_pattern(d)

### 2025-04-29 figures/plot_patterns.R

@@ -1,0 +1,59 @@
+source("../R/generate_datasets.R")
+# pat <- create_patterns(3)
+# d <- dat[1:10,]
+# is.na(d) <- pat == 0
+# ggmice::plot_pattern(d)
+plot_patterns <- function(n_col = 4) {
+  pat <- create_patterns(n_col - 1)
+  pat <- abs(pat - 1)
+  pat <- rbind(pat, 0)
+  pat[pat == 1] <- "missing"
+  pat[pat == 0] <- "observed"
+  rws <- nrow(pat)
+  cls <- ncol(pat)
+  vrb <- colnames(pat)
+  # frq <- row.names(pat)[-rws]
+  # na_row <- pat[-rws, cls]
+  # na_col <- pat[rws, -cls]
+  pat_clean <- cbind(.opacity = 1, pat)
+  long <-
+    as.data.frame(cbind(.y = 1:(rws), pat_clean)) %>%
+    tidyr::pivot_longer(
+      cols = tidyselect::all_of(vrb),
+      names_to = "x",
+      values_to = ".where"
+    ) %>%
+    dplyr::mutate(
+      .x = as.numeric(factor(
+        .data$x, levels = vrb, ordered = TRUE
+      )),
+      .where = factor(
+        .data$.where,
+        levels = c("observed", "missing"),
+        ordered = TRUE
+      ),
+      .opacity = as.numeric(.data$.opacity)
+    )
+  gg <-
+    ggplot2::ggplot(
+      long,
+      ggplot2::aes(
+        .data$.x,
+        .data$.y,
+        fill = .data$.where,
+        alpha = 0.1 + .data$.opacity / 2
+      )
+    ) +
+    ggplot2::geom_tile(color = "black") +
+    ggplot2::scale_fill_manual(values = c(
+      "observed" = "#006CC2B3",
+      "missing" = "#B61A51B3"
+    )) +
+    ggplot2::scale_alpha_continuous(limits = c(0, 1), guide = "none") +
+    ggplot2::scale_x_continuous(breaks = 1:cls, labels = vrb) +
+    ggplot2::scale_y_continuous(breaks = NULL, labels = NULL) +
+    ggmice:::theme_minimice() +
+    labs(fill = "", x = "Variables", y = "Patterns")
+  gg <- gg + ggplot2::coord_cartesian(expand = FALSE)
+  return(gg)
+}

### 2025-04-29 figures/plot_results.R

@@ -6,14 +6,25 @@ plot_results <- function(summarized_results, metric, hline, mis = "MAR") {
   }
   metric_ll <- paste0(metric, "_ll")
   metric_ul <- paste0(metric, "_ul")
+  y_lab <- switch(
+    metric,
+    bias = "Bias",
+    cov = "Coverage rate",
+    ciw = "CI width",
+    ac_mean = "AC in chain means",
+    ac_sd = "AC in chain SDs",
+    psrf_mean = "PSRF in chain means",
+    psrf_sd = "PSRF in chain SDs"
+  )
   summarized_results |>
     filter(method == "MICE") |> 
     filter(mis %in% mech) |>
     ggplot(aes(x = .it, y = .data[[metric]])) +
-    geom_hline(yintercept = hline, alpha = 0.1) + 
-    geom_line() +
-    geom_line(aes(y = .data[[metric_ll]]), linetype = "dashed", alpha = 0.5) +
-    geom_line(aes(y = .data[[metric_ul]]), linetype = "dashed", alpha = 0.5) +
+    geom_hline(yintercept = hline, alpha = 0.1, linewidth = 1) + 
+    geom_line(na.rm = TRUE) +
+    geom_line(aes(y = .data[[metric_ll]]), linetype = "dashed", alpha = 0.3, na.rm = TRUE) +
+    geom_line(aes(y = .data[[metric_ul]]), linetype = "dashed", alpha = 0.3, na.rm = TRUE) +
     facet_grid(term ~ prop + mech) +
+    labs(x = "Iteration number", y = y_lab) +
     theme_classic()
 }

### 2025-04-29 simulation.R

@@ -16,15 +16,15 @@ set.seed(11)
 
 # parameters
 n_sim <- 200
-n_obs <- 200
+n_obs <- 500
 n_col <- 3
 corr <- 0.5
 beta <- 1
 betas <- rep(beta, n_col)
 mis_pat <- create_patterns(n_col)
-mis_mech = c("MCAR", "MAR", "MNAR")
-mis_prop = c(0.25, 0.5, 0.75)
-n_it <- 100
+mis_mech = c("MCAR") #, "MAR", "MNAR")
+mis_prop = c(0.05, 0.25, 0.5, 0.75, 0.95)
+n_it <- 50
 
 # #################################
 # ### TEST LOWER LEVEL FUCTIONS ###
@@ -34,7 +34,7 @@ n_it <- 100
 # dat <- generate_complete(n_obs, n_col, corr, betas)
 # 
 # # ampute data
-# amps <- induce_missingness(dat, mis_pat, mis_mech = "MAR", mis_prop = 0.5)
+# amps <- induce_missingness(dat, mis_pat, mis_mech = "MCAR", mis_prop = 0.5)
 # 
 # ##################################
 # ### TEST HIGHER LEVEL FUCTIONS ###
@@ -47,10 +47,11 @@ n_it <- 100
 ### COMBINE INTO ONE FUCTION ###
 ################################
 
-simulate_once <- function(n_obs, betas, mis_pat, mis_mech, mis_prop) {
+simulate_once <- function(n_obs, corr, betas, mis_pat, mis_mech, mis_prop, n_it) {
   # generate incomplete data
   amps <- create_data(
     sample_size = n_obs,
+    correlations = corr,
     effects = betas,
     patterns = mis_pat,
     mechanisms = mis_mech,
@@ -66,7 +67,7 @@ simulate_once <- function(n_obs, betas, mis_pat, mis_mech, mis_prop) {
 # ### TEST SIMULATION FUNCTION ###
 # ################################
 # 
-# ests <- simulate_once(n_obs, betas, mis_pat, mis_mech, mis_prop)
+# ests <- simulate_once(n_obs, corr, betas, mis_pat, mis_mech, mis_prop, n_it)
 
 ######################
 ### RUN SIMULATION ###
@@ -90,7 +91,8 @@ library(pbapply)
 cl <- parallel::makeCluster(4)
 parallel::clusterExport(
   cl,
-  c("generate_complete",
+  c(
+    "generate_complete",
     "induce_missingness",
     "create_data",
[…]

### 2025-04-30 Manuscript/manuscript.qmd
edit notation, add MICE algorithm

@@
 
-Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample.
-The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing.
-The collection of observed data values in $y$ is denoted by $y_{\mathrm{obs}}$; the missing part of $y$ is referred to as $y_{\mathrm{mis}}$.
-We define the proportion of incomplete cases, $p_{\mathrm{inc}}$, as the number of units in dataset $y$ with at least one missing value divided by the total number of units $n$.
-
-For each datapoint in $y_{\mathrm{mis}}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations in the imputation algorithm ($t = 1, 2, \dots, T$).
+Let $\mathbf{Y}$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample.
+The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $Y_j$ ($j = 1, 2, \dots, p$) may be either observed or missing.
+The collection of observed data values in $\mathbf{Y}$ is denoted by $\mathbf{Y}_{\mathrm{obs}}$; the missing part of $\mathbf{Y}$ is referred to as $\mathbf{Y}_{\mathrm{mis}}$.
+We define the proportion of incomplete cases, $\pi_{\mathrm{inc}}$, as the number of units with at least one missing value divided by the total number of units $n$ in dataset $\mathbf{Y}$.
+
+For each datapoint in $\mathbf{Y}_{\mathrm{mis}}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations in the imputation algorithm ($t = 1, 2, \dots, T$).
 The state-space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values).
@@
 Figure \ref{diagram} provides an overview of the steps involved with MI.
-The missing part $y_{\mathrm{mis}}$ of an incomplete dataset is imputed $m$ times.
-This creates $m$ sets of imputed data $\dot y_{\mathrm{imp}, \ell}$, where $\ell = 1, 2, ..., m$.
-The imputed data is then combined with the observed data $y_{\mathrm{obs}}$ to create $m$ completed datasets.
+The missing part $\mathbf{Y}_{\mathrm{mis}}$ of an incomplete dataset is imputed $m$ times.
+This creates $m$ sets of imputed data $\dot \mathbf{Y}_{\mathrm{imp}, \ell}$, where $\ell = 1, 2, ..., m$.
+The imputed data is then combined with the observed data $\mathbf{Y}_{\mathrm{obs}}$ to create $m$ completed datasets.
 On each of these datasets the analysis of scientific interest is performed to estimate $Q$: the quantity of scientific interest (e.g., a regression coefficient).
@@
 The algorithm is initialized by filling in $m$ sets of starting values, randomly drawn from the observed data.
-The algorithm then iterates over the $k$ variables, generating $m$ sets of imputed values for each variable $j$.
+The algorithm then iterates over the $p$ variables, generating $m$ sets of imputed values for each variable $j$.
 After one full run through the data ($t=1$), the imputed values will be 'updated' in the second iteration ($t=2$).
@@
 -->
+
+**MICE algorithm**
+
+*Initialization phase ($t = 0$)*
+
+1. Specify an imputation model for each variable $Y_j$, where $j = 1,...,p$.
+2. Repeat over variables $j = 1,...,p$.
+    a. Fill in starting imputations ($\dot{Y_j^0}$) by random draws from the observed part of each variable ($Y_{j}^{\mathrm{obs}}$).
+3. End repeat $j$.
+
+*Iteration phase ($t = 1,...,T$)*
+
+4. Repeat over iterations $t = 1,...,T$.
+    a. Repeat over variables $j = 1,...,p$.
+        I.
+Define $\dot{\mathbf{Y}}_{-j}^t$ as the currently complete data *except* variable $Y_j$ at iteration $t$, where $\dot{\mathbf{Y}}_{-j}^t = (\dot Y_1^t,\dots,\dot Y_{j-1}^t,\dot Y_{j+1}^{t-1},\dots,\dot Y_p^{t-1})$.
+        II. Draw imputation model parameters $\dot\phi_j^t$ from the posterior predictive distribution of the observed data, where $\dot\phi_j^t \sim P(\phi_j^t|Y_j^\mathrm{obs}, \dot Y_{-j}^t, R)$.
+        III. Draw imputations $\dot Y_j^t$, where $\dot Y_j^t \sim P(Y_j^\mathrm{mis}|Y_j^\mathrm{obs},  \dot Y_{-j}^t, R, \dot\phi_j^t)$.
+    b. End repeat $j$.
+5. End repeat $t$.
+
+<!-- \begin{enumerate} -->
+<!--    \item First level item -->
+<!--    \item First level item -->
+<!--    \begin{enumerate} -->
+<!--      \item Second level item -->
+<!--      \item Second level item -->
+<!--      \begin{enumerate} -->
+<!--        \item Third level item -->
+<!--        \item Third level item -->
+<!--      \end{enumerate} -->
+<!--    \end{enumerate} -->
+<!--  \end{enumerate} -->
 
@@
 \geq \lambda_{j, \ell}$ be the eigenvalues of $S_\ell$ in each imputation.
-The first eigenvalue, $\lambda_{1,\ell}$, then summarizes the largest possible amount of covariance in each completed dataset {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}.
+The first eigenvalue, $\lambda_{1,\ell}$, then summarizes the largest possible amount of covariance in each completed dataset {$\mathbf{Y}_{\mathrm{obs}}, \dot \mathbf{Y}_{\mathrm{imp}, \ell}$}.
 By definition, $\lambda_{1}$-values are equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on the completed data.
@@
 The proportion of incomplete cases $\pi_{\mathrm{inc}}$ is set to 25%, 50%, and 75%.
-We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\mathrm{obs}}, y_{\mathrm{mis}}$} for every missingness condition.
+We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$\mathbf{Y}_{\mathrm{obs}}, \mathbf{Y}_{\mathrm{mis}}$} for every missingness condition.
 
@@
 <!-- All imputation procedures are performed using `mice` [function `mice()`; @mice], with Bayesian linear regression imputation, and five imputation chains ($m=5$).
-This results in $m=5$ sets of imputations for each of the conditions, $\dot y_{\mathrm{imp}, \ell}$.
--->
-
-<!-- Subsequently, we obtain the completed data per imputation, {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice].
+This results in $m=5$ sets of imputations for each of the conditions, $\dot \mathbf{Y}_{\mathrm{imp}, \ell}$.
+-->
+
+<!-- Subsequently, we obtain the completed data per imputation, {$\mathbf{Y}_{\mathrm{obs}}, \dot \mathbf{Y}_{\mathrm{imp}, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice].
 Estimates of scientific quantities $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R].
@@
 <!-- We consider eight sets of diagnostic methods to identify non-convergence, by applying $\widehat{R}$ and $AC$ on four different $\theta$s.
-Univariate $\theta$s are obtained from $\dot y_{\mathrm{imp}, \ell}$.
-The model-independent multivariate $\theta$, $\lambda_{1}$, is obtained from {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$} as the variance in the first PCA component [function `stats::princomp`; R].
-The model-dependent multivariate $\theta$ is one of the quantities of scientific interest: the estimated regression coefficients in {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}.
+Univariate $\theta$s are obtained from $\dot \mathbf{Y}_{\mathrm{imp}, \ell}$.
+The model-independent multivariate $\theta$, $\lambda_{1}$, is obtained from {$\mathbf{Y}_{\mathrm{obs}}, \dot \mathbf{Y}_{\mathrm{imp}, \ell}$} as the variance in the first PCA component [function `stats::princomp`; R].
+The model-dependent multivariate $\theta$ is one of the quantities of scientific interest: the estimated regression coefficients in {$\mathbf{Y}_{\mathrm{obs}}, \dot \mathbf{Y}_{\mathrm{imp}, \ell}$}.
 We apply the two non-convergence diagnostics on the $\theta$s by implementing the adapted version of $\widehat{R}$ and by manually programming the $AC$ function.
@@
 This implies that approximately unbiased estimates may be obtained with as little as one iteration ($T \geq 1$).
-The bias in $\bar{Q}$ depends solely on the proportion of incomplete cases, with $p_{\mathrm{inc}}\geq.75$ leading to more extreme biases than other conditions.
-Completely unbiased estimates are only obtained in conditions where $p_{\mathrm{inc}}\leq.50$.
+The bias in $\bar{Q}$ depends solely on the proportion of incomplete cases, with $\pi_{\mathrm{inc}}\geq.75$ leading to more extreme biases than other conditions.
+Completely unbiased estimates are only obtained in conditions where $\pi_{\mathrm{inc}}\leq.50$.
 This is curious since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.
@@
 Therefore, no iteration at all is needed to obtain approximately unbiased estimates for $Q=\sigma$.
-Completely unbiased estimates are only obtained in conditions where $p_{\mathrm{inc}}=.05$.
-The other missingness conditions are impacted by non-convergence in the order of increasing $p_{\mathrm{inc}}$.
+Completely unbiased estimates are only obtained in conditions where $\pi_{\mathrm{inc}}=.05$.
+The other missingness conditions are impacted by non-convergence in the order of increasing $\pi_{\mathrm{inc}}$.
 Similar to $Q=\mu$, the magnitude of the bias in $Q=\sigma$ is negligible.
@@
 Therefore, we see that the number of iterations that is necessary to reach stable, non-improving estimates also differs across missingness conditions.
-In conditions where $p_{\mathrm{inc}}=.05$, estimates are unbiased after one iteration.
-The highest number of iterations necessary to reach approximate unbiasedness is seven ($p_{\mathrm{inc}}=.95$; $T\geq7$).
-Completely unbiased estimates are only obtained in conditions where $p_{\mathrm{inc}}\leq.25$.
+In conditions where $\pi_{\mathrm{inc}}=.05$, estimates are unbiased after one iteration.
+The highest number of iterations necessary to reach approximate unbiasedness is seven ($\pi_{\mathrm{inc}}=.95$; $T\geq7$).
+Completely unbiased estimates are only obtained in conditions where $\pi_{\mathrm{inc}}\leq.25$.
 This implies that even for biased estimates, $T\geq7$ would suffice to reach a stable solution.
@@
 As Figure \ref{fig:Qs}D shows, the bias in $\bar{Q}$ is affected by both the proportion of incomplete cases and the number of iterations.
-Similar to $Q=r^2$, we observe approximately unbiased estimates after at least one, and at most seven iterations, depending on $p_{\mathrm{inc}}$.
-Completely unbiased estimates are only obtained in conditions where $T\geq3$ and $p_{\mathrm{inc}}\leq .50$.
-Despite persistent bias in conditions where $p_{\mathrm{inc}}\geq.75$, the coverage rates are non-improving after just three iterations, irrespective of $p_{\mathrm{inc}}$.
+Similar to $Q=r^2$, we observe approximately unbiased estimates after at least one, and at most seven iterations, depending on $\pi_{\mathrm{inc}}$.
+Completely unbiased estimates are only obtained in conditions where $T\geq3$ and $\pi_{\mathrm{inc}}\leq .50$.
+Despite persistent bias in conditions where $\pi_{\mathrm{inc}}\geq.75$, the coverage rates are non-improving after just three iterations, irrespective of $\pi_{\mathrm{inc}}$.
 We conclude that any condition where $T\geq3$ yields approximately nominal coverage rates, and thus confidence-valid estimates.
@@
 The CIWs depicted in Figure \ref{fig:Qs}F are conform the theoretical foundation of MI: CIs are wider in conditions with higher proportions of incomplete cases (i.e., there is less information in the data, and thus more uncertainty due to missingness).
-Since conditions with higher $p_{\mathrm{inc}}$ result in both wider CIs and more severe bias, the true value of $Q$ may be included in the CI, despite of the bias in $\bar{Q}$.
+Since conditions with higher $\pi_{\mathrm{inc}}$ result in both wider CIs and more severe bias, the true value of $Q$ may be included in the CI, despite of the bias in $\bar{Q}$.
 -->
@@
 Multivariate estimates, by contrast, are affected by both the number of iterations and the proportion of missing cases.
-Completely unbiased estimates are only obtained under low to moderate missingness ($p_{\mathrm{inc}}\leq.50$), after at most three iterations.
-We observe approximately unbiased estimates after at most seven iterations (for any $p_{\mathrm{inc}}$ considered).
+Completely unbiased estimates are only obtained under low to moderate missingness ($\pi_{\mathrm{inc}}\leq.50$), after at most three iterations.
+We observe approximately unbiased estimates after at most seven iterations (for any $\pi_{\mathrm{inc}}$ considered).
 This implies that the algorithm produces stable, non-improving estimates when $T\geq7$.
@@
 As we concluded in the last section, substantially biased estimates were only obtained in conditions where $T\leq6$, and we observed non-nominal coverage only when $T\leq3$.
-However, there was some persistent bias in conditions where $p_{\mathrm{inc}}\geq.75$, irrespective of the number of iterations.
+However, there was some persistent bias in conditions where $\pi_{\mathrm{inc}}\geq.75$, irrespective of the number of iterations.
 If the diagnostic methods are appropriate for MI, we will detect non-convergence in these conditions.
@@
 <!-- The $\widehat{R}$-values in Figure \ref{fig:rhats-acs}E show a similar trend across iterations as we observed for chain means and chain variances.
-Only conditions with a very high  proportion of missing cases ($p_{\mathrm{inc}}\geq.75$) diverge from the earlier observations.
+Only conditions with a very high  proportion of missing cases ($\pi_{\mathrm{inc}}\geq.75$) diverge from the earlier observations.
 The $\widehat{R}$-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations.
@@
 
-<!-- The $AC$-values also follow earlier observations, with the same exception as for $\widehat{R}$ (when $p_{\mathrm{inc}}\geq.75$).
+<!-- The $AC$-values also follow earlier observations, with the same exception as for $\widehat{R}$ (when $\pi_{\mathrm{inc}}\geq.75$).
 To reach maximum stationarity, these conditions require at most seven iterations instead of five.
@@
 For this $\theta$, the results differ only in the most extreme missingness condition.
-When $p_{\mathrm{inc}}=.95$, we diagnose non-convergence at $T\leq9$ according to $\widehat{R}>1.2$, $T\leq19$ for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).
+When $\pi_{\mathrm{inc}}=.95$, we diagnose non-convergence at $T\leq9$ according to $\widehat{R}>1.2$, $T\leq19$ for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).
 The 'dip' in $\widehat{R}$-values at $T=3$ is even more pronounced than with other $\theta$s.
@@
 Under low to moderate missingness, it does not seem to matter on which $\theta$ the non-convergence diagnostics are applied.
-If we look at complete unbiasedness, however, we notice that univariate $\theta$s fail to diagnose the persistent bias in conditions where $p_{\mathrm{inc}}\geq.75$.
+If we look at complete unbiasedness, however, we notice that univariate $\theta$s fail to diagnose the persistent bias in conditions where $\pi_{\mathrm{inc}}\geq.75$.
 -->
@@
 <!-- 1.
-As hypothesized, biased, invalid estimates of quantities of scientific interest occurred more often in simulation conditions with a high proportion of incomplete cases $p_{\mathrm{inc}}$ and a low number of iterations $T$.
+As hypothesized, biased, invalid estimates of quantities of scientific interest occurred more often in simulation conditions with a high proportion of incomplete cases $\pi_{\mathrm{inc}}$ and a low number of iterations $T$.
 However, this only holds for multivariate scientific estimands.
@@
 However, some signs of non-convergence were detected in simulation conditions many more iterations.
-Under moderate missingness ($p_{\mathrm{inc}}\leq.50$) $\widehat{R}$-values decreased substantially until 30 to 40 iterations, which implies that mixing in the algorithm still improved with each additional iteration.
+Under moderate missingness ($\pi_{\mathrm{inc}}\leq.50$) $\widehat{R}$-values decreased substantially until 30 to 40 iterations, which implies that mixing in the algorithm still improved with each additional iteration.
 $AC$-values only improved until about six iterations, suggesting minimal improvement in trending was obtained beyond $T=6$ in the current simulation set-up.
@@
 Each line portrays one imputation chain (i.e., the values of a scalar summary $\theta$ across iterations).
-The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in the $m$ sets of imputations $\dot y_{\mathrm{imp}, \ell}$.
+The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in the $m$ sets of imputations $\dot \mathbf{Y}_{\mathrm{imp}, \ell}$.
 
@@
 The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
-It is obvious that non-convergence in this example impacts the distribution of the imputed values per imputation $\dot y_{\mathrm{imp}, \ell}$.
-This effect may translate into the distribution of the $m$ sets of completed data {$y_{\mathrm{obs}}, \dot y_{\mathrm{imp}, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$, and finally the pooled estimate $\bar{Q}$.
+It is obvious that non-convergence in this example impacts the distribution of the imputed values per imputation $\dot \mathbf{Y}_{\mathrm{imp}, \ell}$.
+This effect may translate into the distribution of the $m$ sets of completed data {$\mathbf{Y}_{\mathrm{obs}}, \dot \mathbf{Y}_{\mathrm{imp}, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$, and finally the pooled estimate $\bar{Q}$.
 $\bar{Q}$ is then a  biased, invalid estimate of $Q$.

### 2025-05-01 Manuscript/manuscript.qmd
merge notation sections

@@
 
-For each datapoint in $\mathbf{Y}_{\mathrm{mis}}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations in the imputation algorithm ($t = 1, 2, \dots, T$).
-The state-space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values).
-The collection of $\theta$-values between $t=1$ (summarizing the state-space of the algorithm at initialization) and $t=T$ (summarizing the state-space for the imputed values) will be referred to as an 'imputation chain'.
-
-## The MICE algorithm
-
-Anyone who analyzes person-data may run into a missing data problem.
+<!-- Anyone who analyzes person-data may run into a missing data problem.
 Missing data is not only ubiquitous across science, but treating it can also be a tedious task.
@@
 This technique is known as 'multiple imputation' [MI; @rubin76].
-
-Figure \ref{diagram} provides an overview of the steps involved with MI.
+-->
+
+Figure \ref{fig:diagram} provides an overview of the steps involved with MI.
 The missing part $\mathbf{Y}_{\mathrm{mis}}$ of an incomplete dataset is imputed $m$ times.
@@
 The algorithm then iterates over the $p$ variables, generating $m$ sets of imputed values for each variable $j$.
-After one full run through the data ($t=1$), the imputed values will be 'updated' in the second iteration ($t=2$).
+After one full run through the data, the imputed values will be 'updated' in the second iteration.
 This process should be repeated until convergence is reached.
+
+<!-- Each missing datapoint in $\mathbf{Y}_{\mathrm{mis}}$ may be imputed $m$ times, where $m$ is the number of imputations  ($\ell = 1, 2, \dots, m$). -->
+With iterative imputation, plausible values are sampled at every iteration $t$ of the imputation algorithm.
+But only the imputed values drawn at the final iteration $\tau$ (where $t = 1, 2, \dots, \tau$) are used for further analysis.
+<!-- For each datapoint in $\mathbf{Y}_{\mathrm{mis}}$, we draw $m$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $\tau$ is the number of iterations in the imputation algorithm ($t = 1, 2, \dots, \tau$).
+-->
+The state-space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$.
+A typical $\theta$ to track over iterations is the 'chain mean': the average of the imputed values for a variable $Y_j$ at iteration $t$.
+The collection of $\theta$-values between $t=1$ (the state-space of the algorithm after initialization) and $t=\tau$ (the state-space for the imputed values) will be referred to as an 'imputation chain'.
 
@@
 
-**MICE algorithm**
+**Algorithm 1: MICE algorithm**
 
@@
 1. Specify an imputation model for each variable $Y_j$, where $j = 1,...,p$.
-2. Repeat over variables $j = 1,...,p$.
+2. Repeat over variables ($j = 1,...,p$).
+2. Repeat over imputations ($\ell = 1,...,m$).
     a. Fill in starting imputations ($\dot{Y_j^0}$) by random draws from the observed part of each variable ($Y_{j}^{\mathrm{obs}}$).
@@
 
-*Iteration phase ($t = 1,...,T$)*
-
-4. Repeat over iterations $t = 1,...,T$.
-    a. Repeat over variables $j = 1,...,p$.
+*Iteration phase ($t = 1,...,\tau$)*
+
+4. Repeat over iterations ($t = 1,...,\tau$).
+    a. Repeat over variables ($j = 1,...,p$).
         I.
@@
 In MI, we only use the final estimate.
-Also, computing n_eff assumes infinitely many iterations, which is fine in Bayesian analyses where T is often at least one thousand.
+Also, computing n_eff assumes infinitely many iterations, which is fine in Bayesian analyses where \tau is often at least one thousand.
 But the default in iterative imp is often substantially lower.
@@
 5) to define $\widehat{R}$.
-Let $m$ be the total number of chains, $T$ the number of iterations per chain (where $T\geq2$), and $\theta$ the scalar summary of interest.
+Let $m$ be the total number of chains, $\tau$ the number of iterations per chain (where $\tau\geq2$), and $\theta$ the scalar summary of interest.
 For each chain ($\ell = 1, 2, \dots, m$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.
@@
 \begin{align*}
-W&=\frac{1}{m} \sum_{\ell=1}^{m} s_{\ell}^{2}, \text { where } s_{\ell}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t \ell)}-\bar{\theta}^{(\cdot \ell)}\right)^{2}. 
+W&=\frac{1}{m} \sum_{\ell=1}^{m} s_{\ell}^{2}, \text { where } s_{\ell}^{2}=\frac{1}{\tau-1} \sum_{t=1}^{\tau}\left(\theta^{(t \ell)}-\bar{\theta}^{(\cdot \ell)}\right)^{2}. 
 \end{align*}
@@
 \begin{align*}
-B&=\frac{T}{m-1} \sum_{\ell=1}^{m}\left(\bar{\theta}^{(\cdot \ell)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot \ell)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t \ell)} \text{, } \bar{\theta}^{(\cdot \cdot)}=\frac{1}{m} \sum_{\ell=1}^{m} \bar{\theta}^{(\cdot \ell)}.
+B&=\frac{\tau}{m-1} \sum_{\ell=1}^{m}\left(\bar{\theta}^{(\cdot \ell)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot \ell)}=\frac{1}{\tau} \sum_{t=1}^{\tau} \theta^{(t \ell)} \text{, } \bar{\theta}^{(\cdot \cdot)}=\frac{1}{m} \sum_{\ell=1}^{m} \bar{\theta}^{(\cdot \ell)}.
 \end{align*}
@@
 \widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
-\text{ where } \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{T-1}{T} W+\frac{1}{T} B.
+\text{ where } \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{\tau-1}{\tau} W+\frac{1}{\tau} B.
 \end{equation*}
@@
 We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run [@gelm92].
-The assumption underlying this interpretation is that chains are 'over-dispersed' at $t=1$, and reach convergence as $T \to \infty$.
+The assumption underlying this interpretation is that chains are 'over-dispersed' at $t=1$, and reach convergence as $\tau \to \infty$.
 Over-dispersion implies that the initial values of the chains are 'far away' from the target distribution and each other.
@@
 \begin{equation*}
-AC = \left( \frac{T}{T-1} \right) \frac{\sum_{t=1}^{T-1}(\theta_t - \bar{\theta}^{(\cdot m)})(\theta_{t+1} - \bar{\theta}^{(\cdot m)})}{\sum_{t=1}^{T}(\theta_t - \bar{\theta}^{(\cdot m)})^2}.
+AC = \left( \frac{\tau}{\tau-1} \right) \frac{\sum_{t=1}^{\tau-1}(\theta_t - \bar{\theta}^{(\cdot m)})(\theta_{t+1} - \bar{\theta}^{(\cdot m)})}{\sum_{t=1}^{\tau}(\theta_t - \bar{\theta}^{(\cdot m)})^2}.
 \end{equation*}
@@
 
-**Algorithm 1: simulation set-up (pseudo-code)**
+**Algorithm 2: simulation set-up (pseudo-code)**
 ```
@@
 Since there is just one estimate for $Q=r^2$ to evaluate, we consider $Q=\mu_Y, \sigma_Y,r^2, \beta_1$.
-In the figures, we present all missingness conditions, but only early stopping conditions where $1 \leq T\leq50$.
-That is, the results are more or less stable for conditions where $T \geq 30$.
+In the figures, we present all missingness conditions, but only early stopping conditions where $1 \leq \tau\leq50$.
+That is, the results are more or less stable for conditions where $\tau \geq 30$.
 Full results are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).
@@
 <!-- The estimated univariate means seem unaffected by early stopping.
-This implies that approximately unbiased estimates may be obtained with as little as one iteration ($T \geq 1$).
+This implies that approximately unbiased estimates may be obtained with as little as one iteration ($\tau \geq 1$).
 The bias in $\bar{Q}$ depends solely on the proportion of incomplete cases, with $\pi_{\mathrm{inc}}\geq.75$ leading to more extreme biases than other conditions.
@@
 <!-- The estimated standard deviations are not substantially impacted by early stopping either.
-While the estimates for $T=1$ are not equal to other conditions, the bias is actually *less* severe than for $T\geq2$.
+While the estimates for $\tau=1$ are not equal to other conditions, the bias is actually *less* severe than for $\tau\geq2$.
 Therefore, no iteration at all is needed to obtain approximately unbiased estimates for $Q=\sigma$.
@@
 In conditions where $\pi_{\mathrm{inc}}=.05$, estimates are unbiased after one iteration.
-The highest number of iterations necessary to reach approximate unbiasedness is seven ($\pi_{\mathrm{inc}}=.95$; $T\geq7$).
+The highest number of iterations necessary to reach approximate unbiasedness is seven ($\pi_{\mathrm{inc}}=.95$; $\tau\geq7$).
 Completely unbiased estimates are only obtained in conditions where $\pi_{\mathrm{inc}}\leq.25$.
-This implies that even for biased estimates, $T\geq7$ would suffice to reach a stable solution.
+This implies that even for biased estimates, $\tau\geq7$ would suffice to reach a stable solution.
 -->
@@
 Similar to $Q=r^2$, we observe approximately unbiased estimates after at least one, and at most seven iterations, depending on $\pi_{\mathrm{inc}}$.
-Completely unbiased estimates are only obtained in conditions where $T\geq3$ and $\pi_{\mathrm{inc}}\leq .50$.
+Completely unbiased estimates are only obtained in conditions where $\tau\geq3$ and $\pi_{\mathrm{inc}}\leq .50$.
 Despite persistent bias in conditions where $\pi_{\mathrm{inc}}\geq.75$, the coverage rates are non-improving after just three iterations, irrespective of $\pi_{\mathrm{inc}}$.
-We conclude that any condition where $T\geq3$ yields approximately nominal coverage rates, and thus confidence-valid estimates.
+We conclude that any condition where $\tau\geq3$ yields approximately nominal coverage rates, and thus confidence-valid estimates.
 -->
@@
 -->
-<!-- We conclude that approximately unbiased estimates may be obtained in conditions where $T\geq7$, whereas confidence-valid estimates require at most three iterations.  -->
+<!-- We conclude that approximately unbiased estimates may be obtained in conditions where $\tau\geq7$, whereas confidence-valid estimates require at most three iterations.  -->
 
@@
 We observe approximately unbiased estimates after at most seven iterations (for any $\pi_{\mathrm{inc}}$ considered).
-This implies that the algorithm produces stable, non-improving estimates when $T\geq7$.
+This implies that the algorithm produces stable, non-improving estimates when $\tau\geq7$.
 -->
@@
 We evaluate the performance of the methods by establishing whether they correctly identify conditions in which $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
-As we concluded in the last section, substantially biased estimates were only obtained in conditions where $T\leq6$, and we observed non-nominal coverage only when $T\leq3$.
+As we concluded in the last section, substantially biased estimates were only obtained in conditions where $\tau\leq6$, and we observed non-nominal coverage only when $\tau\leq3$.
 However, there was some persistent bias in conditions where $\pi_{\mathrm{inc}}\geq.75$, irrespective of the number of iterations.
@@
 <!-- As expected, $\widehat{R}$ lowers with the number of iterations.
-After an initial 'dip' in conditions where $T=3$, we observe a gradual decrease in $\widehat{R}$-values with every additional iteration.
+After an initial 'dip' in conditions where $\tau=3$, we observe a gradual decrease in $\widehat{R}$-values with every additional iteration.
 This implies improving convergence, until the decrease tapers off after 30 to 40 iterations.
-Using the threshold $\widehat{R}>1.2$, we diagnose non-convergence in conditions where $T\leq7$, whereas the threshold 1.1 is exceeded when $T\leq13$.
-The very strict threshold 1.01 is surpassed in all conditions $T\leq100$ (not shown), and therefore not informative under the current specifications.
+Using the threshold $\widehat{R}>1.2$, we diagnose non-convergence in conditions where $\tau\leq7$, whereas the threshold 1.1 is exceeded when $\tau\leq13$.
+The very strict threshold 1.01 is surpassed in all conditions $\tau\leq100$ (not shown), and therefore not informative under the current specifications.
 With this set of methods, we fail to identify non-convergence in conditions with persistent bias due to missingness.
@@
 
-<!-- Similarly, $AC$ decreases with higher $T$s.
-$AC$ indicates improving stationarity in conditions where $T\leq6$.
+<!-- Similarly, $AC$ decreases with higher $\tau$s.
+$AC$ indicates improving stationarity in conditions where $\tau\leq6$.
 The $AC$-values do not exceed the threshold defined by statistical significance of the $AC$-values.
@@
 The only apparent difference is in the $\widehat{R}$ threshold 1.1.
-Instead of diagnosing non-convergence in conditions where $T\leq13$, we now diagnose this when $T\leq11$.
+Instead of diagnosing non-convergence in conditions where $\tau\leq13$, we now diagnose this when $\tau\leq11$.
 All other observations are equivalent.
@@
 Interestingly, this only occurs after ten or even thirty iterations, while we would expect more signs of trending early in the iterations.
-Since the $AC$-values are not improving or worsening when $T\geq7$, we conclude that non-convergence is incorrectly diagnosed based on the threshold.
+Since the $AC$-values are not improving or worsening when $\tau\geq7$, we conclude that non-convergence is incorrectly diagnosed based on the threshold.
 -->
@@
 For this $\theta$, the results differ only in the most extreme missingness condition.
-When $\pi_{\mathrm{inc}}=.95$, we diagnose non-convergence at $T\leq9$ according to $\widehat{R}>1.2$, $T\leq19$ for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).
-The 'dip' in $\widehat{R}$-values at $T=3$ is even more pronounced than with other $\theta$s.
+When $\pi_{\mathrm{inc}}=.95$, we diagnose non-convergence at $\tau\leq9$ according to $\widehat{R}>1.2$, $\tau\leq19$ for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).
+The 'dip' in $\widehat{R}$-values at $\tau=3$ is even more pronounced than with other $\theta$s.
 Some $\widehat{R}$-values even overcome the threshold of 1.2.
-If we would terminate the algorithm at $T=3$, we would incorrectly conclude that the algorithm reached approximate convergence, according to this threshold.
+If we would terminate the algorithm at $\tau=3$, we would incorrectly conclude that the algorithm reached approximate convergence, according to this threshold.
 -->
@@
 The threshold to diagnose significant $AC$-values does not appear to be appropriate in at least the current set-up, and perhaps in iterative imputation in general.
-A better heuristic to diagnose non-stationarity with $AC$ may be through evaluation of the $AC$-values across iterations: If the values do not substantially decrease with $T$, approximate stationarity may be concluded.
+A better heuristic to diagnose non-stationarity with $AC$ may be through evaluation of the $AC$-values across iterations: If the values do not substantially decrease with $\tau$, approximate stationarity may be concluded.
 -->
@@
 <!-- 1.
-As hypothesized, biased, invalid estimates of quantities of scientific interest occurred more often in simulation conditions with a high proportion of incomplete cases $\pi_{\mathrm{inc}}$ and a low number of iterations $T$.
+As hypothesized, biased, invalid estimates of quantities of scientific interest occurred more often in simulation conditions with a high proportion of incomplete cases $\pi_{\mathrm{inc}}$ and a low number of iterations $\tau$.
 However, this only holds for multivariate scientific estimands.
@@
 In this study, however, we observed $AC$-values as high as theoretically possible ($AC=1$).
-The second part of the hypothesis is not refuted: the $AC$-values did decrease quickly as a function of $T$.
+The second part of the hypothesis is not refuted: the $AC$-values did decrease quickly as a function of $\tau$.
 -->
@@
 Under moderate missingness ($\pi_{\mathrm{inc}}\leq.50$) $\widehat{R}$-values decreased substantially until 30 to 40 iterations, which implies that mixing in the algorithm still improved with each additional iteration.
-$AC$-values only improved until about six iterations, suggesting minimal improvement in trending was obtained beyond $T=6$ in the current simulation set-up.
+$AC$-values only improved until about six iterations, suggesting minimal improvement in trending was obtained beyond $\tau=6$ in the current simulation set-up.
 -->
@@
 Approximately unbiased, confidence-valid estimates of multivariate $Q$s were obtained after a maximum of seven iterations.
-Continued iterations beyond $T = 7$ did not yield better estimates.
+Continued iterations beyond $\tau = 7$ did not yield better estimates.
 -->
@@
 In practice, this implies that the default number of iterations in MICE is not sufficient.
-$AC$ can only be computed if $T\geq3$, while the current default in the `mice` package is five [@mice].
+$AC$ can only be computed if $\tau\geq3$, while the current default in the `mice` package is five [@mice].
 Identifying a reasonable decrease in the $AC$-values across iterations (an 'elbow') requires more than three observations.
@@
 Moreover, an increased number of iterations would grant the opportunity to exclude the first iterations from evaluation with $\widehat{R}$.
-Excluding $T\leq3$ provides empirical researchers with a less ambiguous heuristic to diagnose non-convergence, because we expect the initial 'dip' in $\widehat{R}$-values to disappear.
+Excluding $\tau\leq3$ provides empirical researchers with a less ambiguous heuristic to diagnose non-convergence, because we expect the initial 'dip' in $\widehat{R}$-values to disappear.
 -->
@@
 ```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence (reproduced from van Buuren; 2018).
-Depicted are chain means of variable $j$ in each imputation ($m=5$; $T=10$).
+Depicted are chain means of variable $j$ in each imputation ($m=5$; $\tau=10$).
 Note that any difference between the two scenarios (see e.g., the magnitude of the chain means on the y-axes) stems from the imputation algorithms---the two scenarios have equal incomplete data $y$."}
@@
 Panel A consists of the two traceplots that may be evaluated through visual inspection.
-Panel B shows two versions of the $AC$: the default calculation with `R` function `stats::acf()` [@R], and manual calculation as the correlation between $\theta$ in iteration $t$ and $\theta$ in iteration $t+1$.
+Panel B shows two versions of the $AC$: the default calculation with `R` function `stats::acf()` [@R], and manual calculation as the correlation between $\theta$ in iteration $\tau$ and $\theta$ in iteration $t+1$.
 Panel C displays the traditional computation of $\widehat{R}$ conform Gelman and Rubin (original), and in panel D we see $\widehat{R}$ as computed by implementing Vehtari et al.'s recommendations (adapted).
 
-```{r diagnostics, fig.cap="Convergence diagnostics applied on the two scenarios from Figure 2 ($\\theta$ = chain mean of variable $j$; $m=5$; $T=10$).
+```{r diagnostics, fig.cap="Convergence diagnostics applied on the two scenarios from Figure 2 ($\\theta$ = chain mean of variable $j$; $m=5$; $\tau=10$).
 Depicted are the imputation chains in panel A, two versions of the autocorrelation in panel B, and two versions of $\\widehat{R}$ in panels C and D."}
@@
 Otherwise, it is not possible to perform all three transformations to the chains of $\theta$-values.
-For low values of $t$, the $\widehat{R}$-values conform Vehtari et al.
+For low values of $\tau$, the $\widehat{R}$-values conform Vehtari et al.
 are, therefore, more similar to the original $\widehat{R}$ by Gelman and Rubin.

### 2025-05-02 Manuscript/manuscript.qmd
update identifiers in background section

@@
 The technique enables researchers to draw valid inferences from incomplete data [@rubin76].
-[TODO: remove the next sentence?] This is in contrast to ad hoc solutions such as list-wise deletion, which may bias scientific estimates.
 Multiple imputation separates the missing data problem from the scientific problem.
@@
 This process is repeated several times to obtain a distribution of plausible values, which reflects the uncertainty in the data due to missingness.
-Finally, the scientific estimates are pooled across the multiple imputations.
+The analysis of scientific interest thus yields one estimate for each imputed dataset.
+Afterwards, the estimates from the different imputations are pooled into a single estimate.
 This pooled estimate is unbiased and confidence-valid if--and only if--both the missing data problem and the scientific problem are appropriately considered.
 
-[TODO: add refs? e.g., rubin after 18 years paper? murray 2018?]
+<!-- [TODO: add refs? e.g., rubin after 18 years paper? murray 2018?] -->
+<!--  This is in contrast to ad hoc solutions such as list-wise deletion, which may bias scientific estimates. -->
 
 To obtain valid scientific estimates from incomplete data, both the missing data problem and the scientific problem should be appropriately considered.
-The popular imputation algorithm 'MICE' has proven to be a powerful tool to draw valid inferences from incomplete data under many circumstances [@buuren2011; @buur18].
+The popular imputation algorithm 'Multivariate Imputation by Chained Equetions' (MICE) has proven to be a powerful tool to draw valid inferences from incomplete data under many circumstances [@buuren2011; @buur18].
 The algorithm is named after its iterative nature: missing data are imputed on a variable-by-variable basis, using a sequence of univariate imputation models.
@@
 Yet, there has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms.
+
+[TODO: move/integrate this section?] There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@zhu15; @takahashi2017].
+Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murray2018].
+Therefore, algorithmic convergence should be monitored carefully---although this is not straightforward.
 
@@
 Diagnosing non-convergence through visual inspection may, however, be undesirable: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18].
-[add argument: "Inspection of such plots is a notoriously unreliable method of assessing convergence and in addition is unwieldy when monitoring a large number of quantities of interest, such as can arise in complicated hierarchical models" (Gelman et al., 2013, p.
+[TODO: add argument: "Inspection of such plots is a notoriously unreliable method of assessing convergence and in addition is unwieldy when monitoring a large number of quantities of interest, such as can arise in complicated hierarchical models" (Gelman et al., 2013, p.
 285).] Therefore, a quantitative diagnostic method to assess convergence would be preferred.
@@
 
-[TODO: add *approximate* convergence as the most converged state that the algorithm can obtain?]
 <!-- It is challenging to arrive upon a single point at which convergence has been reached. -->
@@
 
+[TODO: add *approximate* convergence as the most converged state that the algorithm can obtain?]
+
 <!-- [TODO: integrate this section] In an empirical study with STATA, where $\widehat{R}$ was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold $\widehat{R}>1.2$.
@@
 The imputed data is then combined with the observed data $\mathbf{Y}_{\mathrm{obs}}$ to create $m$ completed datasets.
-On each of these datasets the analysis of scientific interest is performed to estimate $Q$: the quantity of scientific interest (e.g., a regression coefficient).
+On each of these datasets, the analysis of scientific interest is performed to estimate $Q$: the quantity of scientific interest (e.g., a regression coefficient).
 Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}_\ell$-values are obtained.
@@
 
-A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@buuren2011].
+A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE' [@buuren2011].
 With MICE, imputed values are drawn from the posterior predictive distribution of the missing values on a variable-by-variable basis.
@@
 The algorithm then iterates over the $p$ variables, generating $m$ sets of imputed values for each variable $j$.
-After one full run through the data, the imputed values will be 'updated' in the second iteration.
+After one full run through the data, the imputed values are 'updated' in the next iteration.
 This process should be repeated until convergence is reached.
@@
 The values generated by the algorithm will vary even after convergence [@gelm13].
+The most converged state that such an algorithm may reach is *approximate*  convergence.
 Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09].
@@
 
-[TODO: move/integrate this section?] There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@zhu15; @takahashi2017].
-Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murray2018].
-Therefore, algorithmic convergence should be monitored carefully---although this is not straightforward.
-
 <!-- Recommendation @buur18: "the different streams should be freely intermingled with one another, without showing any definite trends". -->
@@
 
-In 2019, Vehtari et al.
-proposed an updated version of the potential scale reduction factor $\widehat{R}$, originally coined in 1992.
-The adapted version would be better suited to detect non-mixing in the tails of distributions, and even identify non-stationarity.
-This version uses three transformations on the scalar summary $\theta$, before computing $\widehat{R}$-values.
-Namely, rank-normalization, folding, and localization.
-We follow Vehtari et al.'s formulation (2019, p.
-5) to define $\widehat{R}$.
+In 2021, Vehtari et al.
+published an updated version of the potential scale reduction factor $\widehat{R}$, originally coined by Gelman and Rubin in 1992.
+This adapted version aims to detect non-mixing even in the tails of distributions, and non-stationarity over iterationsuses.
+To achieve this, the 2021 version uses three transformations on the scalar summary $\theta$ before computing $\widehat{R}$-values: rank-normalization, folding, and localization.
+To define $\widehat{R}$, we largely follow Vehtari et al.'s formulation (2019, p.
+5), with a few exceptions.
 Let $m$ be the total number of chains, $\tau$ the number of iterations per chain (where $\tau\geq2$), and $\theta$ the scalar summary of interest.
@@
 
-It is unlikely that iterative algorithms such as MICE will achieve the ideal values of $\widehat{R}=1$ and $AC=0$.
-[TODO: check this!
-isn't it because of under-dispersion and reliance on observed data?] Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state.
+<!--  [TODO: check this!
+isn't it because of under-dispersion and reliance on observed data?] It is unlikely that iterative algorithms such as MICE will achieve the ideal values of $\widehat{R}=1$ and $AC=0$.
+Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state.
 The aim, therefore, is to reach approximate convergence [@gelm13].
+-->
+
 Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} \gtrapprox 1$), and there is little dependency between subsequent iterations of imputation chains ($AC \gtrapprox 0$).
@@
 The number of simulation repetitions $n_{\mathrm{sim}} = 2000$.
-
-[TODO: add guaranteed convergence because compatible imputation models, equivalent to JM and true Gibbs sampler, @vanbuuren2007]
 
@@
 Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{\mathrm{sim}} = 2000$).
-To evaluate convergence, we terminate the imputation algorithm after a varying number of iterations ($n_{\mathrm{it}} = 1, 2, ..., 100$).
+To evaluate convergence, we terminate the imputation algorithm after a varying number of iterations ($n_{\mathrm{it}} = 1, 2, ..., 100$, corresponding to $\tau$ in each individual imputation run).
 We differentiate between six different missingness scenarios that are defined in the data generating mechanism (see Table XYZ).
@@
 
+<!-- [TODO: add guaranteed convergence because compatible imputation models, equivalent to JM and true Gibbs sampler, @vanbuuren2007] -->
+
 We impute the missing data five times ($m = 5$) using Bayesian linear regression imputation with `mice` [@mice].
-On each imputed dataset, we perform multiple linear regression as our analysis of scientific interest.
-Our estimands are the regression coefficients $\boldsymbol{\beta}$,
+Bayesian linear regression is a compatible imputation model for the set of variables at hand.
+Therefore, the imputation algorithm is guaranteed to converge [@vanbuuren2007].
+
+On each imputed dataset, we perform multiple linear regression as our analysis of scientific interest. Our estimands are the regression coefficients $\boldsymbol{\beta}$,
 
@@
 
-- non-convergence in iterative algorithms is diagnosed using an identifier and a parameter
-
-- the parameter can be any statistic that we track across iterations, for example the average imputed value per imputation (i.e., chain means)
-
-- the identifier is a calculation of some sort that quantifies non-convergence
-
-- identifiers are historically focused on either non-mixing between chains or non-stationarity within chains
-
-- the popular non-mixing identifier rhat has recently been updated by Vehtari et al, and should now work for non-stationarity as well
-
-- just to be sure, we also use autocorrelation to quantify trending within chains
-
-- we apply these identifiers to the two parameters that we typically evaluate after imputation using visual inspection: chain means and chain variances
-
 We use two non-convergence identifiers---autocorrelation and $\widehat{R}$---to diagnose non-convergence in the imputation models of the four incomplete variables.
 For each variable we apply the two identifiers on two parameters---chain means and chain variances---and four variables, resulting in 16 sets of identifier-parameter-variable pairs.
+
+- non-convergence in iterative algorithms is diagnosed using an identifier and a parameter
+
+- the parameter can be any statistic that we track across iterations, for example the average imputed value per imputation (i.e., chain means)
+
+- the identifier is a calculation of some sort that quantifies non-convergence
+
+- identifiers are historically focused on either non-mixing between chains or non-stationarity within chains
+
+- the popular non-mixing identifier rhat has recently been updated by Vehtari et al, and should now work for non-stationarity as well
+
+- just to be sure, we also use autocorrelation to quantify trending within chains
+
+- we apply these identifiers to the two parameters that we typically evaluate after imputation using visual inspection: chain means and chain variances
 

### 2025-05-08 Manuscript/manuscript.qmd
cross-ref figures

@@
 
+<!-- ## Aims -->
+
+<!-- - can we diagnose non-convergence quantitatively? -->
+<!-- - ~~which parameter should we track?~~ -->
+<!-- - how many iterations are needed before converge to a stationary distribution is reached? -->
+<!-- - how many iterations are needed before valid inferences are reached? -->
+
 ```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
@@
 
-## Aims
-
-- can we diagnose non-convergence quantitatively?
-- ~~which parameter should we track?~~
-- how many iterations are needed before converge to a stationary distribution is reached?
-- how many iterations are needed before valid inferences are reached?
-
 # Introduction
@@
 
-[TODO: add *approximate* convergence as the most converged state that the algorithm can obtain?]
+<!-- *approximate* convergence as the most converged state that the algorithm can obtain (gelman, 2013?) -->
 
@@
 
-Figure \ref{fig:diagram} provides an overview of the steps involved with MI.
+Figure @fig-diagram provides an overview of the steps involved with MI.
 The missing part $\mathbf{Y}_{\mathrm{mis}}$ of an incomplete dataset is imputed $m$ times.
@@
 
-```{r diagram, echo=FALSE, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\\hat{Q}$, to a single pooled estimate $\\bar{Q}$."}
+```{r fig-diagram, echo=FALSE, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\\hat{Q}$, to a single pooled estimate $\\bar{Q}$."}
 
@@
 
-**Algorithm 1: MICE algorithm**
-
-*Initialization phase ($t = 0$)*
+::: {#nte-algorithm .callout-note}
+
+## Algorithm 1: MICE algorithm
+
+**Initialization phase ($t = 0$)**
 
@@
 
-*Iteration phase ($t = 1,...,\tau$)*
+**Iteration phase ($t = 1,...,\tau$)**
 
@@
 
+:::
 <!-- \begin{enumerate} -->
@@
 
+<!-- - identifiers are historically focused on either non-mixing between chains or non-stationarity within chains -->
+
+<!-- - the popular non-mixing identifier rhat has recently been updated by Vehtari et al, and should now work for non-stationarity as well -->
+
+<!-- - just to be sure, we also use autocorrelation to quantify trending within chains -->
+
+<!-- - we apply these identifiers to the two parameters that we typically evaluate after imputation using visual inspection: chain means and chain variances -->
+
 There are many diagnostic tools to identify non-convergence in iterative (MCMC) algorithms [@broo98; @elad06].
+Non-convergence is diagnosed using an identifier and a parameter.
+The parameter can be any statistic that we track across iterations, for example the average imputed value per imputation (i.e., chain means).
+The identifier is a calculation of some sort that quantifies non-convergence in the scalar.
 We consider only two of them that may be appropriate for imputation algorithms---one to monitor signs of non-mixing, and one for non-stationarity.
@@
 We investigate non-convergence in iterative imputation through model-based simulation in R (version 4.4.2; @R).
-We provide a summary of the simulation set-up in Algorithm 1; the complete script and technical details are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint) [TODO: make this a Zenodo DOI].
+We provide a summary of the simulation set-up in Algorithm 2; the complete script and technical details are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint) [TODO: make this a Zenodo DOI].
 The number of simulation repetitions $n_{\mathrm{sim}} = 2000$.
@@
 
-**Algorithm 2: simulation set-up (pseudo-code)**
+::: {#nte-simulation .callout-note}
+
+## Algorithm 2: simulation set-up (pseudo-code)
+
 ```
@@
 ```
+:::
 
@@
 \begin{table}[]
+\caption{Missingness scenarios}
 \begin{tabular}{lcr}
@@
 
-[TODO: add left-tailed M(N)AR mechanism, md pattern, and maybe refer to ampute paper or "Dance of..." by Schouten 2018?].
+<!-- [TODO: add left-tailed M(N)AR mechanism, md pattern, and maybe refer to ampute paper or "Dance of..." by Schouten 2018?]. -->
 
@@
 
+We use two non-convergence identifiers---autocorrelation and $\widehat{R}$---to diagnose non-convergence in the imputation algorithm.
+These two identifiers are applied to two scalar summaries of the state-space of the algorithm---chain means and chain variances.
+We compute the two identifiers on four different variables: the outcome variable $Y$, and the three predictors $X_1$, $X_2$, and $X_3$.
+In total, we apply the two identifiers on two parameters and four variables, resulting in 16 sets of identifier-parameter-variable pairs.
+
 \begin{table}[]
@@
 
-We use two non-convergence identifiers---autocorrelation and $\widehat{R}$---to diagnose non-convergence in the imputation models of the four incomplete variables.
-For each variable we apply the two identifiers on two parameters---chain means and chain variances---and four variables, resulting in 16 sets of identifier-parameter-variable pairs.
-
-- non-convergence in iterative algorithms is diagnosed using an identifier and a parameter
-
-- the parameter can be any statistic that we track across iterations, for example the average imputed value per imputation (i.e., chain means)
-
-- the identifier is a calculation of some sort that quantifies non-convergence
-
-- identifiers are historically focused on either non-mixing between chains or non-stationarity within chains
-
-- the popular non-mixing identifier rhat has recently been updated by Vehtari et al, and should now work for non-stationarity as well
-
-- just to be sure, we also use autocorrelation to quantify trending within chains
-
-- we apply these identifiers to the two parameters that we typically evaluate after imputation using visual inspection: chain means and chain variances
-
 <!-- We consider eight sets of diagnostic methods to identify non-convergence, by applying $\widehat{R}$ and $AC$ on four different $\theta$s.
@@
 
-[TODO: add tables, with MC SEs]
-
-## Performance measures
-
-```{r bias}
-plot_results(results, "bias", 0, "MCAR")
-```
-
-- more missingness = more extreme bias, but also steeper decrease over iterations
-
-```{r cov}
-plot_results(results, "cov", 0.95, "MCAR")
-```
-
-- no clear trend
-
-```{r ciw}
-plot_results(results, "ciw", NULL, "MCAR")
-```
-
-- more missingness = wider CIs, with steeper decrease in CIW, but not stababiling at same value between missingness conditions.
-
-In Figure \@ref(fig:perf) we show the performance measures: bias in the regression estimate (panel A), the empirical coverage rate of the regression estimate (panel C) and the average confidence interval width of this estimate (panel D).
-
-We see that within a few iterations the bias in the regression estimate approaches zero (see \@ref(fig:perf)A).
-When $n_{it} \geq 6$, even the worst-performing conditions (e.g., with a proportion of incomplete cases of 75%) produce stable, non-improving estimates [regression coefficient is underestimated because there is less info to estimate the relation??].
-
-Nominal coverage is quickly reached (see \@ref(fig:perf)C).
-After just three iterations, the coverage rates are non-improving in every missingness condition [but MNAR with 5% incomplete cases does not reach nominal coverage --\> due to bias in the estimate in combination with very narrow CI (see CIW!)].
-
-The average confidence interval width decreases quickly with every added iteration until a stable plateau is reached (see \@ref(fig:perf)D).
-Depending on the proportion of incomplete cases this takes up-to $n_{it} \geq 9$.
+Our results show the non-convergence metrics and the performance of the MICE algorithm under different missingness conditions at a varying number of iterations.
+For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias and coverage, $X_3$.
+Full results are available from [TODO: add repo DOI].
+[TODO: add tables, with MC SEs, to appendix.]
 
@@
 
-Figure \@ref(fig:ac) shows the autocorrelations in the chain means (panel A) and chain variances (panel B).
-
-```{r ac_mean}
+### AC
+
+Figures @fig-ac_mean and @fig-ac_sd show the autocorrelations in the chain means and chain variances, respectively.
+
+```{r fig-ac_mean}
 plot_results(results, "ac_mean", 0, "MCAR")
@@
 
-```{r ac_sd}
-plot_results(results, "ac_sd", 0, "MCAR")
-```
-
-Autocorrelation in the chain means decreases rapidly in the first few iterations (see \@ref(fig:ac)A).
+Autocorrelation in the chain means decreases rapidly in the first few iterations (see @fig-ac_mean).
 The decrease is substantive until $n_{it} \geq 6$.
@@
 These results hold irrespective of the missingness condition.
-Autocorrelation in the chain variances show us something similar (see \@ref(fig:ac)B).
+
+```{r fig-ac_sd}
+plot_results(results, "ac_sd", 0, "MCAR")
+```
+
+Autocorrelation in the chain variances show us something similar (see @fig-ac_sd).
 The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous than for chain means, but generally around $n_{it} \geq 10$.
@@
 
-```{r psrf_mean}
+### $\widehat{R}$
+
+Figure @fig-psrf_mean and @fig-psrf_sd show the potential scale reduction factor in the chain means and chain variances respectively.
+
+```{r fig-psrf_mean}
 plot_results(results, "psrf_mean", 1, "MCAR")
@@
 
-```{r psrf_sd}
-plot_results(results, "psrf_sd", 1, "MCAR")
-```
-
-Figure \@ref(fig:rh) shows the potential scale reduction factor in the chain means (panel A) and chain variances (panel B).
-
-We observe that $\widehat{R}$-values of the chain means generally decrease as a function of the number of iterations (see \@ref(fig:rh)A).
+We observe that $\widehat{R}$-values of the chain means generally decrease as a function of the number of iterations (see @fig-psrf_mean).
 An exception to this observation is a steep increase in iterations $3 \leq n_{it} \leq 5$ [TODO: interpret??
@@
 There is no apparent differentiation between the missingness conditions.
-The mixing between chain variances mimics the mixing between chain means almost perfectly (see \@ref(fig:rh)B).
+
+```{r fig-psrf_sd}
+plot_results(results, "psrf_sd", 1, "MCAR")
+```
+
+The mixing between chain variances mimics the mixing between chain means almost perfectly (see @fig-psrf_sd.
 Irrespective of the missingness condition, the $\widehat{R}$-values taper off around $n_{it} \geq 30$.
-[The rhat plots all show some initialization before the fifth iteration: is rhat useful before that??]
-
-```{r, cache=TRUE, eval=FALSE, fig.height=9}
-performance |>
-  filter(method == "MICE") |>
-ggplot(aes(x = .it, y = psrf_mean)) +
-  geom_point(aes(group = .sim), shape = 1, alpha = 0.05) +
-  geom_smooth(method = "loess") +
-  facet_grid(term ~ mech + prop) +
-  labs(
-    title = "Potential Scale reduction Factor(PSRF) applied to chain means",
-    x = "Iteration number",
-    y = "PSRF"
-  ) +
-  theme_classic() 
-```
+
+[TODO: write about: The rhat plots all show some initialization before the fifth iteration: is rhat useful before that??]
+
+## Performance measures
+
+In Figures @fig-bias - @fig-ciw we show the performance measures: bias in the regression estimate, the empirical coverage rate of the regression estimate and the average confidence interval width of this estimate.
+
+```{r fig-bias}
+plot_results(results, "bias", 0, "MCAR")
+```
+
+We see that within a few iterations the bias in the regression estimate approaches zero (see @fig-bias).
+When $n_{it} \geq 6$, even the worst-performing conditions (e.g., with a proportion of incomplete cases of 75%) produce stable, non-improving estimates.
+[TODO: check if regression coefficients are first underestimated because there is less info to estimate the relation??] [TODO: write about: more missingness = more extreme bias, but also steeper decrease over iterations].
+
+```{r fig-cov}
+plot_results(results, "cov", 0.95, "MCAR")
+```
+
+Nominal coverage is quickly reached (see @fig-cov). After just three iterations, the coverage rates are non-improving in every missingness condition. 
+<!-- no clear trend across conditions-->
+
+```{r fig-ciw}
+plot_results(results, "ciw", NULL, "MCAR")
+```
+
+The average confidence interval width decreases quickly with every added iteration until a stable plateau is reached (see @fig-ciw).
+Depending on the proportion of incomplete cases this takes up-to $n_{it} \geq 9$.
+[TODO: write about: more missingness = wider CIs, with steeper decrease in CIW, but not stababiling at same value between missingness conditions.]
+
+<!-- [but MNAR with 5% incomplete cases does not reach nominal coverage; due to bias in the estimate in combination with very narrow CI (see CIW!)]. -->
 
@@
 
+## Summary
+
+Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
+We observe approximately unbiased estimates after at most seven iterations (for any $\pi_{\mathrm{inc}}$ considered).
+This implies that the algorithm produces stable, non-improving estimates when $\tau\geq7$.
+The number of iterations necessary to obtain approximately unbiased, confidence-valid estimates corresponds to a $\widehat{R}$ threshold of $\approx 1.2$.
+
+<!-- The thresholds 1.1 and 1.01 seem too strict compared to the validity of the inferences.
+The threshold to diagnose significant $AC$-values does not appear to be appropriate in at least the current set-up, and perhaps in iterative imputation in general.
+A better heuristic to diagnose non-stationarity with $AC$ may be through evaluation of the $AC$-values across iterations: If the values do not substantially decrease with $\tau$, approximate stationarity may be concluded.
+-->
+
 # Discussion
 
-Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the non-convergence identifiers.
-Of course, it never hurts to iterate longer, but such calculations hardly bring added value.
-
-- Convergence diagnostics keep improving substantially until n_it = 20-30
-
-- Performance measures do not improve after n_it = 9
-
-- [methodological explanation is that rhat and ac have a lag (few it to inform your statistic) --\> will always indicate convergence slower than inferential validity is reached]
+Identifying non-convergence may be a crucial aspect of drawing valid statistical inferences from incomplete data.
+In this simulation study, we have shown that non-convergence in iterative imputation algorithms does indeed go hand in hand with biased, invalid estimates.
+Our study , however, found that inferential validity was achieved after five to ten iterations, much earlier than indicated by non-convergence identifiers.
+While  convergence diagnostics kept improving substantially until $\tau \approx 20-30$, performance measures did not improve after $\tau = 9$.
+Of course, it never hurts to iterate longer, but such calculations hardly brought added value.
+
+The main finding of this study is that valid inferences may be obtained much quicker than approximate algorithmic convergence is reached.
+Under the current specifications, approximately unbiased, confidence-valid estimates were obtained after a maximum of nine iterations.
+Continued iterations beyond $\tau = 9$ did not yield better estimates.
+In contrast, $\widehat{R}$-values decreased substantially until 20 to 30 iterations, which implies that mixing in the algorithm still improved with each additional iteration.
+$AC$-values only improved until about six iterations, suggesting minimal improvement in trending was obtained beyond $\tau=6$ in the current simulation set-up.
+
+[TODO: write about: methodological explanation is that rhat and ac have a lag (too few iterations to inform your statistic) --\> will always indicate convergence slower than inferential validity is reached]
 
@@
 
-From the traceplot of the chain means (see \@ref(fig:case)A) it seems that mixing improves up-to 10 iterations, while trending is only apparent in the first three iterations.
-
-This figure (\@ref(fig:case)B) shows that 7 iterations are required before the $\widehat{R}$-values of the chain means drop below the threshold for non-convergence.
-
-The $\widehat{R}$-values for the scientific estimate reaches the threshold much sooner, when $n_{it}=14$ (see \@ref(fig:case)C).
-
-According to the $\widehat{R}$-values with $\lambda$ as parameter, at least 15 iterations are required (see \@ref(fig:case)D).
+From the traceplot of the chain means (see @fig-case)A) it seems that mixing improves up-to 10 iterations, while trending is only apparent in the first three iterations.
+
+This figure (@fig-case)B) shows that 7 iterations are required before the $\widehat{R}$-values of the chain means drop below the threshold for non-convergence.
+
+The $\widehat{R}$-values for the scientific estimate reaches the threshold much sooner, when $n_{it}=14$ (see @fig-case)C).
+
+According to the $\widehat{R}$-values with $\lambda$ as parameter, at least 15 iterations are required (see @fig-case)D).
 

### 2025-05-08 figures/plot_patterns.R

@@ -50,7 +50,7 @@ plot_patterns <- function(n_col = 4) {
       "missing" = "#B61A51B3"
     )) +
     ggplot2::scale_alpha_continuous(limits = c(0, 1), guide = "none") +
-    ggplot2::scale_x_continuous(breaks = 1:cls, labels = vrb) +
+    ggplot2::scale_x_continuous(breaks = 1:cls, labels = vrb, position = "top") +
     ggplot2::scale_y_continuous(breaks = NULL, labels = NULL) +
     ggmice:::theme_minimice() +
     labs(fill = "", x = "Variables", y = "Patterns")

### 2025-05-20 Manuscript/manuscript.qmd
plot just 1 predictor

@@
     a. Repeat over variables ($j = 1,...,p$).
+
         I.
@@
 
-\noindent We then estimate between-chain variance $B$ (note that we diverge from the typical notation in MI, where $B$ denotes the variance between the estimated quantities of scientific interest $\hat{Q}_{\ell}$).
+\noindent We then estimate between-chain variance $B$ (note that we diverge from the typical notation in MI, where $B$ denotes the variance between the estimated quantities of scientific interest $\hat{Q}_{\ell}$ TODO: add not between imp var cf Rubin).
 $B$ is defined as the variance of the collection of average $\theta$s per chain:
@@
 \end{table}
+
+TODO: table captions
 
@@
 
-The following figures display the simulation results for the sixteen diagnostic identifier-parameter-variable pairs, and sixteen performance measure-variable pairs, contrasted to the number of iterations in the imputation algorithm.
-Within the figures, we split the results according to the missingness conditions [TODO: missingness mechanisms as line types, and proportion of incomplete cases as colors?].
+<!-- The following figures display the simulation results for non-convergence diagnostics and performance measures, contrasted to the number of iterations in the imputation algorithm.
+Within the figures, we split the results according to the missingness conditions.
 Note that these results are averages of the $n_{sim} = 2000$ simulation repetitions.
+-->
+
+<!-- [TODO: missingness mechanisms as line types, and proportion of incomplete cases as colors?] -->
 
@@
 
-```{r fig-ac_mean}
-plot_results(results, "ac_mean", 0, "MCAR")
+```{r fig-ac_mean, fig.cap="Autocorrelation in the chain means.
+The dashed line indicates perfect stationarity, $AC=0$.
+The solid black line indicates the avarage autocorrelation accross simulation repetitions, with grey lines representing the empirical confidence interval."}
+plot_results2(results, "ac_mean", 0, "MCAR")
 ```
@@
 
-```{r fig-ac_sd}
-plot_results(results, "ac_sd", 0, "MCAR")
+```{r fig-ac_sd, fig.cap="Autocorrelation in the chain variances.
+The dashed line indicates perfect stationarity, $AC=0$.
+The solid black line indicates the avarage autocorrelation accross simulation repetitions, with grey lines representing the empirical confidence interval."}
+plot_results2(results, "ac_sd", 0, "MCAR")
 ```
@@
 
-```{r fig-psrf_mean}
-plot_results(results, "psrf_mean", 1, "MCAR")
+```{r fig-psrf_mean, fig.cap="PSRF in the chain means.
+The dashed line indicates perfect mixing, $\\widehat{R}=0$.
+The solid black line indicates the avarage PSRF accross simulation repetitions, with grey lines representing the empirical confidence interval."}
+plot_results2(results, "psrf_mean", 1, "MCAR")
 ```
@@
 
-```{r fig-psrf_sd}
-plot_results(results, "psrf_sd", 1, "MCAR")
+```{r fig-psrf_sd, fig.cap="PSRF in the chain variances.
+The dashed line indicates perfect mixing, $\\widehat{R}=0$.
+The solid black line indicates the avarage PSRF accross simulation repetitions, with grey lines representing the empirical confidence interval."}
+plot_results2(results, "psrf_sd", 1, "MCAR")
 ```
@@
 
-```{r fig-bias}
-plot_results(results, "bias", 0, "MCAR")
-```
+```{r fig-bias, fig.height=8, fig.cap="Bias in regression coefficient $\\beta = 1$.
+The dashed line indicates no bias.
+The solid black line indicates the avarage bias accross simulation repetitions, with grey lines representing the empirical confidence interval."}
+plot_results2(results, "bias", 0, "MCAR")
+```
+
+<!-- TODO: free-y in facet grid? Y, X1 en X3 naar appendix, CI not dashed -->
 
@@
 
-```{r fig-cov}
-plot_results(results, "cov", 0.95, "MCAR")
+```{r fig-cov, fig.cap="Coverage rate of regression coefficient $\\beta = 1$.
+The dashed line indicates confidence-validity.
+The solid black line indicates the avarage coverage rate accross simulation repetitions, with grey lines representing the empirical confidence interval."}
+plot_results2(results, "cov", 0.95, "MCAR")
 ```
@@
 
-```{r fig-ciw}
-plot_results(results, "ciw", NULL, "MCAR")
+```{r fig-ciw, fig.cap="Confidence interval width of regression coefficient $\\beta = 1$.
+The solid black line indicates the avarage CIW accross simulation repetitions, with grey lines representing the empirical confidence interval."}
+plot_results2(results, "ciw", NULL, "MCAR")
 ```
@@
 
+## Conclusion
+
 Identifying non-convergence may be a crucial aspect of drawing valid statistical inferences from incomplete data.
@@
 
+## Limitations
+
+The potential scale reduction factor metric assumes over-dispersed starting values of the iterative algorithm.
+The MICE algorithm does not start in an over-dispersed state.
+MICE does not rely on starting values for parameters at all: instead, the algorithm is initialized based on a 'zeroth' iteration, where missing values are filled in using a random draw from observed values.
+Under MCAR, this would typically not yield an over-dispersed state at all, depending on the parameter of interest: means and variances are not affected, regression coefficients and other multivariate parameters start biased downwards (because sampling starting values distorts multivariate structures in the data).
+Whether this accrues to over-dispersion is questionable.
+Under MAR (or MNAR), some over-dispersion might occur when the observed data and imputed data differ impactfully.
+The algorithm should 'escape' the initial state (based on the starting values drawn from observed data alone).
+Over iterations, all parameters of interest (e.g.
+means, variances, multivariate estimands) should converge towards a stable state.
+Moreover, the MICE algorithm is initialized with more information than a typical MCMC algorithm: parameter estimates depend not only on imputed data, but also on observed data that does not change over iterations.
+Therefore, the initial state of the algorithm is 'too good' to observe a relative decrease in $\widehat{R}$.
+
 [TODO: write about: methodological explanation is that rhat and ac have a lag (too few iterations to inform your statistic) --\> will always indicate convergence slower than inferential validity is reached]
+
+## Implications
 

### 2025-05-20 figures/plot_results.R

@@ -22,9 +22,38 @@ plot_results <- function(summarized_results, metric, hline, mis = "MAR") {
     ggplot(aes(x = .it, y = .data[[metric]])) +
     geom_hline(yintercept = hline, alpha = 0.1, linewidth = 1) + 
     geom_line(na.rm = TRUE) +
-    geom_line(aes(y = .data[[metric_ll]]), linetype = "dashed", alpha = 0.3, na.rm = TRUE) +
-    geom_line(aes(y = .data[[metric_ul]]), linetype = "dashed", alpha = 0.3, na.rm = TRUE) +
+    geom_line(aes(y = .data[[metric_ll]]), alpha = 0.3, na.rm = TRUE) +
+    geom_line(aes(y = .data[[metric_ul]]), alpha = 0.3, na.rm = TRUE) +
     facet_grid(term ~ prop + mech) +
+    labs(x = "Iteration number", y = y_lab) +
+    theme_classic()
+}
+
+plot_results2 <- function(summarized_results, metric, hline, mis = c("MCAR", "MAR", "MNAR")) {
+  # Check if the metric is valid
+  if (!metric %in% c("bias", "cov", "ciw", "ac_mean", "ac_sd", "psrf_mean", "psrf_sd")) {
+    stop("Invalid metric. Choose from 'bias', 'cov', 'ciw', 'ac_mean', 'ac_sd', 'psrf_mean', or 'psrf_sd'.")
+  }
+  metric_ll <- paste0(metric, "_ll")
+  metric_ul <- paste0(metric, "_ul")
+  y_lab <- switch(
+    metric,
+    bias = "Bias",
+    cov = "Coverage rate",
+    ciw = "CI width",
+    ac_mean = "AC in chain means",
+    ac_sd = "AC in chain SDs",
+    psrf_mean = "PSRF in chain means",
+    psrf_sd = "PSRF in chain SDs"
+  )
+  summarized_results |>
+    filter(method == "MICE", term == "X3", mech %in% mis) |> 
+    ggplot(aes(x = .it, y = .data[[metric]])) +
+    geom_hline(yintercept = hline, linewidth = 0.2, linetype = "dashed") + 
+    geom_line(na.rm = TRUE) +
+    geom_line(aes(y = .data[[metric_ll]]), alpha = 0.3, na.rm = TRUE) +
+    geom_line(aes(y = .data[[metric_ul]]), alpha = 0.3, na.rm = TRUE) +
[…]

### 2025-05-23 Manuscript/manuscript.qmd
text flow

@@
 
-Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} \gtrapprox 1$), and there is little dependency between subsequent iterations of imputation chains ($AC \gtrapprox 0$).
+Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} \approx 1$), and there is little dependency between subsequent iterations of imputation chains ($AC \approx 0$).
 In practice, we diagnose non-convergence when approximate convergence is violated, i.e., when $\widehat{R}$ and $AC$ exceed a certain threshold.
@@
 We calculate bias as $\bar{Q} - Q$.
-CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$.
+CR is defined as the proportion of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$.
 Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$.
@@
 
-Autocorrelation in the chain means decreases rapidly in the first few iterations (see @fig-ac_mean).
+[TODO: edit!] Autocorrelation in the chain means decreases rapidly in the first few iterations (see @fig-ac_mean).
 The decrease is substantive until $n_{it} \geq 6$.
@@
 
-## Conclusion
+<!-- ## Conclusion -->
 
@@
 
-## Limitations
+<!-- ## Limitations -->
+
+[TODO: add segway]
+
+[TODO: write about: methodological explanation is that rhat and ac have a lag (too few iterations to inform your statistic) --\> will always indicate convergence slower than inferential validity is reached]
 
@@
 
-[TODO: write about: methodological explanation is that rhat and ac have a lag (too few iterations to inform your statistic) --\> will always indicate convergence slower than inferential validity is reached]
-
 ## Implications
@@
 -->
+
+# References

### 2025-10-21 testrun.R
add non-convergence condition

@@ -1,20 +1,38 @@
-# nonconv
-library(ggplot2)
+# setup via simulation.R file
 
 # generate data
-dat <- generate_complete(n_obs, corr, betas)
-# dat$X5 <- (2 * dat$X3) + dat$X4
+dat <- generate_complete(n_obs, n_col, corr, betas)
+# dat$Z <- dat$X1 / (dat$X1 + dat$X2)
+# dat$Z <- dat$X1 / dat$X2^2
+dat$Z <- dat$X1 + dat$X2
 
 # ampute data
-amps <- induce_missingness(dat, mis_pat = NULL, mis_mech = "MAR", mis_prop = 0.5)
-# amp = amps[[1]]
+mis_pat <- create_patterns(n_col + 1)
+amps <- induce_missingness(dat, mis_pat = mis_pat, mis_mech = "MAR", mis_prop = 0.5)
+ggmice::plot_pattern(amps[[1]]$amp)
 
-# impute data with MICE
-imp <- mice(amps[[1]]$amp, maxit = 10)
-plot(imp)
+# impute data with MICE correctly specified
+meth <- make.method(amps[[1]]$amp[, 1:n_col])
+meth[] <- "norm"
+imp <- mice(amps[[1]]$amp[, 1:n_col], maxit = 10, meth = meth)
+# plot(imp)
 ggmice::plot_trace(imp)
 
 # convergence
 conv <- convergence(imp)
 ggplot(conv, aes(x = .it, y = ac, color = vrb)) + geom_hline(yintercept = 0) + geom_line() + theme_classic()
 ggplot(conv, aes(x = .it, y = psrf, color = vrb)) + geom_hline(yintercept = 1) + geom_line() + theme_classic()
+
+# impute data with MICE incorrectly specified
+meth <- make.method(amps[[1]]$amp)
+# meth["Z"] <- "~I(X1 / (X1 + X2))"
+# meth["Z"] <- "~I(X1 / X2^2)"
+meth[] <- "norm"
+meth["Z"] <- "~I(X1 + X2)"
+imp2 <- mice(amps[[1]]$amp, maxit = 10, meth = meth)
+# plot(imp2)
+ggmice::plot_trace(imp2)
+# non-convergence
+conv2 <- convergence(imp2)
+ggplot(conv2, aes(x = .it, y = ac, color = vrb)) + geom_hline(yintercept = 0) + geom_line() + theme_classic()
[…]

### 2025-10-21 simulation.R
add non-convergence condition

@@ -7,6 +7,8 @@ library(dplyr)
 library(mvtnorm)
 library(mice)
 library(miceadds)
+library(ggplot2)
+library(ggmice)
 
 # functions
 miceadds::source.all("./R")
@@ -15,14 +17,15 @@ miceadds::source.all("./R")
 set.seed(11)
 
 # parameters
-n_sim <- 20
-n_obs <- 500
+n_sim <- 500
+n_obs <- 200
 n_col <- 3
-corr <- 0.5
+composite <- FALSE
+corr <- 0.3
 beta <- 1
 betas <- rep(beta, n_col)
 mis_pat <- create_patterns(n_col)
-mis_mech = c("MCAR") #, "MAR", "MNAR")
+mis_mech = c("MCAR", "MAR") #, "MNAR")
 mis_prop = c(0.05, 0.25, 0.5, 0.75, 0.95)
 n_it <- 50
 
@@ -47,7 +50,7 @@ n_it <- 50
 ### COMBINE INTO ONE FUCTION ###
 ################################
 
-simulate_once <- function(n_obs, corr, betas, mis_pat, mis_mech, mis_prop, n_it) {
+simulate_once <- function(n_obs, corr, betas, mis_pat, mis_mech, mis_prop, n_it, composite) {
   # generate incomplete data
   amps <- create_data(
     sample_size = n_obs,
@@ -55,10 +58,11 @@ simulate_once <- function(n_obs, corr, betas, mis_pat, mis_mech, mis_prop, n_it)
     effects = betas,
     patterns = mis_pat,
     mechanisms = mis_mech,
-    proportions = mis_prop
+    proportions = mis_prop,
+    non_conv = composite
   )
   # estimate regression coefficients
-  ests <- apply_methods(amps, betas, n_it)
+  ests <- apply_methods(amps, betas, n_it, non_conv = composite)
   # output
   return(ests)
 }
@@ -117,12 +121,13 @@ parallel::clusterExport(
     "betas",
     "mis_pat",
     "mis_mech",
-    "mis_prop"
+    "mis_prop",
+    "composite"
   )
 )
 
 out <- pbreplicate(n_sim, 
-                   simulate_once(n_obs, corr, betas, mis_pat, mis_mech, mis_prop, n_it),
+                   simulate_once(n_obs, corr, betas, mis_pat, mis_mech, mis_prop, n_it, composite),
                    cl = cl, 
                    simplify = FALSE)
 

### 2025-10-21 testrun.R

@@ -2,19 +2,16 @@
 
 # generate data
 dat <- generate_complete(n_obs, n_col, corr, betas)
-# dat$Z <- dat$X1 / (dat$X1 + dat$X2)
-# dat$Z <- dat$X1 / dat$X2^2
-dat$Z <- dat$X1 + dat$X2
 
 # ampute data
-mis_pat <- create_patterns(n_col + 1)
+mis_pat <- create_patterns(n_col)
 amps <- induce_missingness(dat, mis_pat = mis_pat, mis_mech = "MAR", mis_prop = 0.5)
 ggmice::plot_pattern(amps[[1]]$amp)
 
 # impute data with MICE correctly specified
-meth <- make.method(amps[[1]]$amp[, 1:n_col])
+meth <- make.method(amps[[1]]$amp)
 meth[] <- "norm"
-imp <- mice(amps[[1]]$amp[, 1:n_col], maxit = 10, meth = meth)
+imp <- mice(amps[[1]]$amp, maxit = 10, meth = meth)
 # plot(imp)
 ggmice::plot_trace(imp)
 
@@ -24,15 +21,19 @@ ggplot(conv, aes(x = .it, y = ac, color = vrb)) + geom_hline(yintercept = 0) + g
 ggplot(conv, aes(x = .it, y = psrf, color = vrb)) + geom_hline(yintercept = 1) + geom_line() + theme_classic()
 
 # impute data with MICE incorrectly specified
+dat <- generate_complete(n_obs, n_col, corr, betas, composite = TRUE)
+# ampute data
+mis_pat <- create_patterns(n_col)
+amps <- induce_missingness(dat, mis_pat = mis_pat, mis_mech = "MAR", mis_prop = 0.5)
 meth <- make.method(amps[[1]]$amp)
 # meth["Z"] <- "~I(X1 / (X1 + X2))"
 # meth["Z"] <- "~I(X1 / X2^2)"
 meth[] <- "norm"
-meth["Z"] <- "~I(X1 + X2)"
-imp2 <- mice(amps[[1]]$amp, maxit = 10, meth = meth)
+meth[n_col + 1] <- "~I(X1 + X2)"
+imp <- mice(amps[[1]]$amp, maxit = 10, meth = meth)
 # plot(imp2)
-ggmice::plot_trace(imp2)
+ggmice::plot_trace(imp)
 # non-convergence
-conv2 <- convergence(imp2)
[…]

### 2025-10-21 R/apply_methods.R
non-convergence arguments

@@ -35,9 +35,15 @@ apply_CCA <- function(amp) {
 }
 
 # MICE imputation
-apply_MICE <- function(amp, n_it) {
+apply_MICE <- function(amp, n_it, composite = FALSE) {
+  # make methods vector, edit for non-convergence condition
+  meth <- rep("norm", n_col + 1)
+  names(meth) <- c("Y", paste0("X", 1:n_col))
+  if (composite) {
+    meth[n_col + 1] <- "~I(X1 + X2)"
+  }
   # first imputation with MICE
-  imp1 <- mice::mice(amp$amp, method = "norm", maxit = 1, printFlag = FALSE)
+  imp1 <- mice::mice(amp$amp, method = meth, maxit = 1, printFlag = FALSE)
   # add regression estimates
   implist <- list(imp1, cbind(.it = 1, estimate_param(imp1)))
   # iterate and estimate
@@ -91,13 +97,13 @@ add_iteration <- function(implist) {
 }
 
 # combine into one function
-apply_methods <- function(amps, betas, n_it) {
+apply_methods <- function(amps, betas, n_it, non_conv) {
   # comparative truth on complete dataset
   full <- purrr::map_dfr(amps, ~{apply_full(.x)})
   # apply CCA to each incomplete dataset
   CCA <- purrr::map_dfr(amps, ~{apply_CCA(.x)})
   # impute with MICE and estimate effects
-  MICE <- purrr::map_dfr(amps, ~{apply_MICE(.x, n_it)})
+  MICE <- purrr::map_dfr(amps, ~{apply_MICE(.x, n_it, non_conv)})
   # combine estimates 
   ests <- rbind(full, CCA, MICE) |> 
     cbind(truth = c(0, betas))

### 2025-10-21 R/generate_datasets.R

@@ -1,12 +1,17 @@
 # functions to create incomplete datasets
 
 # generate complete data
-generate_complete <- function(n_obs, n_col, corr, betas) {
+generate_complete <- function(n_obs, n_col, corr, betas, composite = FALSE) {
+  if (composite & n_col < 3L) {stop("Composite variable condition requires 2 or more predictor variables.")}
   # create variance-covariance matrix with moderate correlations
   vcov <- matrix(corr, nrow = n_col, ncol = n_col)
   diag(vcov) <- 1
   # create predictor space data
   X <- mvtnorm::rmvnorm(n = n_obs, sigma = vcov)
+  # for non-convergence condition, add composite variable
+  if (composite) {
+    X[, n_col] <- X[ ,1] + X[, 2]
+  }
   # multiply each predictor observation by the corresponding beta
   Y <- X %*% betas
   # generate residual error for each observation
@@ -22,11 +27,11 @@ generate_complete <- function(n_obs, n_col, corr, betas) {
 create_patterns <- function(n_col) {
   pat_list <- vector(mode = "list", length = n_col + 1)
   pat_list <- lapply(pat_list, \(x) {x <- c(0, 1)})
-  mis_pat <- expand.grid(pat_list) 
-  names(mis_pat) <- c("Y", paste0("X", 1:n_col))
+  obs_pat <- expand.grid(pat_list) 
+  names(obs_pat) <- c("Y", paste0("X", 1:n_col))
   # omit patterns with all/none missing
-  mis_pat <- mis_pat[rowSums(mis_pat) > 0 & rowSums(mis_pat) < n_col, ]
-  return(mis_pat)
+  obs_pat <- obs_pat[rowSums(obs_pat) > 0 & rowSums(obs_pat) < n_col, ]
+  return(abs(obs_pat - 1))
 }
 
 # ampute the complete data
@@ -53,13 +58,16 @@ create_data <- function(sample_size,
                         effects,
                         patterns,
                         mechanisms,
-                        proportions) {
+                        proportions,
+                        non_conv, 
+                        ...) {
   # create a single complete dataset
   dat <- generate_complete(
     n_obs = sample_size,
     n_col = length(effects),
     corr = correlations,
-    betas = effects
+    betas = effects,
+    composite = non_conv
   )
   # ampute the data with different missingness
   amps <- induce_missingness(dat,

### 2025-10-21 R/compute_metrics.R
simulate 500x

@@ -164,7 +164,7 @@ max(bulk_rhat, tail_rhat)
 rstan:::rhat_rfun(chains)
 # higher Rhat value on 'plain' data
 
-# try to match with my functions
+# try to match with my functions 
 bulk_rhat2 <- z_scale(split_chains(chains)) |> calculate_psrf()
 tail_rhat2 <- z_scale(split_chains(abs(chains - median(chains)))) |> calculate_psrf()
 max(bulk_rhat2[length(bulk_rhat2)], tail_rhat2[length(tail_rhat2)])

### 2025-10-21 Manuscript/manuscript.qmd
plot unique simulation repetitions

@@
 plot_results(results, "psrf_sd", 1, "MCAR")
+```
+
+```{r, eval=FALSE}
+# plot each sim
+performance |>
+  filter(method == "MICE", mech == "MAR", term == "X3", prop == 0.5) |> 
+  ggplot(aes(x = .it, y = psrf_mean, group = .sim)) +
+  geom_hline(yintercept = 1, alpha = 0.1) + 
+  geom_line(alpha = 0.1) +
+  # geom_line(aes(y = psrf_mean_ll), linetype = "dashed", alpha = 0.3) +
+  # geom_line(aes(y = psrf_mean_ul), linetype = "dashed", alpha = 0.3) +
+  # facet_grid(term ~ mech + prop) +
+  theme_classic()
 ```
